### 2017-06-12 paper.Rmd
Create paper.Rmd

---
title: "Noise-induced phenonema in Population Biology"
author: "Carl Boettiger"
---

Outline

## What is noise?

All that you leave out of the model

## Sources of noise

- Demographic stochasticity 
- Environmental stochasticity
- Individual heterogeneity
- Spatial variability
- stochastic chaos

## Noise-induced phenonema

- Oscillations (parametric oscillator, stochastic resonance)
- population inflation
- Critical transitions
- Spatial pattern formation
- Population persistence (bet-hedging)
- Species coexistence

## Noise as information

- Measuring stability
- Early warning signals
- maximum entropy methods

### 2017-06-14 paper.Rmd
update sketch

@@
 ---
-title: "Noise-induced phenonema in Population Biology"
+title: "From noise to knowledge: how randomness generates novel phenomena and information"
 author: "Carl Boettiger"
@@
 
-Outline
+# Outline
+
+In this paper, I provide a brief synthesis of ecological research in stochastic phenomena.
+Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath.
+Yet despite this well-earned reputation, noise is often interesting in its own right: noise can induce novel phenomena that could not be understood from some underlying determinstic model alone.
+Nor is all noise the same, and close examination of differences in in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.
+While an exhaustive review is beyond the scope of this piece, I endeavor to highlight some of the most interesting examples of both noise-induced phenomena and noise-derived information, and point the way towards yet uncharted territory.
 
@@
 
-## Noise-induced phenonema
+## Noise the nuisance: Noise can obscure patterns
+
+- deterministic skeletons, signal-to-noise ratio, model fitting
+
+## Noise the creator: Noise can induced novel phenonema
 
@@
 
-## Noise as information
+## Noise the informer: Noise can provide novel information
 

### 2017-10-10 paper.Rmd
WIP

@@
 
-- Demographic stochasticity 
-- Environmental stochasticity
-- Individual heterogeneity
-- Spatial variability
-- stochastic chaos
+**Demographic stochasticity**.
+Demographic stochasticity refers to fluctuations in population sizes or densities that arise from the fundamentally discrete nature of individual birth and death events.
+Demographic stochasticity is a particularly instructive case for illustrating a mechanism for how noise arises as an aggregate description from a fundamentally lower-level mechanistic process.
+Demographic stochasticity is typically modeled as Markov jump process: in a population of $N$ identical individuals at time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate that can depend on both the population size $N$) and increases the population size, $N$ to $N+1$, and deaths occur with probability $d_t(N_t)$, decreasing the population size, $N$ to $N-1$.
+Assuming each of these events are independent, this is a Poisson process
+
+[@Black2012]
+
+**Environmental stochasticity**
+
+[@Black2012]
+
+**Individual heterogeneity**
+
+**Spatial variability**
+
+### Not noise
+
+In addition to these sources of variation, there are additional sources of variation or uncertainty in a model which we will not consider for the purpose of this review which, while related, operate in a fundamentally different way that does not give rise to the same phenomena and information.
+
+**Observational error**: Observational error acts much like another form of stochasticity, arising from
+sampling effects or limitations of instruments, etc. Yet unlike the sources described above, this 
+uncertainty does not impact the actual state of our system, and thus does not give rise to any
+additional phenomena or information about that process.
+
+**Reducible uncertainty**: In constructing and estimating mathematical models of ecological processes, 
+we encounter uncertainty over our parameter estimates and possible choice of models [@Ellison2004].  In some ways, this
+uncertainty can act like the stochasticity described above: for instance, both introduce variation into
+any forecast of future outcomes. Yet unlike stochasticity, this uncertainty is reducible: with additional
+observations, we can narrow the uncertainty over parameters or models, while stochasticity remains an
+inherent source of variation.
+
+**Chaos**: Deterministic chaos [@Desharnais2001] is another mechanism of introducing significant variation into a simple
+model which can be observationally indistinguishable from a stochastic process [@Werndl2009].
+
+Figure: schematic showing types of noise
 
@@
 
-## Noise the creator: Noise can induced novel phenonema
+Figure:
+
+## Noise the creator: Noise can induce novel phenonema
 
 - Oscillations (parametric oscillator, stochastic resonance)
+
+[@Black2010]
+
 - population inflation
+
+@Read2007 <!-- Stochasticity generates an evolutionary instability for infectious disease -->
+
 - Critical transitions
+
+[@Schreiber2008] decreasing variance before crash
+
 - Spatial pattern formation
@@
 - Species coexistence
+
+Figure:
 
@@
 - maximum entropy methods
+
+Figure:
+
+## Conclusions

### 2017-10-20 code/appendix.Rmd
WIP

---
title: "Appendix"
output: html_notebook
---

```{r}
knitr::opts_chunk(message=FALSE)
```

```{r}
library(nimble)
library(tidyverse)
```

## Stochastic inflation

```{r}

logistic  <- nimble::nimbleCode({

  x[1] <- x0
  for(t in 1:(N-1)){
    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)
    x[t+1] ~ dnorm(mu[t], sd = sigma)
  }

})

p <- list(r = .4, K = 1, sigma = .2, N = 500)

model <- 
nimbleModel(logistic,
            constants = p,
            inits = list(x0 = 1))
cmodel <- compileNimble(model)
set.seed(123)

df <- map_dfr(1:10000, 
              function(rep){
                simulate(cmodel)
                data.frame(t = seq_along(cmodel$x), x = cmodel$x)
              },
              .id = "rep")

```

```{r}
(p$K + sqrt(p$K^2 - 8 * p$sigma^2) )/ 2
```

```{r}
df %>% filter(t > 20, x > 0) %>%  summarize(mean(x))
```

```{r}

df %>% 
  filter(x > 0) %>% 
  group_by(t) %>%
  summarise(ave = mean(x), sd = sd(x)) %>%
  gather(moment, value, -t) %>%
  ggplot(aes(t,value, col=moment)) + geom_line()

```

------------

# Quasi-cycles

Understanding the origin of sustained oscillations has long been a central question in ecology [@Hastings]

```{r}

quasicycle  <- nimble::nimbleCode({

  x[1] <- x0
  y[1] <- y0

  for(t in 1:(N-1)){
    mu_x[t] <- x[t] + x[t] * r * (1 - x[t] / K) - b * x[t] * y[t]
    x[t+1] ~ dnorm(mu_x[t], sd = sigma_x)
    mu_y[t] <- y[t] + c * x[t] * y[t] - d * y[t]
    y[t+1] ~ dnorm(mu_y[t], sd = sigma_y)
  }

})

p <- 
  data.frame(
  data.frame(r = .1, K = 5,  b = .1, c = .1, d = .1, N = 500),
  data.frame(sigma_x = c(.00001,0.01),  sigma_y = c(.00001,0.01)))

f <- function(constants){
  model <- compileNimble(nimbleModel(quasicycle, constants = constants, inits = list(x0 = 1, y0 = 1)))
  set.seed(123)
  simulate(model)
  tibble(t = seq_along(model$x), x = model$x, y = model$y, sigma = constants$sigma_x)
}

df <- p %>% rowwise() %>% do(f(.))

```

```{r}
df %>%
  gather(species, pop, -t, -sigma) %>%
  ggplot(aes(t, pop, col=species)) + geom_line() + facet_wrap(~sigma)
```

# Turing patterns?

------------

```{r}
data(lynx)
lynx <- tibble(year = 1821:1934, lynx = as.numeric(lynx))

lynx %>%
  ggplot(aes(year,lynx)) + geom_point() + geom_line()
```

### 2017-10-20 paper.Rmd
WIP

@@
 
-All that you leave out of the model
+<!-- Our state variables don't "exist" except as averages of a lower level process -->
+
+Where does noise come from?
+It is important to bear in mind that whenever stochasticity is introduced into a model, it is only as a simplification or summary of the effects of a large number of other processes, often occuring at different scales.
+Simply put, *noise is all that we leave out of our model*.
+While ecologists will sometimes assert that stochasticity is inherent: birth and death are inherently stochastic events; or the environment is inherently stochastic, these are only assertions at a particular scale.
+Obviously events such as births and deaths are themselves the results of mechanistic processes, just as fluctuations in rainfall and temperature are the result of deterministic laws of physics and not fundamentally random forces.
+In all of science, only quantum mechanics claims a fundamentally probabilistic view of events, and then at such scales that have no relevance to ecology.
+The term "noise" itself reflects this emergent rather than fundamental origin: the combination of hundreds of voices and sounds each individually distinct blends together into a background of noise.
 
@@
 Demographic stochasticity is a particularly instructive case for illustrating a mechanism for how noise arises as an aggregate description from a fundamentally lower-level mechanistic process.
-Demographic stochasticity is typically modeled as Markov jump process: in a population of $N$ identical individuals at time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate that can depend on both the population size $N$) and increases the population size, $N$ to $N+1$, and deaths occur with probability $d_t(N_t)$, decreasing the population size, $N$ to $N-1$.
-Assuming each of these events are independent, this is a Poisson process
+We summarize the myriad lower-level processes that mechanistically lead to the event of a 'birth' in the population as a probability: in a population of $N$ identical individuals at time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate that can depend on both the population size $N$) and increases the population size, $N$ to $N+1$, and deaths occur with probability $d_t(N_t)$, decreasing the population size, $N$ to $N-1$.
+Assuming each of these events are independent, this is a Poisson process.
+
+<!-- Master equation transform-->
+
+<!-- Demographic noise scaling with population size -->
+
+<!-- Gillespie algorithm -->
 
@@
 
-**Environmental stochasticity**
+**Environmental stochasticity**.
+Whereas demographic noise reflects processes at a lower scale (the individual, rather than the population), environmental stochasticity reflects processes operating at the same scale that we are not explicitly modelling.
 
@@
 
+Colored noise e.g. [@Marshall2015]
+
 **Individual heterogeneity**
+
+[@Hart2016]
 
@@
 uncertainty does not impact the actual state of our system, and thus does not give rise to any
-additional phenomena or information about that process.
+additional phenomena or information about that process.  So while this is indeed a source of noise,
+it falls outside our scope here.
 
@@
 any forecast of future outcomes. Yet unlike stochasticity, this uncertainty is reducible: with additional
-observations, we can narrow the uncertainty over parameters or models, while stochasticity remains an
-inherent source of variation.
+observations, we can narrow the uncertainty over parameters or candidate models, while stochasticity remains an
+inherent source of variation. We will this not consider model and parameter uncertainty, as they are not 
+sources of noise.
 
 **Chaos**: Deterministic chaos [@Desharnais2001] is another mechanism of introducing significant variation into a simple
-model which can be observationally indistinguishable from a stochastic process [@Werndl2009].
+model.
+Chaos can a look a lot like noise: indeed, it can be observationally indistinguishable from a stochastic process [@Werndl2009].
+Despite substantial interest in previous decades and the relative ease of encountering chaos in simple ecological models (particularly in differences equations), and convincing reconstructions of chaotic dynamics in laboratory manipulations [@Constantino1997] the prevelance of chaotic dynamics as a driving factor in most ecological systems remains unclear.
+Meanwhile, the undeniable influences of
 
@@
 
-- deterministic skeletons, signal-to-noise ratio, model fitting
+- deterministic skeletons, 
+- signal-to-noise ratio, 
+- model fitting
 
-Figure:
+Figure: Lynx-hare data
 
 ## Noise the creator: Noise can induce novel phenonema
+
+Stochasticity can give rise to a wide array of interesting phenomena that are often at odds with our intuition.
+
+**Quasi-cycles**
 
@@
 [@Black2010] 
+[@Blarer1999]
 
-- population inflation
+- population inflation / deflation
 
@@
 - Spatial pattern formation
+
+Turning patterms found in plankton-herbivore dynamics [@Levin1976]
+
 - Population persistence (bet-hedging)
-- Species coexistence
+
+- Species coexistence  ([@Hart2016])
 
@@
 ## Noise the informer: Noise can provide novel information
+
+One of the most interesting aspects of noise is as a source of additional information.
+Noise can act like a miniature perturbation experiment, constantly perturbing a system away from an equilibrium to reveal it's behavior elsewhere in state space.
 

### 2017-10-20 code/appendix.Rmd
WIP

@@
 
+# Stochastic oscillator
+
+```{r}
+p <- list(r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 1e4)
+
+f <- function(x) x * p$r * (1 - x / p$K)
+g <- function(x) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
+
+x <- seq(0,2, length.out = 100)
+data.frame(x, f = f(x), g = g(x)) %>%
+  ggplot(aes(x)) + geom_line(aes(y = f-g)) + 
+  geom_hline(aes(yintercept=0), lty=2) +
+  coord_cartesian(xlim = c(0,1.5), ylim=c(-.015,.1))
+
+```
+
+```{r}
+system.time({
+may  <- nimble::nimbleCode({
+
+  x[1] <- x0
+  for(t in 1:(N-1)){
+    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)  - a * x[t] ^ Q / (x[t] ^ Q + H ^ Q)
+    y[t+1] ~ dnorm(mu[t], sd = sigma)
+    x[t+1] <- max(y[t+1],0)
+  }
+
+})
+
+model <- nimbleModel(may,constants = p, inits = list(x0 = 1.2))
+cmodel <- compileNimble(model)
+set.seed(123)
+
+simulate(cmodel)
+df <- tibble(t = seq_along(cmodel$x), x = cmodel$x)
+})
+
+df  %>% ggplot(aes(t,x)) + geom_line()
+```
+
+# Tipping points
+
+Needs a model with non-zero alternate state!
+
+```{r}
+
+tip  <- nimble::nimbleCode({
+
+  x[1] <- x0
+  for(t in 1:(N-1)){
+    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K) * (x[t] - A) / K - h[t]
+    y[t+1] ~ dnorm(mu[t], sd = sigma)
+    x[t+1] <- max(y[t+1],0)
+  }
+
+})
+
+p <- list(r = .2, K = 1, A = 0.2, sigma = .01, N = 500, h = seq(0,.03,length=500))
+
+model <- nimbleModel(tip,constants = p, inits = list(x0 = 1))
+cmodel <- compileNimble(model)
+set.seed(123)
+
+simulate(cmodel)
+df <- tibble(t = seq_along(cmodel$x), x = cmodel$x)
+
+```
+
+```{r}
+df %>% ggplot(aes(t,x)) + geom_line()
+```
+
 # Turing patterns?

### 2017-10-21 code/appendix.Rmd
WIP

@@
 
-p <- list(r = .4, K = 1, sigma = .2, N = 500)
+p <- list(r = .4, K = 1, sigma = .15, N = 500)
 
@@
 
+df %>% 
+  filter(x > 0) %>% 
+  group_by(t) %>%
+  summarise(ave = mean(x), sd = sd(x)) %>%
+  write_csv("inflation.csv")
+
 ```
 
+Analytical predicted population size, as a fraction of $K$, is:
+
 ```{r}
-(p$K + sqrt(p$K^2 - 8 * p$sigma^2) )/ 2
+(1 + sqrt(1 - 8 * p$sigma^2/p$K^2) )/ 2
 ```
+Note this is independent of $r$, and increases with $\sigma/K$, the larger the noise $\sigma_g$ as a fraction of the carrying capacity $K$.
+Population sizes can also be inflated relative to their deterministic equilibrium whenever the second derivative is positive (as in the lower equilibrium in the May model of alternative stable states, considered below.)  Stronger nonlinearities (in particular, a larger second derivative at the equilibrium) can drive larger inflationary effects.
 
@@
 
+```{r}
+p <- list(r = .5, K = 2.1, Q = 5, H = .4, sigma = .04, a = 0.27, N = 1e4)
+
+f <- function(x) x * p$r * (1 - x / p$K)
+g <- function(x) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
+
+x <- seq(0,2, length.out = 100)
+df <- data.frame(x, f = f(x), g = g(x))
+
+ggplot(df, aes(x)) + geom_line(aes(y = f)) + geom_line(aes(y = g), col="red")
+
+ggplot(df, aes(x)) + geom_line(aes(y = f-g)) + 
+  geom_hline(aes(yintercept=0), lty=2) +
+  coord_cartesian(xlim = c(0,1.5), ylim=c(-.1,.1))  
+```
+
 ------------
@@
 
-Understanding the origin of sustained oscillations has long been a central question in ecology [@Hastings]
+Understanding the origin of sustained oscillations has long been a central question in ecology [@Hastings1996]
 
@@
 
-df <- p %>% rowwise() %>% do(f(.))
+quasicycle_df <- p %>% rowwise() %>% do(f(.))        
+write_csv(quasicycle_df, "quasicycles.csv")
 
-```
-
-```{r}
-df %>%
+quasicycle_df %>%
   gather(species, pop, -t, -sigma) %>%
@@
 # Stochastic oscillator
+
+May's model
 
@@
 ```{r}
-system.time({
+
 may  <- nimble::nimbleCode({
@@
 set.seed(123)
+simulate(cmodel)
 
-simulate(cmodel)
-df <- tibble(t = seq_along(cmodel$x), x = cmodel$x)
-})
-
-df  %>% ggplot(aes(t,x)) + geom_line()
+tibble(t = seq_along(cmodel$x), x = cmodel$x) %>% write_csv("noisy_switch.csv")
+read_csv("noisy_switch.csv")  %>% ggplot(aes(t,x)) + geom_line()
 ```
@@
 set.seed(123)
+simulate(cmodel)
 
-simulate(cmodel)
-df <- tibble(t = seq_along(cmodel$x), x = cmodel$x)
+tibble(t = seq_along(cmodel$x), x = cmodel$x) %>% write_csv("tipping.csv")
 
-```
-
-```{r}
-df %>% ggplot(aes(t,x)) + geom_line()
+read_csv("tipping.csv") %>% ggplot(aes(t,x)) + geom_line()
 ```

### 2017-10-21 paper.Rmd
WIP

@@
 
+**Inflation**
+
 **Quasi-cycles**
+
+Understanding the origin of sustained oscillations has long been a central question in ecology [@Hastings1996].
+The simple Lotka Volterra model produces only neutrally stable centers, while the more realistic model that introduces carrying capacity for prey as they compete for resources results in a model that produces damped oscillations, eventually returning to a stable
 

### 2017-10-21 appendix/appendix.Rmd
rename

--- 2017-10-21 code/appendix.Rmd
@@
 library(nimble)
+library(hrbrthemes)
 library(tidyverse)
+ggplot2::theme_set(hrbrthemes::theme_ipsum())
 ```
@@
   data.frame(
-  data.frame(r = .1, K = 5,  b = .1, c = .1, d = .1, N = 500),
+  data.frame(r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800),
   data.frame(sigma_x = c(.00001,0.01),  sigma_y = c(.00001,0.01)))
@@
 
-quasicycle_df %>%
+read_csv("quasicycles.csv") %>%
   gather(species, pop, -t, -sigma) %>%
-  ggplot(aes(t, pop, col=species)) + geom_line() + facet_wrap(~sigma)
+  ggplot(aes(t, pop, col=species)) + geom_line() + facet_wrap(~sigma) + scale_color_ipsum()
 ```

### 2017-10-21 paper.Rmd
rename

--- 2017-10-20 paper.Rmd
@@
 ---
+
+```{r include=FALSE, message=FALSE}
+knitr::opts_chunk$set(include = FALSE)
+library(hrbrthemes)
+library(tidyverse)
+ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
+```
 
@@
 
+Stochasticity is most often thought of as something which obscures patterns we are trying to infer.
+This is a simple consequence of modeling: as discussed above, variables of interest to the processes in question are modeled explicitly (e.g.
+change in population density as it depends on the current density), while anything else known to also impact
+
+<!-- blah blah fixed effects vs random effects -->
+
 - deterministic skeletons, 
@@
 
+Understanding the origin of sustained oscillations has long been a central question in ecology [@Hastings1996].
+The simple Lotka Volterra model produces only neutrally stable centers, while the more realistic model that introduces carrying capacity for prey as they compete for resources results in a model that produces only damped oscillations, eventually settling into a stable state (Figure 1a).
+
+```{r}
+read_csv("appendix/quasicycles.csv") %>%
+  gather(species, pop, -t, -sigma) %>%
+  ggplot(aes(t, pop, col=species)) + geom_line() + 
+  facet_wrap(~sigma) + scale_color_ipsum() + 
+  labs(x="Time", 
+       y="Population density",
+       title="Quasi-cycles created by stochastic resonance",
+       subtitle="Left panel: neglible  noise results only in damped oscillations.
+                 Right panel: small noise creates sustained oscillations.")
+```
+
 - Oscillations (parametric oscillator, stochastic resonance)
@@
 [@Blarer1999]
+
+**Inflation**
 
@@
 - Early warning signals
-- maximum entropy methods
 

### 2017-10-21 appendix/appendix.Rmd
WIP

@@
 ```{r}
-knitr::opts_chunk(message=FALSE)
+knitr::opts_chunk$set(message=FALSE)
 ```
@@
 ggplot2::theme_set(hrbrthemes::theme_ipsum())
-```
-
-## Stochastic inflation
-
-```{r}
-
-logistic  <- nimble::nimbleCode({
-
-  x[1] <- x0
-  for(t in 1:(N-1)){
-    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)
-    x[t+1] ~ dnorm(mu[t], sd = sigma)
-  }
-
-})
-
-p <- list(r = .4, K = 1, sigma = .15, N = 500)
-
-model <- 
-nimbleModel(logistic,
-            constants = p,
-            inits = list(x0 = 1))
-cmodel <- compileNimble(model)
-set.seed(123)
-
-df <- map_dfr(1:10000, 
-              function(rep){
-                simulate(cmodel)
-                data.frame(t = seq_along(cmodel$x), x = cmodel$x)
-              },
-              .id = "rep")
-
-df %>% 
-  filter(x > 0) %>% 
-  group_by(t) %>%
-  summarise(ave = mean(x), sd = sd(x)) %>%
-  write_csv("inflation.csv")
-
-```
-
-Analytical predicted population size, as a fraction of $K$, is:
-
-```{r}
-(1 + sqrt(1 - 8 * p$sigma^2/p$K^2) )/ 2
-```
-Note this is independent of $r$, and increases with $\sigma/K$, the larger the noise $\sigma_g$ as a fraction of the carrying capacity $K$.
-Population sizes can also be inflated relative to their deterministic equilibrium whenever the second derivative is positive (as in the lower equilibrium in the May model of alternative stable states, considered below.)  Stronger nonlinearities (in particular, a larger second derivative at the equilibrium) can drive larger inflationary effects.
-
-```{r}
-df %>% filter(t > 20, x > 0) %>%  summarize(mean(x))
-```
-
-```{r}
-
-df %>% 
-  filter(x > 0) %>% 
-  group_by(t) %>%
-  summarise(ave = mean(x), sd = sd(x)) %>%
-  gather(moment, value, -t) %>%
-  ggplot(aes(t,value, col=moment)) + geom_line()
-
-```
-
-```{r}
-p <- list(r = .5, K = 2.1, Q = 5, H = .4, sigma = .04, a = 0.27, N = 1e4)
-
-f <- function(x) x * p$r * (1 - x / p$K)
-g <- function(x) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
-
-x <- seq(0,2, length.out = 100)
-df <- data.frame(x, f = f(x), g = g(x))
-
-ggplot(df, aes(x)) + geom_line(aes(y = f)) + geom_line(aes(y = g), col="red")
-
-ggplot(df, aes(x)) + geom_line(aes(y = f-g)) + 
-  geom_hline(aes(yintercept=0), lty=2) +
-  coord_cartesian(xlim = c(0,1.5), ylim=c(-.1,.1))  
 ```
@@
 
+## Stochastic inflation
+
+Analytical predicted population size, as a fraction of $K$, is $\tfrac{1}{2} + \tfrac{1}{2} \sqrt{1- 8\sigma^2/K^2}$.
+Note this is independent of $r$, and increases with $\sigma/K$, the larger the noise $\sigma_g$ as a fraction of the carrying capacity $K$.
+Population sizes can also be inflated relative to their deterministic equilibrium whenever the second derivative is positive (as in the lower equilibrium in the May model of alternative stable states, considered below.)  Stronger nonlinearities (in particular, a larger second derivative at the equilibrium) can drive larger inflationary effects.
+
+```{r}
+p <- list(r = .4, K = 1, sigma = .15, N = 500)
+equib <- (1 + sqrt(1 - 8 * p$sigma^2/p$K^2) )/ 2
+
+logistic  <- nimble::nimbleCode({
+
+  x[1] <- x0
+  for(t in 1:(N-1)){
+    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)
+    x[t+1] ~ dnorm(mu[t], sd = sigma)
+  }
+
+})
+
+model <- 
+nimbleModel(logistic,
+            constants = p,
+            inits = list(x0 = 1))
+cmodel <- compileNimble(model)
+set.seed(123)
+
+df <- map_dfr(1:10000, 
+              function(rep){
+                simulate(cmodel)
+                data.frame(t = seq_along(cmodel$x), x = cmodel$x)
+              },
+              .id = "rep")
+
+df %>% 
+  filter(x > 0) %>% 
+  group_by(t) %>%
+  summarise(ave = mean(x), sd = sd(x)) %>%
+  mutate(equib = equib) %>%
+  write_csv("inflation.csv")
+
+```
+
+```{r}
+
+read_csv("inflation.csv") %>% 
+  ggplot(aes(t,value, col=moment)) + geom_line() + 
+  geom_hline(aes(yintercept = equib))
+
+```
+
+```{r}
+p <- list(r = .5, K = 2.1, Q = 5, H = .4, sigma = .04, a = 0.27, N = 1e4)
+
+f <- function(x) x * p$r * (1 - x / p$K)
+g <- function(x) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
+
+x <- seq(0,2, length.out = 100)
+df <- data.frame(x, f = f(x), g = g(x))
+
+ggplot(df, aes(x)) + geom_line(aes(y = f)) + geom_line(aes(y = g), col="red")
+
+ggplot(df, aes(x)) + geom_line(aes(y = f-g)) + 
+  geom_hline(aes(yintercept=0), lty=2) +
+  coord_cartesian(xlim = c(0,1.5), ylim=c(-.1,.1))  
+```
+
+----
+
 # Tipping points
-
-Needs a model with non-zero alternate state!
 
@@
 
-tibble(t = seq_along(cmodel$x), x = cmodel$x) %>% write_csv("tipping.csv")
+tibble(t = seq_along(cmodel$x), x = cmodel$x) %>% 
+  mutate(var = RcppRoll::roll_var(x, n = 5)) %>%
+  write_csv("tipping.csv")
 

### 2017-10-21 cover-letter.Rmd
WIP

---
title: "Cover Letter"
author: "Carl Boettiger"
date: "10/21/2017"
output: pdf_document
---

- Tim Coulson

Noisy phenomena in population biology.

<!--

We hope that these papers will combine a mix of synthetic reviews of the current state of an area, as well as insights into how the field has developed during the last twenty years, discussion of current important questions, and developments and insights that would significantly advance the field. Although certainly not necessary, if there are any past Ecology Letters papers that have had a profound impact on the development of an area – be it beneficial or not – a brief discussion of that impact would be appropriate. We also hope that articles will be written in a thoughtful style, but also provocative if appropriate.

-->

### 2017-10-21 paper.Rmd
WIP

@@
 ```
+
+<!--
+
+We hope that these papers will combine a mix of synthetic reviews of the current state of an area, as well as insights into how the field has developed during the last twenty years, discussion of current important questions, and developments and insights that would significantly advance the field.
+Although certainly not necessary, if there are any past Ecology Letters papers that have had a profound impact on the development of an area – be it beneficial or not – a brief discussion of that impact would be appropriate.
+We also hope that articles will be written in a thoughtful style, but also provocative if appropriate.
+
+Key questions
+
+Key advances
+
+Shift from noise as nuiance to noise as opportunity
+Stochastic phenomena outside of theory -- early warning signals
+
+Need for greater tanslation
+
+Population-level vs ABM vs IBM
+
+Discuss out of scope:  
+ - behavior
+ - evolutionary dynamics examples??
+
+-->
 
@@
 
+$$x_{t+1} = x_t + x_t r \left(1 - \frac{x_t}{K}\right) - b x_t  y_t + \epsilon_t$$
+$$y_{t+1} = y_t + c  x_t  y_t - d  y_t+ \xi_t $$
+
 ```{r}
@@
        title="Quasi-cycles created by stochastic resonance",
-       subtitle="Left panel: neglible  noise results only in damped oscillations.
-                 Right panel: small noise creates sustained oscillations.")
+       caption="Left panel: neglible  noise results only in damped oscillations.\nRight panel: small noise creates sustained oscillations.")
 ```
 
-- Oscillations (parametric oscillator, stochastic resonance)
+- Disease models
+
+- Power spectrum
 
@@
 
+**Stochastic switching**
+
+Stochastic fluctuations can also drive spontaneous shifts between alternative stable states.
+
+```{r}
+read_csv("appendix/noisy_switch.csv")  %>% ggplot(aes(t,x)) + 
+  geom_line() + 
+  labs(title="Stochastic switching", 
+       subtitle = "Noise drives transitions between alternative stable states")
+```
+
+- transition probabilities, waiting times
+
 **Inflation**
 
-- population inflation / deflation
+Stochasticity can
+
+```{r}
+read_csv("appendix/inflation.csv") %>% 
+  ggplot(aes(t,value, col=moment)) + geom_line() + 
+  geom_hline(aes(yintercept = equib))
+```
+
+**Bet hedging**
+
+**Further examples**
 

### 2017-10-23 appendix/appendix.Rmd
WIP

@@
 title: "Appendix"
-output: html_notebook
+output: pdf_document
 ---
@@
 library(hrbrthemes)
+library(ggthemes)
 library(tidyverse)
@@
 write_csv(quasicycle_df, "quasicycles.csv")
+```
 
+```{r}
 read_csv("quasicycles.csv") %>%
   gather(species, pop, -t, -sigma) %>%
-  ggplot(aes(t, pop, col=species)) + geom_line() + facet_wrap(~sigma) + scale_color_ipsum()
+  ggplot(aes(t, pop, col=species)) + geom_line() + facet_wrap(~sigma) + scale_color_ptol()
 ```
@@
               .id = "rep")
+```
 
+```{r}
 df %>% 
@@
 ```{r}
-
 read_csv("inflation.csv") %>% 
-  ggplot(aes(t,value, col=moment)) + geom_line() + 
-  geom_hline(aes(yintercept = equib))
+  gather(statistic, value, -t) %>%
+  ggplot(aes(t,value, col=statistic, lty=statistic)) + 
+ # ggplot(aes(t,ave)) +
+  geom_line(lwd=1) + 
+  scale_color_ptol() + 
+  coord_cartesian(xlim=c(100,500), ylim=c(1.2, .5))
 
@@
 
+### Configuration for Dai model
+
+Numerically determine / illustrate tipping point location using small noise:
+
 ```{r}
+max_days <- 1000
 
-tip  <- nimble::nimbleCode({
+z <- y <- x <- numeric(max_days)
+z[1] <- y[1] <- x[1] <- 7e4 # 1.7e5
+set.seed(123)
+for(day in 1:(max_days-1)){
+  x[day+1] <- dai(x[day], epsilon = rnorm(1,0, 0.001), DF = 1790)
+  y[day+1] <- dai(y[day], epsilon = rnorm(1,0, 0.001), DF=1799)
+  z[day+1] <- dai(z[day], epsilon = rnorm(1,0, 0.001), DF = 1800)
 
-  x[1] <- x0
-  for(t in 1:(N-1)){
-    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K) * (x[t] - A) / K - h[t]
-    y[t+1] ~ dnorm(mu[t], sd = sigma)
-    x[t+1] <- max(y[t+1],0)
 }
 
-})
+df <- data.frame(t = 1:max_days, x = x, y = y, z = z) %>%
+  gather(series, pop, -t)
 
-p <- list(r = .2, K = 1, A = 0.2, sigma = .01, N = 500, h = seq(0,.03,length=500))
+ggplot(df, aes(t, pop, col=series)) + 
+  geom_line()
 
-model <- nimbleModel(tip,constants = p, inits = list(x0 = 1))
-cmodel <- compileNimble(model)
-set.seed(123)
-simulate(cmodel)
-
-tibble(t = seq_along(cmodel$x), x = cmodel$x) %>% 
-  mutate(var = RcppRoll::roll_var(x, n = 5)) %>%
-  write_csv("tipping.csv")
-
-read_csv("tipping.csv") %>% ggplot(aes(t,x)) + geom_line()
 ```
 
-# Turing patterns?
-
-------------
+$$\rho = \frac{1}{n-1} \frac{\sum_{i = 1}^n \left( x_{i,t} - \textrm{E}(x_t) \right)\left( x_{i,t+1} - \textrm{E}(x_{t+1})  \right)}{\sigma_{x_{t}}  \sigma_{x_{t+1}}}$$
 
 ```{r}
-data(lynx)
-lynx <- tibble(year = 1821:1934, lynx = as.numeric(lynx))
+roll_acor <- function(x, lag = 1, ...){
+  x_t1 = lag(x, n = lag)  
+  mu_t = roll_mean(x, ...)
+  mu_t1 = roll_mean(x_t1, ...)
+  s_t = roll_sd(x, ...)
+  s_t1 = roll_sd(x_t1, ...)
+  acor = roll_mean( (x - mu_t) * (x_t1 - mu_t1) / (s_t * s_t1),...)
+  acor
+}
+```
 
-lynx %>%
-  ggplot(aes(year,lynx)) + geom_point() + geom_line()
+## Simulation of Dai model:
+
+```{r}
+# Stepwise changes
+DF <- as.numeric(sapply(seq(0, 2000, length=9), rep, 40))
+
+# continuous linear increase
+DF <- seq(1, 2000, length=2000)
+tip_time <- 1800
+max_days <- length(DF)
+y <- numeric(max_days)
+
+set.seed(1111)
+y[1] <- 1.76e5
+for(day in 1:(max_days-1)){
+  y[day+1] <- dai(y[day], DF = DF[day])
+}
+
+tip_df <- tibble(t = seq_along(y), x = y)
+
+tip_df %>% 
+  mutate(autocorrelation = roll_acor(x, n = 100, fill=NA),
+         variance = roll_var(x, n = 100, fill=NA),
+         tip_time = tip_time) %>%
+  write_csv("tipping.csv")
 ```
+
+```{r}
+tipping <- read_csv("appendix/tipping.csv")
+
+p1 <- tipping %>%
+  select(-x) %>%
+  gather(series, value, -t, -tip_time) %>%
+  ggplot(aes(t, value)) + geom_line() + 
+  geom_vline(aes(xintercept = tip_time), col="red", lty=2) +
+  facet_wrap(~series, ncol = 2, scales="free_y") + theme_ipsum() + 
+  labs(title="A: Early warning signs")
+
+p2 <- tipping %>% 
+  select(t, population = x, tip_time) %>%
+  ggplot(aes(t, population)) + geom_line() + 
+  geom_vline(aes(xintercept = tip_time), col="red", lty=2) + 
+  scale_y_log10() + 
+  theme_ipsum()+ labs(title="B: Population density",
+                      caption = "Vertical red dashed line indicates tipping point location")
+
+library(cowplot)
+ggdraw() +
+  draw_plot(p2, x = 0, y = 0, width = 1, height = 0.5) +
+  draw_plot(p1, x = 0, y = .4, width = 1, height = 0.5)
+
+```

### 2017-10-23 paper.Rmd
WIP

@@
   ggplot(aes(t, pop, col=species)) + geom_line() + 
-  facet_wrap(~sigma) + scale_color_ipsum() + 
+  facet_wrap(~sigma) + scale_color_ptol() + 
   labs(x="Time", 
@@
 
-```{r}
-read_csv("appendix/inflation.csv") %>% 
-  ggplot(aes(t,value, col=moment)) + geom_line() + 
-  geom_hline(aes(yintercept = equib))
-```
+<!-- two concurrent flat lines makes for a poor figure -->
 
@@
 
+Individual heterogeneity: spatial spread? persistence?
+
 - Species coexistence  ([@Hart2016])
-
-Figure:
 
@@
 - Measuring stability
+
 - Early warning signals
 
-Figure:
+<!-- @Holling1996 is responsible for the introduction of the unfortunate term "Engineering Resilience" to mean precisely the same thing as mathematicians and theoretical ecologists have refered to as 'stability' for decades.
+Further confusion has been sown by the introduction of the energy-potential diagram into ecology, where the depth of the potential well is not a local property-->
+
+```{r}
+tipping <- read_csv("appendix/tipping.csv")
+
+p1 <- tipping %>%
+  select(-x) %>%
+  gather(series, value, -t, -tip_time) %>%
+  ggplot(aes(t, value)) + geom_line() + 
+  geom_vline(aes(xintercept = tip_time), col="red", lty=2) +
+  facet_wrap(~series, ncol = 2, scales="free_y") + theme_ipsum() + 
+  labs(title="A: Early warning signs")
+
+p2 <- tipping %>% 
+  select(t, population = x, tip_time) %>%
+  ggplot(aes(t, population)) + geom_line() + 
+  geom_vline(aes(xintercept = tip_time), col="red", lty=2) + 
+  scale_y_log10() + 
+  theme_ipsum()+ labs(title="B: Population density",
+                      caption = "Vertical red dashed line indicates tipping point location")
+
+library(cowplot)
+ggdraw() +
+  draw_plot(p2, x = 0, y = 0, width = 1, height = 0.5) +
+  draw_plot(p1, x = 0, y = .4, width = 1, height = 0.5)
+```
+
+Challenges remain in distinguishing these patterns from other causes [@Boettiger2013a] which will ultimately require more domain-specific information  [@Boettiger2013b].
+
+<!-- FIXME unclear, rambling!-->
+It is important to realize that such approaches can only ever measure local properties such as the return time in response to local perturbation (historically termed *stability*; which @Holling1996 landmark chapter unforunately rebrands as "engineering resilience") and not global properties such as the size of the attractor basin / distance to the nearest tipping point (*ecological resilience*, or simply *resilience*).
+This is illustrated rather vividly in @Dai2015, which illustrates how a system can lose resilience before losing stability, resulting in a scenario that provides less opportunity for early warning.
 

### 2017-10-23 paper.Rmd
WIP

@@
 title: "From noise to knowledge: how randomness generates novel phenomena and information"
-author: "Carl Boettiger"
+author:
+  - name: "Carl Boettiger"
+    affiliation: a, 1
+address:
+  - code: a
+    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
+
+abstract: |
+  In this paper, I provide a brief synthesis of ecological research in stochastic phenomena.
+Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath.
+Yet despite this well-earned reputation, noise is often interesting in its own right: noise can induce novel phenomena that could not be understood from some underlying determinstic model alone.
+Nor is all noise the same, and close examination of differences in in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.
+While an exhaustive review is beyond the scope of this piece, I endeavor to highlight some of the most interesting examples of both noise-induced phenomena and noise-derived information, and point the way towards yet uncharted territory.
+
+keywords:
+  - Stochasticity
+  - Demographic noise
+  - Environmental noise
+  - Colored noise
+  - quasi-cycles
+
+lineno: true
+
+bibliography: refs.bib
+output: rticles::elsevier_article
 ---
@@
 ```{r include=FALSE, message=FALSE}
-knitr::opts_chunk$set(include = FALSE)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
 library(hrbrthemes)
+library(ggthemes)
 library(tidyverse)
-ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
+#ggplot2::theme_set(hrbrthemes::theme_ipsum())
 ```
@@
 
-In this paper, I provide a brief synthesis of ecological research in stochastic phenomena.
-Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath.
-Yet despite this well-earned reputation, noise is often interesting in its own right: noise can induce novel phenomena that could not be understood from some underlying determinstic model alone.
-Nor is all noise the same, and close examination of differences in in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.
-While an exhaustive review is beyond the scope of this piece, I endeavor to highlight some of the most interesting examples of both noise-induced phenomena and noise-derived information, and point the way towards yet uncharted territory.
-
 ## What is noise?
@@
 
+![Figure 1: demographic stochasticity arises from discrete transitions  between states in in a Markov process.
+Here, a simple birth-death process has transition rates determined by the current population size, _n_.](markov.png)
+
 [@Black2012]
@@
 Colored noise e.g. [@Marshall2015]
+
+<!-- Identify as beyond the scope of this paper -->
 
@@
        y="Population density",
-       title="Quasi-cycles created by stochastic resonance",
+       title="Stochasticicity induces quasi-cycles",
        caption="Left panel: neglible  noise results only in damped oscillations.\nRight panel: small noise creates sustained oscillations.")
 ```
-
-- Disease models
 
@@
 <!-- two concurrent flat lines makes for a poor figure -->
+
+**Persistence and coexistence**
+
+One of the most fundamental aspects of stochasticty are the implications for long-term persistence of any given population and the stable coexistence of interacting species.
 
@@
   geom_vline(aes(xintercept = tip_time), col="red", lty=2) +
-  facet_wrap(~series, ncol = 2, scales="free_y") + theme_ipsum() + 
+  facet_wrap(~series, ncol = 2, scales="free_y") +
+  #theme_ipsum() +
   labs(title="A: Early warning signs")
@@
   scale_y_log10() + 
-  theme_ipsum()+ labs(title="B: Population density",
+  #theme_ipsum() + 
+  labs(title="B: Population density",
        caption = "Vertical red dashed line indicates tipping point location")

### 2017-10-24 paper.Rmd
WIP

@@
 knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
-library(hrbrthemes)
 library(ggthemes)
 library(tidyverse)
-#ggplot2::theme_set(hrbrthemes::theme_ipsum())
+library(cowplot)
+#library(hrbrthemes)
+#extrafont::loadfonts()
+#ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
+#hrbrthemes::import_roboto_condensed()
+
+ggplot2::theme_set(theme_cowplot())
 ```
@@
 Chaos can a look a lot like noise: indeed, it can be observationally indistinguishable from a stochastic process [@Werndl2009].
-Despite substantial interest in previous decades and the relative ease of encountering chaos in simple ecological models (particularly in differences equations), and convincing reconstructions of chaotic dynamics in laboratory manipulations [@Constantino1997] the prevelance of chaotic dynamics as a driving factor in most ecological systems remains unclear.
+Despite substantial interest in previous decades and the relative ease of encountering chaos in simple ecological models (particularly in differences equations), and convincing reconstructions of chaotic dynamics in laboratory manipulations [@Costantino1997] the prevelance of chaotic dynamics as a driving factor in most ecological systems remains unclear.
 Meanwhile, the undeniable influences of
@@
 read_csv("appendix/quasicycles.csv") %>%
-  gather(species, pop, -t, -sigma) %>%
-  ggplot(aes(t, pop, col=species)) + geom_line() + 
-  facet_wrap(~sigma) + scale_color_ptol() + 
-  labs(x="Time", 
-       y="Population density",
-       title="Stochasticicity induces quasi-cycles",
-       caption="Left panel: neglible  noise results only in damped oscillations.\nRight panel: small noise creates sustained oscillations.")
+  gather(species, population, -t, -sigma) %>%
+  rename(time = t) %>%
+  ggplot(aes(time, population, col = species)) + geom_line() + 
+  facet_wrap(~ sigma) + scale_color_ptol() #+ 
+#  labs(x="Time", 
+#       y="Population density",
+#       title="Noise induces quasi-cycles",
+#       caption="Left panel: neglible  noise results only in damped oscillations.\nRight panel: small noise creates sustained oscillations.")
+#ggsave("figures/quasicycles.pdf")
 ```
@@
 read_csv("appendix/noisy_switch.csv")  %>% ggplot(aes(t,x)) + 
-  geom_line() + 
-  labs(title="Stochastic switching", 
-       subtitle = "Noise drives transitions between alternative stable states")
+  geom_line() #+ 
+  #labs(title="Stochastic switching", 
+  #     subtitle = "Noise drives transitions between alternative stable states")
+#ggsave("figures/noisy_switch.pdf")
+
 ```
@@
   geom_vline(aes(xintercept = tip_time), col="red", lty=2) +
-  facet_wrap(~series, ncol = 2, scales="free_y") +
-  #theme_ipsum() +
-  labs(title="A: Early warning signs")
+  facet_wrap(~series, ncol = 2, scales="free_y") # +
+#  theme_ipsum_rc() +
+#  labs(title="A: Early warning signs")
 
@@
   ggplot(aes(t, population)) + geom_line() + 
-  geom_vline(aes(xintercept = tip_time), col="red", lty=2) + 
-  scale_y_log10() + 
-  #theme_ipsum() + 
-  labs(title="B: Population density",
-       caption = "Vertical red dashed line indicates tipping point location")
+  geom_vline(aes(xintercept = tip_time), col="red", lty=2) #+ 
+#  scale_y_log10() # + 
+#  theme_ipsum_rc() + 
+#  labs(title="B: Population density",
+#       caption = "Vertical red dashed line indicates tipping point location")
 
@@
   draw_plot(p2, x = 0, y = 0, width = 1, height = 0.5) +
-  draw_plot(p1, x = 0, y = .4, width = 1, height = 0.5)
+  draw_plot(p1, x = 0, y = .5, width = 1, height = 0.5)
+ggsave("figures/tipping.pdf")
 ```
@@
 ## Conclusions
+
+## References

### 2017-10-24 paper.Rmd
passable rendering of fonts on linux, though looks like a different font is getting swapped in.

@@
 library(cowplot)
-#library(hrbrthemes)
-#extrafont::loadfonts()
-#ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
-#hrbrthemes::import_roboto_condensed()
-
-ggplot2::theme_set(theme_cowplot())
+library(hrbrthemes)
+hrbrthemes::import_roboto_condensed()
+extrafont::loadfonts()
+ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
+#ggplot2::theme_set(theme_cowplot())
 ```
@@
   draw_plot(p1, x = 0, y = .5, width = 1, height = 0.5)
-ggsave("figures/tipping.pdf")
+#ggsave("figures/tipping.pdf")
 ```

### 2017-10-24 appendix/appendix.Rmd
WIP

@@
 library(ggthemes)
+library(RcppRoll)
+library(regimeshifts) ## devtools::install_github("cboettig/regimeshifts")
 library(tidyverse)
@@
 ```{r}
-tipping <- read_csv("appendix/tipping.csv")
+tipping <- read_csv("tipping.csv")
 

### 2017-10-24 paper.Rmd

@@
 library(hrbrthemes)
+extrafont::loadfonts()
+extrafont::font_import()
 hrbrthemes::import_roboto_condensed()
-extrafont::loadfonts()
 ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
-#ggplot2::theme_set(theme_cowplot())
+#ggplot2::theme_set(theme_bw())
 ```
@@
 
+Noise -> informer
+
 Shift from noise as nuiance to noise as opportunity
@@
  - evolutionary dynamics examples??
+
+Bialek's rules
+1. Noise isn't negligible: photon counting, chemotaxis, 
+2. No fine tuning: developmental patterning? ion channels/ stochastic switches? 
+3. Efficient representation: entropy, information flow
 
@@
 The term "noise" itself reflects this emergent rather than fundamental origin: the combination of hundreds of voices and sounds each individually distinct blends together into a background of noise.
+
+## Noise the nuisance: Noise can obscure patterns
+
+Stochasticity is most often thought of as something which obscures patterns we are trying to infer.
+This is a simple consequence of modeling: as discussed above, variables of interest to the processes in question are modeled explicitly (e.g.
+change in population density as it depends on the current density), everything else is just noise.
+Most statistical inference in ecology is built solely to deal with this problem: pulling back the veil of noise to reveal processes of interest.
+Unfortunately, given the high levels of variation and relatively limited data typical in ecological analyses, this can be a significant challenge, and even as statistical and computational advances provide more powerful and accessible inference tools, the task only gets harder.
+For instance, few ecological concepts are more central than the notion of density dependence: that competition in some form or other should limit a species ability to grow without bound.
+Yet establishing statistical evidence for density dependence is remarkably difficult, as illustrated by a landmark study of @Knape2011.
+Using time series from the Global Population Dynamics Database (GPDD) @Knape2011 show that once one accounts for the potential for measurement error in addition to environmental stochasticity -- an unavoidable source of noise in most ecological data -- it becomes difficult to find evidence for even the simplest models of density dependence in the majority of avaialble time series.
+Faced with such uncertainty, it is tempting to throw up our hands and wait until some prophesized deluge of streaming microsensors stapled to every organism provide enough signal to beat back the noise.
+Fortunately, this has not been the path taken.
+
+<!-- Should we make explicit or only implicit reference to Ecology Letters? -->
+
+<!-- okay, claiming this for the last 2 decades might be overly provocative?-->
+
+Over the past twenty years, ecological theory, modelling, and methods have increasingly elucidated the role of noise not only to *obscure* patterns but to *create* them.
+This period marked a transistion away from the intense focus on simple models which exhibited deterministic chaos that had enchanted much of theoretical ecology in the decade prior in favor of more complex and stochastic models.
+Of particular interest has been those conceptual advances that have bridged the divide from theoretical to empirical interest -- not merely the introduction of "real data" in a theoretical paper, but rather the concepts that take on a life of their own in the larger research community.
+
+Our first stop will be into research into the origins of noise.
+This will not only let us set up definitions for terms and concepts we will encounter throughout, but also introduce some of the fundamental mathematical and computational tooling needed in our tour of noisy phenemena.
+From origins of noise, we will then explore some of the phenomena that noise can create.
+These examples are central to the conceptual transistion away from viewing noise only as a nusiance that obscures patterns and realize the rich set of patterns and phenomena that can emerge.
+Our tour will be by no means exhaustive, but as with chaos before it, theoretical allure does not always reflect ecological releveance.
+Instead, we will pivot to our third
+
+<!-- Origins -->
+
+<!-- Phenomena -->
+
+<!-- Applications -->
+
+<!-- Connection to empiricists.  Not the same thing as connection to data. -->
+
+## Origins of noise
 
@@
 
-### Not noise
-
 In addition to these sources of variation, there are additional sources of variation or uncertainty in a model which we will not consider for the purpose of this review which, while related, operate in a fundamentally different way that does not give rise to the same phenomena and information.
@@
 uncertainty does not impact the actual state of our system, and thus does not give rise to any
-additional phenomena or information about that process.  So while this is indeed a source of noise,
-it falls outside our scope here.
+additional phenomena or information about that process.
+
+### Not noise
 
@@
 
-**Chaos**: Deterministic chaos [@Desharnais2001] is another mechanism of introducing significant variation into a simple
-model.
+**Chaos**: Deterministic chaos [@Desharnais2001] is another mechanism of introducing significant variation into a simple model.
 Chaos can a look a lot like noise: indeed, it can be observationally indistinguishable from a stochastic process [@Werndl2009].
@@
 Figure: schematic showing types of noise
-
-## Noise the nuisance: Noise can obscure patterns
-
-Stochasticity is most often thought of as something which obscures patterns we are trying to infer.
-This is a simple consequence of modeling: as discussed above, variables of interest to the processes in question are modeled explicitly (e.g.
-change in population density as it depends on the current density), while anything else known to also impact
 
@@
 
-Figure: Lynx-hare data
-
-## Noise the creator: Noise can induce novel phenonema
+## Noise the creator: Noise can induce novel phenomena
 
@@
 
-- Early warning signals
+**Early warning signals.**  The theory of early warning signals, or critical slowing down (csd) [@Wissel1984; @Scheffer2009] is an instructive example
 

### 2017-10-24 paper.Rmd
WIP

@@
 
+layout: preprint
 abstract: |
@@
 ```{r include=FALSE, message=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="CairoPDF", fig.width=3.5, fig.height=3.5)
 library(ggthemes)
@@
 extrafont::loadfonts()
-extrafont::font_import()
-hrbrthemes::import_roboto_condensed()
+#extrafont::font_import()                    # One-time use to set up fonts
+#hrbrthemes::import_roboto_condensed()       # One-time use to set up fonts
 ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
-#ggplot2::theme_set(theme_bw())
+#ggplot2::theme_set(theme_bw()) # simpler theme if you do not have these fonts
 ```
@@
 
-Key questions
+- Key questions
+- Key advances
 
-Key advances
+Moving beyond "everything is important" [@Bjornstad2001; @Kingsland1985]
+
+@Bjornstad2001 future: more MCMC methods. More mechanistic models.  dealing with unobserved dynamics
 
@@
 
-# Outline
+<!-- Should we make explicit or only implicit reference to Ecology Letters? -->
 
-## What is noise?
-
-<!-- Our state variables don't "exist" except as averages of a lower level process -->
-
-Where does noise come from?
-It is important to bear in mind that whenever stochasticity is introduced into a model, it is only as a simplification or summary of the effects of a large number of other processes, often occuring at different scales.
-Simply put, *noise is all that we leave out of our model*.
-While ecologists will sometimes assert that stochasticity is inherent: birth and death are inherently stochastic events; or the environment is inherently stochastic, these are only assertions at a particular scale.
-Obviously events such as births and deaths are themselves the results of mechanistic processes, just as fluctuations in rainfall and temperature are the result of deterministic laws of physics and not fundamentally random forces.
-In all of science, only quantum mechanics claims a fundamentally probabilistic view of events, and then at such scales that have no relevance to ecology.
-The term "noise" itself reflects this emergent rather than fundamental origin: the combination of hundreds of voices and sounds each individually distinct blends together into a background of noise.
-
-## Noise the nuisance: Noise can obscure patterns
+# Introduction: Noise the nuisance
 
@@
 
-<!-- Should we make explicit or only implicit reference to Ecology Letters? -->
+<!-- okay, claiming this for the last 2 decades might be overly provocative?-->
+Over the past twenty years, ecological theory, modelling, and methods have increasingly elucidated the role of noise not only to *obscure* patterns but to *create* them.
+This period marked a transistion away from the intense focus on simple models which exhibited deterministic chaos that had enchanted much of theoretical ecology in the previous two decades [@Bjornstad2001] in favor of more complex and stochastic models.
+Of particular interest has been those conceptual advances that have bridged the divide from theoretical to empirical interest -- not merely the introduction of "real data" in a theoretical paper, but rather the concepts that take on a life of their own in the larger research community.
+Consequently, this paper seeks to be a bridge between two audiences: theoretical researchers already expert in noisy phenomena but in interested in reaching wider empirical application, and also primarily empirical researchers seeking an entrance into noisy phenomena.
 
-<!-- okay, claiming this for the last 2 decades might be overly provocative?-->
+<!-- Regarding simulation examples -->
+In this spirit, each of the examples illustrated here is coupled with consise and efficent R code in the appendix to both reproduce and further explore the simulations.
+While examples of noisy phenomena in natural systems are also crucial, matching the pattern does not necessarily mean we have also matched the proccess behind the data.
+The examples cited from the literature here offer some of the best empirical evidence of these phenomena, but I encourage readers to also experiment with the example code provided.
+Simulations permit poking and prodding of empirical investigation quickly and with little mathematical formalism in the way.
+In the code provided, most stochastic models are expressed in the BUGS language, which may be more familiar to empirical readers than corresponding mathematical formulae.
+This also permits both efficient simulation and potential estimation of parameters given sample data using the R package, NIMBLE [@DeValpine2017].
+A copy of this appendix is maintained at <https://github.com/cboettig/noise-phenomena>, and bug reports, suggestions or help requests are welcome through the issue tracker.
+A second appendix provides more mathematical details for interested readers.
+With these in place, we are ready to begin our tour through three major areas of noisy phenomena: Origins of noise, emergent phenomena, and noise-driven inference.
 
-Over the past twenty years, ecological theory, modelling, and methods have increasingly elucidated the role of noise not only to *obscure* patterns but to *create* them.
-This period marked a transistion away from the intense focus on simple models which exhibited deterministic chaos that had enchanted much of theoretical ecology in the decade prior in favor of more complex and stochastic models.
-Of particular interest has been those conceptual advances that have bridged the divide from theoretical to empirical interest -- not merely the introduction of "real data" in a theoretical paper, but rather the concepts that take on a life of their own in the larger research community.
-
+<!-- MAP -->
 Our first stop will be into research into the origins of noise.
@@
 Our tour will be by no means exhaustive, but as with chaos before it, theoretical allure does not always reflect ecological releveance.
-Instead, we will pivot to our third
+Having looked at some of the phenomena that can be created by noise, our third stop will try to run this in reverse: using noise to infer underyling phenomena.
+Examples here are rarer but potentially the most rewarding, bridging the gap from theoretical construct to empirical investigation.
+The most compelling examples have come from ways in which noisy pheneomena are used to infer stability and forecast changes in resilience of populations or ecosystems.
+Though not without weaknesses or missteps, the resulting dialog between theoretical and empirical aspects have made this particularly a rich area of research.
+We end our tour by considering what features best enable the translation and ecological relevance of noisy phenomena.
 
-<!-- Origins -->
+# Origins of noise
 
-<!-- Phenomena -->
-
-<!-- Applications -->
-
-<!-- Connection to empiricists.  Not the same thing as connection to data. -->
-
-## Origins of noise
-
-## Sources of noise
+What is noise?
+Where does it come from?
+It is important to bear in mind that whenever stochasticity is introduced into a model, it is only as a simplification or summary of the effects of a large number of other processes, often occuring at different scales.
+Simply put, *noise is all that we leave out of our model*.
+While ecologists will sometimes assert that stochasticity is inherent: birth and death are inherently stochastic events; or the environment is inherently stochastic, these are only assertions at a particular scale.
+Obviously events such as births and deaths are themselves the results of mechanistic processes, just as fluctuations in rainfall and temperature are the result of deterministic laws of physics and not fundamentally random forces.
+In all of science, only quantum mechanics claims a fundamentally probabilistic view of events, and then at such scales that have no relevance to ecology.
+The term "noise" itself reflects this emergent rather than fundamental origin: the combination of hundreds of voices and sounds each individually distinct blends together into a background of noise.
+<!-- Our state variables don't "exist" except as averages of a lower level process -->
 
@@
 
-<!-- Demographic noise scaling with population size -->
-
-<!-- Gillespie algorithm -->
+<!-- Gillespie algorithm; Population level (SDE) vs agent-based vs individual  -->
 
@@
 
-[@Black2012]
+<!-- Demographic noise scaling away with population size, but not so for environmental noise -->
 
-Colored noise e.g. [@Marshall2015]
+[@Bjornstad2001] _Science_ Review of noisy fluctuations
+
+[@Melbourne2008] Importance of demographic stochasticity and its interaction with environmental stochasticity in driving extinction risk. 
+[@Bonsall2004] Markov chain environmental and demographic noise predator-prey metapopulation model.
+
+Colored noise:
+- [@Ripa1999]
+- [@Marshall2015]
 
@@
 
-**Individual heterogeneity**
+<!-- non-temporal stochasticity -->
 
-[@Hart2016]
-
-**Spatial variability**
-
-In addition to these sources of variation, there are additional sources of variation or uncertainty in a model which we will not consider for the purpose of this review which, while related, operate in a fundamentally different way that does not give rise to the same phenomena and information.
+**Other forms of stochasticity**.
+Stochasticity frequently enters our models in other ways as well.
+For instance, most population-level models treat all individuals within the population, or at least within a particular age or stage class, as identical.
+*Individual heterogeneity* in relevant traits (growth rates, dispersal ability) can be represented probabilistically as a distribution rather than a fixed value, making it function as another form of random varation in a model.
+*Spatial heterogeneity* can likewise be reflected in stochastic parameters of a spatially implicit model, or as spatially varying environmental stochasticity in a spatially explicit model.
+As with other forms of noise [@Melbourne2008; @Bonsall2004], interaction between individual or spatial variation and other sources of noise can drive patterns including the potential for persistence, coexistence, and dispersal of species [@Schreiber2010; @Hart2016; @Schreiber2017], but are beyond the scope of this paper.
 
@@
 
-- deterministic skeletons, 
-- signal-to-noise ratio, 
-- model fitting
-
-## Noise the creator: Noise can induce novel phenomena
+# Noise the creator: Noise can induce novel phenomena
 
@@
 
-```{r}
+```{r fig.width=7}
 read_csv("appendix/quasicycles.csv") %>%
@@
   ggplot(aes(time, population, col = species)) + geom_line() + 
-  facet_wrap(~ sigma) + scale_color_ptol() #+ 
+  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() #+ 
 #  labs(x="Time", 
@@
 
-## Noise the informer: Noise can provide novel information
+# Noise the informer: Noise can provide novel information
 
@@
 
-```{r}
+```{r fig.width=7, fig.height=7}
 tipping <- read_csv("appendix/tipping.csv")
@@
   geom_vline(aes(xintercept = tip_time), col="red", lty=2) +
-  facet_wrap(~series, ncol = 2, scales="free_y") # +
-#  theme_ipsum_rc() +
-#  labs(title="A: Early warning signs")
-
+  facet_wrap(~series, ncol = 2, scales="free_y")  +
+  theme_ipsum_rc() 
 p2 <- tipping %>% 
@@
   ggplot(aes(t, population)) + geom_line() + 
-  geom_vline(aes(xintercept = tip_time), col="red", lty=2) #+ 
-#  scale_y_log10() # + 
-#  theme_ipsum_rc() + 
-#  labs(title="B: Population density",
-#       caption = "Vertical red dashed line indicates tipping point location")
+  geom_vline(aes(xintercept = tip_time), col="red", lty=2) + 
+  #scale_y_log10()  + 
+  theme_ipsum_rc() + 
+  labs(subtitle = "Yeast population size", 
+       caption = "Vertical red dashed line indicates tipping point location")
 
@@
 ggdraw() +
-  draw_plot(p2, x = 0, y = 0, width = 1, height = 0.5) +
-  draw_plot(p1, x = 0, y = .5, width = 1, height = 0.5)
+  draw_plot(p1, x = 0, y = .5, width = 1, height = 0.5) +
+  draw_plot(p2, x = 0, y = 0, width = 1, height = 0.6)
 #ggsave("figures/tipping.pdf")
@@
 
-## Conclusions
+# Conclusions
 
-## References
+# References

### 2017-10-25 paper.Rmd
WIP

@@
 
-```{r include=FALSE, message=FALSE}
+```{r libraries, include=FALSE, message=FALSE}
 knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="CairoPDF", fig.width=3.5, fig.height=3.5)
@@
 These examples are central to the conceptual transistion away from viewing noise only as a nusiance that obscures patterns and realize the rich set of patterns and phenomena that can emerge.
-Our tour will be by no means exhaustive, but as with chaos before it, theoretical allure does not always reflect ecological releveance.
+Our tour will be by no means exhaustive, but as with chaos before it, theoretical allure does not always reflect ecological relevance.
 Having looked at some of the phenomena that can be created by noise, our third stop will try to run this in reverse: using noise to infer underyling phenomena.
@@
 Simply put, *noise is all that we leave out of our model*.
+Frequently, this additional noise is tacked on to the end of a model, as if as an afterthought,
+
+$$X_{t+1} = f(X_t) + \xi_t$$
+with the assertion that $\xi$ is some normally distributed noise term reflecting the combined effects of any large array of possible factors that also influence the value of our state, $X$.
+While not always wrong, this handwaving over-simplification is increasingly replaced with a more ecologically and mechanistically grounded description for the origins of noise.
+With the exception of quantum machanics (and there only at scales that have no relevance to ecology), stochasticity is not an inherent property of nature but rather an emergent property of many events occurring at a different scale.
+Coins and dice obey the physical laws of motion; only variation in how they are tossed (i.e.
+something complex we do not usually model) creates randomness.
+The term "noise" itself reflects this emergent rather than fundamental origin: the combination of hundreds of voices and sounds each individually distinct blends together into a background of noise.
+Statistical mechanics tells us that temperature is a measure of the average speed of individual molecules in a gas -- any given molecule will experience a range of possible speeds but by itself has no notion of temperature.
+Likewise, stochasticity enters our ecological models because our quantities of interest, such as the density of a population, is but an average over a lower-level, discrete process of individual births and deaths.
+Stochasticity that emerges directly from discrete individual events that underly the process described by our model are considered *intrinsic stochasticity*, commonly referred to as *demographic stochasticity* in ecological literature, as it usually arises in the context scaling from individual demographic events to population level discriptions.
+An important aspect of this demographic stochasicity is that it scales with the square root of the "system size," so that for very large populations the resulting fluctuations are relatively small.
+On the other hand, *extrinsic stochasticity*, usually referred to as *environmental stochasticity* in the ecological literature, arises in processes that are not described by our model, but merely impact the value of some parameter of our model; for instance, fluctuations in temperature may drive variation in a death rate, or fluctuations in rainfall drive variation in carrying capacity.
+
+<!--
+Early ecological models were almost always formulated at the level of the population, described in terms of differential or difference equations reflecting the available tools of the time.
+The advent of fast and inexpensive computing power
+-->
+
+@Black2010 provides an excellent review of the importance of intrinsic stochasticity and the evol
+
+Consider Levins' patch model, where $n$ individuals compete for a finite number of suitable habitats $N$.
+Individuals die a constant rate $e$, and produce offspring at a constant rate $c$ who then have a probability of colonizing an open patch that is simply proportional to the fraction of available patches, $1 - n/N$.
+
+Approximations of the so-called master equation, van Kampen system size approximation [@vanKampen1987] described in @Black2010, (rooted in the theorems of @Kurtz1977),
+
+<!-- Intrinsic vs extrinisic: a matter of perspective.
+demographic noise in one variable is environmental noise in another.
+environmental variation usually implies the environmental variable is simply independent of the -->
+
+Our variables of interest themselves
+
+The behavior of our variables depends on some quantity which itself is a
+
+<!-- Stochasticity is not inherent -->
+
+<!-- 
+$$\textrm{d} X_t = \mu(X_t) \textrm{d}t + \sigma(X_t) \textrm{d}B_t$$
+
+An SDE. Note the separation of the differential elements, which reveal that this is actually shorthand for an integral equation:
+
+$$X_{t+s} - X{t}  = \int_t^{t+s} \mu(X_{\tau}, \tau) \textrm{d}\tau + int_t^{t+s} \sigma(X_{\tau}, \tau) \textrm{d}B\tau$$
+
+The first integral is the ordinary Lebesgue integral, while the second is an It\^o integral, which integrates over the stochastic process $B_t$ (Brownian motion, also termed a Weiner process).
+The It\^o calculus follows different rules from the classic calculus, largely due to the fact that an It\^o integral has non-zero quadratic variation.
+(Crudely put, $\textrm{d}x^2 \sim 0$, but $\textrm{d}B^2 ~ dx$).
+
+-->
+
 While ecologists will sometimes assert that stochasticity is inherent: birth and death are inherently stochastic events; or the environment is inherently stochastic, these are only assertions at a particular scale.
@@
 In all of science, only quantum mechanics claims a fundamentally probabilistic view of events, and then at such scales that have no relevance to ecology.
-The term "noise" itself reflects this emergent rather than fundamental origin: the combination of hundreds of voices and sounds each individually distinct blends together into a background of noise.
 <!-- Our state variables don't "exist" except as averages of a lower level process -->
@@
 
+Given ecologists natural inclination towards modeling the biological rather than environmental aspects of a system, it will be no surprise that mechanistic descriptions of environmental stochasticity have remained far more fuzzy.
+One step in this direction has been recognition that whatever the mechanisms, this variation is unlikely to result in purely white noise, as commonly postulated for mathematical convienence, but rather some degree of correlation, termed red noise, or possibly anti-correlation such as from seasonal or other oscillatory patterns, so called blue noise.
+
+[@Roughgarden; @Ellner]
+
 <!-- Demographic noise scaling away with population size, but not so for environmental noise -->
@@
 <!-- Identify as beyond the scope of this paper -->
+<!-- non-temporal stochasticity -->
 
-<!-- non-temporal stochasticity -->
+Several additional sources of stochasticity or variation more generally fall outside these categories demographic and environmental noise.
+While a detailed treament is beyond the scope of this paper, I enumerate these sources to disambiguate such phenomena from the intrinsic and extrinsic noise sources that comprise the focus here.
 
@@
 *Spatial heterogeneity* can likewise be reflected in stochastic parameters of a spatially implicit model, or as spatially varying environmental stochasticity in a spatially explicit model.
-As with other forms of noise [@Melbourne2008; @Bonsall2004], interaction between individual or spatial variation and other sources of noise can drive patterns including the potential for persistence, coexistence, and dispersal of species [@Schreiber2010; @Hart2016; @Schreiber2017], but are beyond the scope of this paper.
+As with other forms of noise [@Melbourne2008; @Bonsall2004], interaction between individual or spatial variation and other sources of noise can drive patterns including the potential for persistence, coexistence, and dispersal of species [@Schreiber2010; @Hart2016; @Schreiber2017].
 
@@
 uncertainty does not impact the actual state of our system, and thus does not give rise to any
-additional phenomena or information about that process.
+additional phenomena or information about that process, though it can obscure it [@Knape2011; @deValpine2002].
 
-### Not noise
-
-**Reducible uncertainty**: In constructing and estimating mathematical models of ecological processes, 
+Other forms of uncertainty common in models are not stochastic phenomena.  This includes any form of 
+**reducible uncertainty**: In constructing and estimating mathematical models of ecological processes, 
 we encounter uncertainty over our parameter estimates and possible choice of models [@Ellison2004].  In some ways, this
@@
 Despite substantial interest in previous decades and the relative ease of encountering chaos in simple ecological models (particularly in differences equations), and convincing reconstructions of chaotic dynamics in laboratory manipulations [@Costantino1997] the prevelance of chaotic dynamics as a driving factor in most ecological systems remains unclear.
-Meanwhile, the undeniable influences of
+After all, the great variation of natural populations is hardly suprising: the undeniable influences of variation created by all those things left out of our models, all that individual detail from intrinsic, lower level processes and external environmental processes that becomes aggregated in our simple models all but begs for a stochastic description.
 
-Figure: schematic showing types of noise
-
-<!-- blah blah fixed effects vs random effects -->
+Much active research continues to explore the consequences of introducing these more explicitly motivated and interacting sources of noise.
+With some grounding in the origins of different types of noise in place, we are ready to move to our second stop to see how this noise can drive the creation of novel phenonema on a marcroscopic level.
 
@@
 Stochasticity can give rise to a wide array of interesting phenomena that are often at odds with our intuition.
+Such behavior often arises through an interaction between stochastic and non-linear elements of a system, both hallmark traits of ecological processes.
+In this section, I highlight several examples that have shown up frequently in the literature and prove sufficiently tractable to provide a simple intution for the mechanisms involved.
 
@@
 
-Understanding the origin of sustained oscillations has long been a central question in ecology [@Hastings1996].
+Understanding the origin of sustained oscillations has long been a central question in ecology [@Hastings1996; @Bjornstad2001].
 The simple Lotka Volterra model produces only neutrally stable centers, while the more realistic model that introduces carrying capacity for prey as they compete for resources results in a model that produces only damped oscillations, eventually settling into a stable state (Figure 1a).
+However, the introduction of relatively small noise into the system can be sufficient to excite the system into sustained oscillations (Figure 1b.)  In the example shown here, a simple predator-prey system is driven by environmental white noise $\xi_t$, with prey following the equation:
 
-$$x_{t+1} = x_t + x_t r \left(1 - \frac{x_t}{K}\right) - b x_t  y_t + \epsilon_t$$
-$$y_{t+1} = y_t + c  x_t  y_t - d  y_t+ \xi_t $$
+$$x_{t+1} = x_t + x_t r \left(1 - \frac{x_t}{K}\right) - b x_t  y_t + \xi_{x,t}$$
+where $x_t$ is the prey density at time $t$, $r$ the growth rate, $K$ the carrying capacity, $b$ an encounter rate with the predator $y$, who obeys the equation:
 
-```{r fig.width=7}
+$$y_{t+1} = y_t + c  x_t  y_t - d  y_t+ \xi_{y,t} $$
+with $c$ the conversion ratio and $d$ the predator mortality rate.
+Here we have assumed for the prey, $\xi_x \sim \mathcal{N(0, \sigma_x)}$, and similarly for the predator driven by $\xi_y$.
+Figure 1a shows examples for noise $\sigma_x = \sigma_y = 10^{-5}$, while panel 1b shows $\sigma_x = \sigma_y = 0.01$.
+
+```{r quasicycles, fig.width=7}
 read_csv("appendix/quasicycles.csv") %>%
@@
 
-- Power spectrum
+Oscillations arise through the process of stochastic resonance.
+The damped oscillations seen in Figure 1a are analgous to a child sitting still on a swing: friction slowly damps the magnitude of each successive oscillation, while the period of pendulum remains unchanged.
+Driving the pendulum through perturbations timed to match that natural period can quickly create sustained oscillations with large magnitudes, but such perfect timing is not required.
+In the model above, white noise effectively drives the pendulum at all frequencies, though at very low magnitude, since the noise is small relative to the scale of the dynamics.
+However, frequencies that most closely match the natural period of the damped oscillator are amplified, resonating with the natural oscillations.
+This process is sufficient to permit even small noise to drive sustained oscillations in the model indefinitely.
 
-[@Black2010] 
-[@Blarer1999]
+The power spectrum (the squared norm of the Fourier transform, a measure of what frequencies are present; see, e.g.
+@Black2012) provides a convenient way to visualize the effects of these oscillations.
+The power spectrum  for a deterministic limit cycle would create a perfect sinsoidal oscillation and a resulting power spectrum of a delta spike at that single frequency.
+The power spectrum for the quasi-cycle instead shows a distribution of frequencies, centered at the the resonant frequency of the damped oscillator.
+
+Early work by @Blarer1999 identified quasi-cycles as a possible explanation for oscillations and periodic outbreaks in pest ecosystms.
+@Bjornstad2004 analyzes the potential for such stochastic effects to drive cyclic behavior in population abundances of Atlantic bluefin tuna and Atlantic Cod, while @Black2010 illustrates the use of power-sepctrum approaches to examine the behavior of quasicycles in common disease outbreak models, focusing on a case where noise arises explicitly from demographic stochasticity, rather than generic environmental noise as in the model above and elsewhere.
+@Pineda-Krch2007 compares patterns from these stochastically generated cycles to truely stable limit cycles they resemble.
+
+Stochastic resonance highlights one way in which even small noise can play an important role in driving large scale patterns where we had previously only sought deterministic explanations.
+This example also illustrates the interplay between nonlinear dynamics and stochasticity that is the hallmark of stochastically driven phenomena.
+A stable node with linear dynamics, such as an Ornstein-Ulhenbeck process, has no resonant frequency, and could not produce sustained oscillations (unless the environmental driver itself had a periodic pattern).
+
+<!-- variable noise -->
+
+<!--autocorrelated or red noise can also drive stochastic resonance -->
 
@@
 
-Stochastic fluctuations can also drive spontaneous shifts between alternative stable states.
+We turn from stochastic resonance to oscillations of a very different sort, which do not resemble any deterministic process:
 
-```{r}
+```{r noisy_switch}
 read_csv("appendix/noisy_switch.csv")  %>% ggplot(aes(t,x)) + 
@@
 ```
+
+Stochastic fluctuations can also drive spontaneous shifts between alternative stable states.
 
@@
 @Read2007 <!-- Stochasticity generates an evolutionary instability for infectious disease -->
-
-- Critical transitions
 
@@
 
-```{r fig.width=7, fig.height=7}
+```{r tipping, fig.width=7, fig.height=7}
 tipping <- read_csv("appendix/tipping.csv")

### 2017-10-26 appendix/appendix.Rmd
WIP, mostly adding gillespie simulation

--- 2017-10-23 appendix/appendix.Rmd
@@
 library(nimble)
+library(RcppRoll)
+library(regimeshifts) ## devtools::install_github("cboettig/regimeshifts")
+library(adaptivetau)
+library(tidyverse)
+
+## plot themes, non-essential
 library(hrbrthemes)
 library(ggthemes)
-library(tidyverse)
 ggplot2::theme_set(hrbrthemes::theme_ipsum())
+```
+
+# Gillespie algorithm
+
+Levins' Patch model as an individual-based birth-death process:
+
+$$\frac{dn}{dt} = \underbrace{c n \left(1 - n/N \right)}_{\textrm{birth rate, } b(n)} - \underbrace{en}_{\textrm{death rate, } d(n)} $$
+
+We simulate this using the Gillespie's exact stochastic simulation algorithm (SSA), as implemented in the `adaptivetau` R package.
+We declare the transition events: birth increases the state $n$ to $n+1$, and death $n$ to $n-1$, and describe the rates associated with them:
+
+```{r}
+# indicate state variable and the change
+transitions = list(c(n = +1), # birth event
+                   c(n = -1)) # death event
+
+rateF <- function(state, params, t) {
+  c(params$c * state[["n"]] * (1 - state[["n"]] / params$N), # birth rate
+    params$e * state[["n"]])                                 # death rate
+}
+
+```
+
+We compare the stochastic simulation with identical parameters, $c=1$, $e=0.2$ except for the total number of available sites.
+In our "small" system scenario, we have a total of $N=100$ sites, while in the "large" system we have $N=1000$ sites.
+
+The theory predicts an equilibrium population size $\bar n$ at $b(\hat n) = d(\hat n)$, or
+
+$$\left(1 - \tfrac{e}{c}\right) N,$$ 
+and a variance of
+
+$$\left. \frac{b(n) + d (n)}{2 d'(n) - b'(n)} \right\rvert_{n = \hat n} = \tfrac{e}{c}N$$
+
+```{r gillespie}
+set.seed(1234) # set random number generator seed to be reproducible
+
+e <- 0.2
+c <- 1
+
+sim_small <- adaptivetau::ssa.exact(init.values = c(n = 50),
+                       transitions, rateF,
+                       params = list(c=c, e=e, N=100),
+                       tf=30) %>% as_tibble()
+
+sim_large <- adaptivetau::ssa.exact(init.values = c(n = 500),
+                       transitions, rateF,
+                       params = list(c=c, e=e, N=1000),
+                       tf=30) %>% as_tibble()
+
+theory <- 
+list(small = 100,large = 1000) %>%
+  map_dfr(function(N){
+    data.frame(mean = N * (1-e/c)) %>%
+    mutate(plus_sd = mean + sqrt(N * e / c),
+           minus_sd = mean - sqrt(N * e / c))
+  }, .id = "system_size")
+
+gillespie <- bind_rows("large"=sim_large, 
+                       "small"=sim_small,  
+                       .id = "system_size") %>% bind_rows(theory)
+write_csv(gillespie, "gillespie.csv")
+
+read_csv("gillespie.csv", col_types = "cdiddd")  %>% 
+  ggplot(aes(x = time)) + 
+  geom_hline(aes(yintercept = mean), lty=2) + 
+  geom_hline(aes(yintercept = minus_sd), lty=2) + 
+  geom_hline(aes(yintercept = plus_sd), lty=2) + 
+  geom_line(aes(y = n), col="red") +
+  facet_wrap(~system_size, scales = "free_y") + 
+  labs(caption = "theoretical predictions for mean and plus/minus one standard \n deviation shown in horizontal black dashed lines.")
 ```

### 2017-10-26 paper.Rmd
WIP, mostly adding gillespie simulation

@@
 
-<!-- Master equation transform-->
+<!-- Master equation solution of birth-death model, from appendix-->
 
-<!-- Gillespie algorithm; Population level (SDE) vs agent-based vs individual  -->
+<!-- Gillespie algorithm; Population level (SDE) vs agent-based vs individual. Show figure with small and large N  -->
 
-![Figure 1: demographic stochasticity arises from discrete transitions  between states in in a Markov process.
-Here, a simple birth-death process has transition rates determined by the current population size, _n_.](markov.png)
+```{r warning=FALSE}
+read_csv("appendix/gillespie.csv", col_types = "cdiddd")  %>% 
+  ggplot(aes(x = time)) + 
+  geom_hline(aes(yintercept = mean), lty=2) + 
+  geom_hline(aes(yintercept = minus_sd), lty=2) + 
+  geom_hline(aes(yintercept = plus_sd), lty=2) + 
+  geom_line(aes(y = n), col="red") +
+  facet_wrap(~system_size, scales = "free_y") + 
+  labs(caption = 
+"Population dynamics from a Gillespie simulation of the Levins model with large (N=1000) 
+and small (N=100) number of sites (red) show relatively smaller effects of demographic 
+noise in the larger systme. Theoretical predictions for mean and plus/minus one standard
+deviation shown in horizontal black dashed lines.")
+```
 

### 2017-10-26 appendixB/appendixB.Rmd
appendix B

---
title: "Appendix B: System size expansion of the master equation for demographic noise"
author: 
  name: "Carl Boettiger"
  affiliation: a
address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
#layout: 5p  
bibliography: canonical.bib
output: 
  rticles::elsevier_article:
    includes:
      in_header: header.tex

---

<!--
## The Poisson process description

The Poisson process is a common description of demographic noise.  It arises as the limit of a binomial process (coin flipping) when the events are rare and the population large.  This model is commonly applied as follows.  Consider a model of the form:

\begin{align}
N_{t+1} = f(N_t)
\end{align}

This approach chooses with probability $P$ the number in the next generation $N_{t+1}$ at random from a Poisson distribution with mean $f(N)$:

\begin{align}
P(N_t)_{t+1} = \text{Poisson}(N, f(N_t)), 
\end{align}

while the variance of the Poisson distribution will also be $f(N)$.  If the population has some equilibrium size $N^* \neq 0$ such that $f(N^*) = N^*$, then at equilibrium the population will have size $N^*$ and fluctuate around that value with standard deviation $\sqrt{N^*}$.  In the continuous time limit this approaches a Gaussian with the same mean and variance.  Observe that $P(0) > 0$ for any $N$, hence a population can always go extinct.  Further, if we naturally assume the population is isolated, then $f(0) = 0$, and extinction is the only long-term stable state.  This description provides a first pass at describing demographic noise, and captures some of its important phenomena:

- Finite populations fluctuate around their quasi-equilibrium.
- Any isolated population that cannot grow without bound will eventually go extinct.
-  The magnitude of the steady-state fluctuations relative to the equilibrium decreases with increasing population size (as $\sqrt{N^*}/N^*$), decreasing the extinction risk of larger populations proportionally.

While we have started to capture one aspect of demographic noise (populations are finite), we have ignored the second (population dynamics come from birth and death processes).  Too often we stop at this description of demographic noise.  By using only $f(N)$, we have not specified births and deaths separately.  This is quite common in population models, where the choice of $f(N)$ does not obviously suggest which are the birth and which are the death terms.  Our second introductory example illustrates the importance of doing so.  
-->

Consider a population that is of size $n$ with probability $p_n$, with birth and death rates given by $b_n$ and $d_n$ respectively, diagrammed in Fig~\ref{markov}  This representation is known as a single step Markov process.  The rate of transitions to the $n+1$ state $b_n$, the rate of transitions to $n-1$ is $d_n$.  Both of these are transitions away from the state $p_n$, hence the decrease the probability of $p_n$.  The probability $p_n$ increases by transitions into the state $n$, from either side: births enter from the state below at rate $b_{n-1}$ and deaths from the state above, $d_{n+1}$.  Hence the rate of change in $p_n$ is given by

\begin{align}
\dot p_n = b_{n-1} p_{n-1} + p_{n+1} d_{n+1} - [b_n + d_n]p_n
\end{align}

\begin{figure}
\begin{center}
\includegraphics[width=.25\textwidth]{markov}
\caption{The birth-death process as a Markov process.}\label{markov}
\end{center}
\end{figure}

This probability balance is known for historical reasons as a master equation. This equation will form the center of our treatment.  Master equations of this form can be written more concisely by introducing the step operator, $\mathbb E^k$ such that $\mathbb E^k f_n = f_{n+k}$, giving us

\begin{align}
\frac{\ud p_n}{\ud t} = (\mathbb E^{-1} - 1) b_n p_n + (\mathbb E -1) d_n p_n \label{master1}
\end{align}

which for historical reasons is known as a master equation. From this we can directly calculate the mean and variance for this process by multiplying by $n$ or $n^2$ and summing over all $n$. These calculations are greatly simplified by observing the following property of the step operator: for any pair of test functions $f_n$, $g_n$, 
\begin{align}
\sum_{n=0}^{N-1}  g_n \mathbb E f_n = \sum_{n=1}^N f_n \mathbb E^{-1} g_n \nonumber
\end{align}

For example, the mean is:

\begin{align}
\frac{\ud }{\ud t} \langle n \rangle &= \sum n (\mathbb E^{-1} - 1) b_n p_n + \sum n (\mathbb E -1) d_n p_n \nonumber \\ 
&= \sum  b_n p_n(\mathbb E - 1)n  + \sum d_n p_n(\mathbb E^{-1} -1) n \nonumber \\
&= - \langle d_n \rangle +  \langle b_n \rangle \label{mean} \\
\intertext{and similarly}
\frac{\ud }{\ud t} \langle n^2 \rangle &= 2 \langle n (b_n - d_n) \rangle + \langle b_n+d_n \rangle \label{var}
\end{align}

The key observation here is that the variance includes the sum of birth and death, not just the difference.  Intuitively we would expect a population with large numbers of births and deaths that average to a small difference to fluctuate much more than one in which this difference is very small.  Despite this simple observation, countless models purporting to include demographic noise ignore this distinction.  While all moments can be derived as above, it is often impossible to solve these equations for nonlinear models.  Instead, we introduce an approximation of Eq \eqref{master1} that is physically motivated and intuitive.  The method is based on derivation presented by the van Kampen [@vanKampen], though a formal proof can be found in earlier work of Thomas Kurtz [@Kurtz1970; @Kurtz1971; @Kurtz1972].

# The system size expansion for a birth-death process

## Introduction, a change of variables

In this approximation, we expand the master equation \eqref{master1} in terms of a measure of the system size, $\Omega$.  The heart of this approximation is a change of variables, the rest is simply book-keeping.  We begin by explaining this change of variables, which replaces $n$ by some average value $\phi$ and some fluctuations $\xi$.

Observe that \eqref{master1} is written in terms of discrete individuals, represented by the integer $n$.  Many population models permit real-valued variables for the population, usually interpreted as the density of individuals, for which fractional values have meaning.  If we go out to the field and mark off a very large area and count all the individuals within it, we can expect to get the population density, $\phi$.  Over some appropriate region, we expect the population density to be independent of our survey area.  Knowing the density, we can predict how many individuals we'd expect to find in any given area $\Omega$, simply $\phi \Omega$.  We also know that the larger the area, the more accurate our prediction.  We call $\phi$ a macroscopic variable -- it describes what we expect to see over an entire population (the macroscopic level) on average, rather than at the individual level.  It is an intensive (bulk) variable, because it does not depend on the area surveyed, while number $n$ will depend on the area $\Omega$ considered.  We expect $n$ to deviate around an average value of $\phi \Omega$ by some amount that depends on the system size.  For several reasons (such as the error term we found in the simple Poisson process), we will guess that the size of the fluctuations $\xi$ scale with system size as $\Omega^{1/2}$.  Mathematically,

\begin{align}
n = \Omega \phi(t) + \Omega^{1/2} \xi \label{n}
\end{align}

We will change Eq \eqref{master1} into the variables $\phi$ and $\xi$.   We begin with the step operator, which can be approximated by a Taylor expansion.To formulate a Taylor expansion of the step operator (where $k$ can be a positive or negative integer), first consider the steop operator under the original discrete variable $n$:

\begin{align}
\intertext{Take as the definition}
\mathbb E^k f(n) &= f(n+k) \nonumber \\
\intertext{Then}
\mathbb E^k n &=  n+k \nonumber \\
\mathbb E^k n^2 &= \mathbb E^k n n = (n+k)(n+k) = n^2+2kn + k^2 \nonumber \\
\intertext{This suggests we can approximate the step operator by a Taylor series}
\mathbb E^k &= 1+ k\frac{\partial}{\partial n} + \frac{k^2}{2} \frac{\partial^2}{\partial n^2} + \ldots \label{En}
\end{align}
To change variables, recall the chain rule,
\begin{align}
&\frac{\partial}{\partial \xi} f(n(\xi)) =  \frac{\partial n}{\partial \xi} \frac{\partial}{\partial n} f(n(\xi)) \nonumber \\
\intertext{hence} 
&\frac{\partial}{\partial n} =  \left(\frac{\partial n}{\partial \xi}\right)^{-1} \frac{\partial}{\partial \xi} \nonumber 
\end{align}
Making this substitution to \eqref{En} we find:

\begin{align}
& \mathbb E^k = 1+ \Omega^{-1/2} k\frac{\partial}{\partial \xi} + \Omega^{-1} \frac{k^2}{2} \frac{\partial^2}{\partial \xi^2} + \ldots \label{taylor}
\end{align}

Taking $P(n,t) = \Pi(\xi, t)$, we can rewrite the time derivative.  First, note that the derivative of the probability distribution in the master equation, $\tfrac{\partial}{\partial t} P(n,t)$ is taken with $n$ held constant,

\begin{align}
\frac{\ud n}{\ud t} &= \Omega \frac{\ud \phi(t)}{\ud t}  + \Omega^{1/2}\frac{\ud \xi}{\ud t} = 0, \nonumber \\
\intertext{therefore}
\frac{\ud \xi}{\ud t} &= -\Omega^{1/2}\frac{\ud \phi(t)}{\ud t}. \label{dxidt}
\end{align}

Also by the chain rule, we have
\begin{align}
\frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  + \frac{\partial \Pi}{\partial \xi}\frac{\ud \xi}{\ud t} 
\end{align}
Substituting in \eqref{dxidt}, we find:

\begin{align}
\frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  - \Omega^{1/2}\frac{\ud \phi}{\ud t}\frac{\partial \Pi}{\partial \xi} \label{Pi}
\end{align}

To complete the transformation, we put \eqref{Pi} on the left side, replace all the $\mathbb E$'s in \eqref{master1} with \eqref{taylor}, and all the $n$'s appearing in the functions $b_n$ and $d_n$ with \eqref{n}:

\begin{align}
&\frac{\partial \Pi(\xi, t)}{\partial t}  - \Omega^{1/2}\frac{\ud \phi}{\ud t}\frac{\partial \Pi}{\partial \xi}  = \nonumber \\
&\quad \left( -\Omega^{-1/2} \frac{\partial}{\partial \xi} +  \frac{\Omega^{-1}}{2} \frac{\partial^2}{\partial \xi^2} + \ldots\right)b(\phi\Omega+\xi\Omega^{1/2})\Pi(\xi, t)\nonumber \\
+&\quad \left( \Omega^{-1/2} \frac{\partial}{\partial \xi} +  \frac{\Omega^{-1}}{2} \frac{\partial^2}{\partial \xi^2} + \ldots\right)d(\phi\Omega+\xi\Omega^{1/2})\Pi(\xi, t)\end{align}

Collecting terms of order $\Omega^{1/2}$ on both sides, we have:

\begin{align}
\frac{\ud \phi(t)}{\ud t} = b(\phi)-d(\phi) = \alpha_1(\phi) \label{macroscopic}
\end{align}

Where we have the part of the birth and death functions that depend on the macroscopic variable $\phi$ alone.  This is known as the macroscopic equation, and corresponds to the density equations commonly written down.  We call this difference $\alpha_1(\phi)$ as a shorthand\footnote{This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.  It so happens this term will give the macroscopic law for any Markov process, not only a birth-death process.  The second moment of the transition rates, $a_2$, will be for us simply the sum of the birth and death rates.  We use this notation as it generalizes to processes that have a different master equation from \eqref{master1}.}

Collecting terms of order $\Omega^0$ we recover the diffusion equation:

\begin{align}
\frac{\partial \Pi}{\partial t} = - \alpha'_{1,0}(\phi) \frac{\partial}{\partial \xi} \xi \Pi + \tfrac{1}{2}\alpha_{2,0}(\phi) \frac{\partial^2}{\partial \xi^2} \Pi \label{FP}
\end{align}

Where we define $\alpha_2(\phi) = b(\phi)+d(\phi)$ as an analogous short hand.  From this it is a straight forward exercise to calculate the moments of the distribution (multiply by $\xi$, $\xi^2$ and integrate),

\begin{align}
\partial_t \langle \xi \rangle &= \alpha'_{1,0}(\phi) \langle \xi \rangle \\
\partial_t  \langle \xi^2 \rangle &= 2\alpha'_{1,0}(\phi) \langle \xi \rangle + \alpha_{2,0}(\phi) \label{fluc} 
\end{align}

The variance $\sigma_{\xi}^2 = \langle \xi^2 \rangle - \langle \xi \rangle^2$ obeys the same relation as the second moment. These equations must be solved using $\phi(t)$ from the macroscopic solution, solved with the appropriate initial condition $n_0$.  Then we can transform back to the original variables:

\begin{align}
\langle n \rangle &= \phi(t | n_0) \Omega + \Omega^{1/2} \langle \xi \rangle \label{n mean}\\
\sigma^2 &= \Omega \sigma_{\xi}^2 \label{n var}
\end{align}

It is possible to prove that any solution to \eqref{FP} must be Gaussian [@Kurtz1970; @Kurtz1971].  Consequently, knowing these two moments completely determines the distribution of $n$.  This is often assumed for demographic noise.  We have justified this common Gaussian noise assumption by showing it is simply a consequence of expanding the master equation to linear order, $\mathcal O(\Omega^{0})$.  From \eqref{n mean} we can also conclude that the macroscopic (average) variable obeys the deterministic law.

## Fluctuation dissipation theorem

Eq \eqref{fluc} is particularly instructive.  Note that $\alpha_2(\phi)$ is always positive, and hence will try to increase the fluctuations $\langle \xi^2 \rangle$.  As this term increases, it increases the influence of its coefficient $\alpha_1'(\phi)$.  If this term is negative (as it must be near a stable equilibrium) then it serves to dissipate these fluctuations.  Note the dissipation comes from the macroscopic behavior alone, and does not depend on knowing $b$ and $d$ separately.  Since this dissipation is strongest for large fluctuations and small for very small fluctuations, $\xi^2$ will exponentially approach an equilibrium:

\begin{align}
\langle \xi^2 \rangle = \frac{-\alpha_2(\phi)}{2 \alpha_1'(\phi)} \label{fluctuation dissipation}
\end{align}

However, $\alpha_1'(\phi)$ need not be negative everywhere.  This is easy to see by plotting the macroscopic growth rate $\tfrac{\ud \phi}{\ud t} = \alpha_1(\phi)$ against $\phi$, as in Fig~\ref{levins a1}, where $\alpha_1'$ is simply the slope.  Equilibria $\phi_s$ occur wherever this curve crosses zero.  If it does so with a negative slope, $\alpha_1'(\phi^s) < 0$ then this point is stable.  Hence there will be a region around any stable point where this fluctuation-dissipation relationship given by \eqref{fluctuation dissipation} provides a good description of the fluctuations.  Consequently, for any birth-death process $b>d$ we can write down the equilibrium fluctuations as:

\begin{equation}
\sigma^2 = \frac{b+d}{2 [d' - b']} \label{fluc diss} 
\end{equation}

This expression has the wonderful properties of being both simple and general.

## Example: The Levins patch model

The Levins meta-population model is given by

\begin{align}
\dot n = c n \left(1 - \frac{n}{N}\right) - e n \label{levins}
\end{align}

where $n$ is the number of occupied patches, $N$ the total number of patches, $c$ is the colonization rate, and $e$ the extinction rate.  The colonization term provides our birth rate function and the extinction rate the death rate function.  The total number of patches $N$ is an obvious choice for the system size $\Omega$.  From this we can immediately apply the above theory.  The jump moments written in the macroscopic variable are:

\begin{align}
\alpha_1(\phi) &= c \phi(1- \phi) - e\phi \\
\alpha_2(\phi) &= c \phi(1-\phi) + e\phi
\end{align}

From which the macroscopic equation \eqref{macroscopic}, the fluctuations \eqref{fluc} can be solved.  According to \eqref{FP} the distribution is simply the Gaussian with the mean given by \eqref{n mean} and variance \eqref{n var}. The equilibrium of the macroscopic equation is:

\begin{align}
\langle n\rangle_s = N\left(1-\frac{e}{c}\right)
\end{align}
while the steady-state fluctuations are given by \eqref{fluctuation dissipation}, \eqref{n var}:
\begin{align}
\sigma_n^2 = N\frac{e}{c}
\end{align}

### Comparison to other models:

Note that the macroscopic equation for Levins' patch model has the same mathematical formulation as the familiar logistic equation,

$$\frac{\ud n}{\ud t} = r n \left( 1 - \frac{n}{K} \right)$$
though the partition into birth and death rates is not explicit.  Different ways of dividing this equation between birth and death can therefore create different fluctuation patterns, even as the macroscopic average remains unchanged.  For instance, taking $b = rn$ and $d = rn^2/K$ and  we can calculate the fluctuations of the logistic equation around its equilibrium $n^* = K$ as:

\begin{align}
\sigma^2 = \frac{rn^* + rn^{*2}/K}{2(2rn^*/K - r)} = K \nonumber
\end{align}

Which agrees with calculations elsewhere [@NisbetGurney].

# Alternate derivation: the diffusion approximation

A somewhat different approach to deriving a diffusion equation predominates in the literature.  This approach does not make system size explicit, but instead hinges on taking a simultaneous limit of both short time and small step size.  I prefer the van Kampen derivation, since it appeals to system size rather than an arbitrary small parameter going to zero.  It also provides a more natural treatment for nonlinear systems, for which the diffusion equation requires extra care.  With these notes in mind, the diffusion derivation usually goes as follows:

Consider small steps $x = \epsilon n$, and hence $p_n = \epsilon P_x$, where $\epsilon \ll 1$.  To keep the process from slowing down we consequently have to scale up the rates by $\epsilon$ as well:

\begin{align}
	b_n - d_n = \frac{A_x}{\epsilon}
\end{align}

Where $A$ is independent of $\epsilon$.  Similarly we have to scale the sum, which being proportional to the variance must be scaled by
\begin{align}
b_n + d_n = \frac{B_x}{\epsilon^2}
\end{align}

We Taylor expand the step operator $\mathbb E$ in the new variable:
\begin{align}
\mathbb E^k &= 1+ \epsilon k\frac{\partial}{\partial x} + \frac{\epsilon^2 k^2}{2} \frac{\partial^2}{\partial x^2} + \ldots
\end{align}

Then \eqref{master1} becomes:
\begin{align}
\frac{\partial P(x,t)}{\partial t} &= \left( \epsilon \frac{\partial }{\partial x} + \epsilon^2 \tfrac{1}{2}\frac{\partial^2 }{\partial x^2} + \ldots \right)\left(\frac{B_x}{2\epsilon^2}-\frac{A_x}{2\epsilon} \right)P \nonumber \\
+&\left(- \epsilon \frac{\partial }{\partial x} + \epsilon^2 \tfrac{1}{2}\frac{\partial^2 }{\partial x^2} - \ldots \right) \left( \frac{B_x}{2\epsilon^2} + \frac{A_x}{2\epsilon} \right)P 
\intertext{Collecting terms of common order $\epsilon$, }
& = - \frac{\partial A_x P}{\partial x} + \tfrac{1}{2} \frac{\partial^2 B_x P }{\partial x^2} + \mathcal O(\epsilon^2)\label{FP1}
\end{align}

For small $\epsilon$ we can ignore the terms of $\mathcal O(\epsilon^2)$, and we have recovered the diffusion approximation.  Unfortunately, nature does not give us a parameter $\epsilon$ that goes to zero.  We expect that larger populations will have relatively smaller fluctuations (recall the introductory example) and hence this limit should have something to do with increasing the system size.  We make these notions more precise by expanding equation \eqref{master1} explicitly in terms of the system size.  This will lead us to a much richer and more satisfying description of demographic noise than the mathematical limit above.

# References

### 2017-10-26 paper.Rmd
WIP - spelling

@@
 
-Population-level vs ABM vs IBM
-
 Discuss out of scope:  
@@
 
-<!-- Should we make explicit or only implicit reference to Ecology Letters? -->
-
 # Introduction: Noise the nuisance
@@
 Yet establishing statistical evidence for density dependence is remarkably difficult, as illustrated by a landmark study of @Knape2011.
-Using time series from the Global Population Dynamics Database (GPDD) @Knape2011 show that once one accounts for the potential for measurement error in addition to environmental stochasticity -- an unavoidable source of noise in most ecological data -- it becomes difficult to find evidence for even the simplest models of density dependence in the majority of avaialble time series.
-Faced with such uncertainty, it is tempting to throw up our hands and wait until some prophesized deluge of streaming microsensors stapled to every organism provide enough signal to beat back the noise.
+Using time series from the Global Population Dynamics Database (GPDD) @Knape2011 show that once one accounts for the potential for measurement error in addition to environmental stochasticity -- an unavoidable source of noise in most ecological data -- it becomes difficult to find evidence for even the simplest models of density dependence in the majority of available time series.
+Faced with such uncertainty, it is tempting to throw up our hands and wait until some prophesied deluge of streaming micro-sensors stapled to every organism provide enough signal to beat back the noise.
 Fortunately, this has not been the path taken.
@@
 Over the past twenty years, ecological theory, modelling, and methods have increasingly elucidated the role of noise not only to *obscure* patterns but to *create* them.
-This period marked a transistion away from the intense focus on simple models which exhibited deterministic chaos that had enchanted much of theoretical ecology in the previous two decades [@Bjornstad2001] in favor of more complex and stochastic models.
+This period marked a transition away from the intense focus on simple models which exhibited deterministic chaos that had enchanted much of theoretical ecology in the previous two decades [@Bjornstad2001] in favor of more complex and stochastic models.
 Of particular interest has been those conceptual advances that have bridged the divide from theoretical to empirical interest -- not merely the introduction of "real data" in a theoretical paper, but rather the concepts that take on a life of their own in the larger research community.
@@
 <!-- Regarding simulation examples -->
-In this spirit, each of the examples illustrated here is coupled with consise and efficent R code in the appendix to both reproduce and further explore the simulations.
-While examples of noisy phenomena in natural systems are also crucial, matching the pattern does not necessarily mean we have also matched the proccess behind the data.
+In this spirit, each of the examples illustrated here is coupled with concise and efficient R code in the appendix to both reproduce and further explore the simulations.
+While examples of noisy phenomena in natural systems are also crucial, matching the pattern does not necessarily mean we have also matched the process behind the data.
 The examples cited from the literature here offer some of the best empirical evidence of these phenomena, but I encourage readers to also experiment with the example code provided.
 Simulations permit poking and prodding of empirical investigation quickly and with little mathematical formalism in the way.
-In the code provided, most stochastic models are expressed in the BUGS language, which may be more familiar to empirical readers than corresponding mathematical formulae.
+In the code provided, most stochastic models are expressed in the BUGS language, which may be more familiar to empirical readers than corresponding mathematical formulas.
 This also permits both efficient simulation and potential estimation of parameters given sample data using the R package, NIMBLE [@DeValpine2017].
@@
 Our first stop will be into research into the origins of noise.
-This will not only let us set up definitions for terms and concepts we will encounter throughout, but also introduce some of the fundamental mathematical and computational tooling needed in our tour of noisy phenemena.
+This will not only let us set up definitions for terms and concepts we will encounter throughout, but also introduce some of the fundamental mathematical and computational tooling needed in our tour of noisy phenomena.
 From origins of noise, we will then explore some of the phenomena that noise can create.
-These examples are central to the conceptual transistion away from viewing noise only as a nusiance that obscures patterns and realize the rich set of patterns and phenomena that can emerge.
+These examples are central to the conceptual transition away from viewing noise only as a nuisance that obscures patterns and realize the rich set of patterns and phenomena that can emerge.
 Our tour will be by no means exhaustive, but as with chaos before it, theoretical allure does not always reflect ecological relevance.
-Having looked at some of the phenomena that can be created by noise, our third stop will try to run this in reverse: using noise to infer underyling phenomena.
+Having looked at some of the phenomena that can be created by noise, our third stop will try to run this in reverse: using noise to infer underlying phenomena.
 Examples here are rarer but potentially the most rewarding, bridging the gap from theoretical construct to empirical investigation.
-The most compelling examples have come from ways in which noisy pheneomena are used to infer stability and forecast changes in resilience of populations or ecosystems.
+The most compelling examples have come from ways in which noisy phenomena are used to infer stability and forecast changes in resilience of populations or ecosystems.
 Though not without weaknesses or missteps, the resulting dialog between theoretical and empirical aspects have made this particularly a rich area of research.
@@
 Where does it come from?
-It is important to bear in mind that whenever stochasticity is introduced into a model, it is only as a simplification or summary of the effects of a large number of other processes, often occuring at different scales.
+It is important to bear in mind that whenever stochasticity is introduced into a model, it is only as a simplification or summary of the effects of a large number of other processes, often occurring at different scales.
 Simply put, *noise is all that we leave out of our model*.
@@
 with the assertion that $\xi$ is some normally distributed noise term reflecting the combined effects of any large array of possible factors that also influence the value of our state, $X$.
-While not always wrong, this handwaving over-simplification is increasingly replaced with a more ecologically and mechanistically grounded description for the origins of noise.
-With the exception of quantum machanics (and there only at scales that have no relevance to ecology), stochasticity is not an inherent property of nature but rather an emergent property of many events occurring at a different scale.
+While not always wrong, this hand-waving over-simplification is increasingly replaced with a more ecologically and mechanistically grounded description for the origins of noise.
+With the exception of quantum mechanics (and there only at scales that have no relevance to ecology), stochasticity is not an inherent property of nature but rather an emergent property of many events occurring at a different scale.
 Coins and dice obey the physical laws of motion; only variation in how they are tossed (i.e.
@@
 Likewise, stochasticity enters our ecological models because our quantities of interest, such as the density of a population, is but an average over a lower-level, discrete process of individual births and deaths.
-Stochasticity that emerges directly from discrete individual events that underly the process described by our model are considered *intrinsic stochasticity*, commonly referred to as *demographic stochasticity* in ecological literature, as it usually arises in the context scaling from individual demographic events to population level discriptions.
+Stochasticity that emerges directly from discrete individual events that underlie the process described by our model are considered *intrinsic stochasticity*, commonly referred to as *demographic stochasticity* in ecological literature, as it usually arises in the context scaling from individual demographic events to population level descriptions.
 An important aspect of this demographic stochasicity is that it scales with the square root of the "system size," so that for very large populations the resulting fluctuations are relatively small.
@@
 
-@Black2010 provides an excellent review of the importance of intrinsic stochasticity and the evol
+@Black2010 provides an excellent review of the importance of intrinsic stochasticity and the evolution
 
@@
 and small (N=100) number of sites (red) show relatively smaller effects of demographic 
-noise in the larger systme. Theoretical predictions for mean and plus/minus one standard
-deviation shown in horizontal black dashed lines.")
+noise in the larger systme. Models are otherwise identical, with e = 0.2 and c = 1 (code
+in appendix A). Theoretical predictions for mean and plus/minus one standard deviation 
+shown in horizontal black dashed lines.")
 ```
@@
 Given ecologists natural inclination towards modeling the biological rather than environmental aspects of a system, it will be no surprise that mechanistic descriptions of environmental stochasticity have remained far more fuzzy.
-One step in this direction has been recognition that whatever the mechanisms, this variation is unlikely to result in purely white noise, as commonly postulated for mathematical convienence, but rather some degree of correlation, termed red noise, or possibly anti-correlation such as from seasonal or other oscillatory patterns, so called blue noise.
+One step in this direction has been recognition that whatever the mechanisms, this variation is unlikely to result in purely white noise, as commonly postulated for mathematical convenience, but rather some degree of correlation, termed red noise, or possibly anti-correlation such as from seasonal or other oscillatory patterns, so called blue noise.
 
@@
 Several additional sources of stochasticity or variation more generally fall outside these categories demographic and environmental noise.
-While a detailed treament is beyond the scope of this paper, I enumerate these sources to disambiguate such phenomena from the intrinsic and extrinsic noise sources that comprise the focus here.
+While a detailed treatment is beyond the scope of this paper, I enumerate these sources to disambiguate such phenomena from the intrinsic and extrinsic noise sources that comprise the focus here.
 
@@
 For instance, most population-level models treat all individuals within the population, or at least within a particular age or stage class, as identical.
-*Individual heterogeneity* in relevant traits (growth rates, dispersal ability) can be represented probabilistically as a distribution rather than a fixed value, making it function as another form of random varation in a model.
+*Individual heterogeneity* in relevant traits (growth rates, dispersal ability) can be represented probabilistically as a distribution rather than a fixed value, making it function as another form of random variation in a model.
 *Spatial heterogeneity* can likewise be reflected in stochastic parameters of a spatially implicit model, or as spatially varying environmental stochasticity in a spatially explicit model.
@@
 Chaos can a look a lot like noise: indeed, it can be observationally indistinguishable from a stochastic process [@Werndl2009].
-Despite substantial interest in previous decades and the relative ease of encountering chaos in simple ecological models (particularly in differences equations), and convincing reconstructions of chaotic dynamics in laboratory manipulations [@Costantino1997] the prevelance of chaotic dynamics as a driving factor in most ecological systems remains unclear.
-After all, the great variation of natural populations is hardly suprising: the undeniable influences of variation created by all those things left out of our models, all that individual detail from intrinsic, lower level processes and external environmental processes that becomes aggregated in our simple models all but begs for a stochastic description.
+Despite substantial interest in previous decades and the relative ease of encountering chaos in simple ecological models (particularly in differences equations), and convincing reconstructions of chaotic dynamics in laboratory manipulations [@Costantino1997] the prevalence of chaotic dynamics as a driving factor in most ecological systems remains unclear.
+After all, the great variation of natural populations is hardly surprising: the undeniable influences of variation created by all those things left out of our models, all that individual detail from intrinsic, lower level processes and external environmental processes that becomes aggregated in our simple models all but begs for a stochastic description.
 
 Much active research continues to explore the consequences of introducing these more explicitly motivated and interacting sources of noise.
-With some grounding in the origins of different types of noise in place, we are ready to move to our second stop to see how this noise can drive the creation of novel phenonema on a marcroscopic level.
+With some grounding in the origins of different types of noise in place, we are ready to move to our second stop to see how this noise can drive the creation of novel phenomena on a macroscopic level.
 
@@
 Such behavior often arises through an interaction between stochastic and non-linear elements of a system, both hallmark traits of ecological processes.
-In this section, I highlight several examples that have shown up frequently in the literature and prove sufficiently tractable to provide a simple intution for the mechanisms involved.
+In this section, I highlight several examples that have shown up frequently in the literature and prove sufficiently tractable to provide a simple intuition for the mechanisms involved.
 
@@
 @Black2012) provides a convenient way to visualize the effects of these oscillations.
-The power spectrum  for a deterministic limit cycle would create a perfect sinsoidal oscillation and a resulting power spectrum of a delta spike at that single frequency.
+The power spectrum  for a deterministic limit cycle would create a perfect sinusoidal oscillation and a resulting power spectrum of a delta spike at that single frequency.
 The power spectrum for the quasi-cycle instead shows a distribution of frequencies, centered at the the resonant frequency of the damped oscillator.
 
-Early work by @Blarer1999 identified quasi-cycles as a possible explanation for oscillations and periodic outbreaks in pest ecosystms.
-@Bjornstad2004 analyzes the potential for such stochastic effects to drive cyclic behavior in population abundances of Atlantic bluefin tuna and Atlantic Cod, while @Black2010 illustrates the use of power-sepctrum approaches to examine the behavior of quasicycles in common disease outbreak models, focusing on a case where noise arises explicitly from demographic stochasticity, rather than generic environmental noise as in the model above and elsewhere.
-@Pineda-Krch2007 compares patterns from these stochastically generated cycles to truely stable limit cycles they resemble.
+Early work by @Blarer1999 identified quasi-cycles as a possible explanation for oscillations and periodic outbreaks in pest ecosystems.
+@Bjornstad2004 analyzes the potential for such stochastic effects to drive cyclic behavior in population abundances of Atlantic bluefin tuna and Atlantic Cod, while @Black2010 illustrates the use of power-spectrum approaches to examine the behavior of quasi-cycles in common disease outbreak models, focusing on a case where noise arises explicitly from demographic stochasticity, rather than generic environmental noise as in the model above and elsewhere.
+@Pineda-Krch2007 compares patterns from these stochastically generated cycles to truly stable limit cycles they resemble.
 
@@
 This example also illustrates the interplay between nonlinear dynamics and stochasticity that is the hallmark of stochastically driven phenomena.
-A stable node with linear dynamics, such as an Ornstein-Ulhenbeck process, has no resonant frequency, and could not produce sustained oscillations (unless the environmental driver itself had a periodic pattern).
+A stable node with linear dynamics, such as an Ornstein-Uhlenbeck process, has no resonant frequency, and could not produce sustained oscillations (unless the environmental driver itself had a periodic pattern).
 
@@
 
-One of the most fundamental aspects of stochasticty are the implications for long-term persistence of any given population and the stable coexistence of interacting species.
+One of the most fundamental aspects of stochasticity are the implications for long-term persistence of any given population and the stable coexistence of interacting species.
 
@@
 
-Turning patterms found in plankton-herbivore dynamics [@Levin1976]
+Turning patterns found in plankton-herbivore dynamics [@Levin1976]
 
@@
 
-**Early warning signals.**  The theory of early warning signals, or critical slowing down (csd) [@Wissel1984; @Scheffer2009] is an instructive example
+**Early warning signals.**  The theory of early warning signals, or critical slowing down (CSD) [@Wissel1984; @Scheffer2009] is an instructive example
 
@@
 <!-- FIXME unclear, rambling!-->
-It is important to realize that such approaches can only ever measure local properties such as the return time in response to local perturbation (historically termed *stability*; which @Holling1996 landmark chapter unforunately rebrands as "engineering resilience") and not global properties such as the size of the attractor basin / distance to the nearest tipping point (*ecological resilience*, or simply *resilience*).
+It is important to realize that such approaches can only ever measure local properties such as the return time in response to local perturbation (historically termed *stability*; which @Holling1996 landmark chapter unfortunately re-brands as "engineering resilience") and not global properties such as the size of the attractor basin / distance to the nearest tipping point (*ecological resilience*, or simply *resilience*).
 This is illustrated rather vividly in @Dai2015, which illustrates how a system can lose resilience before losing stability, resulting in a scenario that provides less opportunity for early warning.

### 2017-10-26 appendixB/appendixB.Rmd
tidy up

@@
 
-However, $\alpha_1'(\phi)$ need not be negative everywhere.
-This is easy to see by plotting the macroscopic growth rate $\tfrac{\ud \phi}{\ud t} = \alpha_1(\phi)$ against $\phi$, as in Fig~\ref{levins a1}, where $\alpha_1'$ is simply the slope.
-Equilibria $\phi_s$ occur wherever this curve crosses zero.
-If it does so with a negative slope, $\alpha_1'(\phi^s) < 0$ then this point is stable.
+However, $\alpha_1'(\phi)$ need not be negative everywhere, but will be negative for any stable point, $\alpha_1'(\phi^s) < 0$.
 Hence there will be a region around any stable point where this fluctuation-dissipation relationship given by \eqref{fluctuation dissipation} provides a good description of the fluctuations.

### 2017-10-27 appendixA/appendixA.Rmd
appendixA
WIP

--- 2017-10-26 appendix/appendix.Rmd
@@
 ---
-title: "Appendix"
+title: "Appendix A: Stochastic simulations of noisy phenomena"
 output: pdf_document
@@
 ggplot2::theme_set(hrbrthemes::theme_ipsum())
+colours <- ptol_pal()(2)
+
 ```
@@
   gather(species, pop, -t, -sigma) %>%
-  ggplot(aes(t, pop, col=species)) + geom_line() + facet_wrap(~sigma) + scale_color_ptol()
+  mutate(sigma = as.factor(sigma)) %>%
+  mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
+  ggplot(aes(t, pop, col=species)) +
+  geom_line() + 
+  facet_wrap(~sigma) + 
+  scale_color_ptol()
 ```
@@
 
-May's model
+May's model:
+
+$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{Vegitation growth}}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{Vegitation consumption}} + \xi_t,$$
 
@@
 p <- list(r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 1e4)
-
 f <- function(x) x * p$r * (1 - x / p$K)
 g <- function(x) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
+x <- seq(0,2, length.out = 100)
 
-x <- seq(0,2, length.out = 100)
-data.frame(x, f = f(x), g = g(x)) %>%
-  ggplot(aes(x)) + geom_line(aes(y = f-g)) + 
-  geom_hline(aes(yintercept=0), lty=2) +
-  coord_cartesian(xlim = c(0,1.5), ylim=c(-.015,.1))
+p2 <- data.frame(x, f = f(x), g = g(x)) %>%
+  ggplot(aes(x, y)) + 
+  geom_line(aes(y = f), col=colours[1], lwd=1) +
+  geom_line(aes(y = g), col=colours[2], lwd=1) + labs(x = bquote(X[t]), y = bquote(X[t+1]))
 

### 2017-10-27 paper.Rmd
WIP

@@
 keywords:
-  - Stochasticity
-  - Demographic noise
-  - Environmental noise
-  - Colored noise
+  - stochasticity
+  - demographic noise
+  - environmental noise
+  - colored noise
   - quasi-cycles
@@
 ```{r libraries, include=FALSE, message=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="CairoPDF", fig.width=3.5, fig.height=3.5)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="CairoPDF", fig.width=7, fig.height=3.5)
 library(ggthemes)
@@
 #ggplot2::theme_set(theme_bw()) # simpler theme if you do not have these fonts
+colours <- ptol_pal()(2)
 ```
@@
 
-Approximations of the so-called master equation, van Kampen system size approximation [@vanKampen1987] described in @Black2010, (rooted in the theorems of @Kurtz1977),
+Approximations of the so-called master equation, van Kampen system size approximation [@vanKampen2007] described in @Black2010, (rooted in the theorems of [@Kurtz1970; @Kurtz1971]),
 
@@
 
-```{r warning=FALSE}
-read_csv("appendix/gillespie.csv", col_types = "cdiddd")  %>% 
+<!-- Figure 1: Gillespie simulations -->
+```{r gillespie, warning=FALSE, fig.width=7, fig.cap="Population dynamics from a Gillespie simulation of the Levins model with large (N=1000, panel A) and small (N=100, panel B) number of sites (blue) show relatively weaker effects of demographic noise in the bigger system.
+Models are otherwise identical, with e = 0.2 and c = 1 (code in appendix A).
+Theoretical predictions for mean and plus/minus one standard deviation shown in horizontal re dashed lines."}
+read_csv("appendixA/gillespie.csv", col_types = "cdiddd")  %>% 
+  mutate(system_size = recode(system_size, large = "A. 1000 total sites", small= "B. 100 total sites")) %>%
   ggplot(aes(x = time)) + 
-  geom_hline(aes(yintercept = mean), lty=2) + 
-  geom_hline(aes(yintercept = minus_sd), lty=2) + 
-  geom_hline(aes(yintercept = plus_sd), lty=2) + 
-  geom_line(aes(y = n), col="red") +
-  facet_wrap(~system_size, scales = "free_y") + 
-  labs(caption = 
-"Population dynamics from a Gillespie simulation of the Levins model with large (N=1000) 
-and small (N=100) number of sites (red) show relatively smaller effects of demographic 
-noise in the larger systme. Models are otherwise identical, with e = 0.2 and c = 1 (code
-in appendix A). Theoretical predictions for mean and plus/minus one standard deviation 
-shown in horizontal black dashed lines.")
+  geom_hline(aes(yintercept = mean), lty=2, col=colours[[2]]) + 
+  geom_hline(aes(yintercept = minus_sd), lty=2, col=colours[[2]]) + 
+  geom_hline(aes(yintercept = plus_sd), lty=2, col=colours[[2]]) + 
+  geom_line(aes(y = n), col=colours[[1]]) +
+  facet_wrap(~system_size, scales = "free_y") 
 ```
@@
 
-**Quasi-cycles**
+## Quasi-cycles
 
 Understanding the origin of sustained oscillations has long been a central question in ecology [@Hastings1996; @Bjornstad2001].
-The simple Lotka Volterra model produces only neutrally stable centers, while the more realistic model that introduces carrying capacity for prey as they compete for resources results in a model that produces only damped oscillations, eventually settling into a stable state (Figure 1a).
-However, the introduction of relatively small noise into the system can be sufficient to excite the system into sustained oscillations (Figure 1b.)  In the example shown here, a simple predator-prey system is driven by environmental white noise $\xi_t$, with prey following the equation:
+The simple Lotka Volterra model produces only neutrally stable centers, while the more realistic model that introduces carrying capacity for prey as they compete for resources results in a model that produces only damped oscillations, eventually settling into a stable state (Figure 1A).
+However, the introduction of relatively small noise into the system can be sufficient to excite the system into sustained oscillations (Figure 1B).
+In the example shown here, a simple predator-prey system is driven by environmental white noise $\xi_t$, with prey following the equation:
 
@@
 Here we have assumed for the prey, $\xi_x \sim \mathcal{N(0, \sigma_x)}$, and similarly for the predator driven by $\xi_y$.
-Figure 1a shows examples for noise $\sigma_x = \sigma_y = 10^{-5}$, while panel 1b shows $\sigma_x = \sigma_y = 0.01$.
+Figure 1A shows examples for noise $\sigma_x = \sigma_y = 10^{-5}$, while Figure 1B shows $\sigma_x = \sigma_y = 0.01$.
 
-```{r quasicycles, fig.width=7}
-read_csv("appendix/quasicycles.csv") %>%
+<!-- Figure 2: Quasi-cycles -->
+```{r quasicycles, fig.width=7, fig.cap="Panel A: under negligible environmental noise, the introduction of a carrying capacity damps Lotka-Volterra oscillations which eventually settle to a stable state.
+Panel B: small noise induces quasi-cycles in the same model: sustained long-term oscillations predominately around the natural frequency of the original damped oscillations.
+Parameters for models shown are r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800; code in Appendix A."}
+read_csv("appendixA/quasicycles.csv") %>%
   gather(species, population, -t, -sigma) %>%
+  mutate(sigma = as.factor(sigma)) %>%
+  mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
   rename(time = t) %>%
@@
 #       caption="Left panel: neglible  noise results only in damped oscillations.\nRight panel: small noise creates sustained oscillations.")
-#ggsave("figures/quasicycles.pdf")
 ```
@@
 
-**Stochastic switching**
+## Stochastic switching
 
-We turn from stochastic resonance to oscillations of a very different sort, which do not resemble any deterministic process:
+We turn from stochastic resonance to oscillations of a very different sort, which do not resemble any deterministic process, as seen in Figure 3.
 
-```{r noisy_switch}
-read_csv("appendix/noisy_switch.csv")  %>% ggplot(aes(t,x)) + 
-  geom_line() #+ 
-  #labs(title="Stochastic switching", 
-  #     subtitle = "Noise drives transitions between alternative stable states")
-#ggsave("figures/noisy_switch.pdf")
+<!--Figure 3: Noisy switch -->
+```{r noisy_switch, fig.width=7, fig.cap="Stochastic switiching in May's model of alternative stable state dynamics.  r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000"}
 
+panel1 <- read_csv("appendixA/noisy_switch.csv")  %>%
+  ggplot(aes(t,x)) + 
+  geom_line(lwd=.1, col=colours[[1]]) +
+  labs(subtitle = bquote(bold(A.)))
+
+p <- list(r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 1e4)
+f <- function(x) x * p$r * (1 - x / p$K)
+g <- function(x) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
+x <- seq(0,2, length.out = 100)
+
+panel2 <- data.frame(x, f = f(x), g = g(x)) %>%
+  ggplot(aes(x, y)) + 
+  geom_line(aes(y = f), col=colours[1], lwd=1) +
+  geom_line(aes(y = g), col=colours[2], lwd=1) + 
+  labs(x = bquote(X[t]), y = bquote(X[t+1]), subtitle= bquote(bold(B.)))
+
+library(gtable)
+library(grid) # low-level grid functions are required
+g1 <- ggplotGrob(panel1)
+g2 <- ggplotGrob(panel2)
+g <- cbind(g1, g2, size="first") # stack the two plots
+#g$widths <- unit.pmax(g1$widths, g2$widths) # use the largest widths
+grid.newpage()
+grid.draw(g)
 ```
 
+This pattern is characterized by long stretches during which fluctuations in the state variable remain concentrated around a high average value followed by sudden, rapid transistions into fluctuations concentrated around a lower value.
+This behavior arises through the interaction of stochasticity and alternative stable state dynamics.
+The simulation shown in Figure 3 is based on an herbivore consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics, to which I have simply introduced an additional environmental noise term:
+
+$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{Vegitation growth}}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{Vegitation consumption}} + \xi_t,$$
+where $\xi_t$ is normally distributed environmental variation with mean zero and standard deviation $\sigma$.
+
 Stochastic fluctuations can also drive spontaneous shifts between alternative stable states.
+
+[@Boettiger2012]
+
+[@Drake2013]
 
@@
 
+<!-- Figure 4: Early warning signals -->
 ```{r tipping, fig.width=7, fig.height=7}
-tipping <- read_csv("appendix/tipping.csv")
+tipping <- read_csv("appendixA/tipping.csv")
 
@@
   select(-x) %>%
+  rename("A. autocorrelation" = autocorrelation, "B. variance" = variance) %>%
   gather(series, value, -t, -tip_time) %>%
-  ggplot(aes(t, value)) + geom_line() + 
-  geom_vline(aes(xintercept = tip_time), col="red", lty=2) +
+  ggplot(aes(t, value)) + geom_line(col=colours[[1]]) + 
+  geom_vline(aes(xintercept = tip_time), col=colours[[2]], lty=2) +
   facet_wrap(~series, ncol = 2, scales="free_y")  +
@@
   select(t, population = x, tip_time) %>%
-  ggplot(aes(t, population)) + geom_line() + 
-  geom_vline(aes(xintercept = tip_time), col="red", lty=2) + 
+  ggplot(aes(t, population)) + geom_line(col=colours[[1]]) + 
+  geom_vline(aes(xintercept = tip_time), col=colours[[2]], lty=2) + 
   #scale_y_log10()  + 
   theme_ipsum_rc() + 
-  labs(subtitle = "Yeast population size", 
+  labs(subtitle = "C. Yeast population size", 
        caption = "Vertical red dashed line indicates tipping point location")
@@
   draw_plot(p2, x = 0, y = 0, width = 1, height = 0.6)
-#ggsave("figures/tipping.pdf")
 ```

### 2017-10-27 appendixA/appendixA.Rmd
testing CI

@@
 ## plot themes, non-essential
-library(hrbrthemes)
 library(ggthemes)
-ggplot2::theme_set(hrbrthemes::theme_ipsum())
+#library(hrbrthemes)
+#ggplot2::theme_set(hrbrthemes::theme_ipsum())
+ggplot2::theme_set(hrbrthemes::theme_bw())
 colours <- ptol_pal()(2)

### 2017-10-27 appendixA/appendixA.Rmd
typo

--- 2017-10-27 appendixA/appendixA.Rmd
@@
 ## plot themes, non-essential
-library(hrbrthemes)
 library(ggthemes)
-ggplot2::theme_set(hrbrthemes::theme_ipsum())
+#library(hrbrthemes)
+#ggplot2::theme_set(hrbrthemes::theme_ipsum())
+ggplot2::theme_set(theme_bw())
 colours <- ptol_pal()(2)

### 2017-10-27 appendixA/appendixA.Rmd
tweaking graphs

--- 2017-10-27 appendixA/appendixA.Rmd
@@
 ## plot themes, non-essential
-library(hrbrthemes)
 library(ggthemes)
-ggplot2::theme_set(hrbrthemes::theme_ipsum())
+library(gridExtra)
+#library(hrbrthemes)
+#ggplot2::theme_set(hrbrthemes::theme_ipsum())
+ggplot2::theme_set(theme_bw())
 colours <- ptol_pal()(2)
@@
 ```{r}
+
 p <- list(r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 1e4)
@@
 
-p2 <- data.frame(x, f = f(x), g = g(x)) %>%
-  ggplot(aes(x, y)) + 
-  geom_line(aes(y = f), col=colours[1], lwd=1) +
-  geom_line(aes(y = g), col=colours[2], lwd=1) + labs(x = bquote(X[t]), y = bquote(X[t+1]))
-
+data.frame(x, growth = f(x), consumption = g(x)) %>%
+  gather(curve, y, -x) %>%
+  ggplot(aes(x, y, col=curve)) +
+  geom_line(lwd=1) + 
+  scale_color_ptol() + 
+  labs(x = bquote(X[t]), y = bquote(X[t+1]), subtitle= bquote(bold(B.)))
 ```
@@
   geom_vline(aes(xintercept = tip_time), col="red", lty=2) +
-  facet_wrap(~series, ncol = 2, scales="free_y") + theme_ipsum() + 
+  facet_wrap(~series, ncol = 2, scales="free_y") +
+  #theme_ipsum() + 
   labs(title="A: Early warning signs")
@@
   scale_y_log10() + 
-  theme_ipsum()+ labs(title="B: Population density",
+#  theme_ipsum() + 
+  labs(title="B: Population density",
               caption = "Vertical red dashed line indicates tipping point location")

### 2017-10-27 appendixA/appendixA.Rmd
use gridExtra not cowplot

@@
                        tf=30) %>% as_tibble()
+```
 
+```{r}
 theory <- 
@@
   }, .id = "system_size")
+```
 
+```{r}
+## Bind together both simulation results and theory results
 gillespie <- bind_rows("large"=sim_large, 
@@
 write_csv(gillespie, "gillespie.csv")
+```
 
+```{r}
 read_csv("gillespie.csv", col_types = "cdiddd")  %>% 
@@
   write_csv("inflation.csv")
-
 ```
@@
   coord_cartesian(xlim=c(100,500), ylim=c(1.2, .5))
-
-```
-
-```{r}
-p <- list(r = .5, K = 2.1, Q = 5, H = .4, sigma = .04, a = 0.27, N = 1e4)
-
-f <- function(x) x * p$r * (1 - x / p$K)
-g <- function(x) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
-
-x <- seq(0,2, length.out = 100)
-df <- data.frame(x, f = f(x), g = g(x))
-
-ggplot(df, aes(x)) + geom_line(aes(y = f)) + geom_line(aes(y = g), col="red")
-
-ggplot(df, aes(x)) + geom_line(aes(y = f-g)) + 
-  geom_hline(aes(yintercept=0), lty=2) +
-  coord_cartesian(xlim = c(0,1.5), ylim=c(-.1,.1))  
 ```
@@
 
-library(cowplot)
-ggdraw() +
-  draw_plot(p2, x = 0, y = 0, width = 1, height = 0.5) +
-  draw_plot(p1, x = 0, y = .4, width = 1, height = 0.5)
+gridExtra::grid.arrange(p1,p2, layout_matrix = rbind(c(1,1), c(2,2)))
 

### 2017-10-28 paper.Rmd
WIP

@@
 Nor is all noise the same, and close examination of differences in in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.
-While an exhaustive review is beyond the scope of this piece, I endeavor to highlight some of the most interesting examples of both noise-induced phenomena and noise-derived information, and point the way towards yet uncharted territory.
+Yet with each aspect of stochasticity leading to some new or unexpected behavior, the time is right to move beyond the familiar refrain of "everything is important" [@Bjornstad2001].
+Stochastic phenomena can suggest new ways of inferring process from pattern, and thus spark more dialog between theory and empirical perspectives that best advances the field as a whole.
+I highlight a few compelling examples, while observing that the study of stochastic phenomena are only beginning to make this translation into empirical inference.
+There are rich opportunities at this interface in the years ahead.
 
@@
 
-```{r libraries, include=FALSE, message=FALSE}
+```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
 knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="CairoPDF", fig.width=7, fig.height=3.5)
+library(tidyverse)
+library(gridExtra)
 library(ggthemes)
-library(tidyverse)
-library(cowplot)
+colours <- ptol_pal()(2)
+```
+
+```{r suggested_libs}
 library(hrbrthemes)
+library(extrafont)
+
+##extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
+hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
 extrafont::loadfonts()
-#extrafont::font_import()                    # One-time use to set up fonts
-#hrbrthemes::import_roboto_condensed()       # One-time use to set up fonts
 ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
+
 #ggplot2::theme_set(theme_bw()) # simpler theme if you do not have these fonts
-colours <- ptol_pal()(2)
 ```
@@
 
-- Key questions
-- Key advances
-
+Highlight *key questions* & *key advances*
 Moving beyond "everything is important" [@Bjornstad2001; @Kingsland1985]
 
-@Bjornstad2001 future: more MCMC methods. More mechanistic models.  dealing with unobserved dynamics
-
-Noise -> informer
+(@Bjornstad2001 future: more MCMC methods. More mechanistic models.  dealing with unobserved dynamics)
 
@@
 Stochastic phenomena outside of theory -- early warning signals
-
-Need for greater tanslation
-
-Discuss out of scope:  
- - behavior
- - evolutionary dynamics examples??
 
@@
 $$X_{t+1} = f(X_t) + \xi_t$$
+
 with the assertion that $\xi$ is some normally distributed noise term reflecting the combined effects of any large array of possible factors that also influence the value of our state, $X$.
@@
 
-<!-- Intrinsic vs extrinisic: a matter of perspective.
-demographic noise in one variable is environmental noise in another.
-environmental variation usually implies the environmental variable is simply independent of the -->
-
-Our variables of interest themselves
-
-The behavior of our variables depends on some quantity which itself is a
-
-<!-- Stochasticity is not inherent -->
+$$\frac{\mathrm{d} \langle n \rangle }{dt}  = b(\langle n \rangle) - d(\langle n \rangle)$$
+$$\frac{\mathrm{d} \sigma^2 }{dt}  = 2 \left( \frac{\partial b(n) }{\partial n} - \frac{\partial d(n) }{\partial n}\right) \sigma^2 + b(n) + d(n) $$
 
 <!-- 
+SDE notation
+
 $$\textrm{d} X_t = \mu(X_t) \textrm{d}t + \sigma(X_t) \textrm{d}B_t$$
@@
   ggplot(aes(x = time)) + 
-  geom_hline(aes(yintercept = mean), lty=2, col=colours[[2]]) + 
-  geom_hline(aes(yintercept = minus_sd), lty=2, col=colours[[2]]) + 
-  geom_hline(aes(yintercept = plus_sd), lty=2, col=colours[[2]]) + 
-  geom_line(aes(y = n), col=colours[[1]]) +
+  geom_hline(aes(yintercept = mean), lty=2, col=colours[2]) + 
+  geom_hline(aes(yintercept = minus_sd), lty=2, col=colours[2]) + 
+  geom_hline(aes(yintercept = plus_sd), lty=2, col=colours[2]) + 
+  geom_line(aes(y = n), col=colours[1]) +
   facet_wrap(~system_size, scales = "free_y") 
@@
 
+<!-- Say something about how small noise; particularly demographic noise which is usually small, seems like it should thus be negligible,
+
+**Noise need not be large to be important**.
+-->
+
+**Persistence and coexistence**.
+A natural focus of much research into stochastic ecological systems has been on understanding the implications for persistence of individual and coexistence of interacting species.
+Demographic noise in particular has fundamental implications for the long-term equilibrium: any finite population, however big, is guarenteed to eventually go extinct under demographic noise.
+While extinction is certainly a phenonomenon that stands in stark contrast to the long-term behavior of the corresponding deterministic equation, this result is not particularly useful: "eventually" could be longer than the existence of the planet itself.
+Rather, it is a reminder that we must be careful about how we describe the 'long-term' behavior of a model.
+In place of the mathematical notion of a stationary distribution to describe the long-term behavior of a stochastic model, analysis of demographic stochasticity focuses on the concept of quasi-stationary distributions, which in many cases correspond more closely to the equilibria of their deterministic counterparts [@Schreiber2017].
+
+<!--
+[@Schreiber2008] complex life histories
+- Species coexistence  ([@Hart2016])
+-->
+
+**Bet hedging**
+
+**Inflation**
+
+Stochasticity can
+
+<!-- two concurrent flat lines makes for a poor figure -->
+
 ## Quasi-cycles
@@
 
-We turn from stochastic resonance to oscillations of a very different sort, which do not resemble any deterministic process, as seen in Figure 3.
+We turn from stochastic resonance to oscillations of a very different sort, which do not resemble any deterministic process, as seen in Figure 3A.
 
@@
 
-panel1 <- read_csv("appendixA/noisy_switch.csv")  %>%
+p <- list(r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 1e4)
+
+theory <- 
+  tibble(x= seq(0,2, length.out = 100)) %>%
+  mutate(growth = x * p$r * (1 - x / p$K), 
+         consumption = p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)) %>%
+  mutate(potential = - cumsum(growth - consumption)) %>%
+  gather(curve, y, -x, -potential)
+
+p1 <- read_csv("appendixA/noisy_switch.csv")  %>%
   ggplot(aes(t,x)) + 
@@
 
-p <- list(r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 1e4)
-f <- function(x) x * p$r * (1 - x / p$K)
-g <- function(x) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
-x <- seq(0,2, length.out = 100)
-
-panel2 <- data.frame(x, f = f(x), g = g(x)) %>%
-  ggplot(aes(x, y)) + 
-  geom_line(aes(y = f), col=colours[1], lwd=1) +
-  geom_line(aes(y = g), col=colours[2], lwd=1) + 
+p2 <-  theory %>%
+  ggplot(aes(x, y, col=curve)) +
+  geom_line(lwd=1) + 
+  scale_color_ptol() + 
   labs(x = bquote(X[t]), y = bquote(X[t+1]), subtitle= bquote(bold(B.)))
 
-library(gtable)
-library(grid) # low-level grid functions are required
-g1 <- ggplotGrob(panel1)
-g2 <- ggplotGrob(panel2)
-g <- cbind(g1, g2, size="first") # stack the two plots
+p3 <-  theory %>%
+  ggplot(aes(x, potential)) + 
+  geom_line(col=colours[1], lwd=1) + 
+  coord_cartesian(xlim=c(0,1.6), ylim=c(-1.15,-0.9)) + 
+  labs(subtitle = bquote(bold(C.)))
+
+gridExtra::grid.arrange(p1,p2,p3, layout_matrix = rbind(c(1,1), c(2,3)))
+
 #g$widths <- unit.pmax(g1$widths, g2$widths) # use the largest widths
-grid.newpage()
-grid.draw(g)
+#grid.newpage()
+#grid.draw(g)
 ```
@@
 This behavior arises through the interaction of stochasticity and alternative stable state dynamics.
-The simulation shown in Figure 3 is based on an herbivore consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics, to which I have simply introduced an additional environmental noise term:
+The simulation shown in Figure 3A is based on an herbivore consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics, to which I have simply introduced an additional environmental noise term:
 
-$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{Vegitation growth}}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{Vegitation consumption}} + \xi_t,$$
+$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{Vegitation growth, } g(X_t)}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{Vegitation consumption, } c(X_t)} + \xi_t,$$
 where $\xi_t$ is normally distributed environmental variation with mean zero and standard deviation $\sigma$.
+Figure 3B shows separate curves corresponding the parts of the model describing vegetation growth and vegetation consumption, using the same parameters as shown in the simulation panel, 3A.
+Equilibria exist whenever growth balances consumption.
+For the parameters shown, the S-shaped curve of the Type-III consumption pattern crosses the quadratic curve of logistic growth at four separate locations.
+When growth is larger than consumption before the crossing, and thus smaller after, the point is stable, otherwise it is unstable.
+Consequently, we can see two stable points, in the neighborhood of $X_t \approx 1.2$ and $X_t \approx 0.5$, closely matching the average fluctuations seen in the simulation in 3A.
+Stochastic fluctuations drive spontaneous shifts between these alternative stable states.
 
-Stochastic fluctuations can also drive spontaneous shifts between alternative stable states.
+The concept of a potential well, Figure 3C, is often invoked when describing the behavior of alternative stable states.
+The curve for the potential well, $U(x)$ is defined as the negative integral of the population growth rate,
 
-[@Boettiger2012]
+$$U(x) = -\int_0^x f(y) \textrm{d} y $$ 
+where in our model $f(y)$ is vegetation growth minus consumption, $f(y) = g(y) - c(y)$.
+The motivation comes from physics, where $f$ is a force and $U$ the corresponding potential energy.
+This suggests a convenient analogy to gravity, often implied by drawing a ball inside the curve and referring to this as a "ball-in-cup" diagram.
+The potential function is particularly instructive for understanding stochastic switching under small noise, as in our simulation in Figure 3, where it is straight-forward to show [e.g.
+@Feller; @vanKampen2007] that the probability of a transition is proportional the exponential of the depth of the well from the lowest point to the separation barrier, (a result that is familiar as Arrhenius law of activation energy to grade-school students of chemistry).
+This is effect is visible even in the short simulation shown in 3A, which spends longer times in the lower state which has the deeper well.
+(Readers are encouraged to experiment with this by adjusting the parameters of the simulation in the included code).
+Unfortunately, the literature is not always careful in describing how this potential well diagram arises, which has caused it to become a source for as least as much confusion as it is of insight.
+While the curves in Figure 3B can be clearly associated with underlying mechanisms of rates: logistic growth and consumption with handling time, the same is not true of the potential: it is much harder to have a good intutition about the 'negative integral of the difference growth and consumption,' and how it would change in response to, say, a larger carrying capacity $K$ or higher consumption rate $a$.
+While the x-axis of the potential diagram remains the same: the system state, the conceptual problem is clearly evident in the usual ambiguity describing the y-axis in such ball-in-cup models.
 
-[@Drake2013]
+Despite the nearly ubiquitous association of a ball-in-cup model with ecological discussions of resilience and stability, this potential well diagram (Figure 3C) is also a source of confusion on the topic.
+As we have seen, the most salient feature of this diagram is the depth of each well: transistion probabilities (under small noise) depend only on the depth of the well and not the steepness of the well, the distance between the wells or between well and tipping point barrier.
+However, it is important to remember that because the potential function is defined by the above integral, it in fact already combines two more elementary and mechanistically precise quantities from the growth equation: the slope of the growth rate (stability) and the distance to the tipping point (size of the basin of attraction).
+Both a steeper negative slope or a longer distance to integrate over between stable and unstable points will create a deeper well.
 
-- transition probabilities, waiting times
+Empirical evidence for stochastic switching in ecological timeseries is limited by the long time scale usually required.
+Because a single stochastic transisition between alternative stable states appears as such a dramatic deviation from the previous pattern (as illustrated in Figure 3A), it is often tempting to assume the change could only be the result of some sudden external force or change in the environment, so it is worth remembering that given long enough, these transitions can happen entirely by chance stochasticity [@Boettiger2012; @Drake2013].
+However, these transistions are most likely to observed in systems with relatively shallow potential wells.
+This is mostly likely to occur of some external change is slowly making the potential well shallower and thus transitions more likely -- a phenomenon that has become to be known as "flickering" and suggested as a potential early warning sign of an approaching transition [@Brock2010].
+Examples of flickering between alternative stable states have been suggested in data ranging from geological timescales at the end of glaciation in the Pleistocene [@Taylor1993] to ecological scales over a few decades in lake ecosystems [@Wang2012].
+Stochastic transitions between bistable states are thought to play an important role in genetic switches, where among other phenomena this mechanism can explain the emergence of two alternative phenotypes in clonally reproducing microbial populations, which can act to buffer the population in a changing environment  [@Balaban2004].
 
-**Inflation**
-
-Stochasticity can
-
-<!-- two concurrent flat lines makes for a poor figure -->
-
-**Persistence and coexistence**
-
-One of the most fundamental aspects of stochasticity are the implications for long-term persistence of any given population and the stable coexistence of interacting species.
-
-**Bet hedging**
+<!--
 
@@
 
-@Read2007 <!-- Stochasticity generates an evolutionary instability for infectious disease -->
+- @Read2007  Stochasticity generates an evolutionary instability for infectious disease 
+- Spatial pattern formation: Turning patterns found in plankton-herbivore dynamics [@Levin1976]
 
-[@Schreiber2008] decreasing variance before crash
-
-- Spatial pattern formation
-
-Turning patterns found in plankton-herbivore dynamics [@Levin1976]
-
-- Population persistence (bet-hedging)
-
-Individual heterogeneity: spatial spread? persistence?
-
-- Species coexistence  ([@Hart2016])
+-->
 
@@
   select(t, population = x, tip_time) %>%
-  ggplot(aes(t, population)) + geom_line(col=colours[[1]]) + 
-  geom_vline(aes(xintercept = tip_time), col=colours[[2]], lty=2) + 
+  ggplot(aes(t, population)) + geom_line(col=colours[1]) + 
+  geom_vline(aes(xintercept = tip_time), col=colours[2], lty=2) + 
   #scale_y_log10()  + 
@@
 
-library(cowplot)
-ggdraw() +
-  draw_plot(p1, x = 0, y = .5, width = 1, height = 0.5) +
-  draw_plot(p2, x = 0, y = 0, width = 1, height = 0.6)
+gridExtra::grid.arrange(p1,p2, layout_matrix = rbind(c(1,1), c(2,2)))
+
 ```

### 2017-10-31 appendixB/appendixB.Rmd
WIP

@@
     address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
-#layout: 5p  
+layout: 5p  
 bibliography: canonical.bib
-output: 
-  rticles::elsevier_article:
-    includes:
-      in_header: header.tex
-
+output: rticles::elsevier_article
+preamble: "\\newcommand{\\ud}{\\mathrm{d}}"
 ---
-
-<!--
-## The Poisson process description
-
-The Poisson process is a common description of demographic noise.
-It arises as the limit of a binomial process (coin flipping) when the events are rare and the population large.
-This model is commonly applied as follows.
-Consider a model of the form:
-
-\begin{align}
-N_{t+1} = f(N_t)
-\end{align}
-
-This approach chooses with probability $P$ the number in the next generation $N_{t+1}$ at random from a Poisson distribution with mean $f(N)$:
-
-\begin{align}
-P(N_t)_{t+1} = \text{Poisson}(N, f(N_t)), 
-\end{align}
-
-while the variance of the Poisson distribution will also be $f(N)$.
-If the population has some equilibrium size $N^* \neq 0$ such that $f(N^*) = N^*$, then at equilibrium the population will have size $N^*$ and fluctuate around that value with standard deviation $\sqrt{N^*}$.
-In the continuous time limit this approaches a Gaussian with the same mean and variance.
-Observe that $P(0) > 0$ for any $N$, hence a population can always go extinct.
-Further, if we naturally assume the population is isolated, then $f(0) = 0$, and extinction is the only long-term stable state.
-This description provides a first pass at describing demographic noise, and captures some of its important phenomena:
-
-- Finite populations fluctuate around their quasi-equilibrium.
-- Any isolated population that cannot grow without bound will eventually go extinct.
--  The magnitude of the steady-state fluctuations relative to the equilibrium decreases with increasing population size (as $\sqrt{N^*}/N^*$), decreasing the extinction risk of larger populations proportionally.
-
-While we have started to capture one aspect of demographic noise (populations are finite), we have ignored the second (population dynamics come from birth and death processes).
-Too often we stop at this description of demographic noise.
-By using only $f(N)$, we have not specified births and deaths separately.
-This is quite common in population models, where the choice of $f(N)$ does not obviously suggest which are the birth and which are the death terms.
-Our second introductory example illustrates the importance of doing so.
--->
 
@@
 
+This probability balance is known for historical reasons as a master equation.
+This equation will form the center of our treatment.
+Master equations of this form can be written more concisely by introducing the step operator, $\mathbb E^k$ such that $\mathbb E^k f_n = f_{n+k}$, giving us
+
 \begin{figure}
@@
 \includegraphics[width=.25\textwidth]{markov}
-\caption{The birth-death process as a Markov process.}\label{markov}
+\caption{Birth and death as state-dependent transition rates in a Markov process.}\label{markov}
 \end{center}
 \end{figure}
-
-This probability balance is known for historical reasons as a master equation.
-This equation will form the center of our treatment.
-Master equations of this form can be written more concisely by introducing the step operator, $\mathbb E^k$ such that $\mathbb E^k f_n = f_{n+k}$, giving us
 
@@
 &= - \langle d_n \rangle +  \langle b_n \rangle \label{mean} \\
-\intertext{and similarly}
+\intertext{and similarly the second moment is}
 \frac{\ud }{\ud t} \langle n^2 \rangle &= 2 \langle n (b_n - d_n) \rangle + \langle b_n+d_n \rangle \label{var}
@@
 This is known as the macroscopic equation, and corresponds to the density equations commonly written down.
-We call this difference $\alpha_1(\phi)$ as a shorthand\footnote{This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.
+Following @vanKampen, we will call this difference $\alpha_1(\phi)$ as a shorthand\footnote{This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.
 It so happens this term will give the macroscopic law for any Markov process, not only a birth-death process.
@@
 
-Where we define $\alpha_2(\phi) = b(\phi)+d(\phi)$ as an analogous short hand.
+Where we define $\alpha_2(\phi) = b(\phi)+d(\phi)$ as an analogously.
 From this it is a straight forward exercise to calculate the moments of the distribution (multiply by $\xi$, $\xi^2$ and integrate),
@@
 From \eqref{n mean} we can also conclude that the macroscopic (average) variable obeys the deterministic law.
+
+## Stochastic Differential Equations
+
+The partial differential equation (PDE) given by \eqref{FP} is sometimes referred to as the Fokker-Planck equation.
+It is possible to prove this PDE is equivalent to as the stochastic differntial equation (SDE),
+
+\begin{equation}
+\textrm{d} X_t = \mu(X_t) \textrm{d}t + \sigma(X_t) \textrm{d}B_t \label{SDE}
+\end{equation}
+
+which, despite the name, it is important to remember that this is actually an integral equation in the Itô calculus, and not a derivative.
+Note the separation of the differential elements, which are meant as shorthand for implicit integrals:
+
+\begin{equation}
+X_{t+s} - X{t}  = \int_t^{t+s} \mu(X_{\tau}, \tau) \textrm{d}\tau + \int_t^{t+s} \sigma(X_{\tau}, \tau) \textrm{d}B\tau
+\end{equation}
+
+The first integral is the ordinary Lebesgue integral, while the second is an Itô integral, which integrates over the stochastic process $B_t$ (Brownian motion, also termed a Weiner process).
+The Itô calculus follows different rules from the classic calculus, largely due to the fact that an Itô integral has non-zero quadratic variation.
+(Crudely put, $\textrm{d}x^2 \sim 0$, but $\textrm{d}B^2 \sim dx$).
+Though it is common to formulate stochastic models directly as SDEs, it is worth remembering that the SDE is the result of a small noise approximation in the system size expansion, and can break down under larger deviations, even when the mean-field equation remains approximately accurate.
+@Ovaskainen2010 provides an overview literature in that addresses this limit and discusses the use of the WKB^[Wentzel, Kramers & Brillouin.
+Naturally this is KWB in Holland, BWK in France, and JWKB in England (for Jeffreys).
+Note that the physicist Gregor Wentzel is not the same as the mathematician Alexander Wentzell of the Fredilin-Wentzell theorem for large deviations years later, though I believe the former's result can be derived directly from the latter's.] method to address this case.
 
@@
 \begin{align}
-\dot n = c n \left(1 - \frac{n}{N}\right) - e n \label{levins}
+\frac{\ud n}{\ud t} = c n \left(1 - \frac{n}{N}\right) - e n \label{levins}
 \end{align}
@@
 
-# Alternate derivation: the diffusion approximation
+<!-- Include environmental noise example via stochastic K? -->
+
+# An alternate derivation: the diffusion approximation
 

### 2017-10-31 paper.Rmd
WIP

@@
   - name: "Carl Boettiger"
-    affiliation: a, 1
+    affiliation: a
+    email: "cboettig@berkeley.edu"
+
 address:
@@
 
-layout: preprint
+csl: ecology-letters.csl
+bibliography: refs.bib
+output: rticles::elsevier_article  
+layout:  5p # review = doublespace, 3p = singlespace, 5p = two-column
+#preamble: "\\usepackage[nomarkers]{endfloat}"
+
 abstract: |
@@
 
-lineno: true
-
-bibliography: refs.bib
-output: rticles::elsevier_article
 ---
@@
 In the code provided, most stochastic models are expressed in the BUGS language, which may be more familiar to empirical readers than corresponding mathematical formulas.
-This also permits both efficient simulation and potential estimation of parameters given sample data using the R package, NIMBLE [@DeValpine2017].
+This also permits both efficient simulation and potential estimation of parameters given sample data using the R package, NIMBLE [@DeValpine2017]; allowing the deductive models illustrated here to readily function as inductive models with parameters inferred from time-series data.
 A copy of this appendix is maintained at <https://github.com/cboettig/noise-phenomena>, and bug reports, suggestions or help requests are welcome through the issue tracker.
@@
 
-What is noise?
-Where does it come from?
-It is important to bear in mind that whenever stochasticity is introduced into a model, it is only as a simplification or summary of the effects of a large number of other processes, often occurring at different scales.
-Simply put, *noise is all that we leave out of our model*.
-Frequently, this additional noise is tacked on to the end of a model, as if as an afterthought,
+What is noise? Where does it come from?  It is important to bear in mind that whenever stochasticity is introduced into a model,
+it is only as a simplification or summary of the effects of a large number of other processes, often occurring at different scales.
+Simply put, *noise is all that we leave out of our model*.  Early quantiative ecological models first formulated at the population level
+as differential or difference equations: deterministic models that seek to capture the essential features of a system while neglecting
+any random component as mere noise [@Coulson2004]. Advances in both theoretical understanding and the advent of widely available and
+efficient numerical methods have helped drive research towards increasingly more explicit descriptions of how stochastiticity enters
+into quantiative models [@Black2012; @Ovaskainen2010].
 
-$$X_{t+1} = f(X_t) + \xi_t$$
-
-with the assertion that $\xi$ is some normally distributed noise term reflecting the combined effects of any large array of possible factors that also influence the value of our state, $X$.
-While not always wrong, this hand-waving over-simplification is increasingly replaced with a more ecologically and mechanistically grounded description for the origins of noise.
-With the exception of quantum mechanics (and there only at scales that have no relevance to ecology), stochasticity is not an inherent property of nature but rather an emergent property of many events occurring at a different scale.
-Coins and dice obey the physical laws of motion; only variation in how they are tossed (i.e.
-something complex we do not usually model) creates randomness.
-The term "noise" itself reflects this emergent rather than fundamental origin: the combination of hundreds of voices and sounds each individually distinct blends together into a background of noise.
-Statistical mechanics tells us that temperature is a measure of the average speed of individual molecules in a gas -- any given molecule will experience a range of possible speeds but by itself has no notion of temperature.
-Likewise, stochasticity enters our ecological models because our quantities of interest, such as the density of a population, is but an average over a lower-level, discrete process of individual births and deaths.
-Stochasticity that emerges directly from discrete individual events that underlie the process described by our model are considered *intrinsic stochasticity*, commonly referred to as *demographic stochasticity* in ecological literature, as it usually arises in the context scaling from individual demographic events to population level descriptions.
-An important aspect of this demographic stochasicity is that it scales with the square root of the "system size," so that for very large populations the resulting fluctuations are relatively small.
-On the other hand, *extrinsic stochasticity*, usually referred to as *environmental stochasticity* in the ecological literature, arises in processes that are not described by our model, but merely impact the value of some parameter of our model; for instance, fluctuations in temperature may drive variation in a death rate, or fluctuations in rainfall drive variation in carrying capacity.
+With the exception of quantum mechanics (and there only at scales that have no relevance to ecology), stochasticity is not an inherent
+property of nature but rather an emergent property of many events occurring at a different scale.  Coins and dice obey the physical
+laws of motion; only variation in how they are tossed (i.e. something complex we do not usually model) creates randomness.  The term
+"noise" itself reflects this emergent rather than fundamental origin: the combination of hundreds of voices and sounds each individually
+distinct blends together into a background of noise. Statistical mechanics tells us that temperature is a measure of the average speed
+of individual molecules in a gas -- any given molecule will experience a range of possible speeds but by itself has no notion of temperature.
+Likewise, stochasticity enters our ecological models because our quantities of interest, such as the density of a population, is but an
+average over a lower-level, discrete process of individual births and deaths. Stochasticity that emerges directly from discrete individual
+events that underlie the process described by our model are considered *intrinsic stochasticity*, commonly referred to as
+*demographic stochasticity* in ecological literature, as it usually arises in the context scaling from individual demographic events to
+population level descriptions. An important aspect of this demographic stochasicity is that it scales with the square root of the
+"system size," so that for very large populations the resulting fluctuations are relatively small. On the other hand, 
+*extrinsic stochasticity*, usually referred to as *environmental stochasticity* in the ecological literature, arises in processes
+that are not described by our model, but merely impact the value of some parameter of our model; for instance, fluctuations in
+temperature may drive variation in a death rate, or fluctuations in rainfall drive variation in carrying capacity.
 
 <!--
-Early ecological models were almost always formulated at the level of the population, described in terms of differential or difference equations reflecting the available tools of the time.
-The advent of fast and inexpensive computing power
+Motivate the presentation of the van Kampen result rather than a generic review.
 -->
 
-@Black2010 provides an excellent review of the importance of intrinsic stochasticity and the evolution
+## Demographic stochasticity
 
+Demographic stochasticity refers to fluctuations in population sizes or densities that arise from the fundamentally discrete nature of individual birth and death events.
+Demographic stochasticity is a particularly instructive case for illustrating a mechanism for how noise arises as an aggregate description from a lower-level mechanistic process.
+We summarize the myriad lower-level processes that mechanistically lead to the event of a 'birth' in the population as a probability: in a population of $N$ identical individuals at time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate that can depend on both the population size $N$) and increases the population size, $N$ to $N+1$, and deaths occur with probability $d_t(N_t)$, decreasing the population size, $N$ to $N-1$.
+Assuming each of these events are independent, this is a state-dependent Poisson process.
+The change in the probability of being in state $N$ is given by the sum over the ways to enter the state, minus the ways to leave the state: a simple expression of probability balance known as the "master equation."
+
+It is possible to simulate such a process exactly using Gillespie's exact stochastic simulation algorithm [SSA; @Gillespie1977], which simulates
+each individual birth or death as a separate event.
 Consider Levins' patch model, where $n$ individuals compete for a finite number of suitable habitats $N$.
-Individuals die a constant rate $e$, and produce offspring at a constant rate $c$ who then have a probability of colonizing an open patch that is simply proportional to the fraction of available patches, $1 - n/N$.
-
-Approximations of the so-called master equation, van Kampen system size approximation [@vanKampen2007] described in @Black2010, (rooted in the theorems of [@Kurtz1970; @Kurtz1971]),
-
-$$\frac{\mathrm{d} \langle n \rangle }{dt}  = b(\langle n \rangle) - d(\langle n \rangle)$$
-$$\frac{\mathrm{d} \sigma^2 }{dt}  = 2 \left( \frac{\partial b(n) }{\partial n} - \frac{\partial d(n) }{\partial n}\right) \sigma^2 + b(n) + d(n) $$
-
-<!-- 
-SDE notation
-
-$$\textrm{d} X_t = \mu(X_t) \textrm{d}t + \sigma(X_t) \textrm{d}B_t$$
-
-An SDE. Note the separation of the differential elements, which reveal that this is actually shorthand for an integral equation:
-
-$$X_{t+s} - X{t}  = \int_t^{t+s} \mu(X_{\tau}, \tau) \textrm{d}\tau + int_t^{t+s} \sigma(X_{\tau}, \tau) \textrm{d}B\tau$$
-
-The first integral is the ordinary Lebesgue integral, while the second is an It\^o integral, which integrates over the stochastic process $B_t$ (Brownian motion, also termed a Weiner process).
-The It\^o calculus follows different rules from the classic calculus, largely due to the fact that an It\^o integral has non-zero quadratic variation.
-(Crudely put, $\textrm{d}x^2 \sim 0$, but $\textrm{d}B^2 ~ dx$).
-
--->
-
-While ecologists will sometimes assert that stochasticity is inherent: birth and death are inherently stochastic events; or the environment is inherently stochastic, these are only assertions at a particular scale.
-Obviously events such as births and deaths are themselves the results of mechanistic processes, just as fluctuations in rainfall and temperature are the result of deterministic laws of physics and not fundamentally random forces.
-In all of science, only quantum mechanics claims a fundamentally probabilistic view of events, and then at such scales that have no relevance to ecology.
-<!-- Our state variables don't "exist" except as averages of a lower level process -->
-
-**Demographic stochasticity**.
-Demographic stochasticity refers to fluctuations in population sizes or densities that arise from the fundamentally discrete nature of individual birth and death events.
-Demographic stochasticity is a particularly instructive case for illustrating a mechanism for how noise arises as an aggregate description from a fundamentally lower-level mechanistic process.
-We summarize the myriad lower-level processes that mechanistically lead to the event of a 'birth' in the population as a probability: in a population of $N$ identical individuals at time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate that can depend on both the population size $N$) and increases the population size, $N$ to $N+1$, and deaths occur with probability $d_t(N_t)$, decreasing the population size, $N$ to $N-1$.
-Assuming each of these events are independent, this is a Poisson process.
-
-<!-- Master equation solution of birth-death model, from appendix-->
-
-<!-- Gillespie algorithm; Population level (SDE) vs agent-based vs individual. Show figure with small and large N  -->
+Individuals die a constant rate $e$, and produce offspring at a constant rate $c$ who then have a probability
+of colonizing an open patch that is simply proportional to the fraction of available patches, $1 - n/N$. 
+Figure 1 shows the results of two exact SSA simulations of this model with identical parameters except for the
+total number of available sites, $N$.
 
@@
 
-[@Black2012]
+The van Kampen system size expansion provides a convenient way to explicitly approximate the "macroscopic" (i.e.
+population level, see @Black2012) dynamics from an individual-based description of events in a Markov process [@vanKampen2007; based on @Kurtz1970; @Kurtz1971].
+In addition to illustrating how common stochastic models (in particular, Guassian noise) arise from lower-level discriptions, this result will also provide us a useful framework for interpeting the behavior of stochastic models more generally.
+In Appendix B I illustrate the derivation of the following equations for the mean and expected population sizes for a generic density dependent birth-death process:
 
-**Environmental stochasticity**.
-Whereas demographic noise reflects processes at a lower scale (the individual, rather than the population), environmental stochasticity reflects processes operating at the same scale that we are not explicitly modelling.
+\begin{align}
+\frac{\mathrm{d} x }{dt}  &= b(x) - d(x) \\
+\frac{\mathrm{d} \sigma^2 }{dt}  &= 2 \left( \frac{\partial b(x) }{\partial x} - \frac{\partial d(x) }{\partial x}\right) \sigma^2 + b(x) + d(x) \label{bd}
+\end{align}
+
+where $x = \langle n \rangle$, the average or expected population size of an ensemble of replicates.
+We can solve these equations at steady-state by setting both equations zero; e.g.
+for our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$, we find the average number of occupied patches $\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of $\sigma^2_n = N \tfrac{e}{c}$.
+Note that the variance scales proportionally to the system size, $N$, which explains the standard devation increases only as $\sqrt{N}$ while the steady state itself increases in direct propotion to $N$, resulting in the smaller relative size of fluctations in the larger system in Figure 1.
+
+<!-- Master equation solution of birth-death model, from appendix-->
+
+## Environmental stochasticity
+
+Whereas demographic noise reflects processes at a lower scale (the individual, rather than the population), environmental stochasticity reflects processes whose dynamics we are not explicitly modelling.
+A population-level model
 
@@
 [@Roughgarden; @Ellner]
+
+[@Ovaskainen2010] Master equation, break-down of Fokker-Planck approximation under large noise, introduction of the WKB approximation
 
@@
 
-Colored noise:
+Colored noise: [@Lawton]
 - [@Ripa1999]
@@
 
-<!-- Identify as beyond the scope of this paper -->
-<!-- non-temporal stochasticity -->
+### Other forms of stochasticity
 
-Several additional sources of stochasticity or variation more generally fall outside these categories demographic and environmental noise.
-While a detailed treatment is beyond the scope of this paper, I enumerate these sources to disambiguate such phenomena from the intrinsic and extrinsic noise sources that comprise the focus here.
-
-**Other forms of stochasticity**.
 Stochasticity frequently enters our models in other ways as well.
-For instance, most population-level models treat all individuals within the population, or at least within a particular age or stage class, as identical.
-*Individual heterogeneity* in relevant traits (growth rates, dispersal ability) can be represented probabilistically as a distribution rather than a fixed value, making it function as another form of random variation in a model.
-*Spatial heterogeneity* can likewise be reflected in stochastic parameters of a spatially implicit model, or as spatially varying environmental stochasticity in a spatially explicit model.
-As with other forms of noise [@Melbourne2008; @Bonsall2004], interaction between individual or spatial variation and other sources of noise can drive patterns including the potential for persistence, coexistence, and dispersal of species [@Schreiber2010; @Hart2016; @Schreiber2017].
-
-**Observational error**: Observational error acts much like another form of stochasticity, arising from
+For instance, most population-level models treat all individuals within the population, or at least
+within a particular age or stage class, as identical.  *Individual heterogeneity* in relevant traits
+(growth rates, dispersal ability) can be represented probabilistically as a distribution rather than
+a fixed value, making it function as another form of random variation in a model [@Coulson2001; @Schreiber2010].
+*Spatial heterogeneity* can likewise be reflected in stochastic parameters of a spatially implicit
+model, or as spatially varying environmental stochasticity in a spatially explicit model.  As with
+other forms of noise [@Melbourne2008; @Bonsall2004], interaction between individual or spatial
+variation and other sources of noise can drive patterns including the potential for persistence,
+coexistence, and dispersal of species [@Schreiber2010; @Hart2016; @Schreiber2017].
+*Observational error*: Observational error acts much like another form of stochasticity, arising from
 sampling effects or limitations of instruments, etc. Yet unlike the sources described above, this 
@@
 Other forms of uncertainty common in models are not stochastic phenomena.  This includes any form of 
-**reducible uncertainty**: In constructing and estimating mathematical models of ecological processes, 
+*reducible uncertainty*: In constructing and estimating mathematical models of ecological processes, 
 we encounter uncertainty over our parameter estimates and possible choice of models [@Ellison2004].  In some ways, this
@@
 inherent source of variation. We will this not consider model and parameter uncertainty, as they are not 
-sources of noise.
+sources of noise.  *Deterministic chaos* [@Desharnais2001] is another mechanism of introducing significant
+variation into a simple model. Chaos can a look a lot like noise: indeed, it can be observationally
+indistinguishable from a stochastic process [@Ellner2005; @Werndl2009].  Despite substantial interest in previous
+decades and the relative ease of encountering chaos in simple ecological models (particularly in
+differences equations), and convincing reconstructions of chaotic dynamics in laboratory manipulations
+[@Costantino1997] the prevalence of chaotic dynamics as a driving factor in most ecological systems
+remains unclear. After all, the great variation of natural populations is hardly surprising: the
+undeniable influences of variation created by all those things left out of our models, all that
+individual detail from intrinsic, lower level processes and external environmental processes that
+becomes aggregated in our simple models all but begs for a stochastic description.
 
-**Chaos**: Deterministic chaos [@Desharnais2001] is another mechanism of introducing significant variation into a simple model.
-Chaos can a look a lot like noise: indeed, it can be observationally indistinguishable from a stochastic process [@Werndl2009].
-Despite substantial interest in previous decades and the relative ease of encountering chaos in simple ecological models (particularly in differences equations), and convincing reconstructions of chaotic dynamics in laboratory manipulations [@Costantino1997] the prevalence of chaotic dynamics as a driving factor in most ecological systems remains unclear.
-After all, the great variation of natural populations is hardly surprising: the undeniable influences of variation created by all those things left out of our models, all that individual detail from intrinsic, lower level processes and external environmental processes that becomes aggregated in our simple models all but begs for a stochastic description.
-
-Much active research continues to explore the consequences of introducing these more explicitly motivated and interacting sources of noise.
-With some grounding in the origins of different types of noise in place, we are ready to move to our second stop to see how this noise can drive the creation of novel phenomena on a macroscopic level.
+Much active research continues to explore the consequences of introducing these more explicitly
+motivated and interacting sources of noise. With some grounding in the origins of different types
+of noise in place, we are ready to move to our second stop to see how this noise can drive the
+creation of novel phenomena on a macroscopic level.
 
@@
 Stochasticity can give rise to a wide array of interesting phenomena that are often at odds with our intuition.
-Such behavior often arises through an interaction between stochastic and non-linear elements of a system, both hallmark traits of ecological processes.
+As these examples will illustrate, such behavior arises through an interaction between stochastic and non-linear elements of a system, underscoring the importance of explicitly including stochasticity in ecological models [@Coulson2004; @Higgins1997].
+Importantly, the noise itself does not have to be large to give rise to these phenomena.
 In this section, I highlight several examples that have shown up frequently in the literature and prove sufficiently tractable to provide a simple intuition for the mechanisms involved.
 
-<!-- Say something about how small noise; particularly demographic noise which is usually small, seems like it should thus be negligible,
-
-**Noise need not be large to be important**.
+<!-- 
+Noise is "tacked on" in these models... contrasts to previous section
 -->
 
-**Persistence and coexistence**.
+## Persistence and coexistence
+
 A natural focus of much research into stochastic ecological systems has been on understanding the implications for persistence of individual and coexistence of interacting species.
@@
 Rather, it is a reminder that we must be careful about how we describe the 'long-term' behavior of a model.
-In place of the mathematical notion of a stationary distribution to describe the long-term behavior of a stochastic model, analysis of demographic stochasticity focuses on the concept of quasi-stationary distributions, which in many cases correspond more closely to the equilibria of their deterministic counterparts [@Schreiber2017].
+In place of the mathematical notion of a stationary distribution to describe the long-term behavior of a stochastic model, analysis of demographic stochasticity focuses on the concept of quasi-stationary distributions, which in many cases correspond more closely to the equilibria of their deterministic counterparts [@Ovaskainen2010; @Schreiber2017].
 
 <!--
-[@Schreiber2008] complex life histories
-- Species coexistence  ([@Hart2016])
--->
+
+Insufficient space left for any treatment of these...
 
 **Bet hedging**
-
 **Inflation**
 
-Stochasticity can
-
-<!-- two concurrent flat lines makes for a poor figure -->
+-->
 
@@
 This process is sufficient to permit even small noise to drive sustained oscillations in the model indefinitely.
-
 The power spectrum (the squared norm of the Fourier transform, a measure of what frequencies are present; see, e.g.
@@
 @Bjornstad2004 analyzes the potential for such stochastic effects to drive cyclic behavior in population abundances of Atlantic bluefin tuna and Atlantic Cod, while @Black2010 illustrates the use of power-spectrum approaches to examine the behavior of quasi-cycles in common disease outbreak models, focusing on a case where noise arises explicitly from demographic stochasticity, rather than generic environmental noise as in the model above and elsewhere.
-@Pineda-Krch2007 compares patterns from these stochastically generated cycles to truly stable limit cycles they resemble.
-
+@Pineda-Krch2007 compare patterns from these stochastically generated cycles to stable limit cycles they resemble.
 Stochastic resonance highlights one way in which even small noise can play an important role in driving large scale patterns where we had previously only sought deterministic explanations.
@@
 
-<!-- variable noise -->
-
-<!--autocorrelated or red noise can also drive stochastic resonance -->
+<!-- See additional citations in Coulson2004 and others that mention stochastic resonance -->
 
@@
 <!--Figure 3: Noisy switch -->
-```{r noisy_switch, fig.width=7, fig.cap="Stochastic switiching in May's model of alternative stable state dynamics.  r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000"}
+```{r noisy_switch, fig.width=7, fig.height=7, fig.cap="Stochastic switiching in May's model of alternative stable state dynamics.  r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000"}
 
@@
   tibble(x= seq(0,2, length.out = 100)) %>%
-  mutate(growth = x * p$r * (1 - x / p$K), 
-         consumption = p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)) %>%
-  mutate(potential = - cumsum(growth - consumption)) %>%
+  mutate(g = x * p$r * (1 - x / p$K), 
+         c = p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)) %>%
+  mutate(potential = - cumsum(g - c)) %>%
   gather(curve, y, -x, -potential)
@@
   scale_color_ptol() + 
-  labs(x = bquote(X[t]), y = bquote(X[t+1]), subtitle= bquote(bold(B.)))
+  labs(x = bquote(X[t]), y = bquote(X[t+1]), subtitle= bquote(bold(B.))) +
+  theme(legend.position="top")
 
@@
 gridExtra::grid.arrange(p1,p2,p3, layout_matrix = rbind(c(1,1), c(2,3)))
-
-#g$widths <- unit.pmax(g1$widths, g2$widths) # use the largest widths
-#grid.newpage()
-#grid.draw(g)
 ```
@@
 This behavior arises through the interaction of stochasticity and alternative stable state dynamics.
-The simulation shown in Figure 3A is based on an herbivore consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics, to which I have simply introduced an additional environmental noise term:
+The simulation shown in Figure 3A for the is based on an resource consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics.
+The resource grows according to a logistic growth model and is consumed according to a Holling Type-III functional response curve, to which I have simply introduced the addition of environmental noise:
 
-$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{Vegitation growth, } g(X_t)}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{Vegitation consumption, } c(X_t)} + \xi_t,$$
+$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{ growth, } g(X_t)}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{consumption, } c(X_t)} + \xi_t,$$
 where $\xi_t$ is normally distributed environmental variation with mean zero and standard deviation $\sigma$.
@@
 <!--
-
 **Further examples**
-
 - @Read2007  Stochasticity generates an evolutionary instability for infectious disease 
@@
 One of the most interesting aspects of noise is as a source of additional information.
-Noise can act like a miniature perturbation experiment, constantly perturbing a system away from an equilibrium to reveal it's behavior elsewhere in state space.
+This can be realized in a variety of ways.
+For instance, in deriving expressions for the variance created by demographic noise above, we observed that different partitions of a logistic growth model can lead to the same macroscopic dynamics (same values for $b$ - $d$), while having different variance (due to different values for $b$+$d$).
+This suggests that the variation observed in such a population could resolve potentially help resolve how dynamics are partitioned between individual birth and death rates (though in practice one would have to first account for any contribution of environmental noise).
 
-- Measuring stability
+<!-- fixme need more background? -->
 
-**Early warning signals.**  The theory of early warning signals, or critical slowing down (CSD) [@Wissel1984; @Scheffer2009] is an instructive example
+Noise can provide novel information in other ways as well.
+Stochastic fluctuations act like a continual miniature perturbation experiment, constantly pusing a system away from an equilibrium to reveal it's behavior elsewhere in state space.
+Observing the size and autocorrelation of these fluctuations thus provides insight on the stability of an equilibrium point, (as can be seen, for instance in the fluctuation-disspation result we found for generic birth death models).
+This observation has been exploited with particular success as a potential *early warning sign* that a system is loosing stability; possibly due to an approaching catastrophic transition or saddle-node bifucation.
+This pattern, known as critical slowing down (in reference to the autocorrelation time in the physics literature) was first proposed an an early warning sign for ecological systems by @Wissel1984, but has recieved considerable empirical as well as theoretical attention only more recently [@Scheffer2009; @Scheffer2015].
+Increases in two of these early warning indicators, variance and autocorrelation, are illustrated along with the timeseries of a simulation of the dynamics of yeast population growth in Figure 4.
+This simulation is based on a mechanistic model developed by @Dai2012 that closely matches the population dynamics they observed in serial dilutions of budding yeast _Saccharomyces cervisiae_ on a sucrose medium.
+Because the relevant proteins are located in the extra-cellular matrix rather than inside the cells, budding yeast break down sucrose collaboratively, creating an Allee effect below which the population is too small to sustain itself.
+Increasing the serial dilution rate slowly over time drives the stable population towards a fold bifurcation, which is accompanied by the patterns of increasing variance and increasing autocorrelation seen empirically in lab populations [@Dai2012] as well as the numerical simulation shown here.
+
+As we saw in the equations from the system-size approximation, the variance and autocorrelation in question are technically properties of an ensemble, and should be computed over replicate simulations.
+While the laboratory experiments of @Dai2012 do precisely this, conducting identical, independent replicate manipulations in separate beakers, this is not an option in natural populations.
+Instead,
 
@@
 <!-- Figure 4: Early warning signals -->
-```{r tipping, fig.width=7, fig.height=7}
+```{r tipping, fig.width=7, fig.height=7, fig.cap="Early warning signs of a critical transition in a simulation of increasing serial dilution of budding yeast.
+Panel A: Rolling variance   autocorrelation Time series is shown in panel C.
+Model details and simulation parameters in Appendix A."}
 tipping <- read_csv("appendixA/tipping.csv")
@@
   geom_vline(aes(xintercept = tip_time), col=colours[[2]], lty=2) +
-  facet_wrap(~series, ncol = 2, scales="free_y")  +
-  theme_ipsum_rc() 
+  facet_wrap(~series, ncol = 2, scales="free_y")  
 p2 <- tipping %>% 
@@
   #scale_y_log10()  + 
-  theme_ipsum_rc() + 
   labs(subtitle = "C. Yeast population size", 
@@
 
-<!-- FIXME unclear, rambling!-->
+<!--
 It is important to realize that such approaches can only ever measure local properties such as the return time in response to local perturbation (historically termed *stability*; which @Holling1996 landmark chapter unfortunately re-brands as "engineering resilience") and not global properties such as the size of the attractor basin / distance to the nearest tipping point (*ecological resilience*, or simply *resilience*).
@@
 
+-->
+
 # Conclusions
 
+The past two decades have seen widespread recognition of the central role that stochasticity plays through the use of mechanistic models can have in creating and understanding ecological phenomena [@Coulson2004; @Black2012].
+We have seen increasing use of master-equation approaches to individual-based models to reflect demographic stochasticity that is amenable to analytical treatment [@Black2012], and much work in understanding the importance of environmental noise, particularly the more realistic but less mathematically convienient notion of colored or auto-correlated noise on population dynamics, persistence, and coexistence [@Ovaskainen2010; @Schreiber2017].
+These deductive models can generate phenomena that resemble familiar patterns such as population cycles for which we have previously sought determinstic explanations, as well as stochastic phenomena with no such analog such as stochastic switching.
+Yet too many of these advances have remained limited to the domain of theory; limited to deductive models which may seek to mimic observed patterns but are rarely used to confront empirical data directly.
+There is good reason for this: statistical inference of even relatively simple models with multiple sources of noise will always be difficult given limited ecological data [@Knape2011].
+However, many stochastic phenomena may suggest new ways to infer rather than merely obscure underlying processes.
+
+The eruption of interest in early warning signals for critical transitions [@Scheffer2009] provides an excellent illustration of this potential.
+A range of stochastic phenomena have been suggested and a tested for their potential to reveal subtle signs of ecosystem change prior to a dramatic shift, and prompted a rich dialog between further theory and new experiment.
+Changes in stochastic properties such as variance or autocorrelation provide evidence of resilience loss that would be invisible to a determinstic model, and non-linear stochastic phenomena such as stochastic switching are observed as evidence of 'flickering' prior to a critical transistion.
+In this way, our understanding of stochastic dynamics is translated into novel sources of information about a system which has sparked a wealth of emprical investigation and further theoretical work to test the potential of this approach.
+The study and application of stochastic phenomena to detect ecosystem change is unlikely to be limited to this simple scenario of a critical transistion or fold bifurcation.
+
+Precisely because most stochastic phenomena arise through their interaction with nonlinearity, the patterns that result can become a lens into better understanding those nonlinear processes and detecting how they may change.
+Meanwhile, the widespread adoption of computational tools and the ability to express theoretical results in numerical algorithms can make any new approaches more readily accessible to application in empirical research.
+In the past two decades, we have set to rest those deterministic skeletons that see noise only as nuciance and embraced the central role noise plays as the creator of ecological phenomena.
+Going forward, I hope and predict that we will see greater discussion and use of stochastic phenomena to infer underlying processes in empirical patterns in similar interactions between theoretical and empirical research.
+
 # References

### 2017-10-31 paper.Rmd
WIP

@@
 abstract: |
-  In this paper, I provide a brief synthesis of ecological research in stochastic phenomena.
+  I provide a brief synthesis of ecological research in stochastic phenomena.
 Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath.
@@
 
+<!--
 [@Roughgarden; @Ellner]
@@
 [@Ovaskainen2010] Master equation, break-down of Fokker-Planck approximation under large noise, introduction of the WKB approximation
-
-<!-- Demographic noise scaling away with population size, but not so for environmental noise -->
-
 [@Bjornstad2001] _Science_ Review of noisy fluctuations
-
-[@Melbourne2008] Importance of demographic stochasticity and its interaction with environmental stochasticity in driving extinction risk. 
-[@Bonsall2004] Markov chain environmental and demographic noise predator-prey metapopulation model.
 
@@
 - [@Marshall2015]
+-->
 
@@
 In place of the mathematical notion of a stationary distribution to describe the long-term behavior of a stochastic model, analysis of demographic stochasticity focuses on the concept of quasi-stationary distributions, which in many cases correspond more closely to the equilibria of their deterministic counterparts [@Ovaskainen2010; @Schreiber2017].
+
+<!-- FIXME this section is still a fragment
+
+[@Melbourne2008] Importance of demographic stochasticity and its interaction with environmental stochasticity in driving extinction risk. 
+[@Bonsall2004] Markov chain environmental and demographic noise predator-prey metapopulation model.
+
+-->
 

### 2017-10-31 paper.Rmd
WIP

@@
 layout:  5p # review = doublespace, 3p = singlespace, 5p = two-column
+preamble: "\\newcommand{\\ud}{\\mathrm{d}}"
+
 #preamble: "\\usepackage[nomarkers]{endfloat}"
@@
 For instance, few ecological concepts are more central than the notion of density dependence: that competition in some form or other should limit a species ability to grow without bound.
-Yet establishing statistical evidence for density dependence is remarkably difficult, as illustrated by a landmark study of @Knape2011.
-Using time series from the Global Population Dynamics Database (GPDD) @Knape2011 show that once one accounts for the potential for measurement error in addition to environmental stochasticity -- an unavoidable source of noise in most ecological data -- it becomes difficult to find evidence for even the simplest models of density dependence in the majority of available time series.
+Yet establishing statistical evidence for density dependence is remarkably difficult, as illustrated by @Knape2011.
+Using time series from the Global Population Dynamics Database (GPDD) Knape & de Valpine show that once one accounts for the potential for measurement error in addition to environmental stochasticity -- an unavoidable source of noise in most ecological data -- it becomes difficult to find evidence for even the simplest models of density dependence in the majority of available time series.
 Faced with such uncertainty, it is tempting to throw up our hands and wait until some prophesied deluge of streaming micro-sensors stapled to every organism provide enough signal to beat back the noise.
@@
 \begin{align}
-\frac{\mathrm{d} x }{dt}  &= b(x) - d(x) \\
-\frac{\mathrm{d} \sigma^2 }{dt}  &= 2 \left( \frac{\partial b(x) }{\partial x} - \frac{\partial d(x) }{\partial x}\right) \sigma^2 + b(x) + d(x) \label{bd}
+\frac{\mathrm{d} x }{dt}  &= b(x) - d(x) + O(N^{-1})\\
+\frac{\mathrm{d} \sigma^2 }{dt}  &= 2 \left( \partial_x b(x)  - \partial_x d(x) \right) \sigma^2 + b(x) + d(x) \label{bd}
 \end{align}
 
-where $x = \langle n \rangle$, the average or expected population size of an ensemble of replicates.
-We can solve these equations at steady-state by setting both equations zero; e.g.
-for our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$, we find the average number of occupied patches $\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of $\sigma^2_n = N \tfrac{e}{c}$.
-Note that the variance scales proportionally to the system size, $N$, which explains the standard devation increases only as $\sqrt{N}$ while the steady state itself increases in direct propotion to $N$, resulting in the smaller relative size of fluctations in the larger system in Figure 1.
+where $x = \langle n \rangle$, the average or expected population size of an ensemble of replicates, assuming the transitions (+/- 1 in this case) are much smaller than the system size ($N$ in the Levins model).
+Note that the dynamics for the average population match the population-level equation^[up to fluctuations of order $\tfrac{1}{N}$.
+We can actually both observe and successfully approximate this higher-order correction from additional terms in the system size expansion to recover the phenomenon of stochastic inflation [@Gonzalez2002], where the average population size can be larger or smaller than predicted by the deterministic model -- see Appendix A for examples in simulations compared to predictions of the system size expansion].
 
-<!-- Master equation solution of birth-death model, from appendix-->
+\begin{align}
+\frac{\ud n}{\ud t} = \underbrace{c n \left(1 - \frac{n}{N}\right)}_{\textrm{birth}} - \underbrace{e n}_{\textrm{death}}, \label{levins}
+\end{align}
+
+Note that like the equation for the mean, we have a dynamic equation for the time-evolution of the variance in Eq \eqref{bd}.
+We can solve these equations at steady-state by setting both equations zero.
+For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$, we find the average number of occupied patches $\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of $\sigma^2_n = N \tfrac{e}{c}$.
+Observe that the variance scales proportionally to the system size, $N$, which explains the standard devation increases only as $\sqrt{N}$ while the steady state itself increases in direct propotion to $N$, resulting in the smaller relative size of fluctations in the larger system in Figure 1.
+We can also see that different ways of partitioning the same macroscopic equation into components of birth and death will create the same mean dynamics but with different levels of noise.
+For instance, in a logistic model with $b = rn$ and $d = rn^2/N$, we instead find the variance at steady state is $\sigma^2 = K$; significantly larger than in the Levins model.
 
@@
 Whereas demographic noise reflects processes at a lower scale (the individual, rather than the population), environmental stochasticity reflects processes whose dynamics we are not explicitly modelling.
-A population-level model
+Once again, this can best be illustrated through an explicit formulation of the master equation.
+For a generic population model that depends not only on population state $x$ but also on environmental factors $y$, it is possible to derive the steady-state variance (Appendix B) as:
 
-Given ecologists natural inclination towards modeling the biological rather than environmental aspects of a system, it will be no surprise that mechanistic descriptions of environmental stochasticity have remained far more fuzzy.
-One step in this direction has been recognition that whatever the mechanisms, this variation is unlikely to result in purely white noise, as commonly postulated for mathematical convenience, but rather some degree of correlation, termed red noise, or possibly anti-correlation such as from seasonal or other oscillatory patterns, so called blue noise.
+$$\sigma_x^2  = \frac{ \left( \partial_y f(x,y) \right)^2}{\left( \partial_x f(x,y) \right)^2 + \partial_x f(x,y) \tau_c } \sigma_e^2 + \sigma^2_d$$
+
+where $\sigma_e^2$ is the variance in the environmental noise and $\tau_c$ the autocorrelation of that noise, $\sigma_d^2$ the equilibrium variance contributed by demographic noise from before, ($\tfrac{b + d}{-2\partial_x (b - d)}$) and where we have defined $f(x,y) = b(x,y) - d(x,y)$ for notational convenience.
+For example, if extinction rate $e$ in our Levins' model is allowed to vary stochastically as Gaussian white noise with variance $\sigma_e^2$ and mean $e$, we find the expected variance introduced in the population to be:
+
+$$\sigma_n^2 = \frac{e^2}{(c+e)^2} \sigma_e^2 + \frac{e}{c}N $$
+
+Note that the contribution of environmental stochasticity is independent of system size and directly proprotional to the variance in the environment.
+This derivation provides a useful illustration of the difference between demographic noise and environmental noise: the latter simply describes the variation introduced by some other dynamic variable that we are not modelling explicitly, and whose dynamics are independent of the state of our system, $x$.
+The correlation time $\tau_c$, is really a shorthand for $\tfrac{1}{\partial_y g(y)}$.
+Coulpling
 
@@
 
+<!-- Figure 2: Quasi-cycles -->
+```{r quasicycles, fig.width=7, fig.cap="Panel A: under negligible environmental noise, the introduction of a carrying capacity damps Lotka-Volterra oscillations which eventually settle to a stable state.
+Panel B: small noise induces quasi-cycles in the same model: sustained long-term oscillations predominately around the natural frequency of the original damped oscillations.
+Parameters for models shown are r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800; code in Appendix A."}
+read_csv("appendixA/quasicycles.csv") %>%
+  gather(species, population, -t, -sigma) %>%
+  mutate(sigma = as.factor(sigma)) %>%
+  mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
+  rename(time = t) %>%
+  ggplot(aes(time, population, col = species)) + geom_line() + 
+  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() 
+```
+
 ## Quasi-cycles
@@
 Figure 1A shows examples for noise $\sigma_x = \sigma_y = 10^{-5}$, while Figure 1B shows $\sigma_x = \sigma_y = 0.01$.
-
-<!-- Figure 2: Quasi-cycles -->
-```{r quasicycles, fig.width=7, fig.cap="Panel A: under negligible environmental noise, the introduction of a carrying capacity damps Lotka-Volterra oscillations which eventually settle to a stable state.
-Panel B: small noise induces quasi-cycles in the same model: sustained long-term oscillations predominately around the natural frequency of the original damped oscillations.
-Parameters for models shown are r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800; code in Appendix A."}
-read_csv("appendixA/quasicycles.csv") %>%
-  gather(species, population, -t, -sigma) %>%
-  mutate(sigma = as.factor(sigma)) %>%
-  mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
-  rename(time = t) %>%
-  ggplot(aes(time, population, col = species)) + geom_line() + 
-  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() #+ 
-#  labs(x="Time", 
-#       y="Population density",
-#       title="Noise induces quasi-cycles",
-#       caption="Left panel: neglible  noise results only in damped oscillations.\nRight panel: small noise creates sustained oscillations.")
-```
 
@@
 We turn from stochastic resonance to oscillations of a very different sort, which do not resemble any deterministic process, as seen in Figure 3A.
+This pattern is characterized by long stretches during which fluctuations in the state variable remain concentrated around a high average value followed by sudden, rapid transistions into fluctuations concentrated around a lower value.
+This behavior arises through the interaction of stochasticity and alternative stable state dynamics.
+The simulation shown in Figure 3A for the is based on an resource consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics.
+The resource grows according to a logistic growth model and is consumed according to a Holling Type-III functional response curve, to which I have simply introduced the addition of environmental noise:
+
+$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{ growth, } g(X_t)}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{consumption, } c(X_t)} + \xi_t,$$
+where $\xi_t$ is normally distributed environmental variation with mean zero and standard deviation $\sigma$.
+Figure 3B shows separate curves corresponding the parts of the model describing vegetation growth and vegetation consumption, using the same parameters as shown in the simulation panel, 3A.
+Equilibria exist whenever growth balances consumption.
+For the parameters shown, the S-shaped curve of the Type-III consumption pattern crosses the quadratic curve of logistic growth at four separate locations.
+When growth is larger than consumption before the crossing, and thus smaller after, the point is stable, otherwise it is unstable.
+Consequently, we can see two stable points, in the neighborhood of $X_t \approx 1.2$ and $X_t \approx 0.5$, closely matching the average fluctuations seen in the simulation in 3A.
+Stochastic fluctuations drive spontaneous shifts between these alternative stable states.
+
+The concept of a potential well, Figure 3C, is often invoked when describing the behavior of alternative stable states.
+The curve for the potential well, $U(x)$ is defined as the negative integral of the population growth rate,
+
+$$U(x) = -\int_0^x f(y) \textrm{d} y $$
+
+where in our model $f(y)$ is vegetation growth minus consumption, $f(y) = g(y) - c(y)$.
+The motivation comes from physics, where $f$ is a force and $U$ the corresponding potential energy.
+This suggests a convenient analogy to gravity, often implied by drawing a ball inside the curve and referring to this as a "ball-in-cup" diagram.
+The potential function is particularly instructive for understanding stochastic switching under small noise, as in our simulation in Figure 3, where it is straight-forward to show [e.g.
+@Feller; @vanKampen2007] that the probability of a transition is proportional the exponential of the depth of the well from the lowest point to the separation barrier, (a result that is familiar as Arrhenius law of activation energy to grade-school students of chemistry).
+This is effect is visible even in the short simulation shown in 3A, which spends longer times in the lower state which has the deeper well.
+(Readers are encouraged to experiment with this by adjusting the parameters of the simulation in the included code).
+Unfortunately, the literature is not always careful in describing how this potential well diagram arises, which has caused it to become a source for as least as much confusion as it is of insight.
+While the curves in Figure 3B can be clearly associated with underlying mechanisms of rates: logistic growth and consumption with handling time, the same is not true of the potential: it is much harder to have a good intutition about the 'negative integral of the difference growth and consumption,' and how it would change in response to, say, a larger carrying capacity $K$ or higher consumption rate $a$.
+While the x-axis of the potential diagram remains the same: the system state, the conceptual problem is clearly evident in the usual ambiguity describing the y-axis in such ball-in-cup models.
 
@@
 ```
-
-This pattern is characterized by long stretches during which fluctuations in the state variable remain concentrated around a high average value followed by sudden, rapid transistions into fluctuations concentrated around a lower value.
-This behavior arises through the interaction of stochasticity and alternative stable state dynamics.
-The simulation shown in Figure 3A for the is based on an resource consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics.
-The resource grows according to a logistic growth model and is consumed according to a Holling Type-III functional response curve, to which I have simply introduced the addition of environmental noise:
-
-$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{ growth, } g(X_t)}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{consumption, } c(X_t)} + \xi_t,$$
-where $\xi_t$ is normally distributed environmental variation with mean zero and standard deviation $\sigma$.
-Figure 3B shows separate curves corresponding the parts of the model describing vegetation growth and vegetation consumption, using the same parameters as shown in the simulation panel, 3A.
-Equilibria exist whenever growth balances consumption.
-For the parameters shown, the S-shaped curve of the Type-III consumption pattern crosses the quadratic curve of logistic growth at four separate locations.
-When growth is larger than consumption before the crossing, and thus smaller after, the point is stable, otherwise it is unstable.
-Consequently, we can see two stable points, in the neighborhood of $X_t \approx 1.2$ and $X_t \approx 0.5$, closely matching the average fluctuations seen in the simulation in 3A.
-Stochastic fluctuations drive spontaneous shifts between these alternative stable states.
-
-The concept of a potential well, Figure 3C, is often invoked when describing the behavior of alternative stable states.
-The curve for the potential well, $U(x)$ is defined as the negative integral of the population growth rate,
-
-$$U(x) = -\int_0^x f(y) \textrm{d} y $$ 
-where in our model $f(y)$ is vegetation growth minus consumption, $f(y) = g(y) - c(y)$.
-The motivation comes from physics, where $f$ is a force and $U$ the corresponding potential energy.
-This suggests a convenient analogy to gravity, often implied by drawing a ball inside the curve and referring to this as a "ball-in-cup" diagram.
-The potential function is particularly instructive for understanding stochastic switching under small noise, as in our simulation in Figure 3, where it is straight-forward to show [e.g.
-@Feller; @vanKampen2007] that the probability of a transition is proportional the exponential of the depth of the well from the lowest point to the separation barrier, (a result that is familiar as Arrhenius law of activation energy to grade-school students of chemistry).
-This is effect is visible even in the short simulation shown in 3A, which spends longer times in the lower state which has the deeper well.
-(Readers are encouraged to experiment with this by adjusting the parameters of the simulation in the included code).
-Unfortunately, the literature is not always careful in describing how this potential well diagram arises, which has caused it to become a source for as least as much confusion as it is of insight.
-While the curves in Figure 3B can be clearly associated with underlying mechanisms of rates: logistic growth and consumption with handling time, the same is not true of the potential: it is much harder to have a good intutition about the 'negative integral of the difference growth and consumption,' and how it would change in response to, say, a larger carrying capacity $K$ or higher consumption rate $a$.
-While the x-axis of the potential diagram remains the same: the system state, the conceptual problem is clearly evident in the usual ambiguity describing the y-axis in such ball-in-cup models.
 
@@
 While the laboratory experiments of @Dai2012 do precisely this, conducting identical, independent replicate manipulations in separate beakers, this is not an option in natural populations.
-Instead,
+If the environmental change is sufficiently slow relative to the measurements, one can assume that the system is close to its stationary state over a given window in time, computing these averages over a rolling time window as we illustrate here, rather than across an ensemble of replicates.
 

### 2017-10-31 appendixB/appendixB.Rmd
WIP

@@
 The second moment of the transition rates, $a_2$, will be for us simply the sum of the birth and death rates.
-We use this notation as it generalizes to processes that have a different master equation from \eqref{master1}.}
+We use this notation as it generalizes to processes that have a different master equation from \eqref{master1}, to describe a transition of arbitrary step size: $\alpha_{i,j} = \int r^i \Phi_j(r) \ud r$.}
 
@@
 
+# System size expansion in higher-dimensional systems: environmental noise example
+
+We introduce environmental noise into the master equation by allowing the transition rates to depend explicitly on the macroscopic variable $\psi$ as well as the state variable $\phi$.
+It can be helpful to think of fluctuations in $\psi$ as also arising from some lower level transitions, but we need not model these explicitly.
+Following the expansion as before (see @vanKampen for details), the mean dynamics are given by the macroscopic equations:
+
+\begin{align}
+\frac{\ud \phi}{\ud t} &= \alpha_{1,0}(\phi, \psi) \\
+\frac{\ud \psi}{\ud t} &= \beta_{1,0}(\psi) \\
+\end{align}
+
+while the fluctuation dynamics are now given by three coupled differential equations:
+
+\begin{align}
+\frac{\ud \langle \xi^2 \rangle}{\ud t} &= 2 \frac{\partial \alpha_{1,0}}{\partial \phi} \langle \xi^2 \rangle + 2 \frac{\partial \alpha_{1,0} }{\partial \psi} \langle \xi \eta \rangle + \alpha_{2,0} \label{phivar} \\
+\frac{\ud \langle \xi \eta \rangle}{\ud t} &= \left( \frac{\partial \alpha_{1,0}}{\partial \phi} + \frac{\partial \beta_{1,0} }{\partial \psi} \right)\langle \xi \eta \rangle + \frac{\partial \alpha_{1,0} }{\partial \psi} \langle \eta^2 \rangle \label{covar}\\
+\frac{\ud \langle \eta^2 \rangle}{\ud t} &= 2 \frac{\partial \beta_{1,0}}{\partial \psi} \langle \eta^2 \rangle + \beta_{2,0} \label{psivar}
+\end{align}
+
+Without an explicit model for the environmental variation process, we can merely summarize its role in terms of the autocorrelation, $\tau_c := \tfrac{\partial \beta_{1,0}}{\partial \psi}$ and its overall variance at steady-state, $\langle \eta^2 \rangle = \tfrac{\beta_{2,0}}{-2 \partial_{\psi} \beta{1,0}}$, which we will refer to as the environmental variation $\sigma^2_e := \langle \eta^2 \rangle$.
+A more detailed mechanistic description of how the environmental noise arises would permit a time-dependent solution, but the system size expansion tells us that these two summary statistics are all we need to express the steady-state solution above.
+
+Setting the remaining equations to zero to solve for the stationary state we can write down a general formula for the variation in $\phi$:
+
+\begin{equation}
+\langle xi^2 \rangle = \frac{ \left( \partial_{\psi} \alpha_{1,0}(\phi,\psi) \right)^2}{\left( \partial_{\phi} \alpha_{1,0}(\phi,\psi) \right)^2 + \partial_{\phi} \alpha_{1,0} (x,y) \tau_c } \sigma_e^2 + \sigma^2_d
+\end{equation}
+
+From which we recover the environmental variation equation presented in the main text.
+
 # References

### 2017-10-31 paper.Rmd
WIP

@@
 it is only as a simplification or summary of the effects of a large number of other processes, often occurring at different scales.
-Simply put, *noise is all that we leave out of our model*.  Early quantiative ecological models first formulated at the population level
+Simply put, *noise is all that we leave out of our model*.  Early quantitative ecological models first formulated at the population level
 as differential or difference equations: deterministic models that seek to capture the essential features of a system while neglecting
@@
 efficient numerical methods have helped drive research towards increasingly more explicit descriptions of how stochastiticity enters
-into quantiative models [@Black2012; @Ovaskainen2010].
+into quantitative models [@Black2012; @Ovaskainen2010].
 
@@
 population level, see @Black2012) dynamics from an individual-based description of events in a Markov process [@vanKampen2007; based on @Kurtz1970; @Kurtz1971].
-In addition to illustrating how common stochastic models (in particular, Guassian noise) arise from lower-level discriptions, this result will also provide us a useful framework for interpeting the behavior of stochastic models more generally.
+In addition to illustrating how common stochastic models (in particular, Gaussian noise) arise from lower-level descriptions, this result will also provide us a useful framework for interpreting the behavior of stochastic models more generally.
 In Appendix B I illustrate the derivation of the following equations for the mean and expected population sizes for a generic density dependent birth-death process:
@@
 For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$, we find the average number of occupied patches $\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of $\sigma^2_n = N \tfrac{e}{c}$.
-Observe that the variance scales proportionally to the system size, $N$, which explains the standard devation increases only as $\sqrt{N}$ while the steady state itself increases in direct propotion to $N$, resulting in the smaller relative size of fluctations in the larger system in Figure 1.
+Observe that the variance scales proportionally to the system size, $N$, which explains the standard deviation increases only as $\sqrt{N}$ while the steady state itself increases in direct proportion to $N$, resulting in the smaller relative size of fluctuations in the larger system in Figure 1.
 We can also see that different ways of partitioning the same macroscopic equation into components of birth and death will create the same mean dynamics but with different levels of noise.
@@
 
-$$\sigma_x^2  = \frac{ \left( \partial_y f(x,y) \right)^2}{\left( \partial_x f(x,y) \right)^2 + \partial_x f(x,y) \tau_c } \sigma_e^2 + \sigma^2_d$$
+\begin{equation}
+\sigma_x^2  = \frac{ \left( \partial_y f(x,y) \right)^2}{\left( \partial_x f(x,y) \right)^2 + \partial_x f(x,y) \tau_c } \sigma_e^2 + \sigma^2_d \label{env_noise}
+\end{equation}
 
-where $\sigma_e^2$ is the variance in the environmental noise and $\tau_c$ the autocorrelation of that noise, $\sigma_d^2$ the equilibrium variance contributed by demographic noise from before, ($\tfrac{b + d}{-2\partial_x (b - d)}$) and where we have defined $f(x,y) = b(x,y) - d(x,y)$ for notational convenience.
+where $\sigma_e^2$ is the variance in the environmental noise and $\tau_c$ the auto-correlation of that noise, $\sigma_d^2$ the equilibrium variance contributed by demographic noise from before, ($\tfrac{b + d}{-2\partial_x (b - d)}$) and where we have defined $f(x,y) = b(x,y) - d(x,y)$ for notational convenience.
 For example, if extinction rate $e$ in our Levins' model is allowed to vary stochastically as Gaussian white noise with variance $\sigma_e^2$ and mean $e$, we find the expected variance introduced in the population to be:
@@
 
-Note that the contribution of environmental stochasticity is independent of system size and directly proprotional to the variance in the environment.
+Note that the contribution of environmental stochasticity is independent of system size and directly proportional to the variance in the environment.
 This derivation provides a useful illustration of the difference between demographic noise and environmental noise: the latter simply describes the variation introduced by some other dynamic variable that we are not modelling explicitly, and whose dynamics are independent of the state of our system, $x$.
-The correlation time $\tau_c$, is really a shorthand for $\tfrac{1}{\partial_y g(y)}$.
-Coulpling
+This environmental variation no doubt also arise from some more detailed lower level process, but we need not be explicit about those dynamics -- the expansion tells us that it is sufficient to know the overall variance and the degree of auto-correlation.
+Recent work has also highlighted the importance of including both demographic and environmental noise [@Melbourne2008; @Bonsall2004].
+The system size expansion still requires the assumption that the noise is small relative to the system size.
+In the case of large fluctuations, this approximation must be replaced with a different approach, such as the WKB approximation, as discussed in @Ovaskainen2010.
 
-<!--
-[@Roughgarden; @Ellner]
+The importance of $\tau_c$, or the degree of auto-correlation, which emerges from our derivation above, has also been a subject of significant interest in the study of stochastic population dynamics [e.g.@Rougharden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
+Early work rarely often assumed environmental noise would be uncorrelated as a mathematical convenience.
+As the derivation of Eq \eqref{env_noise} makes clear, for a continuous-time process, the relevant metric is the timescale of auto-correlation in the environmental noise process relative to the timescale (auto-correlation) of the population dynamics.
+The Gaussian noise that emerges out of the system-size approximation is always auto-correlated to some degree -- if we measure our environment at consecutive times infinitesimally far apart, we expect nearly identical values.
+At intervals much longer than the correlation time, consecutive measurements are effectively independent: auto-correlated noise is all about _relative_ timescales.
 
-[@Ovaskainen2010] Master equation, break-down of Fokker-Planck approximation under large noise, introduction of the WKB approximation
-[@Bjornstad2001] _Science_ Review of noisy fluctuations
-
-Colored noise: [@Lawton]
-- [@Ripa1999]
-- [@Marshall2015]
--->
-
-### Other forms of stochasticity
+## Other forms of stochasticity
 
@@
 A natural focus of much research into stochastic ecological systems has been on understanding the implications for persistence of individual and coexistence of interacting species.
-Demographic noise in particular has fundamental implications for the long-term equilibrium: any finite population, however big, is guarenteed to eventually go extinct under demographic noise.
-While extinction is certainly a phenonomenon that stands in stark contrast to the long-term behavior of the corresponding deterministic equation, this result is not particularly useful: "eventually" could be longer than the existence of the planet itself.
+Demographic noise in particular has fundamental implications for the long-term equilibrium: any finite population, however big, is guaranteed to eventually go extinct under demographic noise.
+While extinction is certainly a phenomenon that stands in stark contrast to the long-term behavior of the corresponding deterministic equation, this result is not particularly useful: "eventually" could be longer than the existence of the planet itself.
 Rather, it is a reminder that we must be careful about how we describe the 'long-term' behavior of a model.
-In place of the mathematical notion of a stationary distribution to describe the long-term behavior of a stochastic model, analysis of demographic stochasticity focuses on the concept of quasi-stationary distributions, which in many cases correspond more closely to the equilibria of their deterministic counterparts [@Ovaskainen2010; @Schreiber2017].
+In place of the mathematical notion of a stationary distribution to describe the long-term behavior of a stochastic model, analysis of demographic stochasticity focuses on the concept of quasi-stationary distributions, which in many cases correspond more closely to the equilibrium of their deterministic counterparts [@Ovaskainen2010; @Schreiber2017].
 
@@
 We turn from stochastic resonance to oscillations of a very different sort, which do not resemble any deterministic process, as seen in Figure 3A.
-This pattern is characterized by long stretches during which fluctuations in the state variable remain concentrated around a high average value followed by sudden, rapid transistions into fluctuations concentrated around a lower value.
+This pattern is characterized by long stretches during which fluctuations in the state variable remain concentrated around a high average value followed by sudden, rapid transitions into fluctuations concentrated around a lower value.
 This behavior arises through the interaction of stochasticity and alternative stable state dynamics.
@@
 Unfortunately, the literature is not always careful in describing how this potential well diagram arises, which has caused it to become a source for as least as much confusion as it is of insight.
-While the curves in Figure 3B can be clearly associated with underlying mechanisms of rates: logistic growth and consumption with handling time, the same is not true of the potential: it is much harder to have a good intutition about the 'negative integral of the difference growth and consumption,' and how it would change in response to, say, a larger carrying capacity $K$ or higher consumption rate $a$.
+While the curves in Figure 3B can be clearly associated with underlying mechanisms of rates: logistic growth and consumption with handling time, the same is not true of the potential: it is much harder to have a good intuition about the 'negative integral of the difference growth and consumption,' and how it would change in response to, say, a larger carrying capacity $K$ or higher consumption rate $a$.
 While the x-axis of the potential diagram remains the same: the system state, the conceptual problem is clearly evident in the usual ambiguity describing the y-axis in such ball-in-cup models.
@@
 Despite the nearly ubiquitous association of a ball-in-cup model with ecological discussions of resilience and stability, this potential well diagram (Figure 3C) is also a source of confusion on the topic.
-As we have seen, the most salient feature of this diagram is the depth of each well: transistion probabilities (under small noise) depend only on the depth of the well and not the steepness of the well, the distance between the wells or between well and tipping point barrier.
+As we have seen, the most salient feature of this diagram is the depth of each well: transition probabilities (under small noise) depend only on the depth of the well and not the steepness of the well, the distance between the wells or between well and tipping point barrier.
 However, it is important to remember that because the potential function is defined by the above integral, it in fact already combines two more elementary and mechanistically precise quantities from the growth equation: the slope of the growth rate (stability) and the distance to the tipping point (size of the basin of attraction).
@@
 Empirical evidence for stochastic switching in ecological timeseries is limited by the long time scale usually required.
-Because a single stochastic transisition between alternative stable states appears as such a dramatic deviation from the previous pattern (as illustrated in Figure 3A), it is often tempting to assume the change could only be the result of some sudden external force or change in the environment, so it is worth remembering that given long enough, these transitions can happen entirely by chance stochasticity [@Boettiger2012; @Drake2013].
-However, these transistions are most likely to observed in systems with relatively shallow potential wells.
+Because a single stochastic transition between alternative stable states appears as such a dramatic deviation from the previous pattern (as illustrated in Figure 3A), it is often tempting to assume the change could only be the result of some sudden external force or change in the environment, so it is worth remembering that given long enough, these transitions can happen entirely by chance stochasticity [@Boettiger2012; @Drake2013].
+However, these transitions are most likely to observed in systems with relatively shallow potential wells.
 This is mostly likely to occur of some external change is slowly making the potential well shallower and thus transitions more likely -- a phenomenon that has become to be known as "flickering" and suggested as a potential early warning sign of an approaching transition [@Brock2010].
 Examples of flickering between alternative stable states have been suggested in data ranging from geological timescales at the end of glaciation in the Pleistocene [@Taylor1993] to ecological scales over a few decades in lake ecosystems [@Wang2012].
-Stochastic transitions between bistable states are thought to play an important role in genetic switches, where among other phenomena this mechanism can explain the emergence of two alternative phenotypes in clonally reproducing microbial populations, which can act to buffer the population in a changing environment  [@Balaban2004].
+Stochastic transitions between bi-stable states are thought to play an important role in genetic switches, where among other phenomena this mechanism can explain the emergence of two alternative phenotypes in clonal microbial populations, which can act to buffer the population in a changing environment  [@Balaban2004].
 
@@
 Noise can provide novel information in other ways as well.
-Stochastic fluctuations act like a continual miniature perturbation experiment, constantly pusing a system away from an equilibrium to reveal it's behavior elsewhere in state space.
-Observing the size and autocorrelation of these fluctuations thus provides insight on the stability of an equilibrium point, (as can be seen, for instance in the fluctuation-disspation result we found for generic birth death models).
-This observation has been exploited with particular success as a potential *early warning sign* that a system is loosing stability; possibly due to an approaching catastrophic transition or saddle-node bifucation.
-This pattern, known as critical slowing down (in reference to the autocorrelation time in the physics literature) was first proposed an an early warning sign for ecological systems by @Wissel1984, but has recieved considerable empirical as well as theoretical attention only more recently [@Scheffer2009; @Scheffer2015].
-Increases in two of these early warning indicators, variance and autocorrelation, are illustrated along with the timeseries of a simulation of the dynamics of yeast population growth in Figure 4.
+Stochastic fluctuations act like a continual miniature perturbation experiment, constantly pushing a system away from an equilibrium to reveal its behavior elsewhere in state space.
+Observing the size and auto-correlation of these fluctuations thus provides insight on the stability of an equilibrium point, (as can be seen, for instance in the fluctuation-dissipation result we found for generic birth death models).
+This observation has been exploited with particular success as a potential *early warning sign* that a system is loosing stability; possibly due to an approaching catastrophic transition or saddle-node bifurcation.
+This pattern, known as critical slowing down (in reference to the auto-correlation time in the physics literature) was first proposed an an early warning sign for ecological systems by @Wissel1984, but has received considerable empirical as well as theoretical attention only more recently [@Scheffer2009; @Scheffer2015].
+Increases in two of these early warning indicators, variance and auto-correlation, are illustrated along with the timeseries of a simulation of the dynamics of yeast population growth in Figure 4.
 This simulation is based on a mechanistic model developed by @Dai2012 that closely matches the population dynamics they observed in serial dilutions of budding yeast _Saccharomyces cervisiae_ on a sucrose medium.
 Because the relevant proteins are located in the extra-cellular matrix rather than inside the cells, budding yeast break down sucrose collaboratively, creating an Allee effect below which the population is too small to sustain itself.
-Increasing the serial dilution rate slowly over time drives the stable population towards a fold bifurcation, which is accompanied by the patterns of increasing variance and increasing autocorrelation seen empirically in lab populations [@Dai2012] as well as the numerical simulation shown here.
+Increasing the serial dilution rate slowly over time drives the stable population towards a fold bifurcation, which is accompanied by the patterns of increasing variance and increasing auto-correlation seen empirically in lab populations [@Dai2012] as well as the numerical simulation shown here.
 
-As we saw in the equations from the system-size approximation, the variance and autocorrelation in question are technically properties of an ensemble, and should be computed over replicate simulations.
+As we saw in the equations from the system-size approximation, the variance and auto-correlation in question are technically properties of an ensemble, and should be computed over replicate simulations.
 While the laboratory experiments of @Dai2012 do precisely this, conducting identical, independent replicate manipulations in separate beakers, this is not an option in natural populations.
@@
 The past two decades have seen widespread recognition of the central role that stochasticity plays through the use of mechanistic models can have in creating and understanding ecological phenomena [@Coulson2004; @Black2012].
-We have seen increasing use of master-equation approaches to individual-based models to reflect demographic stochasticity that is amenable to analytical treatment [@Black2012], and much work in understanding the importance of environmental noise, particularly the more realistic but less mathematically convienient notion of colored or auto-correlated noise on population dynamics, persistence, and coexistence [@Ovaskainen2010; @Schreiber2017].
-These deductive models can generate phenomena that resemble familiar patterns such as population cycles for which we have previously sought determinstic explanations, as well as stochastic phenomena with no such analog such as stochastic switching.
+We have seen increasing use of master-equation approaches to individual-based models to reflect demographic stochasticity that is amenable to analytical treatment [@Black2012], and much work in understanding the importance of environmental noise, particularly the more realistic but less mathematically convenient notion of colored or auto-correlated noise on population dynamics, persistence, and coexistence [@Ovaskainen2010; @Schreiber2017].
+These deductive models can generate phenomena that resemble familiar patterns such as population cycles for which we have previously sought deterministic explanations, as well as stochastic phenomena with no such analog such as stochastic switching.
 Yet too many of these advances have remained limited to the domain of theory; limited to deductive models which may seek to mimic observed patterns but are rarely used to confront empirical data directly.
@@
 A range of stochastic phenomena have been suggested and a tested for their potential to reveal subtle signs of ecosystem change prior to a dramatic shift, and prompted a rich dialog between further theory and new experiment.
-Changes in stochastic properties such as variance or autocorrelation provide evidence of resilience loss that would be invisible to a determinstic model, and non-linear stochastic phenomena such as stochastic switching are observed as evidence of 'flickering' prior to a critical transistion.
-In this way, our understanding of stochastic dynamics is translated into novel sources of information about a system which has sparked a wealth of emprical investigation and further theoretical work to test the potential of this approach.
-The study and application of stochastic phenomena to detect ecosystem change is unlikely to be limited to this simple scenario of a critical transistion or fold bifurcation.
+Changes in stochastic properties such as variance or auto-correlation provide evidence of resilience loss that would be invisible to a deterministic model, and non-linear stochastic phenomena such as stochastic switching are observed as evidence of 'flickering' prior to a critical transition.
+In this way, our understanding of stochastic dynamics is translated into novel sources of information about a system which has sparked a wealth of empirical investigation and further theoretical work to test the potential of this approach.
+The study and application of stochastic phenomena to detect ecosystem change is unlikely to be limited to this simple scenario of a critical transition or fold bifurcation.
 
-Precisely because most stochastic phenomena arise through their interaction with nonlinearity, the patterns that result can become a lens into better understanding those nonlinear processes and detecting how they may change.
+Precisely because most stochastic phenomena arise through their interaction with non-linearity, the patterns that result can become a lens into better understanding those nonlinear processes and detecting how they may change.
 Meanwhile, the widespread adoption of computational tools and the ability to express theoretical results in numerical algorithms can make any new approaches more readily accessible to application in empirical research.
-In the past two decades, we have set to rest those deterministic skeletons that see noise only as nuciance and embraced the central role noise plays as the creator of ecological phenomena.
+In the past two decades, we have set to rest those deterministic skeletons that see noise only as nuisance and embraced the central role noise plays as the creator of ecological phenomena.
 Going forward, I hope and predict that we will see greater discussion and use of stochastic phenomena to infer underlying processes in empirical patterns in similar interactions between theoretical and empirical research.

### 2017-11-01 paper.Rmd
hard-wrap lines

@@
 ---
-title: "From noise to knowledge: how randomness generates novel phenomena and information"
+title: "From noise to knowledge: how randomness generates novel phenomena and reveals information"
 author:
@@
 <!--
+Directions:
 
@@
 
-Highlight *key questions* & *key advances*
-Moving beyond "everything is important" [@Bjornstad2001; @Kingsland1985]
-
-(@Bjornstad2001 future: more MCMC methods. More mechanistic models.  dealing with unobserved dynamics)
-
-Shift from noise as nuiance to noise as opportunity
-Stochastic phenomena outside of theory -- early warning signals
-
-Bialek's rules
-1. Noise isn't negligible: photon counting, chemotaxis, 
-2. No fine tuning: developmental patterning? ion channels/ stochastic switches? 
-3. Efficient representation: entropy, information flow
-
 -->
@@
 
-Stochasticity is most often thought of as something which obscures patterns we are trying to infer.
-This is a simple consequence of modeling: as discussed above, variables of interest to the processes in question are modeled explicitly (e.g.
-change in population density as it depends on the current density), everything else is just noise.
-Most statistical inference in ecology is built solely to deal with this problem: pulling back the veil of noise to reveal processes of interest.
-Unfortunately, given the high levels of variation and relatively limited data typical in ecological analyses, this can be a significant challenge, and even as statistical and computational advances provide more powerful and accessible inference tools, the task only gets harder.
-For instance, few ecological concepts are more central than the notion of density dependence: that competition in some form or other should limit a species ability to grow without bound.
-Yet establishing statistical evidence for density dependence is remarkably difficult, as illustrated by @Knape2011.
-Using time series from the Global Population Dynamics Database (GPDD) Knape & de Valpine show that once one accounts for the potential for measurement error in addition to environmental stochasticity -- an unavoidable source of noise in most ecological data -- it becomes difficult to find evidence for even the simplest models of density dependence in the majority of available time series.
-Faced with such uncertainty, it is tempting to throw up our hands and wait until some prophesied deluge of streaming micro-sensors stapled to every organism provide enough signal to beat back the noise.
-Fortunately, this has not been the path taken.
+Stochasticity is most often thought of as something which obscures
+patterns we are trying to infer.  This is a simple consequence of
+modeling: as discussed above, variables of interest to the processes
+in question are modeled explicitly (e.g. change in population density
+as it depends on the current density), everything else is just
+noise.  Most statistical inference in ecology is built solely to
+deal with this problem: pulling back the veil of noise to reveal
+processes of interest.  Unfortunately, given the high levels of
+variation and relatively limited data typical in ecological analyses,
+this can be a significant challenge, and even as statistical and
+computational advances provide more powerful and accessible inference
+tools, the task only gets harder.  For instance, few ecological
+concepts are more central than the notion of density dependence:
+that competition in some form or other should limit a species ability
+to grow without bound.  Yet establishing statistical evidence for
+density dependence is remarkably difficult, as illustrated by
+@Knape2011.  Using time series from the Global Population Dynamics
+Database (GPDD) Knape & de Valpine show that once one accounts for
+the potential for measurement error in addition to environmental
+stochasticity -- an unavoidable source of noise in most ecological
+data -- it becomes difficult to find evidence for even the simplest
+models of density dependence in the majority of available time
+series.  Faced with such uncertainty, it is tempting to throw up
+our hands and wait until some prophesied deluge of streaming
+micro-sensors stapled to every organism provide enough signal to
+beat back the noise. Fortunately, this has not been the path taken.
 
 <!-- okay, claiming this for the last 2 decades might be overly provocative?-->
-Over the past twenty years, ecological theory, modelling, and methods have increasingly elucidated the role of noise not only to *obscure* patterns but to *create* them.
-This period marked a transition away from the intense focus on simple models which exhibited deterministic chaos that had enchanted much of theoretical ecology in the previous two decades [@Bjornstad2001] in favor of more complex and stochastic models.
-Of particular interest has been those conceptual advances that have bridged the divide from theoretical to empirical interest -- not merely the introduction of "real data" in a theoretical paper, but rather the concepts that take on a life of their own in the larger research community.
-Consequently, this paper seeks to be a bridge between two audiences: theoretical researchers already expert in noisy phenomena but in interested in reaching wider empirical application, and also primarily empirical researchers seeking an entrance into noisy phenomena.
+Over the past twenty years, ecological theory, modelling, and methods
+have increasingly elucidated the role of noise not only to *obscure*
+patterns but to *create* them.  This period marked a transition
+away from the intense focus on simple models which exhibited
+deterministic chaos that had enchanted much of theoretical ecology
+in the previous two decades [@Bjornstad2001] in favor of more complex
+and stochastic models.  Of particular interest has been those
+conceptual advances that have bridged the divide from theoretical
+to empirical interest -- not merely the introduction of "real data"
+in a theoretical paper, but rather the concepts that take on a life
+of their own in the larger research community.  Consequently, this
+paper seeks to be a bridge between two audiences: theoretical
+researchers already expert in noisy phenomena but in interested in
+reaching wider empirical application, and also primarily empirical
+researchers seeking an entrance into noisy phenomena.
 
 <!-- Regarding simulation examples -->
-In this spirit, each of the examples illustrated here is coupled with concise and efficient R code in the appendix to both reproduce and further explore the simulations.
-While examples of noisy phenomena in natural systems are also crucial, matching the pattern does not necessarily mean we have also matched the process behind the data.
-The examples cited from the literature here offer some of the best empirical evidence of these phenomena, but I encourage readers to also experiment with the example code provided.
-Simulations permit poking and prodding of empirical investigation quickly and with little mathematical formalism in the way.
-In the code provided, most stochastic models are expressed in the BUGS language, which may be more familiar to empirical readers than corresponding mathematical formulas.
-This also permits both efficient simulation and potential estimation of parameters given sample data using the R package, NIMBLE [@DeValpine2017]; allowing the deductive models illustrated here to readily function as inductive models with parameters inferred from time-series data.
-A copy of this appendix is maintained at <https://github.com/cboettig/noise-phenomena>, and bug reports, suggestions or help requests are welcome through the issue tracker.
-A second appendix provides more mathematical details for interested readers.
-With these in place, we are ready to begin our tour through three major areas of noisy phenomena: Origins of noise, emergent phenomena, and noise-driven inference.
+In this spirit, each of the examples illustrated here is coupled
+with concise and efficient R code in the appendix to both reproduce
+and further explore the simulations.  While examples of noisy
+phenomena in natural systems are also crucial, matching the pattern
+does not necessarily mean we have also matched the process behind
+the data.  The examples cited from the literature here offer some
+of the best empirical evidence of these phenomena, but I encourage
+readers to also experiment with the example code provided.  Simulations
+permit poking and prodding of empirical investigation quickly and
+with little mathematical formalism in the way.  In the code provided,
+most stochastic models are expressed in the BUGS language, which
+may be more familiar to empirical readers than corresponding
+mathematical formulas. This also permits both efficient simulation
+and potential estimation of parameters given sample data using the
+R package, NIMBLE [@DeValpine2017]; allowing the deductive models
+illustrated here to readily function as inductive models with
+parameters inferred from time-series data. A copy of this appendix
+is maintained at <https://github.com/cboettig/noise-phenomena>, and
+bug reports, suggestions or help requests are welcome through the
+issue tracker.  A second appendix provides more mathematical details
+for interested readers.  With these in place, we are ready to begin
+our tour through three major areas of noisy phenomena: Origins of
+noise, emergent phenomena, and noise-driven inference.
 
@@
 Our first stop will be into research into the origins of noise.
-This will not only let us set up definitions for terms and concepts we will encounter throughout, but also introduce some of the fundamental mathematical and computational tooling needed in our tour of noisy phenomena.
-From origins of noise, we will then explore some of the phenomena that noise can create.
-These examples are central to the conceptual transition away from viewing noise only as a nuisance that obscures patterns and realize the rich set of patterns and phenomena that can emerge.
-Our tour will be by no means exhaustive, but as with chaos before it, theoretical allure does not always reflect ecological relevance.
-Having looked at some of the phenomena that can be created by noise, our third stop will try to run this in reverse: using noise to infer underlying phenomena.
-Examples here are rarer but potentially the most rewarding, bridging the gap from theoretical construct to empirical investigation.
-The most compelling examples have come from ways in which noisy phenomena are used to infer stability and forecast changes in resilience of populations or ecosystems.
-Though not without weaknesses or missteps, the resulting dialog between theoretical and empirical aspects have made this particularly a rich area of research.
-We end our tour by considering what features best enable the translation and ecological relevance of noisy phenomena.
+This will not only let us set up definitions for terms and concepts
+we will encounter throughout, but also introduce some of the
+fundamental mathematical and computational tooling needed in our
+tour of noisy phenomena.  From origins of noise, we will then explore
+some of the phenomena that noise can create.  These examples are
+central to the conceptual transition away from viewing noise only
+as a nuisance that obscures patterns and realize the rich set of
+patterns and phenomena that can emerge.  Our tour will be by no
+means exhaustive, but as with chaos before it, theoretical allure
+does not always reflect ecological relevance.  Having looked at
+some of the phenomena that can be created by noise, our third stop
+will try to run this in reverse: using noise to infer underlying
+phenomena.  Examples here are rarer but potentially the most
+rewarding, bridging the gap from theoretical construct to empirical
+investigation.  The most compelling examples have come from ways
+in which noisy phenomena are used to infer stability and forecast
+changes in resilience of populations or ecosystems. Though not
+without weaknesses or missteps, the resulting dialog between
+theoretical and empirical aspects have made this particularly a
+rich area of research.  We end our tour by considering what features
+best enable the translation and ecological relevance of noisy
+phenomena.
 
@@
 
-What is noise? Where does it come from?  It is important to bear in mind that whenever stochasticity is introduced into a model,
-it is only as a simplification or summary of the effects of a large number of other processes, often occurring at different scales.
-Simply put, *noise is all that we leave out of our model*.  Early quantitative ecological models first formulated at the population level
-as differential or difference equations: deterministic models that seek to capture the essential features of a system while neglecting
-any random component as mere noise [@Coulson2004]. Advances in both theoretical understanding and the advent of widely available and
-efficient numerical methods have helped drive research towards increasingly more explicit descriptions of how stochastiticity enters
-into quantitative models [@Black2012; @Ovaskainen2010].
+What is noise? Where does it come from?  It is important to bear
+in mind that whenever stochasticity is introduced into a model, it
+is only as a simplification or summary of the effects of a large
+number of other processes, often occurring at different scales.
+Simply put, *noise is all that we leave out of our model*.  Early
+quantitative ecological models first formulated at the population
+level as differential or difference equations: deterministic models
+that seek to capture the essential features of a system while
+neglecting any random component as mere noise [@Coulson2004].
+Advances in both theoretical understanding and the advent of widely
+available and efficient numerical methods have helped drive research
+towards increasingly more explicit descriptions of how stochastiticity
+enters into quantitative models [@Black2012; @Ovaskainen2010].
 
-With the exception of quantum mechanics (and there only at scales that have no relevance to ecology), stochasticity is not an inherent
-property of nature but rather an emergent property of many events occurring at a different scale.  Coins and dice obey the physical
-laws of motion; only variation in how they are tossed (i.e. something complex we do not usually model) creates randomness.  The term
-"noise" itself reflects this emergent rather than fundamental origin: the combination of hundreds of voices and sounds each individually
-distinct blends together into a background of noise. Statistical mechanics tells us that temperature is a measure of the average speed
-of individual molecules in a gas -- any given molecule will experience a range of possible speeds but by itself has no notion of temperature.
-Likewise, stochasticity enters our ecological models because our quantities of interest, such as the density of a population, is but an
-average over a lower-level, discrete process of individual births and deaths. Stochasticity that emerges directly from discrete individual
-events that underlie the process described by our model are considered *intrinsic stochasticity*, commonly referred to as
-*demographic stochasticity* in ecological literature, as it usually arises in the context scaling from individual demographic events to
-population level descriptions. An important aspect of this demographic stochasicity is that it scales with the square root of the
-"system size," so that for very large populations the resulting fluctuations are relatively small. On the other hand, 
-*extrinsic stochasticity*, usually referred to as *environmental stochasticity* in the ecological literature, arises in processes
-that are not described by our model, but merely impact the value of some parameter of our model; for instance, fluctuations in
-temperature may drive variation in a death rate, or fluctuations in rainfall drive variation in carrying capacity.
-
-<!--
-Motivate the presentation of the van Kampen result rather than a generic review.
--->
+With the exception of quantum mechanics (and there only at scales
+that have no relevance to ecology), stochasticity is not an inherent
+property of nature but rather an emergent property of many events
+occurring at a different scale.  Coins and dice obey the physical
+laws of motion; only variation in how they are tossed (i.e. something
+complex we do not usually model) creates randomness.  The term
+"noise" itself reflects this emergent rather than fundamental origin:
+the combination of hundreds of voices and sounds each individually
+distinct blends together into a background of noise. Statistical
+mechanics tells us that temperature is a measure of the average
+speed of individual molecules in a gas -- any given molecule will
+experience a range of possible speeds but by itself has no notion
+of temperature.  Likewise, stochasticity enters our ecological
+models because our quantities of interest, such as the density of
+a population, is but an average over a lower-level, discrete process
+of individual births and deaths. Stochasticity that emerges directly
+from discrete individual events that underlie the process described
+by our model are considered *intrinsic stochasticity*, commonly
+referred to as *demographic stochasticity* in ecological literature,
+as it usually arises in the context scaling from individual demographic
+events to population level descriptions. An important aspect of
+this demographic stochasicity is that it scales with the square
+root of the "system size," so that for very large populations the
+resulting fluctuations are relatively small. On the other hand,
+*extrinsic stochasticity*, usually referred to as *environmental
+stochasticity* in the ecological literature, arises in processes
+that are not described by our model, but merely impact the value
+of some parameter of our model; for instance, fluctuations in
+temperature may drive variation in a death rate, or fluctuations
+in rainfall drive variation in carrying capacity.
 
@@
 
-Demographic stochasticity refers to fluctuations in population sizes or densities that arise from the fundamentally discrete nature of individual birth and death events.
-Demographic stochasticity is a particularly instructive case for illustrating a mechanism for how noise arises as an aggregate description from a lower-level mechanistic process.
-We summarize the myriad lower-level processes that mechanistically lead to the event of a 'birth' in the population as a probability: in a population of $N$ identical individuals at time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate that can depend on both the population size $N$) and increases the population size, $N$ to $N+1$, and deaths occur with probability $d_t(N_t)$, decreasing the population size, $N$ to $N-1$.
-Assuming each of these events are independent, this is a state-dependent Poisson process.
-The change in the probability of being in state $N$ is given by the sum over the ways to enter the state, minus the ways to leave the state: a simple expression of probability balance known as the "master equation."
+Demographic stochasticity refers to fluctuations in population sizes
+or densities that arise from the fundamentally discrete nature of
+individual birth and death events.  Demographic stochasticity is a
+particularly instructive case for illustrating a mechanism for how
+noise arises as an aggregate description from a lower-level mechanistic
+process.  We summarize the myriad lower-level processes that
+mechanistically lead to the event of a 'birth' in the population
+as a probability: in a population of $N$ identical individuals at
+time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate
+that can depend on both the population size $N$) and increases the
+population size, $N$ to $N+1$, and deaths occur with probability
+$d_t(N_t)$, decreasing the population size, $N$ to $N-1$.  Assuming
+each of these events are independent, this is a state-dependent
+Poisson process. The change in the probability of being in state
+$N$ is given by the sum over the ways to enter the state, minus the
+ways to leave the state: a simple expression of probability balance
+known as the "master equation."
 
-It is possible to simulate such a process exactly using Gillespie's exact stochastic simulation algorithm [SSA; @Gillespie1977], which simulates
-each individual birth or death as a separate event.
-Consider Levins' patch model, where $n$ individuals compete for a finite number of suitable habitats $N$.
-Individuals die a constant rate $e$, and produce offspring at a constant rate $c$ who then have a probability
-of colonizing an open patch that is simply proportional to the fraction of available patches, $1 - n/N$.
-Figure 1 shows the results of two exact SSA simulations of this model with identical parameters except for the
-total number of available sites, $N$.
+It is possible to simulate such a process exactly using Gillespie's
+exact stochastic simulation algorithm [SSA; @Gillespie1977], which
+simulates each individual birth or death as a separate event.
+Consider Levins' patch model, where $n$ individuals compete for a
+finite number of suitable habitats $N$.  Individuals die a constant
+rate $e$, and produce offspring at a constant rate $c$ who then
+have a probability of colonizing an open patch that is simply
+proportional to the fraction of available patches, $1 - n/N$.  Figure
+1 shows the results of two exact SSA simulations of this model with
+identical parameters except for the total number of available sites,
+$N$.
 
@@
 
-The van Kampen system size expansion provides a convenient way to explicitly approximate the "macroscopic" (i.e.
-population level, see @Black2012) dynamics from an individual-based description of events in a Markov process [@vanKampen2007; based on @Kurtz1970; @Kurtz1971].
-In addition to illustrating how common stochastic models (in particular, Gaussian noise) arise from lower-level descriptions, this result will also provide us a useful framework for interpreting the behavior of stochastic models more generally.
-In Appendix B I illustrate the derivation of the following equations for the mean and expected population sizes for a generic density dependent birth-death process:
+The van Kampen system size expansion provides a convenient way to
+explicitly approximate the "macroscopic" (i.e. population level,
+see @Black2012) dynamics from an individual-based description of
+events in a Markov process [@vanKampen2007; based on @Kurtz1970;
+@Kurtz1971].  In addition to illustrating how common stochastic
+models (in particular, Gaussian noise) arise from lower-level
+descriptions, this result will also provide us a useful framework
+for interpreting the behavior of stochastic models more generally.
+In Appendix B I illustrate the derivation of the following equations
+for the mean and expected population sizes for a generic density
+dependent birth-death process:
 
@@
 
-where $x = \langle n \rangle$, the average or expected population size of an ensemble of replicates, assuming the transitions (+/- 1 in this case) are much smaller than the system size ($N$ in the Levins model).
-Note that the dynamics for the average population match the population-level equation^[up to fluctuations of order $\tfrac{1}{N}$.
-We can actually both observe and successfully approximate this higher-order correction from additional terms in the system size expansion to recover the phenomenon of stochastic inflation [@Gonzalez2002], where the average population size can be larger or smaller than predicted by the deterministic model -- see Appendix A for examples in simulations compared to predictions of the system size expansion].
+where $x = \langle n \rangle$, the average or expected population
+size of an ensemble of replicates, assuming the transitions (+/- 1
+in this case) are much smaller than the system size ($N$ in the
+Levins model).  Note that the dynamics for the average population
+match the population-level equation^[up to fluctuations of order
+$\tfrac{1}{N}$. We can actually both observe and successfully
+approximate this higher-order correction from additional terms in
+the system size expansion to recover the phenomenon of stochastic
+inflation [@Gonzalez2002], where the average population size can
+be larger or smaller than predicted by the deterministic model --
+see Appendix A for examples in simulations compared to predictions
+of the system size expansion].
 
@@
 
-Note that like the equation for the mean, we have a dynamic equation for the time-evolution of the variance in Eq \eqref{bd}.
-We can solve these equations at steady-state by setting both equations zero.
-For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$, we find the average number of occupied patches $\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of $\sigma^2_n = N \tfrac{e}{c}$.
-Observe that the variance scales proportionally to the system size, $N$, which explains the standard deviation increases only as $\sqrt{N}$ while the steady state itself increases in direct proportion to $N$, resulting in the smaller relative size of fluctuations in the larger system in Figure 1.
-We can also see that different ways of partitioning the same macroscopic equation into components of birth and death will create the same mean dynamics but with different levels of noise.
-For instance, in a logistic model with $b = rn$ and $d = rn^2/N$, we instead find the variance at steady state is $\sigma^2 = K$; significantly larger than in the Levins model.
+Note that like the equation for the mean, we have a dynamic equation
+for the time-evolution of the variance in Eq \eqref{bd}.  We can
+solve these equations at steady-state by setting both equations
+zero. For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$,
+we find the average number of occupied patches 
+$\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of 
+$\sigma^2_n = N \tfrac{e}{c}$.  Observe that the variance scales proportionally
+to the system size, $N$, which explains the standard deviation
+increases only as $\sqrt{N}$ while the steady state itself increases
+in direct proportion to $N$, resulting in the smaller relative size
+of fluctuations in the larger system in Figure 1.  We can also see
+that different ways of partitioning the same macroscopic equation
+into components of birth and death will create the same mean dynamics
+but with different levels of noise.  For instance, in a logistic
+model with $b = rn$ and $d = rn^2/N$, we instead find the variance
+at steady state is $\sigma^2 = K$; significantly larger than in the
+Levins model.
 
@@
 
-Whereas demographic noise reflects processes at a lower scale (the individual, rather than the population), environmental stochasticity reflects processes whose dynamics we are not explicitly modelling.
-Once again, this can best be illustrated through an explicit formulation of the master equation.
-For a generic population model that depends not only on population state $x$ but also on environmental factors $y$, it is possible to derive the steady-state variance (Appendix B) as:
+Whereas demographic noise reflects processes at a lower scale (the
+individual, rather than the population), environmental stochasticity
+reflects processes whose dynamics we are not explicitly modelling.
+Once again, this can best be illustrated through an explicit
+formulation of the master equation.  For a generic population model
+that depends not only on population state $x$ but also on environmental
+factors $y$, it is possible to derive the steady-state variance
+(Appendix B) as:
 
@@
 
-where $\sigma_e^2$ is the variance in the environmental noise and $\tau_c$ the auto-correlation of that noise, $\sigma_d^2$ the equilibrium variance contributed by demographic noise from before, ($\tfrac{b + d}{-2\partial_x (b - d)}$) and where we have defined $f(x,y) = b(x,y) - d(x,y)$ for notational convenience.
-For example, if extinction rate $e$ in our Levins' model is allowed to vary stochastically as Gaussian white noise with variance $\sigma_e^2$ and mean $e$, we find the expected variance introduced in the population to be:
+where $\sigma_e^2$ is the variance in the environmental noise and
+$\tau_c$ the auto-correlation of that noise, $\sigma_d^2$ the
+equilibrium variance contributed by demographic noise from before,
+($\tfrac{b + d}{-2\partial_x (b - d)}$) and where we have defined
+$f(x,y) = b(x,y) - d(x,y)$ for notational convenience.  For example,
+if extinction rate $e$ in our Levins' model is allowed to vary
+stochastically as Gaussian white noise with variance $\sigma_e^2$
+and mean $e$, we find the expected variance introduced in the
+population to be:
 
@@
 
-Note that the contribution of environmental stochasticity is independent of system size and directly proportional to the variance in the environment.
-This derivation provides a useful illustration of the difference between demographic noise and environmental noise: the latter simply describes the variation introduced by some other dynamic variable that we are not modelling explicitly, and whose dynamics are independent of the state of our system, $x$.
-This environmental variation no doubt also arise from some more detailed lower level process, but we need not be explicit about those dynamics -- the expansion tells us that it is sufficient to know the overall variance and the degree of auto-correlation.
-Recent work has also highlighted the importance of including both demographic and environmental noise [@Melbourne2008; @Bonsall2004].
-The system size expansion still requires the assumption that the noise is small relative to the system size.
-In the case of large fluctuations, this approximation must be replaced with a different approach, such as the WKB approximation, as discussed in @Ovaskainen2010.
+Note that the contribution of environmental stochasticity is
+independent of system size and directly proportional to the variance
+in the environment.  This derivation provides a useful illustration
+of the difference between demographic noise and environmental noise:
+the latter simply describes the variation introduced by some other
+dynamic variable that we are not modelling explicitly, and whose
+dynamics are independent of the state of our system, $x$. This
+environmental variation no doubt also arise from some more detailed
+lower level process, but we need not be explicit about those dynamics
+-- the expansion tells us that it is sufficient to know the overall
+variance and the degree of auto-correlation.  Recent work has also
+highlighted the importance of including both demographic and
+environmental noise [@Melbourne2008; @Bonsall2004].  The system
+size expansion still requires the assumption that the noise is small
+relative to the system size.  In the case of large fluctuations,
+this approximation must be replaced with a different approach, such
+as the WKB approximation, as discussed in @Ovaskainen2010.
 
-The importance of $\tau_c$, or the degree of auto-correlation, which emerges from our derivation above, has also been a subject of significant interest in the study of stochastic population dynamics [e.g.@Rougharden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
-Early work rarely often assumed environmental noise would be uncorrelated as a mathematical convenience.
-As the derivation of Eq \eqref{env_noise} makes clear, for a continuous-time process, the relevant metric is the timescale of auto-correlation in the environmental noise process relative to the timescale (auto-correlation) of the population dynamics.
-The Gaussian noise that emerges out of the system-size approximation is always auto-correlated to some degree -- if we measure our environment at consecutive times infinitesimally far apart, we expect nearly identical values.
-At intervals much longer than the correlation time, consecutive measurements are effectively independent: auto-correlated noise is all about _relative_ timescales.
+The importance of $\tau_c$, or the degree of auto-correlation, which
+emerges from our derivation above, has also been a subject of
+significant interest in the study of stochastic population dynamics
+[e.g.@Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
+Early work rarely often assumed environmental noise would be
+uncorrelated as a mathematical convenience.  As the derivation of
+Eq \eqref{env_noise} makes clear, for a continuous-time process,
+the relevant metric is the timescale of auto-correlation in the
+environmental noise process relative to the timescale (auto-correlation)
+of the population dynamics.  The Gaussian noise that emerges out
+of the system-size approximation is always auto-correlated to some
+degree -- if we measure our environment at consecutive times
+infinitesimally far apart, we expect nearly identical values.  At
+intervals much longer than the correlation time, consecutive
+measurements are effectively independent: auto-correlated noise is
+all about _relative_ timescales.
 
@@
 Stochasticity frequently enters our models in other ways as well.
-For instance, most population-level models treat all individuals within the population, or at least
-within a particular age or stage class, as identical.  *Individual heterogeneity* in relevant traits
-(growth rates, dispersal ability) can be represented probabilistically as a distribution rather than
-a fixed value, making it function as another form of random variation in a model [@Coulson2001; @Schreiber2010].
-*Spatial heterogeneity* can likewise be reflected in stochastic parameters of a spatially implicit
-model, or as spatially varying environmental stochasticity in a spatially explicit model.  As with
-other forms of noise [@Melbourne2008; @Bonsall2004], interaction between individual or spatial
-variation and other sources of noise can drive patterns including the potential for persistence,
-coexistence, and dispersal of species [@Schreiber2010; @Hart2016; @Schreiber2017].
-*Observational error*: Observational error acts much like another form of stochasticity, arising from
-sampling effects or limitations of instruments, etc. Yet unlike the sources described above, this 
-uncertainty does not impact the actual state of our system, and thus does not give rise to any
-additional phenomena or information about that process, though it can obscure it [@Knape2011; @deValpine2002].
+For instance, most population-level models treat all individuals
+within the population, or at least within a particular age or stage
+class, as identical.  *Individual heterogeneity* in relevant traits
+(growth rates, dispersal ability) can be represented probabilistically
+as a distribution rather than a fixed value, making it function as
+another form of random variation in a model [@Coulson2001;
+@Schreiber2010].  *Spatial heterogeneity* can likewise be reflected
+in stochastic parameters of a spatially implicit model, or as
+spatially varying environmental stochasticity in a spatially explicit
+model.  As with other forms of noise [@Melbourne2008; @Bonsall2004],
+interaction between individual or spatial variation and other sources
+of noise can drive patterns including the potential for persistence,
+coexistence, and dispersal of species [@Schreiber2010; @Hart2016;
+@Schreiber2017].  *Observational error*: Observational error acts
+much like another form of stochasticity, arising from sampling
+effects or limitations of instruments, etc. Yet unlike the sources
+described above, this uncertainty does not impact the actual state
+of our system, and thus does not give rise to any additional phenomena
+or information about that process, though it can obscure it
+[@Knape2011; @deValpine2002].
 
-Other forms of uncertainty common in models are not stochastic phenomena.  This includes any form of 
-*reducible uncertainty*: In constructing and estimating mathematical models of ecological processes, 
-we encounter uncertainty over our parameter estimates and possible choice of models [@Ellison2004].  In some ways, this
-uncertainty can act like the stochasticity described above: for instance, both introduce variation into
-any forecast of future outcomes. Yet unlike stochasticity, this uncertainty is reducible: with additional
-observations, we can narrow the uncertainty over parameters or candidate models, while stochasticity remains an
-inherent source of variation. We will this not consider model and parameter uncertainty, as they are not 
-sources of noise.  *Deterministic chaos* [@Desharnais2001] is another mechanism of introducing significant
-variation into a simple model. Chaos can a look a lot like noise: indeed, it can be observationally
-indistinguishable from a stochastic process [@Ellner2005; @Werndl2009].  Despite substantial interest in previous
-decades and the relative ease of encountering chaos in simple ecological models (particularly in
-differences equations), and convincing reconstructions of chaotic dynamics in laboratory manipulations
-[@Costantino1997] the prevalence of chaotic dynamics as a driving factor in most ecological systems
-remains unclear. After all, the great variation of natural populations is hardly surprising: the
-undeniable influences of variation created by all those things left out of our models, all that
-individual detail from intrinsic, lower level processes and external environmental processes that
-becomes aggregated in our simple models all but begs for a stochastic description.
+Other forms of uncertainty common in models are not stochastic
+phenomena.  This includes any form of *reducible uncertainty*: In
+constructing and estimating mathematical models of ecological
+processes, we encounter uncertainty over our parameter estimates
+and possible choice of models [@Ellison2004].  In some ways, this
+uncertainty can act like the stochasticity described above: for
+instance, both introduce variation into any forecast of future
+outcomes. Yet unlike stochasticity, this uncertainty is reducible:
+with additional observations, we can narrow the uncertainty over
+parameters or candidate models, while stochasticity remains an
+inherent source of variation. We will this not consider model and
+parameter uncertainty, as they are not sources of noise.  *Deterministic
+chaos* [@Desharnais2001] is another mechanism of introducing
+significant variation into a simple model. Chaos can a look a lot
+like noise: indeed, it can be observationally indistinguishable
+from a stochastic process [@Ellner2005; @Werndl2009].  Despite
+substantial interest in previous decades and the relative ease of
+encountering chaos in simple ecological models (particularly in
+differences equations), and convincing reconstructions of chaotic
+dynamics in laboratory manipulations [@Costantino1997] the prevalence
+of chaotic dynamics as a driving factor in most ecological systems
+remains unclear. After all, the great variation of natural populations
+is hardly surprising: the undeniable influences of variation created
+by all those things left out of our models, all that individual
+detail from intrinsic, lower level processes and external environmental
+processes that becomes aggregated in our simple models all but begs
+for a stochastic description.
 
-Much active research continues to explore the consequences of introducing these more explicitly
-motivated and interacting sources of noise. With some grounding in the origins of different types
-of noise in place, we are ready to move to our second stop to see how this noise can drive the
-creation of novel phenomena on a macroscopic level.
+Much active research continues to explore the consequences of
+introducing these more explicitly motivated and interacting sources
+of noise. With some grounding in the origins of different types of
+noise in place, we are ready to move to our second stop to see how
+this noise can drive the creation of novel phenomena on a macroscopic
+level.
 
@@
 
-Stochasticity can give rise to a wide array of interesting phenomena that are often at odds with our intuition.
-As these examples will illustrate, such behavior arises through an interaction between stochastic and non-linear elements of a system, underscoring the importance of explicitly including stochasticity in ecological models [@Coulson2004; @Higgins1997].
-Importantly, the noise itself does not have to be large to give rise to these phenomena.
-In this section, I highlight several examples that have shown up frequently in the literature and prove sufficiently tractable to provide a simple intuition for the mechanisms involved.
-
-<!-- 
-Noise is "tacked on" in these models... contrasts to previous section
--->
+Stochasticity can give rise to a wide array of interesting phenomena
+that are often at odds with our intuition. As these examples will
+illustrate, such behavior arises through an interaction between
+stochastic and non-linear elements of a system, underscoring the
+importance of explicitly including stochasticity in ecological
+models [@Coulson2004; @Higgins1997].  Importantly, the noise itself
+does not have to be large to give rise to these phenomena.  In this
+section, I highlight several examples that have shown up frequently
+in the literature and prove sufficiently tractable to provide a
+simple intuition for the mechanisms involved.
 
@@
 
-A natural focus of much research into stochastic ecological systems has been on understanding the implications for persistence of individual and coexistence of interacting species.
-Demographic noise in particular has fundamental implications for the long-term equilibrium: any finite population, however big, is guaranteed to eventually go extinct under demographic noise.
-While extinction is certainly a phenomenon that stands in stark contrast to the long-term behavior of the corresponding deterministic equation, this result is not particularly useful: "eventually" could be longer than the existence of the planet itself.
-Rather, it is a reminder that we must be careful about how we describe the 'long-term' behavior of a model.
-In place of the mathematical notion of a stationary distribution to describe the long-term behavior of a stochastic model, analysis of demographic stochasticity focuses on the concept of quasi-stationary distributions, which in many cases correspond more closely to the equilibrium of their deterministic counterparts [@Ovaskainen2010; @Schreiber2017].
+A natural focus of much research into stochastic ecological systems
+has been on understanding the implications for persistence of
+individual and coexistence of interacting species. Demographic noise
+in particular has fundamental implications for the long-term
+equilibrium: any finite population, however big, is guaranteed to
+eventually go extinct under demographic noise.  While extinction
+is certainly a phenomenon that stands in stark contrast to the
+long-term behavior of the corresponding deterministic equation,
+this result is not particularly useful: "eventually" could be longer
+than the existence of the planet. Rather, it is a reminder that we
+must be careful about how we describe the 'long-term' behavior of
+a model.  In place of the mathematical notion of a stationary
+distribution to describe the long-term behavior of a stochastic
+model, analysis of demographic stochasticity focuses on the concept
+of quasi-stationary distributions, which in many cases correspond
+more closely to the equilibrium of their deterministic counterparts
+[@Ovaskainen2010; @Schreiber2017].
 
@@
 **Bet hedging**
-**Inflation**
 
@@
 
-Understanding the origin of sustained oscillations has long been a central question in ecology [@Hastings1996; @Bjornstad2001].
-The simple Lotka Volterra model produces only neutrally stable centers, while the more realistic model that introduces carrying capacity for prey as they compete for resources results in a model that produces only damped oscillations, eventually settling into a stable state (Figure 1A).
-However, the introduction of relatively small noise into the system can be sufficient to excite the system into sustained oscillations (Figure 1B).
-In the example shown here, a simple predator-prey system is driven by environmental white noise $\xi_t$, with prey following the equation:
+Understanding the origin of sustained oscillations has long been a
+central question in ecology [@Hastings1996; @Bjornstad2001].  The
+simple Lotka Volterra model produces only neutrally stable centers,
+while the more realistic model that introduces carrying capacity
+for prey as they compete for resources results in a model that
+produces only damped oscillations, eventually settling into a stable
+state (Figure 1A).  However, the introduction of relatively small
+noise into the system can be sufficient to excite the system into
+sustained oscillations (Figure 1B).  In the example shown here, a
+simple predator-prey system is driven by environmental white noise
+$\xi_t$, with prey following the equation:
 
@@
 Oscillations arise through the process of stochastic resonance.
-The damped oscillations seen in Figure 1a are analgous to a child sitting still on a swing: friction slowly damps the magnitude of each successive oscillation, while the period of pendulum remains unchanged.
-Driving the pendulum through perturbations timed to match that natural period can quickly create sustained oscillations with large magnitudes, but such perfect timing is not required.
-In the model above, white noise effectively drives the pendulum at all frequencies, though at very low magnitude, since the noise is small relative to the scale of the dynamics.
-However, frequencies that most closely match the natural period of the damped oscillator are amplified, resonating with the natural oscillations.
-This process is sufficient to permit even small noise to drive sustained oscillations in the model indefinitely.
-The power spectrum (the squared norm of the Fourier transform, a measure of what frequencies are present; see, e.g.
-@Black2012) provides a convenient way to visualize the effects of these oscillations.
-The power spectrum  for a deterministic limit cycle would create a perfect sinusoidal oscillation and a resulting power spectrum of a delta spike at that single frequency.
-The power spectrum for the quasi-cycle instead shows a distribution of frequencies, centered at the the resonant frequency of the damped oscillator.
+The damped oscillations seen in Figure 1a are analgous to a child
+sitting still on a swing: friction slowly damps the magnitude of
+each successive oscillation, while the period of pendulum remains
+unchanged.  Driving the pendulum through perturbations timed to
+match that natural period can quickly create sustained oscillations
+with large magnitudes, but such perfect timing is not required.  In
+the model above, white noise effectively drives the pendulum at all
+frequencies, though at very low magnitude, since the noise is small
+relative to the scale of the dynamics.  However, frequencies that
+most closely match the natural period of the damped oscillator are
+amplified, resonating with the natural oscillations.  This process
+is sufficient to permit even small noise to drive sustained
+oscillations in the model indefinitely.  The power spectrum (the
+squared norm of the Fourier transform, a measure of what frequencies
+are present; see, e.g. @Black2012) provides a convenient way to
+visualize the effects of these oscillations.  The power spectrum
+for a deterministic limit cycle would create a perfect sinusoidal
+oscillation and a resulting power spectrum of a delta spike at that
+single frequency.  The power spectrum for the quasi-cycle instead
+shows a distribution of frequencies, centered at the the resonant
+frequency of the damped oscillator.
 
-Early work by @Blarer1999 identified quasi-cycles as a possible explanation for oscillations and periodic outbreaks in pest ecosystems.
-@Bjornstad2004 analyzes the potential for such stochastic effects to drive cyclic behavior in population abundances of Atlantic bluefin tuna and Atlantic Cod, while @Black2010 illustrates the use of power-spectrum approaches to examine the behavior of quasi-cycles in common disease outbreak models, focusing on a case where noise arises explicitly from demographic stochasticity, rather than generic environmental noise as in the model above and elsewhere.
-@Pineda-Krch2007 compare patterns from these stochastically generated cycles to stable limit cycles they resemble.
-Stochastic resonance highlights one way in which even small noise can play an important role in driving large scale patterns where we had previously only sought deterministic explanations.
-This example also illustrates the interplay between nonlinear dynamics and stochasticity that is the hallmark of stochastically driven phenomena.
-A stable node with linear dynamics, such as an Ornstein-Uhlenbeck process, has no resonant frequency, and could not produce sustained oscillations (unless the environmental driver itself had a periodic pattern).
+Early work by @Blarer1999 identified quasi-cycles as a possible
+explanation for oscillations and periodic outbreaks in pest ecosystems.
+@Bjornstad2004 analyzes the potential for such stochastic effects
+to drive cyclic behavior in population abundances of Atlantic bluefin
+tuna and Atlantic Cod, while @Black2010 illustrates the use of
+power-spectrum approaches to examine the behavior of quasi-cycles
+in common disease outbreak models, focusing on a case where noise
+arises explicitly from demographic stochasticity, rather than generic
+environmental noise as in the model above and elsewhere.  @Pineda-Krch2007
+compare patterns from these stochastically generated cycles to
+stable limit cycles they resemble.  Stochastic resonance highlights
+one way in which even small noise can play an important role in
+driving large scale patterns where we had previously only sought
+deterministic explanations. This example also illustrates the
+interplay between nonlinear dynamics and stochasticity that is the
+hallmark of stochastically driven phenomena.  A stable node with
+linear dynamics, such as an Ornstein-Uhlenbeck process, has no
+resonant frequency, and could not produce sustained oscillations
+(unless the environmental driver itself had a periodic pattern).
 
@@
 
-We turn from stochastic resonance to oscillations of a very different sort, which do not resemble any deterministic process, as seen in Figure 3A.
-This pattern is characterized by long stretches during which fluctuations in the state variable remain concentrated around a high average value followed by sudden, rapid transitions into fluctuations concentrated around a lower value.
-This behavior arises through the interaction of stochasticity and alternative stable state dynamics.
-The simulation shown in Figure 3A for the is based on an resource consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics.
-The resource grows according to a logistic growth model and is consumed according to a Holling Type-III functional response curve, to which I have simply introduced the addition of environmental noise:
+We turn from stochastic resonance to oscillations of a very different
+sort, which do not resemble any deterministic process, as seen in
+Figure 3A.  This pattern is characterized by long stretches during
+which fluctuations in the state variable remain concentrated around
+a high average value followed by sudden, rapid transitions into
+fluctuations concentrated around a lower value.  This behavior
+arises through the interaction of stochasticity and alternative
+stable state dynamics.  The simulation shown in Figure 3A for the
+is based on an resource consumption model first proposed by @May1977
+as one of the early examples of the potential for alternative stable
+state dynamics.  The resource grows according to a logistic growth
+model and is consumed according to a Holling Type-III functional
+response curve, to which I have simply introduced the addition of
+environmental noise:
 
 $$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{ growth, } g(X_t)}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{consumption, } c(X_t)} + \xi_t,$$
-where $\xi_t$ is normally distributed environmental variation with mean zero and standard deviation $\sigma$.
-Figure 3B shows separate curves corresponding the parts of the model describing vegetation growth and vegetation consumption, using the same parameters as shown in the simulation panel, 3A.
-Equilibria exist whenever growth balances consumption.
-For the parameters shown, the S-shaped curve of the Type-III consumption pattern crosses the quadratic curve of logistic growth at four separate locations.
-When growth is larger than consumption before the crossing, and thus smaller after, the point is stable, otherwise it is unstable.
-Consequently, we can see two stable points, in the neighborhood of $X_t \approx 1.2$ and $X_t \approx 0.5$, closely matching the average fluctuations seen in the simulation in 3A.
-Stochastic fluctuations drive spontaneous shifts between these alternative stable states.
 
-The concept of a potential well, Figure 3C, is often invoked when describing the behavior of alternative stable states.
-The curve for the potential well, $U(x)$ is defined as the negative integral of the population growth rate,
+where $\xi_t$ is normally distributed environmental variation with
+mean zero and standard deviation $\sigma$. Figure 3B shows separate
+curves corresponding the parts of the model describing vegetation
+growth and vegetation consumption, using the same parameters as
+shown in the simulation panel, 3A.  Equilibria exist whenever growth
+balances consumption.  For the parameters shown, the S-shaped curve
+of the Type-III consumption pattern crosses the quadratic curve of
+logistic growth at four separate locations.  When growth is larger
+than consumption before the crossing, and thus smaller after, the
+point is stable, otherwise it is unstable.  Consequently, we can
+see two stable points, in the neighborhood of $X_t \approx 1.2$ and
+$X_t \approx 0.5$, closely matching the average fluctuations seen
+in the simulation in 3A.  Stochastic fluctuations drive spontaneous
+shifts between these alternative stable states.
+
+The concept of a potential well, Figure 3C, is often invoked when
+describing the behavior of alternative stable states.  The curve
+for the potential well, $U(x)$ is defined as the negative integral
+of the population growth rate,
 
@@
 
-where in our model $f(y)$ is vegetation growth minus consumption, $f(y) = g(y) - c(y)$.
-The motivation comes from physics, where $f$ is a force and $U$ the corresponding potential energy.
-This suggests a convenient analogy to gravity, often implied by drawing a ball inside the curve and referring to this as a "ball-in-cup" diagram.
-The potential function is particularly instructive for understanding stochastic switching under small noise, as in our simulation in Figure 3, where it is straight-forward to show [e.g.
-@Feller; @vanKampen2007] that the probability of a transition is proportional the exponential of the depth of the well from the lowest point to the separation barrier, (a result that is familiar as Arrhenius law of activation energy to grade-school students of chemistry).
-This is effect is visible even in the short simulation shown in 3A, which spends longer times in the lower state which has the deeper well.
-(Readers are encouraged to experiment with this by adjusting the parameters of the simulation in the included code).
-Unfortunately, the literature is not always careful in describing how this potential well diagram arises, which has caused it to become a source for as least as much confusion as it is of insight.
-While the curves in Figure 3B can be clearly associated with underlying mechanisms of rates: logistic growth and consumption with handling time, the same is not true of the potential: it is much harder to have a good intuition about the 'negative integral of the difference growth and consumption,' and how it would change in response to, say, a larger carrying capacity $K$ or higher consumption rate $a$.
-While the x-axis of the potential diagram remains the same: the system state, the conceptual problem is clearly evident in the usual ambiguity describing the y-axis in such ball-in-cup models.
+where in our model $f(y)$ is vegetation growth minus consumption,
+$f(y) = g(y) - c(y)$.  The motivation comes from physics, where $f$
+is a force and $U$ the corresponding potential energy.  This suggests
+a convenient analogy to gravity, often implied by drawing a ball
+inside the curve and referring to this as a "ball-in-cup" diagram.
+The potential function is particularly instructive for understanding
+stochastic switching under small noise, as in our simulation in
+Figure 3, where it is straight-forward to show [e.g. @Feller;
+@vanKampen2007] that the probability of a transition is proportional
+the exponential of the depth of the well from the lowest point to
+the separation barrier, (a result that is familiar as Arrhenius law
+of activation energy to grade-school students of chemistry).  This
+is effect is visible even in the short simulation shown in 3A, which
+spends longer times in the lower state which has the deeper well.
+(Readers are encouraged to experiment with this by adjusting the
+parameters of the simulation in the included code).  Unfortunately,
+the literature is not always careful in describing how this potential
+well diagram arises, which has caused it to become a source for as
+least as much confusion as it is of insight.  While the curves in
+Figure 3B can be clearly associated with underlying mechanisms of
+rates: logistic growth and consumption with handling time, the same
+is not true of the potential: it is much harder to have a good
+intuition about the 'negative integral of the difference growth and
+consumption,' and how it would change in response to, say, a larger
+carrying capacity $K$ or higher consumption rate $a$.  While the
+x-axis of the potential diagram remains the same: the system state,
+the conceptual problem is clearly evident in the usual ambiguity
+describing the y-axis in such ball-in-cup models.
 
@@
 
-Despite the nearly ubiquitous association of a ball-in-cup model with ecological discussions of resilience and stability, this potential well diagram (Figure 3C) is also a source of confusion on the topic.
-As we have seen, the most salient feature of this diagram is the depth of each well: transition probabilities (under small noise) depend only on the depth of the well and not the steepness of the well, the distance between the wells or between well and tipping point barrier.
-However, it is important to remember that because the potential function is defined by the above integral, it in fact already combines two more elementary and mechanistically precise quantities from the growth equation: the slope of the growth rate (stability) and the distance to the tipping point (size of the basin of attraction).
-Both a steeper negative slope or a longer distance to integrate over between stable and unstable points will create a deeper well.
+Despite the nearly ubiquitous association of a ball-in-cup model
+with ecological discussions of resilience and stability, this
+potential well diagram (Figure 3C) is also a source of confusion
+on the topic.  As we have seen, the most salient feature of this
+diagram is the depth of each well: transition probabilities (under
+small noise) depend only on the depth of the well and not the
+steepness of the well, the distance between the wells or between
+well and tipping point barrier.  However, it is important to remember
+that because the potential function is defined by the above integral,
+it in fact already combines two more elementary and mechanistically
+precise quantities from the growth equation: the slope of the growth
+rate (stability) and the distance to the tipping point (size of the
+basin of attraction).  Both a steeper negative slope or a longer
+distance to integrate over between stable and unstable points will
+create a deeper well.
 
-Empirical evidence for stochastic switching in ecological timeseries is limited by the long time scale usually required.
-Because a single stochastic transition between alternative stable states appears as such a dramatic deviation from the previous pattern (as illustrated in Figure 3A), it is often tempting to assume the change could only be the result of some sudden external force or change in the environment, so it is worth remembering that given long enough, these transitions can happen entirely by chance stochasticity [@Boettiger2012; @Drake2013].
-However, these transitions are most likely to observed in systems with relatively shallow potential wells.
-This is mostly likely to occur of some external change is slowly making the potential well shallower and thus transitions more likely -- a phenomenon that has become to be known as "flickering" and suggested as a potential early warning sign of an approaching transition [@Brock2010].
-Examples of flickering between alternative stable states have been suggested in data ranging from geological timescales at the end of glaciation in the Pleistocene [@Taylor1993] to ecological scales over a few decades in lake ecosystems [@Wang2012].
-Stochastic transitions between bi-stable states are thought to play an important role in genetic switches, where among other phenomena this mechanism can explain the emergence of two alternative phenotypes in clonal microbial populations, which can act to buffer the population in a changing environment  [@Balaban2004].
+Empirical evidence for stochastic switching in ecological timeseries
+is limited by the long time scale usually required. Because a single
+stochastic transition between alternative stable states appears as
+such a dramatic deviation from the previous pattern (as illustrated
+in Figure 3A), it is often tempting to assume the change could only
+be the result of some sudden external force or change in the
+environment, so it is worth remembering that given long enough,
+these transitions can happen entirely by chance stochasticity
+[@Boettiger2012; @Drake2013]. However, these transitions are most
+likely to observed in systems with relatively shallow potential
+wells.  This is mostly likely to occur of some external change is
+slowly making the potential well shallower and thus transitions
+more likely -- a phenomenon that has become to be known as "flickering"
+and suggested as a potential early warning sign of an approaching
+transition [@Brock2010].  Examples of flickering between alternative
+stable states have been suggested in data ranging from geological
+timescales at the end of glaciation in the Pleistocene [@Taylor1993]
+to ecological scales over a few decades in lake ecosystems [@Wang2012].
+Stochastic transitions between bi-stable states are thought to play
+an important role in genetic switches, where among other phenomena
+this mechanism can explain the emergence of two alternative phenotypes
+in clonal microbial populations, which can act to buffer the
+population in a changing environment  [@Balaban2004].
 
@@
 
-One of the most interesting aspects of noise is as a source of additional information.
-This can be realized in a variety of ways.
-For instance, in deriving expressions for the variance created by demographic noise above, we observed that different partitions of a logistic growth model can lead to the same macroscopic dynamics (same values for $b$ - $d$), while having different variance (due to different values for $b$+$d$).
-This suggests that the variation observed in such a population could resolve potentially help resolve how dynamics are partitioned between individual birth and death rates (though in practice one would have to first account for any contribution of environmental noise).
+One of the most interesting aspects of noise is as a source of
+additional information.  This can be realized in a variety of ways.
+For instance, in deriving expressions for the variance created by
+demographic noise above, we observed that different partitions of
+a logistic growth model can lead to the same macroscopic dynamics
+(same values for $b$ - $d$), while having different variance (due
+to different values for $b$+$d$).  This suggests that the variation
+observed in such a population could resolve potentially help resolve
+how dynamics are partitioned between individual birth and death
+rates (though in practice one would have to first account for any
+contribution of environmental noise).
 
@@
 
-Noise can provide novel information in other ways as well.
-Stochastic fluctuations act like a continual miniature perturbation experiment, constantly pushing a system away from an equilibrium to reveal its behavior elsewhere in state space.
-Observing the size and auto-correlation of these fluctuations thus provides insight on the stability of an equilibrium point, (as can be seen, for instance in the fluctuation-dissipation result we found for generic birth death models).
-This observation has been exploited with particular success as a potential *early warning sign* that a system is loosing stability; possibly due to an approaching catastrophic transition or saddle-node bifurcation.
-This pattern, known as critical slowing down (in reference to the auto-correlation time in the physics literature) was first proposed an an early warning sign for ecological systems by @Wissel1984, but has received considerable empirical as well as theoretical attention only more recently [@Scheffer2009; @Scheffer2015].
-Increases in two of these early warning indicators, variance and auto-correlation, are illustrated along with the timeseries of a simulation of the dynamics of yeast population growth in Figure 4.
-This simulation is based on a mechanistic model developed by @Dai2012 that closely matches the population dynamics they observed in serial dilutions of budding yeast _Saccharomyces cervisiae_ on a sucrose medium.
-Because the relevant proteins are located in the extra-cellular matrix rather than inside the cells, budding yeast break down sucrose collaboratively, creating an Allee effect below which the population is too small to sustain itself.
-Increasing the serial dilution rate slowly over time drives the stable population towards a fold bifurcation, which is accompanied by the patterns of increasing variance and increasing auto-correlation seen empirically in lab populations [@Dai2012] as well as the numerical simulation shown here.
+Noise can provide novel information in other ways as well.  Stochastic
+fluctuations act like a continual miniature perturbation experiment,
+constantly pushing a system away from an equilibrium to reveal its
+behavior elsewhere in state space. Observing the size and
+auto-correlation of these fluctuations thus provides insight on the
+stability of an equilibrium point, (as can be seen, for instance
+in the fluctuation-dissipation result we found for generic birth
+death models).  This observation has been exploited with particular
+success as a potential *early warning sign* that a system is loosing
+stability; possibly due to an approaching catastrophic transition
+or saddle-node bifurcation.  This pattern, known as critical slowing
+down (in reference to the auto-correlation time in the physics
+literature) was first proposed an an early warning sign for ecological
+systems by @Wissel1984, but has received considerable empirical as
+well as theoretical attention only more recently [@Scheffer2009;
+@Scheffer2015].  Increases in two of these early warning indicators,
+variance and auto-correlation, are illustrated along with the
+timeseries of a simulation of the dynamics of yeast population
+growth in Figure 4.  This simulation is based on a mechanistic model
+developed by @Dai2012 that closely matches the population dynamics
+they observed in serial dilutions of budding yeast _Saccharomyces
+cervisiae_ on a sucrose medium. Because the relevant proteins are
+located in the extra-cellular matrix rather than inside the cells,
+budding yeast break down sucrose collaboratively, creating an Allee
+effect below which the population is too small to sustain itself.
+Increasing the serial dilution rate slowly over time drives the
+stable population towards a fold bifurcation, which is accompanied
+by the patterns of increasing variance and increasing auto-correlation
+seen empirically in lab populations [@Dai2012] as well as the
+numerical simulation shown here.
 
-As we saw in the equations from the system-size approximation, the variance and auto-correlation in question are technically properties of an ensemble, and should be computed over replicate simulations.
-While the laboratory experiments of @Dai2012 do precisely this, conducting identical, independent replicate manipulations in separate beakers, this is not an option in natural populations.
-If the environmental change is sufficiently slow relative to the measurements, one can assume that the system is close to its stationary state over a given window in time, computing these averages over a rolling time window as we illustrate here, rather than across an ensemble of replicates.
+As we saw in the equations from the system-size approximation, the
+variance and auto-correlation in question are technically properties
+of an ensemble, and should be computed over replicate simulations.
+While the laboratory experiments of @Dai2012 do precisely this,
+conducting identical, independent replicate manipulations in separate
+beakers, this is not an option in natural populations.  If the
+environmental change is sufficiently slow relative to the measurements,
+one can assume that the system is close to its stationary state
+over a given window in time, computing these averages over a rolling
+time window as we illustrate here, rather than across an ensemble
+of replicates.
 
@@
 
-The past two decades have seen widespread recognition of the central role that stochasticity plays through the use of mechanistic models can have in creating and understanding ecological phenomena [@Coulson2004; @Black2012].
-We have seen increasing use of master-equation approaches to individual-based models to reflect demographic stochasticity that is amenable to analytical treatment [@Black2012], and much work in understanding the importance of environmental noise, particularly the more realistic but less mathematically convenient notion of colored or auto-correlated noise on population dynamics, persistence, and coexistence [@Ovaskainen2010; @Schreiber2017].
-These deductive models can generate phenomena that resemble familiar patterns such as population cycles for which we have previously sought deterministic explanations, as well as stochastic phenomena with no such analog such as stochastic switching.
-Yet too many of these advances have remained limited to the domain of theory; limited to deductive models which may seek to mimic observed patterns but are rarely used to confront empirical data directly.
-There is good reason for this: statistical inference of even relatively simple models with multiple sources of noise will always be difficult given limited ecological data [@Knape2011].
-However, many stochastic phenomena may suggest new ways to infer rather than merely obscure underlying processes.
+The past two decades have seen widespread recognition of the central
+role that stochasticity plays through the use of mechanistic models
+can have in creating and understanding ecological phenomena
+[@Coulson2004; @Black2012]. We have seen increasing use of
+master-equation approaches to individual-based models to reflect
+demographic stochasticity that is amenable to analytical treatment
+[@Black2012], and much work in understanding the importance of
+environmental noise, particularly the more realistic but less
+mathematically convenient notion of colored or auto-correlated noise
+on population dynamics, persistence, and coexistence [@Ovaskainen2010;
+@Schreiber2017].  These deductive models can generate phenomena
+that resemble familiar patterns such as population cycles for which
+we have previously sought deterministic explanations, as well as
+stochastic phenomena with no such analog such as stochastic switching.
+Yet too many of these advances have remained limited to the domain
+of theory; limited to deductive models which may seek to mimic
+observed patterns but are rarely used to confront empirical data
+directly.  There is good reason for this: statistical inference of
+even relatively simple models with multiple sources of noise will
+always be difficult given limited ecological data [@Knape2011].
+However, many stochastic phenomena may suggest new ways to infer
+rather than merely obscure underlying processes.
 
-The eruption of interest in early warning signals for critical transitions [@Scheffer2009] provides an excellent illustration of this potential.
-A range of stochastic phenomena have been suggested and a tested for their potential to reveal subtle signs of ecosystem change prior to a dramatic shift, and prompted a rich dialog between further theory and new experiment.
-Changes in stochastic properties such as variance or auto-correlation provide evidence of resilience loss that would be invisible to a deterministic model, and non-linear stochastic phenomena such as stochastic switching are observed as evidence of 'flickering' prior to a critical transition.
-In this way, our understanding of stochastic dynamics is translated into novel sources of information about a system which has sparked a wealth of empirical investigation and further theoretical work to test the potential of this approach.
-The study and application of stochastic phenomena to detect ecosystem change is unlikely to be limited to this simple scenario of a critical transition or fold bifurcation.
+The eruption of interest in early warning signals for critical
+transitions [@Scheffer2009] provides an excellent illustration of
+this potential.  A range of stochastic phenomena have been suggested
+and a tested for their potential to reveal subtle signs of ecosystem
+change prior to a dramatic shift, and prompted a rich dialog between
+further theory and new experiment. Changes in stochastic properties
+such as variance or auto-correlation provide evidence of resilience
+loss that would be invisible to a deterministic model, and non-linear
+stochastic phenomena such as stochastic switching are observed as
+evidence of 'flickering' prior to a critical transition.  In this
+way, our understanding of stochastic dynamics is translated into
+novel sources of information about a system which has sparked a
+wealth of empirical investigation and further theoretical work to
+test the potential of this approach.  The study and application of
+stochastic phenomena to detect ecosystem change is unlikely to be
+limited to this simple scenario of a critical transition or fold
+bifurcation.
 
-Precisely because most stochastic phenomena arise through their interaction with non-linearity, the patterns that result can become a lens into better understanding those nonlinear processes and detecting how they may change.
-Meanwhile, the widespread adoption of computational tools and the ability to express theoretical results in numerical algorithms can make any new approaches more readily accessible to application in empirical research.
-In the past two decades, we have set to rest those deterministic skeletons that see noise only as nuisance and embraced the central role noise plays as the creator of ecological phenomena.
-Going forward, I hope and predict that we will see greater discussion and use of stochastic phenomena to infer underlying processes in empirical patterns in similar interactions between theoretical and empirical research.
+Precisely because most stochastic phenomena arise through their
+interaction with non-linearity, the patterns that result can become
+a lens into better understanding those nonlinear processes and
+detecting how they may change.  Meanwhile, the widespread adoption
+of computational tools and the ability to express theoretical results
+in numerical algorithms can make any new approaches more readily
+accessible to application in empirical research. In the past two
+decades, we have set to rest those deterministic skeletons that see
+noise only as nuisance and embraced the central role noise plays
+as the creator of ecological phenomena. Going forward, I hope and
+predict that we will see greater discussion and use of stochastic
+phenomena to infer underlying processes in empirical patterns in
+similar interactions between theoretical and empirical research.
 

### 2017-11-01 appendixA/appendixA.Rmd
WIP

@@
 ```{r}
-knitr::opts_chunk$set(message=FALSE)
+knitr::opts_chunk$set(message=FALSE, cache = TRUE)
 ```
 
-```{r}
+```{r libraries}
 library(nimble)
@@
 
-------------
-
 # Quasi-cycles
@@
 
-## Stochastic inflation
+# Stochastic inflation
 
@@
 Note this is independent of $r$, and increases with $\sigma/K$, the larger the noise $\sigma_g$ as a fraction of the carrying capacity $K$.
-Population sizes can also be inflated relative to their deterministic equilibrium whenever the second derivative is positive (as in the lower equilibrium in the May model of alternative stable states, considered below.)  Stronger nonlinearities (in particular, a larger second derivative at the equilibrium) can drive larger inflationary effects.
+Population sizes can also be inflated relative to their deterministic equilibrium whenever the second derivative is positive (as in the lower equilibrium in the May model of alternative stable states, considered below.)  As predicted by the system size expansion, the deviation from the deterministic result is of order $N^{-1}$.
 
-```{r}
+Because a substantial number of replicate simulations are required to derive a precise mean state, this example will be significantly slower to run than others.
+
+```{r eval=FALSE}
 p <- list(r = .4, K = 1, sigma = .15, N = 500)
@@
 
+## Perform the simulations
 df <- map_dfr(1:10000, 
@@
               .id = "rep")
-```
 
-```{r}
+# Tidy up the result data and write to file, along with analytic prediction for equilibrium
 df %>% 
@@
 ```
-
-----
 

### 2017-11-01 appendixB/appendixB.Rmd
WIP

@@
 Also by the chain rule, we have
-\begin{align}
+\begin{align*}
 \frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  + \frac{\partial \Pi}{\partial \xi}\frac{\ud \xi}{\ud t} 
-\end{align}
+\end{align*}
 Substituting in \eqref{dxidt}, we find:
@@
 &\quad \left( -\Omega^{-1/2} \frac{\partial}{\partial \xi} +  \frac{\Omega^{-1}}{2} \frac{\partial^2}{\partial \xi^2} + \ldots\right)b(\phi\Omega+\xi\Omega^{1/2})\Pi(\xi, t)\nonumber \\
-+&\quad \left( \Omega^{-1/2} \frac{\partial}{\partial \xi} +  \frac{\Omega^{-1}}{2} \frac{\partial^2}{\partial \xi^2} + \ldots\right)d(\phi\Omega+\xi\Omega^{1/2})\Pi(\xi, t)\end{align}
++&\quad \left( \Omega^{-1/2} \frac{\partial}{\partial \xi} +  \frac{\Omega^{-1}}{2} \frac{\partial^2}{\partial \xi^2} + \ldots\right)d(\phi\Omega+\xi\Omega^{1/2})\Pi(\xi, t)
+\end{align}
 
@@
 
-\begin{align}
+\begin{align*}
 \langle n\rangle_s = N\left(1-\frac{e}{c}\right)
-\end{align}
+\end{align*}
 while the steady-state fluctuations are given by \eqref{fluctuation dissipation}, \eqref{n var}:
-\begin{align}
+\begin{align*}
 \sigma_n^2 = N\frac{e}{c}
-\end{align}
+\end{align*}
 
@@
 
-\begin{align}
+\begin{align*}
 	b_n - d_n = \frac{A_x}{\epsilon}
-\end{align}
+\end{align*}
 
 Where $A$ is independent of $\epsilon$.  Similarly we have to scale the sum, which being proportional to the variance must be scaled by
-\begin{align}
+\begin{align*}
 b_n + d_n = \frac{B_x}{\epsilon^2}
-\end{align}
+\end{align*}
 
 We Taylor expand the step operator $\mathbb E$ in the new variable:
-\begin{align}
+\begin{align*}
 \mathbb E^k &= 1+ \epsilon k\frac{\partial}{\partial x} + \frac{\epsilon^2 k^2}{2} \frac{\partial^2}{\partial x^2} + \ldots
-\end{align}
+\end{align*}
 
@@
 
-\begin{align}
+\begin{align*}
 \frac{\ud \phi}{\ud t} &= \alpha_{1,0}(\phi, \psi) \\
-\frac{\ud \psi}{\ud t} &= \beta_{1,0}(\psi) \\
-\end{align}
+\frac{\ud \psi}{\ud t} &= \beta_{1,0}(\psi) 
+\end{align*}
 
@@
 
-\begin{equation}
+\begin{equation*}
 \langle xi^2 \rangle = \frac{ \left( \partial_{\psi} \alpha_{1,0}(\phi,\psi) \right)^2}{\left( \partial_{\phi} \alpha_{1,0}(\phi,\psi) \right)^2 + \partial_{\phi} \alpha_{1,0} (x,y) \tau_c } \sigma_e^2 + \sigma^2_d
-\end{equation}
+\end{equation*}
 

### 2017-11-01 paper.Rmd
WIP

@@
 computational advances provide more powerful and accessible inference
-tools, the task only gets harder.  For instance, few ecological
+tools, the task only gets harder.
+
+For instance, few ecological
 concepts are more central than the notion of density dependence:
@@
 exact stochastic simulation algorithm [SSA; @Gillespie1977], which
-simulates each individual birth or death as a separate event.
-Consider Levins' patch model, where $n$ individuals compete for a
+simulates each individual birth or death as a separate event. Figure 1
+uses this algorithm to compare simulations of the Levins' patch model;
+see Appendix A for example code.  In the Levins'  model,
+$n$ individuals compete for a
 finite number of suitable habitats $N$.  Individuals die a constant
@@
 
+<!-- Figure 4: Early warning signals -->
+```{r tipping, fig.width=7, fig.height=7, fig.cap="Early warning signs of a critical transition in a simulation of increasing serial dilution of budding yeast.
+Panel A: Rolling variance   autocorrelation Time series is shown in panel C.
+Model details and simulation parameters in Appendix A."}
+tipping <- read_csv("appendixA/tipping.csv")
+
+p1 <- tipping %>%
+  select(-x) %>%
+  rename("A. autocorrelation" = autocorrelation, "B. variance" = variance) %>%
+  gather(series, value, -t, -tip_time) %>%
+  ggplot(aes(t, value)) + geom_line(col=colours[[1]]) + 
+  geom_vline(aes(xintercept = tip_time), col=colours[[2]], lty=2) +
+  facet_wrap(~series, ncol = 2, scales="free_y")  
+p2 <- tipping %>% 
+  select(t, population = x, tip_time) %>%
+  ggplot(aes(t, population)) + geom_line(col=colours[1]) + 
+  geom_vline(aes(xintercept = tip_time), col=colours[2], lty=2) + 
+  #scale_y_log10()  + 
+  labs(subtitle = "C. Yeast population size", 
+       caption = "Vertical red dashed line indicates tipping point location")
+
+gridExtra::grid.arrange(p1,p2, layout_matrix = rbind(c(1,1), c(2,2)))
+
+```
+
 Noise can provide novel information in other ways as well.  Stochastic
@@
 Further confusion has been sown by the introduction of the energy-potential diagram into ecology, where the depth of the potential well is not a local property-->
-
-<!-- Figure 4: Early warning signals -->
-```{r tipping, fig.width=7, fig.height=7, fig.cap="Early warning signs of a critical transition in a simulation of increasing serial dilution of budding yeast.
-Panel A: Rolling variance   autocorrelation Time series is shown in panel C.
-Model details and simulation parameters in Appendix A."}
-tipping <- read_csv("appendixA/tipping.csv")
-
-p1 <- tipping %>%
-  select(-x) %>%
-  rename("A. autocorrelation" = autocorrelation, "B. variance" = variance) %>%
-  gather(series, value, -t, -tip_time) %>%
-  ggplot(aes(t, value)) + geom_line(col=colours[[1]]) + 
-  geom_vline(aes(xintercept = tip_time), col=colours[[2]], lty=2) +
-  facet_wrap(~series, ncol = 2, scales="free_y")  
-p2 <- tipping %>% 
-  select(t, population = x, tip_time) %>%
-  ggplot(aes(t, population)) + geom_line(col=colours[1]) + 
-  geom_vline(aes(xintercept = tip_time), col=colours[2], lty=2) + 
-  #scale_y_log10()  + 
-  labs(subtitle = "C. Yeast population size", 
-       caption = "Vertical red dashed line indicates tipping point location")
-
-gridExtra::grid.arrange(p1,p2, layout_matrix = rbind(c(1,1), c(2,2)))
-
-```
 

### 2017-11-01 paper.Rmd
WIP

@@
 The importance of $\tau_c$, or the degree of auto-correlation, which
-emerges from our derivation above, has also been a subject of
+emerges from our derivation above, has also been and continues to be a subject of
 significant interest in the study of stochastic population dynamics
@@
 intervals much longer than the correlation time, consecutive
-measurements are effectively independent: auto-correlated noise is
-all about _relative_ timescales.
+measurements are effectively independent (white noise). 
+The degree of auto-correlation in  noise is all about _relative_ timescales.
 
@@
 equilibrium: any finite population, however big, is guaranteed to
-eventually go extinct under demographic noise.  While extinction
-is certainly a phenomenon that stands in stark contrast to the
-long-term behavior of the corresponding deterministic equation,
-this result is not particularly useful: "eventually" could be longer
-than the existence of the planet. Rather, it is a reminder that we
-must be careful about how we describe the 'long-term' behavior of
-a model.  In place of the mathematical notion of a stationary
-distribution to describe the long-term behavior of a stochastic
-model, analysis of demographic stochasticity focuses on the concept
-of quasi-stationary distributions, which in many cases correspond
-more closely to the equilibrium of their deterministic counterparts
-[@Ovaskainen2010; @Schreiber2017].
-
-<!-- FIXME this section is still a fragment
-
-[@Melbourne2008] Importance of demographic stochasticity and its interaction with environmental stochasticity in driving extinction risk. 
-[@Bonsall2004] Markov chain environmental and demographic noise predator-prey metapopulation model.
-
--->
+eventually go extinct under demographic noise.  However, in most
+models it is possible to define a notion of quasi-stationary 
+distribution concentrated around the deterministic attractor,
+and to prove the probability of extinction decreases exponentially with 
+system size (e.g. see @Ovaskainen2010; @Schreiber2017 also reviews 
+more recent literature which has established these results more generally). 
+Generalizations about persistence and stable coexistence of interacting
+species in stochastic models are difficult due to complex interactions
+between demographic and environmental noise, population structure,
+spatial heterogeneity, magnitude of noise, degree of autocorrelation and so forth
+[e.g. @Higgins1997; @Coulson2001; @Melbourne2008; @Hart2016];
+indeed care must be taken even in defining the approriate notion of 
+persistence [e.g. @Schreiber2006].  Instead of these long-term limits,
+here we will focus on a few examples of stochastic phenomena 
+that can emerge on shorter time scales and provide some intuition for 
+how the interaction between stochastic an nonlinear dynamics can 
+create these surprising results.
 
@@
 Insufficient space left for any treatment of these...
+Permenance, robust permenance
 
-**Bet hedging**
+**Bet hedging**, Lewontin & Cohen 1969; Levins 1969;
 
@@
 seen empirically in lab populations [@Dai2012] as well as the
-numerical simulation shown here.
+numerical simulation shown in Figure 4.
 
@@
 
-<!-- @Holling1996 is responsible for the introduction of the unfortunate term "Engineering Resilience" to mean precisely the same thing as mathematicians and theoretical ecologists have refered to as 'stability' for decades.
-Further confusion has been sown by the introduction of the energy-potential diagram into ecology, where the depth of the potential well is not a local property-->
+<!-- FIXME Fragment, also tie back to flickering results, move some of that here-->
 
 Challenges remain in distinguishing these patterns from other causes [@Boettiger2013a] which will ultimately require more domain-specific information  [@Boettiger2013b].
-
-<!--
-It is important to realize that such approaches can only ever measure local properties such as the return time in response to local perturbation (historically termed *stability*; which @Holling1996 landmark chapter unfortunately re-brands as "engineering resilience") and not global properties such as the size of the attractor basin / distance to the nearest tipping point (*ecological resilience*, or simply *resilience*).
-This is illustrated rather vividly in @Dai2015, which illustrates how a system can lose resilience before losing stability, resulting in a scenario that provides less opportunity for early warning.
-
--->
 

### 2017-11-01 paper/paper.Rmd
update paths for data files too

--- 2017-11-01 paper.Rmd
@@
 Theoretical predictions for mean and plus/minus one standard deviation shown in horizontal re dashed lines."}
-read_csv("appendixA/gillespie.csv", col_types = "cdiddd")  %>% 
+read_csv("../appendixA/gillespie.csv", col_types = "cdiddd")  %>% 
   mutate(system_size = recode(system_size, large = "A. 1000 total sites", small= "B. 100 total sites")) %>%
@@
 Parameters for models shown are r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800; code in Appendix A."}
-read_csv("appendixA/quasicycles.csv") %>%
+read_csv("../appendixA/quasicycles.csv") %>%
   gather(species, population, -t, -sigma) %>%
@@
 
-p1 <- read_csv("appendixA/noisy_switch.csv")  %>%
+p1 <- read_csv("../appendixA/noisy_switch.csv")  %>%
   ggplot(aes(t,x)) + 
@@
 Model details and simulation parameters in Appendix A."}
-tipping <- read_csv("appendixA/tipping.csv")
+tipping <- read_csv("../appendixA/tipping.csv")
 

### 2017-11-01 paper/paper.Rmd
here we go

@@
 output: rticles::elsevier_article  
-layout:  5p # review = doublespace, 3p = singlespace, 5p = two-column
-preamble: "\\newcommand{\\ud}{\\mathrm{d}}"
+layout:  review # review = doublespace, 3p = singlespace, 5p = two-column
+preamble: |
+  \newcommand{\ud}{\mathrm{d}}
+  \linenumbers
 
@@
 available and efficient numerical methods have helped drive research
-towards increasingly more explicit descriptions of how stochastiticity
+towards increasingly more explicit descriptions of how stochasticity
 enters into quantitative models [@Black2012; @Ovaskainen2010].
@@
 have a probability of colonizing an open patch that is simply
-proportional to the fraction of available patches, $1 - n/N$.  Figure
-1 shows the results of two exact SSA simulations of this model with
+proportional to the fraction of available patches, $1 - n/N$.
+
+\begin{align}
+\frac{\ud n}{\ud t} = \underbrace{c n \left(1 - \frac{n}{N}\right)}_{\textrm{birth}} - \underbrace{e n}_{\textrm{death}}, \label{levins}
+\end{align}
+
+Figure 1 shows the results of two exact SSA simulations of this model with
 identical parameters except for the total number of available sites,
@@
 in this case) are much smaller than the system size ($N$ in the
-Levins model).  Note that the dynamics for the average population
-match the population-level equation^[up to fluctuations of order
+Levins model).  The dynamics for the average population
+match the population-level equation.^[up to fluctuations of order
 $\tfrac{1}{N}$. We can actually both observe and successfully
@@
 
-\begin{align}
-\frac{\ud n}{\ud t} = \underbrace{c n \left(1 - \frac{n}{N}\right)}_{\textrm{birth}} - \underbrace{e n}_{\textrm{death}}, \label{levins}
-\end{align}
+The change in variance can be seen
+as a balance between the stochastic contribution, $b(x) + d(x)$
+and the deterministic contribution, $\partial_x \left(b(x) - d(x)\right)$. 
+The deterministic contribution is the slope of the macroscopic
+equation, that is, the eigenvalue indicating the stability of the
+equilibrium point.  For stable points, this value must be negative
+and precisely balance the strictly positive stochastic contribution 
+of $b(x) + d(x)$ (the greater the rate of stochastic events involved, 
+the greater the fluctuations).  This balance at steady-state is known as the
+fluctuation-dissipation theorem [@vanKampen2007], an will prove a useful
+expression in understanding stochastic dynamics.  Because Eq \eqref{bd}
+also describes the time-evolution of these fluctuations, we can similarly
+solve for the correlation between two points separated by time $\tau$.
 
-Note that like the equation for the mean, we have a dynamic equation
-for the time-evolution of the variance in Eq \eqref{bd}.  We can
-solve these equations at steady-state by setting both equations
-zero. For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$,
+For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$,
 we find the average number of occupied patches 
@@
 the relevant metric is the timescale of auto-correlation in the
-environmental noise process relative to the timescale (auto-correlation)
+environmental noise process relative to the auto-correlation timescale
 of the population dynamics.  The Gaussian noise that emerges out
@@
 intervals much longer than the correlation time, consecutive
-measurements are effectively independent (white noise). 
+measurements are effectively independent (i.e. white noise). 
 The degree of auto-correlation in  noise is all about _relative_ timescales.
@@
 coexistence, and dispersal of species [@Schreiber2010; @Hart2016;
-@Schreiber2017].  *Observational error*: Observational error acts
+@Schreiber2017].  *Observational error* acts
 much like another form of stochasticity, arising from sampling
@@
 between demographic and environmental noise, population structure,
-spatial heterogeneity, magnitude of noise, degree of autocorrelation and so forth
+spatial heterogeneity, magnitude of noise, degree of auto-correlation and so forth
 [e.g. @Higgins1997; @Coulson2001; @Melbourne2008; @Hart2016];
-indeed care must be taken even in defining the approriate notion of 
+indeed care must be taken even in defining the appropriate notion of 
 persistence [e.g. @Schreiber2006].  Instead of these long-term limits,
@@
 they observed in serial dilutions of budding yeast _Saccharomyces
-cervisiae_ on a sucrose medium. Because the relevant proteins are
+cerevisiae_ on a sucrose medium. Because the relevant proteins are
 located in the extra-cellular matrix rather than inside the cells,
@@
 
-<!-- FIXME Fragment, also tie back to flickering results, move some of that here-->
+The challenge of identifying warning signs that a system is approaching
+a tipping point has proven a rich area for applications of stochastic
+processes.  We have already seen how in addition to the patterns of
+increasing variance and auto-correlation, our survey of stochastic phenomena
+has highlighted the increased potential to observe stochastic switching,
+or flickering, as the system approaches the tipping point.  Challenges remain
+in distinguishing these patterns from other causes [@Boettiger2013a].
+For instance, as we have seen, variance in a system can increase due to
+external drivers without changing the auto-correlation, and both variance
+and auto-correlation can increase if a system loses stability without
+approaching a saddle-node bifurcation, but will not show flickering.
+The greater our suite of stochastic indicators, the better our ability
+to pin down the processes involved.  Such examples suggest
+that stochastic processes could play a similar role in identifying
+other critical transitions. For instance, stochastic resonance may be 
+visible in advance of a Hopf bifurcation, in which a stable node becomes
+a limit cycle.
 
-Challenges remain in distinguishing these patterns from other causes [@Boettiger2013a] which will ultimately require more domain-specific information  [@Boettiger2013b].
+Most importantly, the framing of stochastic processes as indicators of 
+changes in an ecosystem that might otherwise go unnoticed has proven a
+rich catalyst for bringing these concepts beyond the realm of abstract
+theory. A rich array of experimental and observational tests of early
+warning signals [_i.e._ @Carpenter2011; @Dai2012; and many others, _e.g._ see @Scheffer2015 review] has served no merely to illustrate these phenomena can be seen
+in laboratory and field ecosystems, but also to spark suggestions of
+additional possible indicators and surface a wide array of 
+practical [@Boettiger2013b; @Scheffer2015], statistical [e.g. @Boettiger2012d; @Boettiger2012], and theoretical [e.g. @Dai2015; @Wysham2010] challenges to such application.  This
+creates a two-way dialog between empirical and theoretical work that does
+much to advance the field as a whole.
 

### 2017-11-01 appendixA/appendixA.Rmd
test on drone

@@
 title: "Appendix A: Stochastic simulations of noisy phenomena"
-output: pdf_document
+output: 
+  pdf_document:
+    toc: true
+bibliography: "../paper/refs.bib"
 ---
 
-```{r}
-knitr::opts_chunk$set(message=FALSE, cache = TRUE)
+```{r include=FALSE}
+knitr::opts_chunk$set(message=FALSE, cache = TRUE, warning = FALSE)
 ```
@@
 library(tidyverse)
-
-## plot themes, non-essential
 library(ggthemes)
 library(gridExtra)
-#library(hrbrthemes)
-#ggplot2::theme_set(hrbrthemes::theme_ipsum())
-ggplot2::theme_set(theme_bw())
 colours <- ptol_pal()(2)
+```
 
+```{r include=FALSE}
+## plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
+library(hrbrthemes)
+library(extrafont)
+hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
+extrafont::loadfonts()
+ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
 ```
@@
 ```
+
+# References

### 2017-11-01 appendixA/appendixA.Rmd
tidy up appendixA.Rmd

@@
 ---
+
+# Getting started
+
+This appendix includes the code to create all simulated data shown in the paper (and one additional example involving stochastic inflation that is not shown.)  We must first load a variety of packages we use to create and display the simulations.
+The `regimeshifts` package provides an example of the model of @Dai2012 and must be installed separately from GitHub,
+all other packages can be found on CRAN.
+A reproducible Dockerfile for building an image containing all necessary software to run the examples is also provided in the associated GitHub repository for this paper: <https://github.com/cboettig/noise-phenomena>.
 
@@
 library(extrafont)
+extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
 hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
@@
 
+Exact Gillespie simulations can be run efficiently in R using the `adaptivetau::ssa.exact()` function:
+
 ```{r gillespie}
@@
 
+We will also create a separate data frame (`tibble`) corresponding to our theoretical predictions from the system size expansion:
+
 ```{r}
@@
 
+Bind together both simulation results and theory results and store output data in a plain text `.csv` file.
+
 ```{r}
-## Bind together both simulation results and theory results
 gillespie <- bind_rows("large"=sim_large, 
@@
 
-```{r}
-read_csv("gillespie.csv", col_types = "cdiddd")  %>% 
+Example plotting command to generate the figure as shown in main text:
+
+```{r gillespie, warning=FALSE, fig.width=7, fig.cap="Population dynamics from a Gillespie simulation of the Levins model with large (N=1000, panel A) and small (N=100, panel B) number of sites (blue) show relatively weaker effects of demographic noise in the bigger system.
+Models are otherwise identical, with e = 0.2 and c = 1 (code in appendix A).
+Theoretical predictions for mean and plus/minus one standard deviation shown in horizontal re dashed lines."}
+read_csv("../appendixA/gillespie.csv", col_types = "cdiddd")  %>% 
+  mutate(system_size = recode(system_size, large = "A. 1000 total sites", small= "B. 100 total sites")) %>%
   ggplot(aes(x = time)) + 
-  geom_hline(aes(yintercept = mean), lty=2) + 
-  geom_hline(aes(yintercept = minus_sd), lty=2) + 
-  geom_hline(aes(yintercept = plus_sd), lty=2) + 
-  geom_line(aes(y = n), col="red") +
-  facet_wrap(~system_size, scales = "free_y") + 
-  labs(caption = "theoretical predictions for mean and plus/minus one standard \n deviation shown in horizontal black dashed lines.")
+  geom_hline(aes(yintercept = mean), lty=2, col=colours[2]) + 
+  geom_hline(aes(yintercept = minus_sd), lty=2, col=colours[2]) + 
+  geom_hline(aes(yintercept = plus_sd), lty=2, col=colours[2]) + 
+  geom_line(aes(y = n), col=colours[1]) +
+  facet_wrap(~system_size, scales = "free_y") 
 ```
@@
 
-Understanding the origin of sustained oscillations has long been a central question in ecology [@Hastings1996]
+Simulations of the quasi-cycles can easily be performed in NIMBLE, which allows us to express the stochastic models in the familiar BUGS syntax.
+Change the values of the model and noise parameters below to explore these results.
 
@@
 
-```{r}
+From this quasicycle simulation data we can recreate the plot shown in Figure 2 as follows:
+
+```{r quasicycles, fig.width=7, fig.cap="Panel A: under negligible environmental noise, the introduction of a carrying capacity damps Lotka-Volterra oscillations which eventually settle to a stable state.
+Panel B: small noise induces quasi-cycles in the same model: sustained long-term oscillations predominately around the natural frequency of the original damped oscillations.
+Parameters for models shown are r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800; code in Appendix A."}
 read_csv("quasicycles.csv") %>%
-  gather(species, pop, -t, -sigma) %>%
+  gather(species, population, -t, -sigma) %>%
   mutate(sigma = as.factor(sigma)) %>%
   mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
-  ggplot(aes(t, pop, col=species)) +
-  geom_line() + 
-  facet_wrap(~sigma) + 
-  scale_color_ptol()
+  rename(time = t) %>%
+  ggplot(aes(time, population, col = species)) + geom_line() + 
+  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() 
 ```
@@
 
-May's model:
+Our example of a stochastic osccilato is based on May's model:
 
@@
 ```{r}
-
 p <- list(r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 1e4)
-f <- function(x) x * p$r * (1 - x / p$K)
-g <- function(x) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
-x <- seq(0,2, length.out = 100)
-
-data.frame(x, growth = f(x), consumption = g(x)) %>%
-  gather(curve, y, -x) %>%
-  ggplot(aes(x, y, col=curve)) +
-  geom_line(lwd=1) + 
-  scale_color_ptol() + 
-  labs(x = bquote(X[t]), y = bquote(X[t+1]), subtitle= bquote(bold(B.)))
-```
-
-```{r}
 
@@
 
-tibble(t = seq_along(cmodel$x), x = cmodel$x) %>% write_csv("noisy_switch.csv")
-read_csv("noisy_switch.csv")  %>% ggplot(aes(t,x)) + geom_line()
+tibble(t = seq_along(cmodel$x), x = cmodel$x) %>% 
+  write_csv("noisy_switch.csv")
 ```
 
+Seperately, we will also compute the growth and consumption curves and the potential function for comparison with theory:
+
+```{r}
+theory <- 
+  tibble(x= seq(0,2, length.out = 100)) %>%
+  mutate(g = x * p$r * (1 - x / p$K), 
+         c = p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)) %>%
+  mutate(potential = - cumsum(g - c)) %>%
+  gather(curve, y, -x, -potential)
+
+```
+
+We can now create each of the subplots shown in the paper in Figure 3:
+
+```{r noisy_switch, fig.width=7, fig.height=7, fig.cap="Stochastic switiching in May's model of alternative stable state dynamics.  r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000"}
+
+p1 <- read_csv("../appendixA/noisy_switch.csv")  %>%
+  ggplot(aes(t,x)) + 
+  geom_line(lwd=.1, col=colours[[1]]) +
+  labs(subtitle = bquote(bold(A.)))
+
+p2 <-  theory %>%
+  ggplot(aes(x, y, col=curve)) +
+  geom_line(lwd=1) + 
+  scale_color_ptol() + 
+  labs(x = bquote(X[t]), y = bquote(X[t+1]), subtitle= bquote(bold(B.))) +
+  theme(legend.position="top")
+
+p3 <-  theory %>%
+  ggplot(aes(x, potential)) + 
+  geom_line(col=colours[1], lwd=1) + 
+  coord_cartesian(xlim=c(0,1.6), ylim=c(-1.15,-0.9)) + 
+  labs(subtitle = bquote(bold(C.)))
+
+gridExtra::grid.arrange(p1,p2,p3, layout_matrix = rbind(c(1,1), c(2,3)))
+```
+
+# Tipping points
+
+### Configuration for Dai model
+
+Though not shown in the paper, it is useful to illustrate tipping point location using very small noise in a constant environment and varying only the initial condition.
+Starting just above the tipping point leads to a steady stable state, starting just below the tipping point leads to a sudden crash.
+
+Details of the Dai model are found in the appendix of @Dai2012, my best interpretation of that description can be found in R code in the `regimeshifts` package, <https://github.com/cboettig/regimeshifts>, with archival version available at <https://doi.org/10.5281/zenodo.1013975>.
+
+```{r}
+max_days <- 1000
+
+z <- y <- x <- numeric(max_days)
+z[1] <- y[1] <- x[1] <- 7e4 # 1.7e5
+set.seed(123)
+for(day in 1:(max_days-1)){
+  x[day+1] <- dai(x[day], epsilon = rnorm(1,0, 0.001), DF = 1790)
+  y[day+1] <- dai(y[day], epsilon = rnorm(1,0, 0.001), DF=1799)
+  z[day+1] <- dai(z[day], epsilon = rnorm(1,0, 0.001), DF = 1800)
+
+}
+
+df <- data.frame(t = 1:max_days, x = x, y = y, z = z) %>%
+  gather(series, pop, -t)
+
+ggplot(df, aes(t, pop, col=series)) + 
+  geom_line()
+
+```
+
+We will also need a function for rolling autocorrelation for our warning sign calculation:
+
+$$\rho = \frac{1}{n-1} \frac{\sum_{i = 1}^n \left( x_{i,t} - \textrm{E}(x_t) \right)\left( x_{i,t+1} - \textrm{E}(x_{t+1})  \right)}{\sigma_{x_{t}}  \sigma_{x_{t+1}}}$$
+
+```{r}
+roll_acor <- function(x, lag = 1, ...){
+  x_t1 = lag(x, n = lag)  
+  mu_t = roll_mean(x, ...)
+  mu_t1 = roll_mean(x_t1, ...)
+  s_t = roll_sd(x, ...)
+  s_t1 = roll_sd(x_t1, ...)
+  acor = roll_mean( (x - mu_t) * (x_t1 - mu_t1) / (s_t * s_t1),...)
+  acor
+}
+```
+
+## Simulation of Dai model:
+
+```{r}
+# Slow environmental degredation through small stepwise changes (increasing the serial dilution factor, as in the Dai et al experiment.)
+DF <- as.numeric(sapply(seq(0, 2000, length=9), rep, 40))
+
+# continuous linear increase
+DF <- seq(1, 2000, length=2000)
+tip_time <- 1800
+max_days <- length(DF)
+y <- numeric(max_days)
+
+set.seed(1111)
+y[1] <- 1.76e5
+for(day in 1:(max_days-1)){
+  y[day+1] <- dai(y[day], DF = DF[day])
+}
+
+tip_df <- tibble(t = seq_along(y), x = y)
+
+## also compute warning signs, include in single table
+tip_df %>% 
+  mutate(autocorrelation = roll_acor(x, n = 100, fill=NA),
+         variance = roll_var(x, n = 100, fill=NA),
+         tip_time = tip_time) %>%
+  write_csv("tipping.csv")
+```
+
+We are now ready to create the panels shown in Figre 4:
+
+```{r tipping, fig.width=7, fig.height=7, fig.cap="Early warning signs of a critical transition in a simulation of increasing serial dilution of budding yeast.
+Panel A: Rolling variance   autocorrelation Time series is shown in panel C.
+Model details and simulation parameters in Appendix A."}
+tipping <- read_csv("../appendixA/tipping.csv")
+
+p1 <- tipping %>%
+  select(-x) %>%
+  rename("A. autocorrelation" = autocorrelation, "B. variance" = variance) %>%
+  gather(series, value, -t, -tip_time) %>%
+  ggplot(aes(t, value)) + geom_line(col=colours[[1]]) + 
+  geom_vline(aes(xintercept = tip_time), col=colours[[2]], lty=2) +
+  facet_wrap(~series, ncol = 2, scales="free_y")  
+p2 <- tipping %>% 
+  select(t, population = x, tip_time) %>%
+  ggplot(aes(t, population)) + geom_line(col=colours[1]) + 
+  geom_vline(aes(xintercept = tip_time), col=colours[2], lty=2) + 
+  #scale_y_log10()  + 
+  labs(subtitle = "C. Yeast population size", 
+       caption = "Vertical red dashed line indicates tipping point location")
+
+gridExtra::grid.arrange(p1,p2, layout_matrix = rbind(c(1,1), c(2,2)))
+
+```
+
+------
+
 # Stochastic inflation
+
+While stochastic inflation is mentioned only in a footnote of the manuscript, I include this example to illustrate simulations of an ensemble of replicates, commonly needed to get a better picture of stochastic phenomena.
+This example code is written for clarity rather than performance, and will take longer to run than the other examples.
 
@@
 
-# Tipping points
-
-### Configuration for Dai model
-
-Numerically determine / illustrate tipping point location using small noise:
-
-```{r}
-max_days <- 1000
-
-z <- y <- x <- numeric(max_days)
-z[1] <- y[1] <- x[1] <- 7e4 # 1.7e5
-set.seed(123)
-for(day in 1:(max_days-1)){
-  x[day+1] <- dai(x[day], epsilon = rnorm(1,0, 0.001), DF = 1790)
-  y[day+1] <- dai(y[day], epsilon = rnorm(1,0, 0.001), DF=1799)
-  z[day+1] <- dai(z[day], epsilon = rnorm(1,0, 0.001), DF = 1800)
-
-}
-
-df <- data.frame(t = 1:max_days, x = x, y = y, z = z) %>%
-  gather(series, pop, -t)
-
-ggplot(df, aes(t, pop, col=series)) + 
-  geom_line()
-
-```
-
-$$\rho = \frac{1}{n-1} \frac{\sum_{i = 1}^n \left( x_{i,t} - \textrm{E}(x_t) \right)\left( x_{i,t+1} - \textrm{E}(x_{t+1})  \right)}{\sigma_{x_{t}}  \sigma_{x_{t+1}}}$$
-
-```{r}
-roll_acor <- function(x, lag = 1, ...){
-  x_t1 = lag(x, n = lag)  
-  mu_t = roll_mean(x, ...)
-  mu_t1 = roll_mean(x_t1, ...)
-  s_t = roll_sd(x, ...)
-  s_t1 = roll_sd(x_t1, ...)
-  acor = roll_mean( (x - mu_t) * (x_t1 - mu_t1) / (s_t * s_t1),...)
-  acor
-}
-```
-
-## Simulation of Dai model:
-
-```{r}
-# Stepwise changes
-DF <- as.numeric(sapply(seq(0, 2000, length=9), rep, 40))
-
-# continuous linear increase
-DF <- seq(1, 2000, length=2000)
-tip_time <- 1800
-max_days <- length(DF)
-y <- numeric(max_days)
-
-set.seed(1111)
-y[1] <- 1.76e5
-for(day in 1:(max_days-1)){
-  y[day+1] <- dai(y[day], DF = DF[day])
-}
-
-tip_df <- tibble(t = seq_along(y), x = y)
-
-tip_df %>% 
-  mutate(autocorrelation = roll_acor(x, n = 100, fill=NA),
-         variance = roll_var(x, n = 100, fill=NA),
-         tip_time = tip_time) %>%
-  write_csv("tipping.csv")
-```
-
-```{r}
-tipping <- read_csv("tipping.csv")
-
-p1 <- tipping %>%
-  select(-x) %>%
-  gather(series, value, -t, -tip_time) %>%
-  ggplot(aes(t, value)) + geom_line() + 
-  geom_vline(aes(xintercept = tip_time), col="red", lty=2) +
-  facet_wrap(~series, ncol = 2, scales="free_y") +
-  #theme_ipsum() + 
-  labs(title="A: Early warning signs")
-
-p2 <- tipping %>% 
-  select(t, population = x, tip_time) %>%
-  ggplot(aes(t, population)) + geom_line() + 
-  geom_vline(aes(xintercept = tip_time), col="red", lty=2) + 
-  scale_y_log10() + 
-#  theme_ipsum() + 
-  labs(title="B: Population density",
-              caption = "Vertical red dashed line indicates tipping point location")
-
-gridExtra::grid.arrange(p1,p2, layout_matrix = rbind(c(1,1), c(2,2)))
-
-```
-
 # References

### 2017-11-01 appendixA/appendixA.Rmd
avoid duplicate label
attempt to fix fonts on drone

@@
 
+```{r include=FALSE}
+## plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
+d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
+d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
+write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)
+extrafont::loadfonts()
+
+library(hrbrthemes)
+library(extrafont)
+extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
+hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
+extrafont::loadfonts()
+ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
+```
+
 ```{r libraries}
@@
 colours <- ptol_pal()(2)
-```
-
-```{r include=FALSE}
-## plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
-library(hrbrthemes)
-library(extrafont)
-extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
-hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
-extrafont::loadfonts()
-ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
 ```
@@
 
-```{r gillespie}
+```{r }
 set.seed(1234) # set random number generator seed to be reproducible
@@
                        "small"=sim_small,  
-                       .id = "system_size") %>% bind_rows(theory)
+                       .id = "system_size") %>% 
+  bind_rows(theory)
 write_csv(gillespie, "gillespie.csv")
@@
 read_csv("../appendixA/gillespie.csv", col_types = "cdiddd")  %>% 
-  mutate(system_size = recode(system_size, large = "A. 1000 total sites", small= "B. 100 total sites")) %>%
+  mutate(system_size = recode(system_size, 
+                              large = "A. 1000 total sites", 
+                              small= "B. 100 total sites")) %>%
   ggplot(aes(x = time)) + 
@@
 f <- function(constants){
-  model <- compileNimble(nimbleModel(quasicycle, constants = constants, inits = list(x0 = 1, y0 = 1)))
+  model <- compileNimble(nimbleModel(quasicycle, 
+                                     constants = constants, 
+                                     inits = list(x0 = 1, y0 = 1)))
   set.seed(123)
   simulate(model)
-  tibble(t = seq_along(model$x), x = model$x, y = model$y, sigma = constants$sigma_x)
+  tibble(t = seq_along(model$x), 
+         x = model$x, 
+         y = model$y, 
+         sigma = constants$sigma_x)
 }
@@
 ```{r}
-# Slow environmental degredation through small stepwise changes (increasing the serial dilution factor, as in the Dai et al experiment.)
+## Slow environmental degredation through small stepwise changes
+## (increasing the serial dilution factor, as in the Dai et al experiment.)
 DF <- as.numeric(sapply(seq(0, 2000, length=9), rep, 40))
@@
 
-# Tidy up the result data and write to file, along with analytic prediction for equilibrium
+# Tidy up the result data and write to file, 
+# along with analytic prediction for equilibrium
 df %>% 

### 2017-11-01 appendixA/appendixA.Rmd
more troubleshooting on fonts

@@
 
-```{r include=FALSE}
+```{r echo=FALSE}
 ## plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
-d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
-d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
-write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)
-extrafont::loadfonts()
-
 library(hrbrthemes)
@@
 hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts
+
+d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
+d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
+write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)
+
 extrafont::loadfonts()

### 2017-11-01 appendixA/appendixA.Rmd
try without cache please
try without bquote text?

@@
 
-```{r echo=FALSE}
-## plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
-library(hrbrthemes)
-library(extrafont)
-extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
-hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts
-
-d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
-d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
-write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)
-
-extrafont::loadfonts()
-ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
-```
-
 ```{r libraries}
@@
 colours <- ptol_pal()(2)
+```
+
+```{r echo=FALSE}
+## plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
+library(hrbrthemes)
+library(extrafont)
+#extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
+hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts
+
+#d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
+#d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
+#write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)
+
+extrafont::loadfonts()
+ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
 ```
@@
   geom_line(lwd=.1, col=colours[[1]]) +
-  labs(subtitle = bquote(bold(A.)))
+  labs(subtitle = "A.")
 
@@
   scale_color_ptol() + 
-  labs(x = bquote(X[t]), y = bquote(X[t+1]), subtitle= bquote(bold(B.))) +
+ labs(
+ #  x = bquote(X[t]), y = bquote(X[t+1]), 
+   subtitle= "B.") +
   theme(legend.position="top")
@@
   coord_cartesian(xlim=c(0,1.6), ylim=c(-1.15,-0.9)) + 
-  labs(subtitle = bquote(bold(C.)))
+  labs(subtitle = "C.")
 

### 2017-11-01 appendixA/appendixA.Rmd
try with Cairo

--- 2017-11-01 appendixA/appendixA.Rmd
@@
 ```{r include=FALSE}
-knitr::opts_chunk$set(message=FALSE, cache = TRUE, warning = FALSE)
-```
-
-```{r echo=FALSE}
-## plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
-library(hrbrthemes)
-library(extrafont)
-extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
-hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts
-
-d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
-d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
-write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)
-
-extrafont::loadfonts()
-ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
+## note: needs library Cairo
+knitr::opts_chunk$set(message=FALSE, cache = TRUE, warning = FALSE,  dev="CairoPDF")
 ```
@@
 colours <- ptol_pal()(2)
+```
+
+```{r echo=FALSE}
+## plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
+library(Cairo)
+library(hrbrthemes)
+library(extrafont)
+#extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
+hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts
+
+#d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
+#d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
+#write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)
+
+extrafont::loadfonts()
+ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
 ```

### 2017-11-01 paper/paper.Rmd
fix citation

@@
   - quasi-cycles
+  - tipping points
 
@@
 additional possible indicators and surface a wide array of 
-practical [@Boettiger2013b; @Scheffer2015], statistical [e.g. @Boettiger2012d; @Boettiger2012], and theoretical [e.g. @Dai2015; @Wysham2010] challenges to such application.  This
+practical [@Boettiger2013b; @Scheffer2015], statistical [e.g. @Boettiger2012d; @Boettiger2012], and theoretical [e.g. @Dai2015; @Hastings2010] challenges to such application.  This
 creates a two-way dialog between empirical and theoretical work that does

### 2017-11-01 appendixA/appendixA.Rmd
submitted for review

@@
 ## note: needs library Cairo
-knitr::opts_chunk$set(message=FALSE, cache = TRUE, warning = FALSE,  dev="CairoPDF")
+knitr::opts_chunk$set(message=FALSE, cache = 1, warning = FALSE,  dev="CairoPDF")
 ```
 
-```{r libraries}
+```{r libraries, cache=FALSE}
 library(nimble)
@@
 
-```{r echo=FALSE}
+```{r echo=FALSE, cache=FALSE}
 ## plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
@@
 
-```{r gillespie, warning=FALSE, fig.width=7, fig.cap="Population dynamics from a Gillespie simulation of the Levins model with large (N=1000, panel A) and small (N=100, panel B) number of sites (blue) show relatively weaker effects of demographic noise in the bigger system.
-Models are otherwise identical, with e = 0.2 and c = 1 (code in appendix A).
-Theoretical predictions for mean and plus/minus one standard deviation shown in horizontal re dashed lines."}
+```{r gillespie, warning=FALSE, fig.width=7}
 read_csv("../appendixA/gillespie.csv", col_types = "cdiddd")  %>% 
@@
 
-```{r quasicycles, fig.width=7, fig.cap="Panel A: under negligible environmental noise, the introduction of a carrying capacity damps Lotka-Volterra oscillations which eventually settle to a stable state.
-Panel B: small noise induces quasi-cycles in the same model: sustained long-term oscillations predominately around the natural frequency of the original damped oscillations.
-Parameters for models shown are r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800; code in Appendix A."}
+```{r quasicycles, fig.width=7}
 read_csv("quasicycles.csv") %>%
@@
 
-```{r noisy_switch, fig.width=7, fig.height=7, fig.cap="Stochastic switiching in May's model of alternative stable state dynamics.  r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000"}
+```{r noisy_switch, fig.width=7, fig.height=7}
 
@@
 
-```{r tipping, fig.width=7, fig.height=7, fig.cap="Early warning signs of a critical transition in a simulation of increasing serial dilution of budding yeast.
-Panel A: Rolling variance   autocorrelation Time series is shown in panel C.
-Model details and simulation parameters in Appendix A."}
+```{r tipping, fig.width=7, fig.height=7}
 tipping <- read_csv("../appendixA/tipping.csv")

### 2017-11-01 appendixA/appendixA.Rmd
test fonts on Drone
one more go at registering roboto font embedding in Drone

@@
 library(gridExtra)
+ggplot2::theme_set(theme_bw())
 colours <- ptol_pal()(2)
@@
 
-```{r echo=FALSE, cache=FALSE}
+```{r echo=FALSE, cache=FALSE, eval=TRUE}
 ## plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
@@
 library(extrafont)
-#extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
+extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
 hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts
 
-#d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
-#d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
-#write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)
+d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
+d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
+write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)
 
+extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
+hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
 extrafont::loadfonts()

### 2017-11-01 appendixA/appendixA.Rmd
try in separate R sessions

@@
 library(extrafont)
-extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
-hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts
 
-d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
-d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
-write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)
-
-extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
-hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
 extrafont::loadfonts()

### 2017-11-01 appendixA/appendixA.Rmd
streamline font setup?

@@
 library(extrafont)
+#extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
+
+## One-time setup
+hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
+d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
+d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
+write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)
+
+#extrafont::font_import(prompt=FALSE)    # One-time use to set up fonts, can be slow.
+#hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts
 

### 2017-11-01 appendixA/appendixA.Rmd
drop commented calls

--- 2017-11-01 appendixA/appendixA.Rmd
@@
 
+## One-time setup
+hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
+d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
+d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
+write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)
+
+## load fonts (needed each time)
 extrafont::loadfonts()

### 2017-11-06 paper/paper.Rmd
add required title page

@@
   \linenumbers
-
-#preamble: "\\usepackage[nomarkers]{endfloat}"
+  \usepackage[nomarkers]{endfloat}
 
 abstract: |
+  # Article Type: Review & Synthesis
+  # Running title: Noise phenomena
+  # Figures: 4
+  # Tables: 0
+  # Word count: 7488
+  # Word count in abstract: 175
+  # Number of References: 54
+  # Archive statement:
+
+  All data (simulated data and code) is made available on GitHub (<https://github.com/cboettig/noise-phenomena>), and if accepted, will be archived with provided DOI in an appropriate scientific data repositoriy.
+
+  # Keywords: stochasticity, demographic noise, environmental noise, colored noise, quasi-cycles, tipping points
+
+  \newpage
+
+  # Abstract
+
   I provide a brief synthesis of ecological research in stochastic phenomena.
@@
 
-keywords:
-  - stochasticity
-  - demographic noise
-  - environmental noise
-  - colored noise
-  - quasi-cycles
-  - tipping points
+#keywords:
+#  - stochasticity
+#  - demographic noise
+#  - environmental noise
+#  - colored noise
+#  - quasi-cycles
+#  - tipping points
 
@@
 -->
+
+\newpage
 

### 2018-02-06 paper/paper.Rmd
Revisions in progress

@@
 output: rticles::elsevier_article  
-layout:  review # review = doublespace, 3p = singlespace, 5p = two-column
+layout:  3p # review = doublespace, 3p = singlespace, 5p = two-column
 preamble: |
   \newcommand{\ud}{\mathrm{d}}
+  \usepackage[nomarkers]{endfloat}
+  \journal{Ecology Letters}
   \linenumbers
-  \usepackage[nomarkers]{endfloat}
+  \usepackage{setspace}
+  \doublespacing
 
@@
 Yet despite this well-earned reputation, noise is often interesting in its own right: noise can induce novel phenomena that could not be understood from some underlying determinstic model alone.
-Nor is all noise the same, and close examination of differences in in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.
+Nor is all noise the same, and close examination of differences in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.
 Yet with each aspect of stochasticity leading to some new or unexpected behavior, the time is right to move beyond the familiar refrain of "everything is important" [@Bjornstad2001].
@@
 ```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="CairoPDF", fig.width=7, fig.height=3.5)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="cairo_pdf", fig.width=7, fig.height=3.5)
 library(tidyverse)
@@
 ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
-
-#ggplot2::theme_set(theme_bw()) # simpler theme if you do not have these fonts
 ```
@@
 patterns we are trying to infer.  This is a simple consequence of
-modeling: as discussed above, variables of interest to the processes
+modeling: variables of interest to the processes
 in question are modeled explicitly (e.g. change in population density
@@
 models of density dependence in the majority of available time
-series.  Faced with such uncertainty, it is tempting to throw up
-our hands and wait until some prophesied deluge of streaming
-micro-sensors stapled to every organism provide enough signal to
-beat back the noise. Fortunately, this has not been the path taken.
+series.  Faced with such uncertainty, it is tempting to abandon
+any such attempts until much longer and higher resolution data 
+becomes available. Fortunately, this has not been the path taken.
 
@@
 this demographic stochasicity is that it scales with the square
-root of the "system size," so that for very large populations the
+root of the "system size," [@vanKampen2007] so that for very large populations the
 resulting fluctuations are relatively small. On the other hand,
@@
 mechanistically lead to the event of a 'birth' in the population
-as a probability: in a population of $N$ identical individuals at
+as a probability: In a population of $N$ identical individuals at
 time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate
-that can depend on both the population size $N$) and increases the
-population size, $N$ to $N+1$, and deaths occur with probability
-$d_t(N_t)$, decreasing the population size, $N$ to $N-1$.  Assuming
+that can depend on the population size, $N$), which increases the
+population size to $N+1$. Similarly, death events occur with probability
+$d_t(N_t)$, decreasing the population size by one individual, to $N-1$.  Assuming
 each of these events are independent, this is a state-dependent
@@
 ways to leave the state: a simple expression of probability balance
-known as the "master equation."
+known as the "master equation" [@vanKampen2007].
 
@@
 exact stochastic simulation algorithm [SSA; @Gillespie1977], which
-simulates each individual birth or death as a separate event. Figure 1
-uses this algorithm to compare simulations of the Levins' patch model;
-see Appendix A for example code.  In the Levins'  model,
+simulates each individual birth or death as a separate event. Figure \ref{gillespie}
+uses this algorithm to compare simulations of @Levins1969 patch model;
+(see Appendix A for example code).  In the Levins'  model,
 $n$ individuals compete for a
@@
 
-Figure 1 shows the results of two exact SSA simulations of this model with
+Figure \ref{gillespie} shows the results of two exact SSA simulations of this model with
 identical parameters except for the total number of available sites,
@@
 Models are otherwise identical, with e = 0.2 and c = 1 (code in appendix A).
-Theoretical predictions for mean and plus/minus one standard deviation shown in horizontal re dashed lines."}
+Theoretical predictions for mean and plus/minus one standard deviation shown in horizontal re dashed lines.
+\\label{gillespie}"}
 read_csv("../appendixA/gillespie.csv", col_types = "cdiddd")  %>% 
@@
 in direct proportion to $N$, resulting in the smaller relative size
-of fluctuations in the larger system in Figure 1.  We can also see
+of fluctuations in the larger system in Figure \ref{gillespie}.  We can also see
 that different ways of partitioning the same macroscopic equation
@@
 dynamics are independent of the state of our system, $x$. This
-environmental variation no doubt also arise from some more detailed
+environmental variation no doubt also arises from some more detailed
 lower level process, but we need not be explicit about those dynamics
@@
 [e.g.@Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
-Early work rarely often assumed environmental noise would be
+Early work often assumed environmental noise would be
 uncorrelated as a mathematical convenience.  As the derivation of
@@
 another form of random variation in a model [@Coulson2001;
-@Schreiber2010].  *Spatial heterogeneity* can likewise be reflected
+@Schreiber2010; @Vindenes2015].  *Spatial heterogeneity* can likewise be reflected
 in stochastic parameters of a spatially implicit model, or as
@@
 with additional observations, we can narrow the uncertainty over
-parameters or candidate models, while stochasticity remains an
-inherent source of variation. We will this not consider model and
-parameter uncertainty, as they are not sources of noise.  *Deterministic
-chaos* [@Desharnais2001] is another mechanism of introducing
+parameters or candidate models. Thus we will not consider model and
+parameter uncertainty, as they do not reflect intrinsic phenomena of 
+the dynamics but only our own ignorance about those dynamics.  
+*Deterministic chaos* [@Desharnais2001] is another mechanism of introducing
 significant variation into a simple model. Chaos can a look a lot
@@
 equilibrium: any finite population, however big, is guaranteed to
-eventually go extinct under demographic noise.  However, in most
+eventually go extinct under demographic noise [e.g. @Kendall1949; @Lande2003].  However, in most
 models it is possible to define a notion of quasi-stationary 
@@
 Panel B: small noise induces quasi-cycles in the same model: sustained long-term oscillations predominately around the natural frequency of the original damped oscillations.
-Parameters for models shown are r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800; code in Appendix A."}
+Parameters for models shown are r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800; code in Appendix A.
+\\label{quasicycles}"}
 read_csv("../appendixA/quasicycles.csv") %>%
@@
 produces only damped oscillations, eventually settling into a stable
-state (Figure 1A).  However, the introduction of relatively small
+state (Figure \ref{quasicycles}A).  However, the introduction of relatively small
 noise into the system can be sufficient to excite the system into
-sustained oscillations (Figure 1B).  In the example shown here, a
+sustained oscillations (Figure \ref{quasicycles}B).  In the example shown here, a
 simple predator-prey system is driven by environmental white noise
@@
 Here we have assumed for the prey, $\xi_x \sim \mathcal{N(0, \sigma_x)}$, and similarly for the predator driven by $\xi_y$.
-Figure 1A shows examples for noise $\sigma_x = \sigma_y = 10^{-5}$, while Figure 1B shows $\sigma_x = \sigma_y = 0.01$.
+Figure \ref{quasicycles}A shows examples for noise $\sigma_x = \sigma_y = 10^{-5}$, while Figure \ref{quasicycles}B shows $\sigma_x = \sigma_y = 0.01$.
 
 Oscillations arise through the process of stochastic resonance.
-The damped oscillations seen in Figure 1a are analgous to a child
+The damped oscillations seen in Figure \ref{quasicycles}A are analgous to a child
 sitting still on a swing: friction slowly damps the magnitude of
@@
 resonant frequency, and could not produce sustained oscillations
-(unless the environmental driver itself had a periodic pattern).
+(unless the environmental driver itself had a periodic pattern)
+[e.g. @Gardiner2009; @Black2010].
 
 <!-- See additional citations in Coulson2004 and others that mention stochastic resonance --> 
+<!-- [@Keeling2001; @Coulson2004;]  -->
 
@@
 sort, which do not resemble any deterministic process, as seen in
-Figure 3A.  This pattern is characterized by long stretches during
+Figure \ref{noisy_switch}A.  This pattern is characterized by long stretches during
 which fluctuations in the state variable remain concentrated around
@@
 arises through the interaction of stochasticity and alternative
-stable state dynamics.  The simulation shown in Figure 3A for the
+stable state dynamics.  The simulation shown in Figure \ref{noisy_switch}A for the
 is based on an resource consumption model first proposed by @May1977
@@
 where $\xi_t$ is normally distributed environmental variation with
-mean zero and standard deviation $\sigma$. Figure 3B shows separate
+mean zero and standard deviation $\sigma$. Figure \ref{noisy_switch}B shows separate
 curves corresponding the parts of the model describing vegetation
@@
 
-The concept of a potential well, Figure 3C, is often invoked when
+The concept of a potential well, Figure \ref{noisy_switch}C, is often invoked when
 describing the behavior of alternative stable states.  The curve
@@
 stochastic switching under small noise, as in our simulation in
-Figure 3, where it is straight-forward to show [e.g. @Feller;
+Figure \ref{noisy_switch}, where it is straight-forward to show [e.g. @Feller;
 @vanKampen2007] that the probability of a transition is proportional
-the exponential of the depth of the well from the lowest point to
+to the exponential of the depth of the well from the lowest point to
 the separation barrier, (a result that is familiar as Arrhenius law
 of activation energy to grade-school students of chemistry).  This
-is effect is visible even in the short simulation shown in 3A, which
+effect is visible even in the short simulation shown in 3A, which
 spends longer times in the lower state which has the deeper well.
@@
 least as much confusion as it is of insight.  While the curves in
-Figure 3B can be clearly associated with underlying mechanisms of
+Figure \ref{noisy_switch}B can be clearly associated with underlying mechanisms of
 rates: logistic growth and consumption with handling time, the same
@@
 
-<!--Figure 3: Noisy switch -->
-```{r noisy_switch, fig.width=7, fig.height=7, fig.cap="Stochastic switiching in May's model of alternative stable state dynamics.  r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000"}
+<!--Figure \ref{noisy_switch}: Noisy switch -->
+```{r noisy_switch, fig.width=7, fig.height=7, fig.cap="Stochastic switiching in May's model of alternative stable state dynamics.
+r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000.
+\\label{noisy_switch}"}
 
@@
 with ecological discussions of resilience and stability, this
-potential well diagram (Figure 3C) is also a source of confusion
+potential well diagram (Figure \ref{noisy_switch}C) is also a source of confusion
 on the topic.  As we have seen, the most salient feature of this
@@
 Empirical evidence for stochastic switching in ecological timeseries
-is limited by the long time scale usually required. Because a single
+is limited by the long time scale usually required 
+[but see @Keeling2001; @Clements2016; @Clements2017]. Because a single
 stochastic transition between alternative stable states appears as
 such a dramatic deviation from the previous pattern (as illustrated
-in Figure 3A), it is often tempting to assume the change could only
+in Figure \ref{noisy_switch}A), it is often tempting to assume the change could only
 be the result of some sudden external force or change in the
@@
 to different values for $b$+$d$).  This suggests that the variation
-observed in such a population could resolve potentially help resolve
+observed in such a population could potentially help resolve
 how dynamics are partitioned between individual birth and death
@@
 Panel A: Rolling variance   autocorrelation Time series is shown in panel C.
-Model details and simulation parameters in Appendix A."}
+Model details and simulation parameters in Appendix A.
+\\label{tipping}"}
 tipping <- read_csv("../appendixA/tipping.csv")
@@
 timeseries of a simulation of the dynamics of yeast population
-growth in Figure 4.  This simulation is based on a mechanistic model
+growth in Figure \ref{tipping}.  This simulation is based on a mechanistic model
 developed by @Dai2012 that closely matches the population dynamics
@@
 seen empirically in lab populations [@Dai2012] as well as the
-numerical simulation shown in Figure 4.
+numerical simulation shown in Figure \ref{tipping}.
 
@@
 theory. A rich array of experimental and observational tests of early
-warning signals [_i.e._ @Carpenter2011; @Dai2012; and many others, _e.g._ see @Scheffer2015 review] has served no merely to illustrate these phenomena can be seen
-in laboratory and field ecosystems, but also to spark suggestions of
-additional possible indicators and surface a wide array of 
-practical [@Boettiger2013b; @Scheffer2015], statistical [e.g. @Boettiger2012d; @Boettiger2012], and theoretical [e.g. @Dai2015; @Hastings2010] challenges to such application.  This
-creates a two-way dialog between empirical and theoretical work that does
-much to advance the field as a whole.
+warning signals [_i.e._ @Carpenter2011; @Dai2012; @Clements2016; @Clements2017;
+and many others, _e.g._ see @Scheffer2015 review] has served no merely
+to illustrate these phenomena can be seen in laboratory and field ecosystems,
+but also to spark suggestions of additional possible indicators and surface
+a wide array of practical [@Boettiger2013b; @Scheffer2015], statistical
+[e.g. @Boettiger2012d; @Boettiger2012], and theoretical [e.g. @Dai2015;
+@Hastings2010] challenges to such application.  This creates a two-way dialog
+between empirical and theoretical work that does much to advance the field
+as a whole.
 
@@
 this potential.  A range of stochastic phenomena have been suggested
-and a tested for their potential to reveal subtle signs of ecosystem
+and tested for their potential to reveal subtle signs of ecosystem
 change prior to a dramatic shift, and prompted a rich dialog between

### 2018-02-07 appendixA/appendixA.Rmd
WIP

--- 2017-11-01 appendixA/appendixA.Rmd
@@
 ## note: needs library Cairo
-knitr::opts_chunk$set(message=FALSE, cache = 1, warning = FALSE,  dev="CairoPDF")
+knitr::opts_chunk$set(message=FALSE, cache = 1, warning = FALSE,  dev="cairo_pdf")
 ```
@@
 library(RcppRoll)
-library(regimeshifts) ## devtools::install_github("cboettig/regimeshifts")
-library(adaptivetau)
+## devtools::install_github("cboettig/regimeshifts")
+library(regimeshifts) 
+library(adaptivetau)  ## For Gillespie simulation only
 library(tidyverse)
@@
 library(gridExtra)
-ggplot2::theme_set(theme_bw())
+
+## Set a default color pallete
 colours <- ptol_pal()(2)
@@
 
+## One-time setup
+hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
+#d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
+#d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
+#write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)
+
+## load fonts (needed each time)
 extrafont::loadfonts()
@@
 
-Levins' Patch model as an individual-based birth-death process:
+Levins' Patch model [@Levins1969] as an individual-based birth-death process:
 
@@
 
-We simulate this using the Gillespie's exact stochastic simulation algorithm (SSA), as implemented in the `adaptivetau` R package.
+We simulate this using the Gillespie's exact stochastic simulation algorithm (SSA) [@Gillespie1977; see overview in @Hastings2012], as implemented in the `adaptivetau` R package.
 We declare the transition events: birth increases the state $n$ to $n+1$, and death $n$ to $n-1$, and describe the rates associated with them:
@@
 
+## Run simulations
 sim_small <- adaptivetau::ssa.exact(init.values = c(n = 50),
@@
                        transitions, rateF,
-                       params = list(c=c, e=e, N=1000),
+                       params = list(c = c, e = e, N = 1000),
                        tf=30) %>% as_tibble()
@@
 
-We will also create a separate data frame (`tibble`) corresponding to our theoretical predictions from the system size expansion:
+We will also create a separate data frame (`tibble`) corresponding to our theoretical predictions from the system size expansion.
+We use popular functions from the `tidyverse` suite of R packages for the manipulation of `data.frames`.
+Below, the `mutate` function is used to add new columns for plus and minuse one standard deviation based on theoretical calculation of $\sqrt{N\tfrac{e}{c}}$, and the `map_dfr` function is used to apply this to both `small` and `large` cases.
+We use `tidyverse` syntax  [see @Ross2017] instead of base R in these examples for the following reasons:
+
+- the resulting code is more concise, due to use of higher-level relational data abstractions
+- the resulting code is closer to natural semantics.
+This is in part because consecutive steps are connected by pipes `%>%`rather than temporary variable assignment, and in part by the use of functions named by verbs corresponding to standard relational data concepts.
+- `tidyverse` is widely used and well documented, see <http://r4ds.had.co.nz/> for an accessible and thorough introduction.
 
@@
 ```{r}
-gillespie <- bind_rows("large"=sim_large, 
-                       "small"=sim_small,  
+## combine data.frames, using column called "system_size"
+## to indicate large noise or small noise simulations
+gillespie <- 
+  bind_rows("large" = sim_large, 
+            "small" = sim_small,  
             .id = "system_size") %>% 
   bind_rows(theory)
+
+## Write data to file
 write_csv(gillespie, "gillespie.csv")
@@
 Change the values of the model and noise parameters below to explore these results.
+While the models below can easily be implemented in base R, the BUGS syntax for expressing a stochastic model provides a more generic formulation that can easily be used in parameter estimation as well; see @nimble.
+This can also imporve computational performance because @nimble compiles these models into C++ code.
 
 ```{r}
-
+## Define the BUGS model:
 quasicycle  <- nimble::nimbleCode({
-
   x[1] <- x0
   y[1] <- y0
-
   for(t in 1:(N-1)){
@@
   }
-
 })
 
+## creates a data.frame of parameter combinations for each noise level
 p <- 
@@
 
+## define a function to perform a simulation given a set of parameters (constants)
 f <- function(constants){
@@
                                      inits = list(x0 = 1, y0 = 1)))
+  ## set random seed for reproducibility
   set.seed(123)
+  ## Run the actual simulation, using nimble
   simulate(model)
+
+  ## Assemble the results into a data.frame (tibble).  
   tibble(t = seq_along(model$x), 
@@
 
+## Do this simulation for each set of parameters, given 
 quasicycle_df <- p %>% rowwise() %>% do(f(.))
+
+## Write results to file for later reference and plotting
 write_csv(quasicycle_df, "quasicycles.csv")
@@
 read_csv("quasicycles.csv") %>%
+  ## Data in long from
   gather(species, population, -t, -sigma) %>%
+  ## Re-arrange and and re-lable for the plots:
   mutate(sigma = as.factor(sigma)) %>%
@@
   tibble(x= seq(0,2, length.out = 100)) %>%
+  ## As before, we add columns with mutate
   mutate(g = x * p$r * (1 - x / p$K), 
@@
   mutate(potential = - cumsum(g - c)) %>%
+  ## assemble resulting data into "long" form with the
+  ## column headings as named here:
   gather(curve, y, -x, -potential)
@@
 max_days <- 1000
+## Perform three simultaneous simulations, x,y,z, using
+## different dilution factors DF
 
+## Intialize variables
 z <- y <- x <- numeric(max_days)
 z[1] <- y[1] <- x[1] <- 7e4 # 1.7e5
+
+## set random seed for reproduciblity
 set.seed(123)
+
+## simulate for max_days
 for(day in 1:(max_days-1)){
@@
 
+## gather results into long data.frame 
 df <- data.frame(t = 1:max_days, x = x, y = y, z = z) %>%
@@
 
+## and plot results
 ggplot(df, aes(t, pop, col=series)) + 
@@
 
-We are now ready to create the panels shown in Figre 4:
+We are now ready to create the panels shown in Figure 4:
 
@@
 ```{r eval=FALSE}
+
+## Parameters
 p <- list(r = .4, K = 1, sigma = .15, N = 500)
@@
 
+## BUGS model
 logistic  <- nimble::nimbleCode({
-
   x[1] <- x0
@@
   }
-
 })
 
+## Compile and run model
 model <- 
@@
   ggplot(aes(t,value, col=statistic, lty=statistic)) + 
- # ggplot(aes(t,ave)) +
   geom_line(lwd=1) + 

### 2018-02-09 paper/paper.Rmd
adding refs

@@
   # Tables: 0
-  # Word count: 7488
-  # Word count in abstract: 175
-  # Number of References: 54
+  # Word count: 
+  # Word count in abstract: 
+  # Number of References: 
   # Archive statement:
@@
 
-  I provide a brief synthesis of ecological research in stochastic phenomena.
    Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath.
@@
 tools, the task only gets harder.
+
+<!-- Need concise overview of stochasticity literature -->
 
@@
 # Origins of noise
+
+<!-- Shorten this, more citations, more focus -->
 
@@
 other critical transitions. For instance, stochastic resonance may be 
-visible in advance of a Hopf bifurcation, in which a stable node becomes
-a limit cycle.
+visible in advance of a Hopf bifurcation, where a stable state gives
+way to stable oscillations (limit cycle) of growing magnitude
+[e.g. @Fussman2000; @Strogatz2001].
 

### 2018-02-09 paper/paper.Rmd
WIP

@@
 
-Stochasticity is most often thought of as something which obscures
-patterns we are trying to infer.  This is a simple consequence of
-modeling: variables of interest to the processes
-in question are modeled explicitly (e.g. change in population density
-as it depends on the current density), everything else is just
-noise.  Most statistical inference in ecology is built solely to
-deal with this problem: pulling back the veil of noise to reveal
-processes of interest.  Unfortunately, given the high levels of
-variation and relatively limited data typical in ecological analyses,
-this can be a significant challenge, and even as statistical and
-computational advances provide more powerful and accessible inference
-tools, the task only gets harder.
+To the empricical ecologist, stochasticity, or more simply, "noise,"
+is just that -- something which obscures patterns we are trying to infer [@Knape2011];
+and an ever richer batteries of stastical methods are developed largely in an
+attempt to strip away this undesirable randomness to reveal the patterns
+beneath [@Coulson2001].  Over the past several decades, theoretical ecology has transitioned
+from thinking of stochasticity in such terms; where noise is a nuciance
+that obscures the "deterministic skeleton" of the underlying mechanisms,
+to the recognition that stochasticity can itself be a mechanism for 
+driving many interesting phenomena.  Yet this transistion from "noise the nuisance"
+to "noise the creator" of ecological phenomena has had, with a few notable exceptions,
+relatively little impact in empricial thinking about stochasticity.  One of the
+most provocative of those exceptions has turned the classical notion of noise
+the nuisance on its head: recognizing that noise driven phenomena can become a tool
+to reveal underlying processes: to become "noise the informer."  Here I argue that
+this third shift in perspective offers an opportunity to better bridge the divide
+between respective primarily theoretical and primarily empricial communities by 
+seeing noise not as mathematical curiousity or statistical bugbear, but as a
+source for new opportunities for inference.
+I approach this synthesis in three parts.  First, I will sketch the ground-work
+for the origins of noise in any model as statistical summary of large number of
+additional processes from some lower level that are modeled only in aggregate.
+This will provide the fundamentals we need to illustrate several key examples
+of noise driven phenomena in ecological dynamics: noise the creator.  Finally,
+we will be able to invert some of these examples; instead of going from process
+to pattern, we will go from pattern to process: noise the informer.
 
-<!-- Need concise overview of stochasticity literature -->
-
-For instance, few ecological
-concepts are more central than the notion of density dependence:
-that competition in some form or other should limit a species ability
-to grow without bound.  Yet establishing statistical evidence for
-density dependence is remarkably difficult, as illustrated by
-@Knape2011.  Using time series from the Global Population Dynamics
-Database (GPDD) Knape & de Valpine show that once one accounts for
-the potential for measurement error in addition to environmental
-stochasticity -- an unavoidable source of noise in most ecological
-data -- it becomes difficult to find evidence for even the simplest
-models of density dependence in the majority of available time
-series.  Faced with such uncertainty, it is tempting to abandon
-any such attempts until much longer and higher resolution data 
-becomes available. Fortunately, this has not been the path taken.
-
-<!-- okay, claiming this for the last 2 decades might be overly provocative?-->
-Over the past twenty years, ecological theory, modelling, and methods
-have increasingly elucidated the role of noise not only to *obscure*
-patterns but to *create* them.  This period marked a transition
-away from the intense focus on simple models which exhibited
-deterministic chaos that had enchanted much of theoretical ecology
-in the previous two decades [@Bjornstad2001] in favor of more complex
-and stochastic models.  Of particular interest has been those
-conceptual advances that have bridged the divide from theoretical
-to empirical interest -- not merely the introduction of "real data"
-in a theoretical paper, but rather the concepts that take on a life
-of their own in the larger research community.  Consequently, this
-paper seeks to be a bridge between two audiences: theoretical
-researchers already expert in noisy phenomena but in interested in
-reaching wider empirical application, and also primarily empirical
-researchers seeking an entrance into noisy phenomena.
-
-<!-- Regarding simulation examples -->
-In this spirit, each of the examples illustrated here is coupled
-with concise and efficient R code in the appendix to both reproduce
-and further explore the simulations.  While examples of noisy
-phenomena in natural systems are also crucial, matching the pattern
-does not necessarily mean we have also matched the process behind
-the data.  The examples cited from the literature here offer some
-of the best empirical evidence of these phenomena, but I encourage
-readers to also experiment with the example code provided.  Simulations
-permit poking and prodding of empirical investigation quickly and
-with little mathematical formalism in the way.  In the code provided,
-most stochastic models are expressed in the BUGS language, which
+I will anchor each of these sections using concrete examples from simple models.
+It is easy to critique such simple models: ecology is complex, and much effort
+has been spent on illustrating the importance that complexity: age and stage
+structure, spatial structure, individual heterogeneity, evolution, non-stationary environments,
+species network interactions, and much more [e.g. @Durrett1994; @Bjornstad2001].
+While acknowledging more realistic examples and more general results where avialable,
+using specific and familiar simple models that -- just complex enough to illustrate
+the phenomenon but no more -- will provide us with a platform that is more accessible 
+to both intution and exploration and extension either through mathematical manipulation or 
+computer simulation.  In this spririt, I provide concise, detailed and commented
+code for simulating each of the models presented in Appendix A.
+Simulations permit poking and prodding of empirically-minded
+investigation unencumbered by either experimental design or mathematical formalism.
+Most stochastic models are expressed in the BUGS language, which
 may be more familiar to empirical readers than corresponding
@@
 
-<!-- MAP -->
-Our first stop will be into research into the origins of noise.
-This will not only let us set up definitions for terms and concepts
-we will encounter throughout, but also introduce some of the
-fundamental mathematical and computational tooling needed in our
-tour of noisy phenomena.  From origins of noise, we will then explore
-some of the phenomena that noise can create.  These examples are
-central to the conceptual transition away from viewing noise only
-as a nuisance that obscures patterns and realize the rich set of
-patterns and phenomena that can emerge.  Our tour will be by no
-means exhaustive, but as with chaos before it, theoretical allure
-does not always reflect ecological relevance.  Having looked at
-some of the phenomena that can be created by noise, our third stop
-will try to run this in reverse: using noise to infer underlying
-phenomena.  Examples here are rarer but potentially the most
-rewarding, bridging the gap from theoretical construct to empirical
-investigation.  The most compelling examples have come from ways
-in which noisy phenomena are used to infer stability and forecast
-changes in resilience of populations or ecosystems. Though not
-without weaknesses or missteps, the resulting dialog between
-theoretical and empirical aspects have made this particularly a
-rich area of research.  We end our tour by considering what features
-best enable the translation and ecological relevance of noisy
-phenomena.
+# Origins of noise
 
-# Origins of noise
+This does not attempt to summarize all of the literature on the origins and
+types of noise, which have been reviewed elsewhere [e.g. @Saether1997; @Coulson2004; @Ovaskienen2010; @Black2012; @Schrieber2012; @Lenormand2009]
 

### 2018-02-13 paper/paper.Rmd
mostly refs

@@
 
+## Use trends style to quickly count refs
+#csl: trendsjournals.csl
 csl: ecology-letters.csl
 bibliography: refs.bib
-output: rticles::elsevier_article  
+output: 
+  rticles::elsevier_article:
+    keep_tex: true
 layout:  3p # review = doublespace, 3p = singlespace, 5p = two-column
@@
 ```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="cairo_pdf", fig.width=7, fig.height=3.5)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="cairo_pdf", fig.width=7, fig.height=3.5, cache = TRUE)
 library(tidyverse)
@@
 and potential estimation of parameters given sample data using the
-R package, NIMBLE [@DeValpine2017]; allowing the deductive models
+R package, NIMBLE [@deValpine2017]; allowing the deductive models
 illustrated here to readily function as inductive models with
@@
 
-This does not attempt to summarize all of the literature on the origins and
-types of noise, which have been reviewed elsewhere [e.g. @Saether1997; @Coulson2004; @Ovaskienen2010; @Black2012; @Schrieber2012; @Lenormand2009]
-
 <!-- Shorten this, more citations, more focus -->
 
-What is noise? Where does it come from?  It is important to bear
-in mind that whenever stochasticity is introduced into a model, it
+What is noise? Where does it come from?  Simply put, *noise is all that we leave out of our model*. 
+Whenever stochasticity is introduced into a model, it
 is only as a simplification or summary of the effects of a large
 number of other processes, often occurring at different scales.
-Simply put, *noise is all that we leave out of our model*.  Early
+A central theme of stochastic modeling research over the past
+several decades has been an emphasis on making this relationship
+between stochastic summary and underlying process more precise:
+that is, on understanding the origins of noise [@Coulson2004].
+
+Over the past several decades, research into stochastic modeling
+has
+
+This does not attempt to summarize all of the literature on the origins and
+types of noise, which have been reviewed elsewhere [e.g. @Saether1997; @Coulson2004; @Ovaskainen2010; @Black2012; @Schreiber2012; @Lenormand2009]
+
+justify simple models, 
+citations adapting this to complex models:
+
+Early
 quantitative ecological models first formulated at the population
@@
 
+Few topics have recieved greater attention then the question of
+the persistence of a given species [@Kendall1949;  @Bartlett1959; @Lande1993; @Saether1998; @Coulson2001; @Schreiber2010]
+
+and coexistence of multiple species
+
+[@Caswell1978; @Tuljapurkar1980; @Chesson1981; @Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]
+
+Structured populations 
+[@Cohen1979; @Tuljapurkar1989; @Metcalf2015; @Vindenes2017b]
+
+See in particular reviews by @Ovaskainen2010 and @Schreiber2012
+
+Individual heterogeneity 
+@Vindenes2015; @Hart2016
+
+@Lee2017
+
 A natural focus of much research into stochastic ecological systems
@@
 
+<!--  -->
+
 Understanding the origin of sustained oscillations has long been a
@@
 noise into the system can be sufficient to excite the system into
-sustained oscillations (Figure \ref{quasicycles}B).  In the example shown here, a
+sustained oscillations (Figure \ref{quasicycles}B) 
+[e.g. empirical examples in @Blarer1999; @Greenman2003; @Black2010].  
+In the example shown here, a
 simple predator-prey system is driven by environmental white noise
@@
 arises explicitly from demographic stochasticity, rather than generic
-environmental noise as in the model above and elsewhere.  @Pineda-Krch2007
-compare patterns from these stochastically generated cycles to
+environmental noise as in the model above and elsewhere [@Kendall1999; @Pascual2003].  
+@Pineda-Krch2007 compare patterns from these stochastically generated cycles to
 stable limit cycles they resemble.  Stochastic resonance highlights
@@
 way to stable oscillations (limit cycle) of growing magnitude
-[e.g. @Fussman2000; @Strogatz2001].
+[e.g. @Fussmann2000; @Strogatz2001].
 

### 2018-02-19 paper/paper.Rmd
WIP

@@
 ## Use trends style to quickly count refs
-#csl: trendsjournals.csl
-csl: ecology-letters.csl
+csl: trendsjournals.csl
+#csl: ecology-letters.csl
 bibliography: refs.bib
@@
 to the recognition that stochasticity can itself be a mechanism for 
-driving many interesting phenomena.  Yet this transistion from "noise the nuisance"
+driving many interesting phenomena [@Coulson2004].  Yet this transistion from "noise the nuisance"
 to "noise the creator" of ecological phenomena has had, with a few notable exceptions,
@@
 seeing noise not as mathematical curiousity or statistical bugbear, but as a
-source for new opportunities for inference.
+source for new opportunities for inference.  This shift in how we think about noise
+is nonetheless easy to miss in the extensive literature
+
+Origins of noise and motivation for studying stochastic models @Kendall1949 and @Bartlett1960
+
+This trend from nuisance to creator
+to informer is of course not the only or the most obvious trend in the 
+extensive literature on stochastic modeling over the past few decades, but rather,
+an instructive lens in which to draw out a lar
+
+@Bjornstad2001 @Coulson2001 @Coulson2004
+
+I will focus on examples that
+illustrate this larger arc in how we are thinking about stochasticity as the role
+it plays.
+
+(e.g. see in particular
+@Couslon2004, @Ovaskainen2010 on population extinction with 
+a focus on environmental and colored noise, @Black2012 on intrinsic / demographic noise,
+
+Few topics have recieved greater attention then the question of
+the persistence of a given species [@Kendall1949;  @Bartlett1960; @Lande1993; @Saether1998; @Coulson2001; @Schreiber2010]
+
+Excellent reviews and textbook introductions
+
+@Bartlett1960 @Bjornstad2001 @Coulson2001 @Lande2003
+
+See in particular reviews by @Ovaskainen2010 and @Schreiber2012
+
+@Lee2017
+
+The extensive literature on demographic and environmental noise has been
+frequently reviewed elsewhere [@Black2012]
+
+@Saether1997
+
 I approach this synthesis in three parts.  First, I will sketch the ground-work
@@
 I will anchor each of these sections using concrete examples from simple models.
-It is easy to critique such simple models: ecology is complex, and much effort
+It is easy to critique such simple models: after all, one might summarize decades
+of research in noting the twin observations that (1)  ecology is complex and (2)
+that complexity is important. In the stochasticty literature alone, much work 
+has been done to understand and extend fundamental results to more complex models,
+including the importance of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
+@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g. @Durrett1994; @Dieckmann2000;
+@Kerr2002; @Schreiber2010], Individual heterogeneity [e.g. @Vindenes2015; @Hart2016]
+or the coexistence of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
+@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]; species traits
+and the interaction of ecological and evolutionary processes [@Ozgul2009; @Lenormand2009; @Vindesens2015; @Schreiber2015],
+not to mention the extensive literature in stochastic evolution and popluation genetics
+[e.g. Ewens, @Lewontin1969]
+
+and the role of stochasticity in more complex models, particularly in structured populations (matrix models),
+
+in
+
 has been spent on illustrating the importance that complexity: age and stage
@@
 
-<!-- Shorten this, more citations, more focus -->
-
-What is noise? Where does it come from?  Simply put, *noise is all that we leave out of our model*. 
+What is noise? Where does it come from?  Simply put, noise is all that we leave out of our model. 
 Whenever stochasticity is introduced into a model, it
@@
 several decades has been an emphasis on making this relationship
-between stochastic summary and underlying process more precise:
-that is, on understanding the origins of noise [@Coulson2004].
+between stochasticity and underlying processes more precise:
+that is, on the origins of noise [e.g. @Coulson2004]. This literature
+has largely divided the origins of stochasticity between 
+*intrinsic* factors: where stochasticity is introduced to reflect
+the reality that our variables of interest are intriniscally 
+statistical averages of lower-level processes, and *extrinsic*
+factors: where stochasticity reflects the impact of other variables
+we are not explicitly modeling at all [e.g. @Keeling2012].  **Demographic stochasticity**
+is the primary (though not the only) example of an intrinsic factor,
+which reflects the fact that the continuous variable we often think of a
+"population size" or "population density" is an average over a 
+lower level process of births and deaths among a finite number of 
+individuals [@Melbourne2012].  **Environmental stochasticity** is the primary example 
+of an extinsic factor, reflecting the fact that parameters we may
+treat as constants in a model in fact depend on variables such as 
+temperature or soil moisture that are not explicitly modeled by our 
+equations, but simplified into statistical summaries of those processes
+[@Ripa2012].
 
-Over the past several decades, research into stochastic modeling
-has
+The Gillespie simulation provides a simple and exact algorithm
+for simulating demographic stochasticity at an individual level.  
+Free from both the approximations and mathematical complexity of more common
+formulations using stochastic differential equations (SDEs), 
+(or their partial differential equation counterparts the Fokker-Planck equation),
+the Gillespie algorithm offers a simple and powerful approach to simulating
+stochastic phenomena that remains underutilized.
+
+The mathematical proofs @Kurtz1970 & @Kurtz1971, popularized by
+@vanKampen's intuitive "system size" expansion, provide a far more 
+explicit framework to interpret demographic stochasticity than the
+better known diffusion approximation [@Nisbet1982 @Lande2003; 
+see @Gardiner2009].
+
+It is often suggested that stochasticity is "inherent" or "fundamental,"
+e.g. "demographic stochasticity arises because the birth of an individual is probabilistic"
+[@Melbourne2012]. It is easy to forget that this is merely an assertion about a 
+model, not an assertion about reality.  Only in quantum mechanics do we find 
+inherently probabilistic properties: births and deaths of organisms all have
+far more mechanistic explanations which we simply summarize in statistical terms:
+on average -- that is, summing over a wide range of possible mechanisms we have chosen
+not to explicitly model -- an individual death occurs at rate $\lambda$.  
+This is in fact an extrinisic factor (variables not modeled explicitly),
+which makes birth appear probabilistic; (just as we describe the toss of a coin
+as probabilistic when in fact it is a mechanistic outcome of Newton's laws).  
+Demographic stochasticity arises
+only subsequently, when we change scales from a description at the individual
+level to one at the population level.
+
+@Keeling2012; @Melbourne2012; @Ripa2012
 
@@
 enters into quantitative models [@Black2012; @Ovaskainen2010].
-
-With the exception of quantum mechanics (and there only at scales
-that have no relevance to ecology), stochasticity is not an inherent
-property of nature but rather an emergent property of many events
-occurring at a different scale.  Coins and dice obey the physical
-laws of motion; only variation in how they are tossed (i.e. something
-complex we do not usually model) creates randomness.  The term
-"noise" itself reflects this emergent rather than fundamental origin:
-the combination of hundreds of voices and sounds each individually
-distinct blends together into a background of noise. Statistical
-mechanics tells us that temperature is a measure of the average
-speed of individual molecules in a gas -- any given molecule will
-experience a range of possible speeds but by itself has no notion
-of temperature.  Likewise, stochasticity enters our ecological
-models because our quantities of interest, such as the density of
-a population, is but an average over a lower-level, discrete process
-of individual births and deaths. Stochasticity that emerges directly
-from discrete individual events that underlie the process described
-by our model are considered *intrinsic stochasticity*, commonly
-referred to as *demographic stochasticity* in ecological literature,
-as it usually arises in the context scaling from individual demographic
-events to population level descriptions. An important aspect of
-this demographic stochasicity is that it scales with the square
-root of the "system size," [@vanKampen2007] so that for very large populations the
-resulting fluctuations are relatively small. On the other hand,
-*extrinsic stochasticity*, usually referred to as *environmental
-stochasticity* in the ecological literature, arises in processes
-that are not described by our model, but merely impact the value
-of some parameter of our model; for instance, fluctuations in
-temperature may drive variation in a death rate, or fluctuations
-in rainfall drive variation in carrying capacity.
 

### 2018-02-20 paper/paper.Rmd
wip

@@
 or the coexistence of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
-@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]; species traits
-and the interaction of ecological and evolutionary processes [@Ozgul2009; @Lenormand2009; @Vindesens2015; @Schreiber2015],
+@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]; 
+correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010; @Lee2017; @Spanio2017] or periodic
+[@Coulson2001; @Bjornstad2001; @Keeling2001] structure in environmental noise, or
+the interaction of ecological and evolutionary processes [@Saether1997; @Ozgul2009; @Lenormand2009; @Vindenes2015; @Schreiber2015],
 not to mention the extensive literature in stochastic evolution and popluation genetics
-[e.g. Ewens, @Lewontin1969]
+[e.g. @Ewens2004, @Lewontin1969].
 
-and the role of stochasticity in more complex models, particularly in structured populations (matrix models),
-
-in
-
-has been spent on illustrating the importance that complexity: age and stage
-structure, spatial structure, individual heterogeneity, evolution, non-stationary environments,
-species network interactions, and much more [e.g. @Durrett1994; @Bjornstad2001].
-While acknowledging more realistic examples and more general results where avialable,
-using specific and familiar simple models that -- just complex enough to illustrate
+While acknowledging more realistic examples and more general results where available,
+using specific and familiar simple models -- just complex enough to illustrate
 the phenomenon but no more -- will provide us with a platform that is more accessible 
@@
 Few topics have recieved greater attention then the question of
-the persistence of a given species [@Kendall1949;  @Bartlett1959; @Lande1993; @Saether1998; @Coulson2001; @Schreiber2010]
+the persistence of a given species [@Kendall1949;  @Bartlett1955; @Bartlett1960; @Lande1993; @Saether1998; @Coulson2001; @Schreiber2010]
 

### 2018-02-21 paper/paper.Rmd
WIP

@@
 seeing noise not as mathematical curiousity or statistical bugbear, but as a
-source for new opportunities for inference.  This shift in how we think about noise
-is nonetheless easy to miss in the extensive literature
+source for new opportunities for inference.
 
-Origins of noise and motivation for studying stochastic models @Kendall1949 and @Bartlett1960
-
-This trend from nuisance to creator
-to informer is of course not the only or the most obvious trend in the 
-extensive literature on stochastic modeling over the past few decades, but rather,
-an instructive lens in which to draw out a lar
-
-@Bjornstad2001 @Coulson2001 @Coulson2004
-
-I will focus on examples that
-illustrate this larger arc in how we are thinking about stochasticity as the role
-it plays.
-
-(e.g. see in particular
-@Couslon2004, @Ovaskainen2010 on population extinction with 
-a focus on environmental and colored noise, @Black2012 on intrinsic / demographic noise,
-
-Few topics have recieved greater attention then the question of
-the persistence of a given species [@Kendall1949;  @Bartlett1960; @Lande1993; @Saether1998; @Coulson2001; @Schreiber2010]
-
-Excellent reviews and textbook introductions
-
-@Bartlett1960 @Bjornstad2001 @Coulson2001 @Lande2003
-
-See in particular reviews by @Ovaskainen2010 and @Schreiber2012
-
-@Lee2017
-
-The extensive literature on demographic and environmental noise has been
-frequently reviewed elsewhere [@Black2012]
-
-@Saether1997
-
-I approach this synthesis in three parts.  First, I will sketch the ground-work
-for the origins of noise in any model as statistical summary of large number of
-additional processes from some lower level that are modeled only in aggregate.
-This will provide the fundamentals we need to illustrate several key examples
-of noise driven phenomena in ecological dynamics: noise the creator.  Finally,
-we will be able to invert some of these examples; instead of going from process
-to pattern, we will go from pattern to process: noise the informer.
-
-I will anchor each of these sections using concrete examples from simple models.
-It is easy to critique such simple models: after all, one might summarize decades
+As the literature on stochastic modeling in ecology is both extensive and often
+technical, and so rather than attempt a comprehensive review, this synthesis
+will focus on the simplest examples necessary to illustrate this broader trend
+in the changing way we think about the role of noise and the opportunity
+that shift can present to better bridge theoretical and empirical communities.
+To that end, I will focus on examples from simple population models, echoing
+the motivation of @Bartlett1960 to focus on basic theoretical aspects common
+across models. In so doing, we will ignore the equally rich literature of 
+stochastic models in evolution [@Lenormand2009] and population genetics
+[e.g. @Ewens2004]. In an effort to make these themes accessible to more emprical
+readers, I will also focus on simplest examples necessary, with only passing
+mention of the rich literature extending these concepts to more realistic
+models.  It is easy to critique this choice: indeed, one might summarize decades
 of research in noting the twin observations that (1)  ecology is complex and (2)
-that complexity is important. In the stochasticty literature alone, much work 
+that complexity is important. In the stochasticity literature alone, much work 
 has been done to understand and extend fundamental results to more complex models,
@@
 [@Coulson2001; @Bjornstad2001; @Keeling2001] structure in environmental noise, or
-the interaction of ecological and evolutionary processes [@Saether1997; @Ozgul2009; @Lenormand2009; @Vindenes2015; @Schreiber2015],
-not to mention the extensive literature in stochastic evolution and popluation genetics
-[e.g. @Ewens2004, @Lewontin1969].
+the interaction of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
+@Lenormand2009; @Vindenes2015; @Schreiber2015]. In this we follow @Levins1966 in
+sacrificing realism for generality.
 
-While acknowledging more realistic examples and more general results where available,
-using specific and familiar simple models -- just complex enough to illustrate
-the phenomenon but no more -- will provide us with a platform that is more accessible 
-to both intution and exploration and extension either through mathematical manipulation or 
-computer simulation.  In this spririt, I provide concise, detailed and commented
+In this spririt, I provide concise, detailed and commented
 code for simulating each of the models presented in Appendix A.
@@
 bug reports, suggestions or help requests are welcome through the
-issue tracker.  A second appendix provides more mathematical details
+issue tracker.  Appendix B provides more mathematical details
 for interested readers.  With these in place, we are ready to begin
@@
 noise, emergent phenomena, and noise-driven inference.
+
+<!--
+I approach this synthesis in three parts.  First, I will sketch the ground-work
+for the origins of noise in any model as statistical summary of large number of
+additional processes from some lower level that are modeled only in aggregate.
+This will provide the fundamentals we need to illustrate several key examples
+of noise driven phenomena in ecological dynamics: noise the creator.  Finally,
+we will be able to invert some of these examples; instead of going from process
+to pattern, we will go from pattern to process: noise the informer.  
+-->
 
@@
 
+Many stochastic models are not explicit about whether the noise terms
+they introduce are intended to reflect intrinsic or extrinsic factors.
+Early quantitative ecological models first formulated at the population
+level as differential or difference equations: deterministic models
+that seek to capture the essential features of a system while
+neglecting any random component as mere noise [@Coulson2004].
+Advances in both theoretical understanding and the advent of widely
+available and efficient numerical methods have helped drive research
+towards increasingly more explicit descriptions of how stochasticity
+enters into quantitative models.
+
+@Black2012 provides an important 
+review of intrinsic noise and approaches for addressing it, focusing
+in particular on use of the van Kampen system size expansion as a
+mathematical means to properly bridge between individual-based models
+(IBMs, usually implemented only in simulations) and familiar population-level
+dynamics (so called macroscopic dynamics).
+
+; @Ovaskainen2010 the canonical on the highly influential @Leigh1981, @Lande1993, @Foley1994
+
+This division between intrinsic or demographic noise and extrinsic or
+environmental noise
+
 The Gillespie simulation provides a simple and exact algorithm
@@
 [@Melbourne2012]. It is easy to forget that this is merely an assertion about a 
-model, not an assertion about reality.  Only in quantum mechanics do we find 
+model, not an assertion about reality [@Bartlett1960].  Only in quantum mechanics do we find 
 inherently probabilistic properties: births and deaths of organisms all have
@@
 as probabilistic when in fact it is a mechanistic outcome of Newton's laws).  
-Demographic stochasticity arises
-only subsequently, when we change scales from a description at the individual
-level to one at the population level.
-
-@Keeling2012; @Melbourne2012; @Ripa2012
-
-This does not attempt to summarize all of the literature on the origins and
-types of noise, which have been reviewed elsewhere [e.g. @Saether1997; @Coulson2004; @Ovaskainen2010; @Black2012; @Schreiber2012; @Lenormand2009]
-
-justify simple models, 
-citations adapting this to complex models:
-
-Early
-quantitative ecological models first formulated at the population
-level as differential or difference equations: deterministic models
-that seek to capture the essential features of a system while
-neglecting any random component as mere noise [@Coulson2004].
-Advances in both theoretical understanding and the advent of widely
-available and efficient numerical methods have helped drive research
-towards increasingly more explicit descriptions of how stochasticity
-enters into quantitative models [@Black2012; @Ovaskainen2010].
+Demographic stochasticity arises only subsequently, when we change scales from
+a description at the individual level to one at the population level.
 
@@
 exact stochastic simulation algorithm [SSA; @Gillespie1977], which
-simulates each individual birth or death as a separate event. Figure \ref{gillespie}
-uses this algorithm to compare simulations of @Levins1969 patch model;
-(see Appendix A for example code).  In the Levins'  model,
-$n$ individuals compete for a
+simulates each individual birth or death as a separate event.
+
+We can illustrate both the Gillespie algorithm and the van Kampen
+system size expansion using @Levins1969 classic patch model,
+
+\begin{align}
+\frac{\ud n}{\ud t} = \underbrace{c n \left(1 - \frac{n}{N}\right)}_{\textrm{birth}} - \underbrace{e n}_{\textrm{death}}, \label{levins}
+\end{align}
+
+where $n$ individuals compete for a
 finite number of suitable habitats $N$.  Individuals die a constant
@@
 proportional to the fraction of available patches, $1 - n/N$.  
+We choose this model because it has precisely same functional form
+as the ubiquitous logistic growth model, $\frac{\ud n}{\ud t} = rn (1 - \tfrac{n}{K})$
+but is explicit about how this net growth rate is divided between
+birth and death rate.  As we shall see, this is not merely a notational
+convience but leads to real differences in observed stochastic properties
+of the model.  Though we focus on this simple model for clarity of 
+presentation, both computational and mathematical framework presented here 
+can easily be applied to alternative models to reach the same conclusions.
 
-\begin{align}
-\frac{\ud n}{\ud t} = \underbrace{c n \left(1 - \frac{n}{N}\right)}_{\textrm{birth}} - \underbrace{e n}_{\textrm{death}}, \label{levins}
-\end{align}
+Figure \ref{gillespie}
+uses this algorithm to compare simulations of @Levins1969 patch model;
+(see Appendix A for example code).  In the Levins'  model,
 
@@
 [@Knape2011; @deValpine2002].
-
-Other forms of uncertainty common in models are not stochastic
-phenomena.  This includes any form of *reducible uncertainty*: In
-constructing and estimating mathematical models of ecological
-processes, we encounter uncertainty over our parameter estimates
-and possible choice of models [@Ellison2004].  In some ways, this
-uncertainty can act like the stochasticity described above: for
-instance, both introduce variation into any forecast of future
-outcomes. Yet unlike stochasticity, this uncertainty is reducible:
-with additional observations, we can narrow the uncertainty over
-parameters or candidate models. Thus we will not consider model and
-parameter uncertainty, as they do not reflect intrinsic phenomena of 
-the dynamics but only our own ignorance about those dynamics.  
 *Deterministic chaos* [@Desharnais2001] is another mechanism of introducing
@@
 
-Much active research continues to explore the consequences of
-introducing these more explicitly motivated and interacting sources
-of noise. With some grounding in the origins of different types of
-noise in place, we are ready to move to our second stop to see how
-this noise can drive the creation of novel phenomena on a macroscopic
-level.
-
 # Noise the creator: Noise can induce novel phenomena
@@
 
-Few topics have recieved greater attention then the question of
-the persistence of a given species [@Kendall1949;  @Bartlett1955; @Bartlett1960; @Lande1993; @Saether1998; @Coulson2001; @Schreiber2010]
+The loss of a species is without a doubt one of the most important
+possible consequences of noise, and unsurprisingly few topics have
+recieved greater attention then the question of the persistence of
+a given species [e.g. @Kendall1949;  @Bartlett1955; @Bartlett1960;
+@Lande1993; @Saether1998; @Coulson2001; @Schreiber2010] and coexistence
+of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
+@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017]
+in stochastic models.  Two important trends in this literature have
+been establishing existing results in more general and precise language
+and theorems [e.g. see @Schreiber2012 and @Schreiber2017 for an excellent
+overview and synthesis of persistence and coexistence respectively],
+and continuing to extend such results to more complex contexts 
+[@Ovaskainen2010 provides one excellent review], in
+particular addressing autocorrelated noise [e.g. @Ripa; @Ripa1999; 
+@Petchey, @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
+[e.g. @Durrett1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
+[e.g. @Cohen1979; @Tuljapurkar1989; @Caswell2009; @Metcalf2015; @Vindenes2017b], 
+among other issues.
 
-and coexistence of multiple species
-
-[@Caswell1978; @Tuljapurkar1980; @Chesson1981; @Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]
-
-Structured populations 
-[@Cohen1979; @Tuljapurkar1989; @Metcalf2015; @Vindenes2017b]
-
-See in particular reviews by @Ovaskainen2010 and @Schreiber2012
-
-Individual heterogeneity 
-@Vindenes2015; @Hart2016
-
-@Lee2017
-
-A natural focus of much research into stochastic ecological systems
-has been on understanding the implications for persistence of
-individual and coexistence of interacting species. Demographic noise
-in particular has fundamental implications for the long-term
+Demographic noise in particular has fundamental implications for the long-term
 equilibrium: any finite population, however big, is guaranteed to
-eventually go extinct under demographic noise [e.g. @Kendall1949; @Lande2003].  However, in most
-models it is possible to define a notion of quasi-stationary 
+eventually go extinct under demographic noise [e.g. @Kendall1949; @Lande2003].  
+In most models it is possible [@Schreiber2006] to define a notion of quasi-stationary 
 distribution concentrated around the deterministic attractor,
 and to prove the probability of extinction decreases exponentially with 
-system size (e.g. see @Ovaskainen2010; @Schreiber2017 also reviews 
-more recent literature which has established these results more generally). 
-Generalizations about persistence and stable coexistence of interacting
-species in stochastic models are difficult due to complex interactions
-between demographic and environmental noise, population structure,
-spatial heterogeneity, magnitude of noise, degree of auto-correlation and so forth
-[e.g. @Higgins1997; @Coulson2001; @Melbourne2008; @Hart2016];
-indeed care must be taken even in defining the appropriate notion of 
-persistence [e.g. @Schreiber2006].  Instead of these long-term limits,
+system size [@Ovaskainen2010; @Schreiber2017]. We will turn our attention
+away from the well-trodden ground of persistence and coexistence to focus
+on other phenomena that
+
 here we will focus on a few examples of stochastic phenomena 

### 2018-02-21 paper/paper.Rmd
WIP

@@
 
-What is noise? Where does it come from?  Simply put, noise is all that we leave out of our model. 
+What is noise? Where does it come from? 
+Many stochastic models are not explicit about whether the noise terms
+they introduce are intended to reflect intrinsic or extrinsic factors.
+Early quantitative ecological models first formulated at the population
+level in deterministic settings. The notion of a deterministic
+"skeleton" -- responsible for the phenomena of interest -- that is obscured
+from view by the nusiance of noise has slowly given way to recognition
+that stochastic and deterministic elements not only interact in 
+interesting ways [@Coulson2004] but also share a common origin [e.g. @Black2012].
+A central theme of stochastic modeling research over the past
+several decades has been an emphasis on making this relationship
+between stochasticity and underlying processes more precise:
+that is, on the origins of noise.
+
+Simply put, noise is all that we leave out of our model. 
 Whenever stochasticity is introduced into a model, it
@@
 number of other processes, often occurring at different scales.
-A central theme of stochastic modeling research over the past
-several decades has been an emphasis on making this relationship
-between stochasticity and underlying processes more precise:
-that is, on the origins of noise [e.g. @Coulson2004]. This literature
-has largely divided the origins of stochasticity between 
+This literature has largely divided the origins of stochasticity between 
 *intrinsic* factors: where stochasticity is introduced to reflect
@@
 
-Many stochastic models are not explicit about whether the noise terms
-they introduce are intended to reflect intrinsic or extrinsic factors.
-Early quantitative ecological models first formulated at the population
-level as differential or difference equations: deterministic models
-that seek to capture the essential features of a system while
-neglecting any random component as mere noise [@Coulson2004].
+In the extensive literature developing the consequences of both demographic
+and environmental noise in models of increasing complexity, it can 
+be easy to miss the equally important developments in making the origins
+of these sources of noise more precise -- developments which have important
+consquences for when and how the resulting models can be applied.
+
+the canonical model on the highly influential @Leigh1981, @Lande1993, @Foley1994
+
+$$ \frac{\textrm{d} n}{\textrm{d} t} = \underbrace{f(n)}_{\textrm{det skeleton}} + \underbrace{\sigma_d \xi_d(t) \sqrt{n}}_{\textrm{demographic noise}} + \underbrace{\sigma_e \xi_e(t) n}_{\textrm{environmental noise}}$$
+
+Stochastic differential equations of this form are so-called "Langevin equation," popular in the physics literature.
+This notation is rather sloppy, as the standard calculus does not apply to stochastic equations.
+The Langevin formulation, most popular in physics, is usually associated with a Stratonovich integral.
+Mathematicians prefer to write stochastic differential equations as $\textrm{d} n = f(x) \textrm{d} t + g(x) \textrm{d} B_t$, where the differentials hint at the underlying definition that this is the sum of a classic Reimann integral (over $\textrm{d}t$) and a stochastic It\^o integral.
+
 Advances in both theoretical understanding and the advent of widely
@@
 
-; @Ovaskainen2010 the canonical on the highly influential @Leigh1981, @Lande1993, @Foley1994
+; @Ovaskainen2010
 
@@
 coexistence, and dispersal of species [@Schreiber2010; @Hart2016;
-@Schreiber2017].  *Observational error* acts
+@Schreiber2017].
+
+*Observational error* acts
 much like another form of stochasticity, arising from sampling
@@
 @Petchey, @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
-[e.g. @Durrett1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
+[e.g. @Durrett1994; @Hastings1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
 [e.g. @Cohen1979; @Tuljapurkar1989; @Caswell2009; @Metcalf2015; @Vindenes2017b], 

### 2018-02-22 appendixB/appendixB.Rmd
WIP

@@
 layout: 5p  
-bibliography: canonical.bib
+bibliography: ../paper/refs.bib
 output: rticles::elsevier_article
@@
 ---
+
+## Stochastic Differential Equations
+
+@Ovaskainen2010 writes the canonical equation, Eq (1) in the main text, in the form:
+
+$$ \frac{\textrm{d} n}{\textrm{d} t} = \underbrace{f(n)}_{\textrm{det skeleton}} + \underbrace{\sigma_d \xi_d(t) \sqrt{n}}_{\textrm{demographic noise}} + \underbrace{\sigma_e \xi_e(t) n}_{\textrm{environmental noise}}$$
+
+rather than the more formal stochastic differential equation notation shown in the text.
+Equations of this form are so-called "Langevin equation," popular in the physics literature [@vanKampen2007].
+This notation looks like a more familiar differential equation, but the resemblence is misleading as the standard calculus does not apply to stochastic equations.
+
+A stochastic differntial equation (SDE) of the form
+
+\begin{equation}
+\textrm{d} X_t = \mu(X_t) \textrm{d}t + \sigma(X_t) \textrm{d}B_t \label{SDE}
+\end{equation}
+
+separates out the differential elements$\textrm{d}X_t$, $\textrm{d}t$ and $\textrm{d}B_t$ intentionally to remind the reader that the expression is short-hand not for a derivative expression, but for integrals:
+
+\begin{equation}
+X_{t+s} - X{t}  = \int_t^{t+s} \mu(X_{\tau}, \tau) \textrm{d}\tau + \int_t^{t+s} \sigma(X_{\tau}, \tau) \textrm{d}B\tau
+\end{equation}
+
+The first integral (over $\textrm{d}t$) is the familiar Reimann integral of the classic calculus, but the second is an Itô integral, which integrates over the stochastic process $B_t$ (Brownian motion, also termed a Weiner process).
+The Itô calculus follows different rules from the classic calculus, largely due to the fact that an Itô integral has non-zero quadratic variation.
+(Crudely put, $\textrm{d}x^2 \sim 0$, but $\textrm{d}B^2 \sim dx$).
+
+The intepretation of Langevin equation is less precise, as it can be read to correspond to one of two possible stochastic integrals, and is often associated with a Stratonovich integral.
+(Unlike the Itô integral, the Stratonovich integral has zero quadratic variation and so obeys the same chain rule as the classic calculus, but it is not a martingale, e.g.
+an unbiased random walk, which is a convenient for property for proving theorems).
+Fortunately, the notion as a partial differential equation over probabilities does not have this ambiguity, which shows that @Ovaskainen2010 is referring to an Itô expression, just as I have written it more explicitly.
+See @vanKampen2007 for an excellent discussion of the Itô-Stratonovich dillema and its intepretation with respect to intrinsic vs environmental noise factors.
+
+## Master equations and the step operator
 
@@
 &= \sum  b_n p_n(\mathbb E - 1)n  + \sum d_n p_n(\mathbb E^{-1} -1) n \nonumber \\
-&= - \langle d_n \rangle +  \langle b_n \rangle \label{mean} \\
-\intertext{and similarly the second moment is}
+&= - \langle d_n \rangle +  \langle b_n \rangle \label{mean}
+\end{align}
+
+and similarly the second moment is
+
+\begin{align}
 \frac{\ud }{\ud t} \langle n^2 \rangle &= 2 \langle n (b_n - d_n) \rangle + \langle b_n+d_n \rangle \label{var}
@@
 
-The key observation here is that the variance includes the sum of birth and death, not just the difference.
-Intuitively we would expect a population with large numbers of births and deaths that average to a small difference to fluctuate much more than one in which this difference is very small.
-Despite this simple observation, countless models purporting to include demographic noise ignore this distinction.
 While all moments can be derived as above, it is often impossible to solve these equations for nonlinear models.
-Instead, we introduce an approximation of Eq \eqref{master1} that is physically motivated and intuitive.
-The method is based on derivation presented by the van Kampen [@vanKampen], though a formal proof can be found in earlier work of Thomas Kurtz [@Kurtz1970; @Kurtz1971; @Kurtz1972].
+Moreover, we have yet to make any moment closure argument that solving for the first two moments in this manner will be sufficient.
+We will illustrate how this is done in both the traditional diffusion approximation and then in the van Kampen system size expansion.
 
-# The system size expansion for a birth-death process
+## A traditional derivation
 
-## Introduction, a change of variables
+A derivation of the above formulation can be found, e.g.
+in @Nisbet1982's classic textbook on fluctuating populations, or more recently in @Lande2003.
+This diffusion approximation is also given for a more mathematically inclined audience in @Gardiner2009 and in @vanKampen2007.
+The approach starts with the same master equation that defines the underlying dynamics of births and deaths (matching the process implemented in the exact Gillespie algorithm, @Gillespie1977, see Appendix A).
+
+The essence of the approach is to argue that in the limit of large populations, the discrete state for the number of indivduals in a population, $N$, can be replaced with a continuous variable, $n$, in which we can expand the master equation as a Taylor series and then truncate at second order.
+In contrast to the system size expansion of van Kampen, this approach does not make system size explicit, but instead hinges on taking a simultaneous limit of both short time and small step size.
+Consider small steps $x = \epsilon n$, and hence $p_n = \epsilon P_x$, where $\epsilon \ll 1$.
+To keep the process from slowing down we consequently have to scale up the rates by $\epsilon$ as well:
+
+\begin{align*}
+	b_n - d_n = \frac{A_x}{\epsilon}
+\end{align*}
+
+Where $A$ is independent of $\epsilon$.  Similarly we have to scale the sum, which being proportional to the variance must be scaled by
+\begin{align*}
+b_n + d_n = \frac{B_x}{\epsilon^2}
+\end{align*}
+
+We Taylor expand the step operator $\mathbb E$ in the new variable:
+\begin{align*}
+\mathbb E^k &= 1+ \epsilon k\frac{\partial}{\partial x} + \frac{\epsilon^2 k^2}{2} \frac{\partial^2}{\partial x^2} + \ldots
+\end{align*}
+
+Then \eqref{master1} becomes:
+\begin{align}
+\frac{\partial P(x,t)}{\partial t} &= \left( \epsilon \frac{\partial }{\partial x} + \epsilon^2 \tfrac{1}{2}\frac{\partial^2 }{\partial x^2} + \ldots \right)\left(\frac{B_x}{2\epsilon^2}-\frac{A_x}{2\epsilon} \right)P \nonumber \\
++&\left(- \epsilon \frac{\partial }{\partial x} + \epsilon^2 \tfrac{1}{2}\frac{\partial^2 }{\partial x^2} - \ldots \right) \left( \frac{B_x}{2\epsilon^2} + \frac{A_x}{2\epsilon} \right)P 
+\intertext{Collecting terms of common order $\epsilon$, }
+& = - \frac{\partial A_x P}{\partial x} + \tfrac{1}{2} \frac{\partial^2 B_x P }{\partial x^2} + \mathcal O(\epsilon^2)\label{FP1}
+\end{align}
+
+For small $\epsilon$ we can ignore the terms of $\mathcal O(\epsilon^2)$, and we have recovered the diffusion approximation.
+Unfortunately, we have not been very precise about what we mean by $\epsilon$ -- nature gives us no such obvious small parameter that we can use to "scale down" step sizes and "scale up" rates.
+The van Kampen expansion makes this notion more precise by introducing the concept of system size.
+Instead of counts we consider a change of units into counts per system size, e.g.
+population density.
+We can imagine drawing ever larger rings in our landscape of individuals, defining an ever bigger and bigger "system."
+
+## The System size expansion
+
+ The method is based on derivation presented by the van Kampen [@vanKampen2007], though a formal proof can be found in earlier work of Thomas Kurtz [@Kurtz1970; @Kurtz1971; @Kurtz1972].
+
+### Introduction: a change of variables
 
@@
 This is known as the macroscopic equation, and corresponds to the density equations commonly written down.
-Following @vanKampen, we will call this difference $\alpha_1(\phi)$ as a shorthand\footnote{This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.
+Following @vanKampen2007, we will call this difference $\alpha_1(\phi)$ as a shorthand\footnote{This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.
 It so happens this term will give the macroscopic law for any Markov process, not only a birth-death process.
@@
 From \eqref{n mean} we can also conclude that the macroscopic (average) variable obeys the deterministic law.
-
-## Stochastic Differential Equations
-
-The partial differential equation (PDE) given by \eqref{FP} is sometimes referred to as the Fokker-Planck equation.
-It is possible to prove this PDE is equivalent to as the stochastic differntial equation (SDE),
-
-\begin{equation}
-\textrm{d} X_t = \mu(X_t) \textrm{d}t + \sigma(X_t) \textrm{d}B_t \label{SDE}
-\end{equation}
-
-which, despite the name, it is important to remember that this is actually an integral equation in the Itô calculus, and not a derivative.
-Note the separation of the differential elements, which are meant as shorthand for implicit integrals:
-
-\begin{equation}
-X_{t+s} - X{t}  = \int_t^{t+s} \mu(X_{\tau}, \tau) \textrm{d}\tau + \int_t^{t+s} \sigma(X_{\tau}, \tau) \textrm{d}B\tau
-\end{equation}
-
-The first integral is the ordinary Lebesgue integral, while the second is an Itô integral, which integrates over the stochastic process $B_t$ (Brownian motion, also termed a Weiner process).
-The Itô calculus follows different rules from the classic calculus, largely due to the fact that an Itô integral has non-zero quadratic variation.
-(Crudely put, $\textrm{d}x^2 \sim 0$, but $\textrm{d}B^2 \sim dx$).
-Though it is common to formulate stochastic models directly as SDEs, it is worth remembering that the SDE is the result of a small noise approximation in the system size expansion, and can break down under larger deviations, even when the mean-field equation remains approximately accurate.
-@Ovaskainen2010 provides an overview literature in that addresses this limit and discusses the use of the WKB^[Wentzel, Kramers & Brillouin.
-Naturally this is KWB in Holland, BWK in France, and JWKB in England (for Jeffreys).
-Note that the physicist Gregor Wentzel is not the same as the mathematician Alexander Wentzell of the Fredilin-Wentzell theorem for large deviations years later, though I believe the former's result can be derived directly from the latter's.] method to address this case.
 
@@
 
-Which agrees with calculations elsewhere [@NisbetGurney].
+Which agrees with calculations elsewhere [@Nisbet1982].
 
@@
 
-A somewhat different approach to deriving a diffusion equation predominates in the literature.
-This approach does not make system size explicit, but instead hinges on taking a simultaneous limit of both short time and small step size.
-I prefer the van Kampen derivation, since it appeals to system size rather than an arbitrary small parameter going to zero.
-It also provides a more natural treatment for nonlinear systems, for which the diffusion equation requires extra care.
-With these notes in mind, the diffusion derivation usually goes as follows:
-
-Consider small steps $x = \epsilon n$, and hence $p_n = \epsilon P_x$, where $\epsilon \ll 1$.  To keep the process from slowing down we consequently have to scale up the rates by $\epsilon$ as well:
-
-\begin{align*}
-	b_n - d_n = \frac{A_x}{\epsilon}
-\end{align*}
-
-Where $A$ is independent of $\epsilon$.  Similarly we have to scale the sum, which being proportional to the variance must be scaled by
-\begin{align*}
-b_n + d_n = \frac{B_x}{\epsilon^2}
-\end{align*}
-
-We Taylor expand the step operator $\mathbb E$ in the new variable:
-\begin{align*}
-\mathbb E^k &= 1+ \epsilon k\frac{\partial}{\partial x} + \frac{\epsilon^2 k^2}{2} \frac{\partial^2}{\partial x^2} + \ldots
-\end{align*}
-
-Then \eqref{master1} becomes:
-\begin{align}
-\frac{\partial P(x,t)}{\partial t} &= \left( \epsilon \frac{\partial }{\partial x} + \epsilon^2 \tfrac{1}{2}\frac{\partial^2 }{\partial x^2} + \ldots \right)\left(\frac{B_x}{2\epsilon^2}-\frac{A_x}{2\epsilon} \right)P \nonumber \\
-+&\left(- \epsilon \frac{\partial }{\partial x} + \epsilon^2 \tfrac{1}{2}\frac{\partial^2 }{\partial x^2} - \ldots \right) \left( \frac{B_x}{2\epsilon^2} + \frac{A_x}{2\epsilon} \right)P 
-\intertext{Collecting terms of common order $\epsilon$, }
-& = - \frac{\partial A_x P}{\partial x} + \tfrac{1}{2} \frac{\partial^2 B_x P }{\partial x^2} + \mathcal O(\epsilon^2)\label{FP1}
-\end{align}
-
-For small $\epsilon$ we can ignore the terms of $\mathcal O(\epsilon^2)$, and we have recovered the diffusion approximation.
-Unfortunately, nature does not give us a parameter $\epsilon$ that goes to zero.
-We expect that larger populations will have relatively smaller fluctuations (recall the introductory example) and hence this limit should have something to do with increasing the system size.
-We make these notions more precise by expanding equation \eqref{master1} explicitly in terms of the system size.
-This will lead us to a much richer and more satisfying description of demographic noise than the mathematical limit above.
-
 # System size expansion in higher-dimensional systems: environmental noise example
@@
 It can be helpful to think of fluctuations in $\psi$ as also arising from some lower level transitions, but we need not model these explicitly.
-Following the expansion as before (see @vanKampen for details), the mean dynamics are given by the macroscopic equations:
+Following the expansion as before (see @vanKampen2007 for details), the mean dynamics are given by the macroscopic equations:
 
@@
 
+## WKB
+
+@Ovaskainen2010 provides an overview literature in that addresses this limit and discusses the use of the WKB^[Wentzel, Kramers & Brillouin.
+Naturally this is KWB in Holland, BWK in France, and JWKB in England (for Jeffreys).
+Note that the physicist Gregor Wentzel is not the same as the mathematician Alexander Wentzell of the Fredilin-Wentzell theorem for large deviations years later, though I believe the former's result can be derived directly from the latter's.] method to address this case.
+
 # References

### 2018-02-22 paper/paper.Rmd
WIP

@@
 
-In the extensive literature developing the consequences of both demographic
-and environmental noise in models of increasing complexity, it can 
-be easy to miss the equally important developments in making the origins
-of these sources of noise more precise -- developments which have important
-consquences for when and how the resulting models can be applied.
-
-the canonical model on the highly influential @Leigh1981, @Lande1993, @Foley1994
-
-$$ \frac{\textrm{d} n}{\textrm{d} t} = \underbrace{f(n)}_{\textrm{det skeleton}} + \underbrace{\sigma_d \xi_d(t) \sqrt{n}}_{\textrm{demographic noise}} + \underbrace{\sigma_e \xi_e(t) n}_{\textrm{environmental noise}}$$
-
-Stochastic differential equations of this form are so-called "Langevin equation," popular in the physics literature.
-This notation is rather sloppy, as the standard calculus does not apply to stochastic equations.
-The Langevin formulation, most popular in physics, is usually associated with a Stratonovich integral.
-Mathematicians prefer to write stochastic differential equations as $\textrm{d} n = f(x) \textrm{d} t + g(x) \textrm{d} B_t$, where the differentials hint at the underlying definition that this is the sum of a classic Reimann integral (over $\textrm{d}t$) and a stochastic It\^o integral.
-
-Advances in both theoretical understanding and the advent of widely
-available and efficient numerical methods have helped drive research
-towards increasingly more explicit descriptions of how stochasticity
-enters into quantitative models.
-
-@Black2012 provides an important 
-review of intrinsic noise and approaches for addressing it, focusing
-in particular on use of the van Kampen system size expansion as a
-mathematical means to properly bridge between individual-based models
-(IBMs, usually implemented only in simulations) and familiar population-level
-dynamics (so called macroscopic dynamics).
-
-; @Ovaskainen2010
-
-This division between intrinsic or demographic noise and extrinsic or
-environmental noise
-
-The Gillespie simulation provides a simple and exact algorithm
-for simulating demographic stochasticity at an individual level.  
-Free from both the approximations and mathematical complexity of more common
-formulations using stochastic differential equations (SDEs), 
-(or their partial differential equation counterparts the Fokker-Planck equation),
-the Gillespie algorithm offers a simple and powerful approach to simulating
-stochastic phenomena that remains underutilized.
-
-The mathematical proofs @Kurtz1970 & @Kurtz1971, popularized by
-@vanKampen's intuitive "system size" expansion, provide a far more 
-explicit framework to interpret demographic stochasticity than the
-better known diffusion approximation [@Nisbet1982 @Lande2003; 
-see @Gardiner2009].
-
 It is often suggested that stochasticity is "inherent" or "fundamental,"
@@
 a description at the individual level to one at the population level.
+
+In the extensive literature developing the consequences of both demographic
+and environmental noise in models of increasing complexity, it can 
+be easy to miss the equally important developments in making the origins
+of these sources of noise more precise. Models in which stochasticity was
+essentially "tacked on" to a deterministic skeleton can be substituted for 
+approaches that permit both the skeleton and noise to emerge from underlying
+processes.  This more mechanistically specific approach to the origins of 
+noise can indeed recover the classic formulations, but it can also 
+help highlight where those models break down.
+Much of the literature [e.g. see @Ovaskainen2010] has focused on the following
+model, established in highly influential papers of @Leigh1981, @Lande1993,
+@Foley1994:
+
+$$ \textrm{d} N_t = \underbrace{f(N_t) \textrm{d} t}_{\textrm{det skeleton}} + \underbrace{\sigma_d \sqrt{N_t} \phantom{\cdot}\textrm{d}B_t^{(d)} }_{\textrm{demographic noise}} + \underbrace{\sigma_e N_t \phantom{\cdot}\textrm{d}B_t^{(e)}}_{\textrm{environmental noise}}$$
+
+Where $\textrm{d}B_t^{(d)}$ and $\textrm{d}B_t^{(e)}$ refer to Brownian processes, (i.e.
+Gaussian white noise process) for demographic and environmental stochasticity respectively.
+When $f(n)$ is simply a Verhulst logistic function this is often referred to as the canonical model [e.g.
+@@Ovaskainen2010], and provides a clear illustration of the partition of separate deterministic and stochastic elements.
+Significant research continues to focus on extending the study of this model to consider more complex and higher-dimensional $f(n)$ (coexistence models, stage structured models) and consider the case of autocorrelated, rather than than white noise of $\textrm{d}B_t^{(e)}$.
+
+Before we follow that path, I will review how a shift towards more mechanistic descriptions of the origins of noise provide a more nuianced view of when the canonical model does an does not hold in the context of both demographic and environmental stochasticity.
+My approach will highlight two fundamental tools: an ability for efficient, exact simulation using @Gillespie1977's stochastic simulation algorithm (SSA), and the van Kampen system size expansion [@vanKampen2007], which is rooted in the mathematical theorems of @Kurtz1970 & @Kurtz1971, in place of the more popular diffusion approximation derived by Taylor series [e.g.
+@Nisbet1982 @Lande2003; see comparisons in @vanKampen2007 and @Gardiner2009].
 
@@
 
-It is possible to simulate such a process exactly using Gillespie's
-exact stochastic simulation algorithm [SSA; @Gillespie1977], which
-simulates each individual birth or death as a separate event.
+The @Gillespie1977 provides a simple and exact algorithm
+for simulating demographic stochasticity at an individual level.  
+Free from both the approximations and mathematical complexity,
+the Gillespie algorithm offers a simple and powerful approach to simulating
+stochastic phenomena that remains underutilized.
 
@@
 
-Figure \ref{gillespie}
-uses this algorithm to compare simulations of @Levins1969 patch model;
-(see Appendix A for example code).  In the Levins'  model,
-
 Figure \ref{gillespie} shows the results of two exact SSA simulations of this model with
@@
 
-The van Kampen system size expansion provides a convenient way to
-explicitly approximate the "macroscopic" (i.e. population level,
-see @Black2012) dynamics from an individual-based description of
-events in a Markov process [@vanKampen2007; based on @Kurtz1970;
-@Kurtz1971].  In addition to illustrating how common stochastic
-models (in particular, Gaussian noise) arise from lower-level
-descriptions, this result will also provide us a useful framework
-for interpreting the behavior of stochastic models more generally.
-In Appendix B I illustrate the derivation of the following equations
-for the mean and expected population sizes for a generic density
-dependent birth-death process:
+The van Kampen system size expansion, shown for a general birth-death
+process in Appendix B, derives an ordinary differential equation (ODE)
+for the macroscopic process and a *linear* Fokker-Planck equation which
+governs the *deviations* from the macroscopic equation:
 
@@
 of the system size expansion].
+
+Note that the van Kampen expansion is not identical to the SDE 
+postulated by the canonical equation, which corresponds to a 
+non-linear Fokker-Planck equation for the probability density.  
+As shown in Appendix B, the popular expansion
 

### 2018-02-22 appendixB/appendixB.Rmd
WIP

@@
 bibliography: ../paper/refs.bib
-output: rticles::elsevier_article
+output: 
+  rticles::elsevier_article:
+    toc: true
+
 preamble: "\\newcommand{\\ud}{\\mathrm{d}}"
 ---
+
+\renewcommand{\thefigure}{B\arabic{figure}}
+
+# Preliminary comments on notation
 
@@
 (Crudely put, $\textrm{d}x^2 \sim 0$, but $\textrm{d}B^2 \sim dx$).
+The SDE notation also traditionally adopts capital letters such as $X_t$ to denote stochastic variables.
 
@@
 See @vanKampen2007 for an excellent discussion of the Itô-Stratonovich dillema and its intepretation with respect to intrinsic vs environmental noise factors.
+To avoid this ambiguity, mathematical literature will frequently write the Statonovich-type SDE explicitly as $\mathrm{d} X_t = f(X_t) \mathrm{d}t +  g(X_t) \circ  B_t$
 
-## Master equations and the step operator
+### Partial Differential Equation formulation
 
-Consider a population that is of size $n$ with probability $p_n$, with birth and death rates given by $b_n$ and $d_n$ respectively, diagrammed in Fig~\ref{markov}  This representation is known as a single step Markov process.
+An SDE can also be written in terms of an equivalent partial differential equation for the probability distribution of a given variable.  The SDE
+
+\begin{equation}
+\textrm{d} X_t = \mu(X_t) \textrm{d}t + \sigma(X_t) \textrm{d}B_t
+\end{equation}
+
+is equivalent to the PDE formulation [e.g. @vanKampen2007 or @Oksendal1985].
+
+\begin{equation}
+\frac{\partial}{\partial t} P(x,t) = - \frac{\partial}{\partial x} \mu(x) P(x,t) + \frac{1}{2} \frac{\partial^2}{\partial x^2} \sigma(x)^2 
+\end{equation}
+
+# Master equations and the step operator
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=.25\textwidth]{markov}
+\caption{Birth and death as state-dependent transition rates in a Markov process.}\label{markov}
+\end{center}
+\end{figure}
+
+Consider a population that is of size $n$ with probability $p_n$, with birth and death rates given by $b_n$ and $d_n$ respectively, diagrammed in Fig \ref{markov}  This representation is known as a single step Markov process.
 The rate of transitions to the $n+1$ state $b_n$, the rate of transitions to $n-1$ is $d_n$.
@@
 Master equations of this form can be written more concisely by introducing the step operator, $\mathbb E^k$ such that $\mathbb E^k f_n = f_{n+k}$, giving us
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=.25\textwidth]{markov}
-\caption{Birth and death as state-dependent transition rates in a Markov process.}\label{markov}
-\end{center}
-\end{figure}
 
@@
 
-## A traditional derivation
+## The system-size expansion vs the diffusion approximation
 
-A derivation of the above formulation can be found, e.g.
-in @Nisbet1982's classic textbook on fluctuating populations, or more recently in @Lande2003.
+@Kramers1940 @Moyal1949, @Kurtz1978
+
+@vanKampen1961 @vanKampen1976 @Kurtz1972 @Kurtz1978
+
+A derivation of diffusion approximation in @Nisbet1982's classic textbook on fluctuating populations, or more recently in @Lande2003.
 This diffusion approximation is also given for a more mathematically inclined audience in @Gardiner2009 and in @vanKampen2007.
@@
 
-## WKB
+## Large deviations
 
+Notably, both the SDE derived from the Kramer-Moyal diffusion approximation and the central limit theorem result arising from the system size expansion break down in the case of large deviations.
 @Ovaskainen2010 provides an overview literature in that addresses this limit and discusses the use of the WKB^[Wentzel, Kramers & Brillouin.
-Naturally this is KWB in Holland, BWK in France, and JWKB in England (for Jeffreys).
-Note that the physicist Gregor Wentzel is not the same as the mathematician Alexander Wentzell of the Fredilin-Wentzell theorem for large deviations years later, though I believe the former's result can be derived directly from the latter's.] method to address this case.
+Note that the physicist Gregor Wentzel is not the same as the mathematician Alexander Wentzell of the Fredilin-Wentzell theorem for large deviations years later, though I believe the former's result can be established more formally by the theorem from the latter.] method to address this case.
 

### 2018-02-22 paper/paper.Rmd
WIP

@@
 When $f(n)$ is simply a Verhulst logistic function this is often referred to as the canonical model [e.g.
-@@Ovaskainen2010], and provides a clear illustration of the partition of separate deterministic and stochastic elements.
+@Ovaskainen2010], and provides a clear illustration of the partition of separate deterministic and stochastic elements.
 Significant research continues to focus on extending the study of this model to consider more complex and higher-dimensional $f(n)$ (coexistence models, stage structured models) and consider the case of autocorrelated, rather than than white noise of $\textrm{d}B_t^{(e)}$.
 
-Before we follow that path, I will review how a shift towards more mechanistic descriptions of the origins of noise provide a more nuianced view of when the canonical model does an does not hold in the context of both demographic and environmental stochasticity.
-My approach will highlight two fundamental tools: an ability for efficient, exact simulation using @Gillespie1977's stochastic simulation algorithm (SSA), and the van Kampen system size expansion [@vanKampen2007], which is rooted in the mathematical theorems of @Kurtz1970 & @Kurtz1971, in place of the more popular diffusion approximation derived by Taylor series [e.g.
-@Nisbet1982 @Lande2003; see comparisons in @vanKampen2007 and @Gardiner2009].
+Before we follow that path, I will highlight
+
+I will review how a shift towards more mechanistic descriptions of the origins of noise provide a more nuianced view of when the canonical model does an does not hold in the context of both demographic and environmental stochasticity.
+My approach will highlight two fundamental tools: an ability for efficient, exact simulation using @Gillespie1977's stochastic simulation algorithm (SSA), and the van Kampen system size expansion [@vanKampen1976].
+
+which is rooted in the mathematical theorems of Thomas Kurtz [@Kurtz1970 & @Kurtz1971, @Kurtz1972 @Kurtz1978], in place of the more popular diffusion approximation derived by Taylor series [e.g.
+@Nisbet1982, @Lande2003; rooted in the earlier work of @Kramers1940 and @Moyal1949, see comparisons in @vanKampen2007 and @Gardiner2009].
 
@@
 The van Kampen system size expansion, shown for a general birth-death
-process in Appendix B, derives an ordinary differential equation (ODE)
-for the macroscopic process and a *linear* Fokker-Planck equation which
-governs the *deviations* from the macroscopic equation:
+process in Appendix B, derives an *ordinary* differential equation (ODE)
+for the macroscopic process, coupled to a *linear* SDE which
+governs the *deviations* ($\xi$)  from the macroscopic equation:
 
 \begin{align}
-\frac{\mathrm{d} x }{dt}  &= b(x) - d(x) + O(N^{-1})\\
-\frac{\mathrm{d} \sigma^2 }{dt}  &= 2 \left( \partial_x b(x)  - \partial_x d(x) \right) \sigma^2 + b(x) + d(x) \label{bd}
+\frac{\mathrm{d} x }{dt}  &= b(x) - d(x) + \mathcal{O}(N^{-1}) \\
+\mathrm{d} \xi &=  \partial_{x}  \left( b(x)  - d(x) \right)  \xi \mathrm{d}t + \sqrt{b(x) + d(x)} \mathrm{d} B_t   + \mathcal{O}(N^{-\tfrac{1}{2}}
 \end{align}
 
-where $x = \langle n \rangle$, the average or expected population
-size of an ensemble of replicates, assuming the transitions (+/- 1
-in this case) are much smaller than the system size ($N$ in the
-Levins model).  The dynamics for the average population
+\begin{align}
+\frac{\mathrm{d} x }{\mathrm{d} t}  &= b(x) - d(x) + \mathcal{O}(N^{-1})\\
+\frac{\mathrm{d} \sigma^2 }{\mathrm{d} t}  &= 2 \left( \partial_x b(x)  - \partial_x d(x) \right) \sigma^2 + b(x) + d(x)  + \mathcal{O}(N^{-\tfrac{1}{2}}) \label{bd}
+\end{align}
+
+The dynamics for the average population
 match the population-level equation.^[up to fluctuations of order
@@
 Note that the van Kampen expansion is not identical to the SDE 
-postulated by the canonical equation, which corresponds to a 
-non-linear Fokker-Planck equation for the probability density.  
+postulated by the-so called canonical equation.  
 As shown in Appendix B, the popular expansion
@@
 
+$$
 \begin{equation}
-\sigma_x^2  = \frac{ \left( \partial_y f(x,y) \right)^2}{\left( \partial_x f(x,y) \right)^2 + \partial_x f(x,y) \tau_c } \sigma_e^2 + \sigma^2_d \label{env_noise}
+\sigma_x^2  = \frac{ \left( \partial_y f(x,y) \right)^2}{\left( \partial_x f(x,y) \right)^2 + \partial_x f(x,y) \tau_c } \sigma_e^2 + \sigma^2_d 
+\label{env_noise}
 \end{equation}
+$$
 

### 2018-02-23 appendixB/appendixB.Rmd
WIP
van Kampen

@@
 
-which for historical reasons is known as a master equation.
+which for historical reasons [see @vanKampen2007] is known as a master equation.
 From this we can directly calculate the mean and variance for this process by multiplying by $n$ or $n^2$ and summing over all $n$.
@@
 For small $\epsilon$ we can ignore the terms of $\mathcal O(\epsilon^2)$, and we have recovered the diffusion approximation.
+As noted above, we can write this PDE in the more compact notation of an SDE, substituting in our original terms for birth and death:
+
+\begin{equation}
+\ud X_t = \left[ b(x) - d(x) \right] \ud t + \sqrt{b(X_t) + d(X_t)} \ud W_t
+\end{equation}
+
 Unfortunately, we have not been very precise about what we mean by $\epsilon$ -- nature gives us no such obvious small parameter that we can use to "scale down" step sizes and "scale up" rates.
@@
  The method is based on derivation presented by the van Kampen [@vanKampen2007], though a formal proof can be found in earlier work of Thomas Kurtz [@Kurtz1970; @Kurtz1971; @Kurtz1972].
+
+The van Kampen system size expansion, shown for a general birth-death
+process in Appendix B, derives an *ordinary* differential equation (ODE)
+for the macroscopic process, coupled to a *linear* SDE which
+governs the *deviations* ($\xi$)  from the macroscopic equation:
+
+\begin{align}
+\frac{\mathrm{d} x }{dt}  &= b(x) - d(x) + \mathcal{O}(N^{-1}) \\
+\mathrm{d} \xi &=  \partial_{x}  \left( b(x)  - d(x) \right)  \xi \mathrm{d}t + \sqrt{b(x) + d(x)} \mathrm{d} B_t   + \mathcal{O}(N^{-\tfrac{1}{2}}
+\end{align}
 

### 2018-02-23 paper/paper.Rmd
WIP
van Kampen

@@
 \newpage
+
+<!-- FIXME look at Coulson, Bjornstead, Ovaskeinen and others and see how they define scope to focus on "population ecology" -->
 
@@
 
-$$ \textrm{d} N_t = \underbrace{f(N_t) \textrm{d} t}_{\textrm{det skeleton}} + \underbrace{\sigma_d \sqrt{N_t} \phantom{\cdot}\textrm{d}B_t^{(d)} }_{\textrm{demographic noise}} + \underbrace{\sigma_e N_t \phantom{\cdot}\textrm{d}B_t^{(e)}}_{\textrm{environmental noise}}$$
+\begin{equation}
+\textrm{d} N_t = \underbrace{f(N_t) \textrm{d} t}_{\textrm{det skeleton}} + \underbrace{\sigma_d \sqrt{N_t} \phantom{\cdot}\textrm{d}B_t^{(d)} }_{\textrm{demographic noise}} + \underbrace{\sigma_e N_t \phantom{\cdot}\textrm{d}B_t^{(e)}}_{\textrm{environmental noise}}
+\label{canonical}
+\end{equation}
 
@@
 
-Before we follow that path, I will highlight
-
-I will review how a shift towards more mechanistic descriptions of the origins of noise provide a more nuianced view of when the canonical model does an does not hold in the context of both demographic and environmental stochasticity.
-My approach will highlight two fundamental tools: an ability for efficient, exact simulation using @Gillespie1977's stochastic simulation algorithm (SSA), and the van Kampen system size expansion [@vanKampen1976].
-
-which is rooted in the mathematical theorems of Thomas Kurtz [@Kurtz1970 & @Kurtz1971, @Kurtz1972 @Kurtz1978], in place of the more popular diffusion approximation derived by Taylor series [e.g.
-@Nisbet1982, @Lande2003; rooted in the earlier work of @Kramers1940 and @Moyal1949, see comparisons in @vanKampen2007 and @Gardiner2009].
+Before we follow that path, we will examine how this canonical model can emerge from a lower-level description and how it can deviate from that description.
+This approach will also set the stage to see how our understanding of stochastic models can be extended to better reflect natural processes that shape not only for the so-called deterministic skeleton, but the noise as well.
+In so doing, I will highlight two tools fundamental to this analysis, one computational and one analytical: the exact simulation method of @Gillespie1977's stochastic simulation algorithm (SSA), and the van Kampen system size expansion [@vanKampen1976].
 
@@
 ways to leave the state: a simple expression of probability balance
-known as the "master equation" [@vanKampen2007].
+known as the "master equation" [@vanKampen2007].  Note that in general
+this approach is equally applicable to stochastic transitions of any sort,
+not just step sizes of +/- 1 and not just birth and death events, but
+can include transitions between stage classes or trait values, including
+mutations to continuously-valued traits in evolutionary dynamics [e.g. @Boettiger2009].
 
-The @Gillespie1977 provides a simple and exact algorithm
+The @Gillespie1977 provides an exact algorithm
 for simulating demographic stochasticity at an individual level.  
+The algorithm is merely a simple and direct implemtentation of the master
+equation, progressing in random step sizes deterimed
+by the waiting time until the next event.
 Free from both the approximations and mathematical complexity,
-the Gillespie algorithm offers a simple and powerful approach to simulating
-stochastic phenomena that remains underutilized.
+the Gillespie algorithm is an interesting example of where we rely on a
+numerical implementation to check the accuracy of an analytic approximation,
+even in the case of simple models such as we will discuss.  Though the 
+algorithm is often maligned as numerically demanding, it can be run much
+more effectively even on large models on today's computers than when
+it was first developed in the 70s, and remains an underutilized approach
+for writing simple and approximation-free[^2] stochastic ecological models.
 
-We can illustrate both the Gillespie algorithm and the van Kampen
-system size expansion using @Levins1969 classic patch model,
+[^2]: that is, free from the approximation made by SDE models as we see in the
+van Kampen example.  All models are, of course, only approixmations.
+
+As our objective is to tie the origins of noise more closely to biological
+processes, it will be helpful to make the notion of a master equation concrete
+with a spefiic example.  We will focus on the classic case of @Levins1969 patch model,
+to illustrate the Gillespie algorithm and the van Kampen system size expansion
 
@@
 proportional to the fraction of available patches, $1 - n/N$.
+
 We choose this model because it has precisely same functional form
@@
 convience but leads to real differences in observed stochastic properties
-of the model.  Though we focus on this simple model for clarity of 
-presentation, both computational and mathematical framework presented here 
-can easily be applied to alternative models to reach the same conclusions.
+of the model. Both the numerical approach (details in Appendix A) and the 
+analytical approach (Appendix B) illustrate that it is straight-forward
+to use this same approach in alternate formuluations.
 
@@
 identical parameters except for the total number of available sites,
-$N$.
+$N$, illustrating the magnitude of these fluctuations indeed scales in proportion
+to $\sqrt{N}$ as postulated by the canonical equation, Eq \eqref{canoical}.
 
@@
 
-The van Kampen system size expansion, shown for a general birth-death
-process in Appendix B, derives an *ordinary* differential equation (ODE)
-for the macroscopic process, coupled to a *linear* SDE which
-governs the *deviations* ($\xi$)  from the macroscopic equation:
+The van Kampen system size expansion provides a mechanism to approximate the individual
+dynamics represented by the master equation (equivalently, by the Gillespie simulation)
+with a population-level description of the process.  This is important not only because
+it will illustrate both origins and approximations that lie behind the population level
+models that predominate this literature, but will give our syntheses a chance to point
+out some more subtle misconceptions that have also arisen. Details of this expansion
+for a generic birth death process and comparison with the more familiar diffusion 
+approximation are found in Appendix B.
 
-\begin{align}
-\frac{\mathrm{d} x }{dt}  &= b(x) - d(x) + \mathcal{O}(N^{-1}) \\
-\mathrm{d} \xi &=  \partial_{x}  \left( b(x)  - d(x) \right)  \xi \mathrm{d}t + \sqrt{b(x) + d(x)} \mathrm{d} B_t   + \mathcal{O}(N^{-\tfrac{1}{2}}
-\end{align}
+van Kampen's approximation follows a central-limit theorem style argument established rigorously in
+the theorems of Kurtz at about the same time [@Kurtz1970 & @Kurtz1971, @Kurtz1972, in particular, see Thereom 4.4. @Kurtz1978].
+The heart of the approximation is to imagine a population on some large landscape of which
+we sample only a fixed area to estimate the population density.  That area is our 
+system size, which as it expands includes more and more total individuals and
+experiences proportionally smaller fluctionations in population density.  This 
+results in the conclusion that the population density obeys a simple Gaussian
+distribution, whose mean size $x$ and variance  $\sigma^2$ are given by *ordinary* differential equations:
 
@@
 \frac{\mathrm{d} x }{\mathrm{d} t}  &= b(x) - d(x) + \mathcal{O}(N^{-1})\\
-\frac{\mathrm{d} \sigma^2 }{\mathrm{d} t}  &= 2 \left( \partial_x b(x)  - \partial_x d(x) \right) \sigma^2 + b(x) + d(x)  + \mathcal{O}(N^{-\tfrac{1}{2}}) \label{bd}
+\frac{\mathrm{d} \sigma^2 }{\mathrm{d} t}  &= 2 \partial_x  \left[ b(x)  - d(x) \right] \sigma^2 + b(x) + d(x)  + \mathcal{O}(N^{-\tfrac{1}{2}}) \label{vanKampen}
 \end{align}
 
-The dynamics for the average population
-match the population-level equation.^[up to fluctuations of order
-$\tfrac{1}{N}$. We can actually both observe and successfully
-approximate this higher-order correction from additional terms in
-the system size expansion to recover the phenomenon of stochastic
-inflation [@Gonzalez2002], where the average population size can
-be larger or smaller than predicted by the deterministic model --
-see Appendix A for examples in simulations compared to predictions
-of the system size expansion].
+This result is striking for a variety of reasons.  First, we see the recovery of the 
+"deterministic skeleton," in which we naively replace population density $x/N$ for
+population count $n$ describes the *expected* density of the population
+(i.e. averaged over an esemble of replicate realizations of the Gillespie simulations),
+and with remarkable accuracy, up to order $1/N}$, i.e.
+a single individual.^[We can actually extend the van Kampen expansion to higher order and successfully approximate even this higher-order correction to the mean dynamics, recovering the phenomenon of stochastic inflation [@Gonzalez2002], where the average population size can be larger or smaller than predicted by the deterministic model depending on the curvature of the macroscopic equation].
+The variance term is equally interesting.  First, we note that noise arises from
+the interaction of individual birth and death events that defined our process,
+illustrating the origins of this noise.  Because both individual events of births
+and deaths add randomness, they both add to the growth of varation observed, i.e.
+$b(x) + d(x)$.  Second, we see that dynamics for the variance $\sigma_^2$ are *non-linear* whenever
+$b(x)$ or $d(x)$ are non-linear, so that different values of the population density
+will experience different noise intensities! In fact, this variance can only
+have an equilibrium size if the macroscopic equation has an equilibrium ($\partial_x  \left[ b(x)  - d(x) \right] = 0$)
+Third, this expansion shows the variance
+equation is also less precise, accurate only to order $1/\sqrt{N}$.
 
-Note that the van Kampen expansion is not identical to the SDE 
-postulated by the-so called canonical equation.  
-As shown in Appendix B, the popular expansion
+In the interests of a theoretical synthesis, I observe that this result
+is *different* than the canonical equation, Eq \eqref{canonical}, or more generally
+the diffusion approximation [or Kramer-Moyal expansion @Kramers1940; @Moyal1949; @Gardiner2009]
+which is much more typical in the ecological literature [e.g. the books of @Nisbet1982 and @Lande2003
+or the review in @Ovaskeinen2010]. @Black2012 gives an excellent introduction 
+to the van Kampen expansion as a connection between individual-based and population-level
+model descriptions and applications, but fails to point out these differences,
+mistakenly characterizing the van Kampen expansion as synonomous with the diffusion approximation.
+In brief, the diffusion approximation does not lead to a Gaussian distribution
+(equivalently, an SDE or PDE with linear noise term, see Appendix B), but rather
+a *non-linear* SDE of the form (e.g. @Ovaskeinen2010)
+
+\begin{equation}
+\ud X_t = \left[ b(x) - d(x) \right] \ud t + \sqrt{b(X_t) + d(X_t)} \ud W_t \label{diffusion}
+\end{equation}
+
+This result has several potential draw-backs.  First, as noted in @vanKampen2007 or
+@Gardiner2009 (see Appendix B), the Kramer-Moyal expansion approach rests on rather
+a more dubious double limit that often has no clear biological intepretation and
+can restrict its validity (though I believe theorem 3.3 of @Kurtz1978 establishes
+this approximation on firmer grounds).  Second, the non-linear SDE formulation can
+prove more difficult to work with than the convenient ODE expressions from van Kampen,
+Eq \eqref{\vanKampen} (which can be expressed as a coupled ODE for the mean and 
+a linear SDE for deviations).  Third, while both approximations break down
+in the case of large deviations [see @Ovaskeinen2010], the van Kampen expansion
+shows that the macroscopic equation will continue to hold even when the 
+variance does not; in contrast to the non-linear SDE from the diffusion approximation.
+
+- Link back to Levins, show where same
+- IMPORTANCE of fluctuation dissipation theorem to the tranisition probability discussion!!
+- ENV noise and colored noise
 
@@
 fluctuation-dissipation theorem [@vanKampen2007], an will prove a useful
-expression in understanding stochastic dynamics.  Because Eq \eqref{bd}
+expression in understanding stochastic dynamics.  Because Eq \eqref{vanKampen}
 also describes the time-evolution of these fluctuations, we can similarly

### 2018-02-26 appendixB/appendixB.Rmd
WIP

@@
 which for historical reasons [see @vanKampen2007] is known as a master equation.
+Note that we can define the master equation more generally in terms of a generic transition probability $w(y|x)$ of going into any state $y$ given that the system is currently in state $x$:
+
+$$
+\begin{equation}
+\frac{\mathrm{d}}{\mathrm{d}t} P(x,t) = \int w(x|y) P(y,t) - w(y|x)P(x,t) \mathrm{d} x
+\end{equation}
+$$
+
+The master equation can be thought of as a more process-based description of the Chapman-Kolmorogov equation defining a stochastic process [@vanKampen2007].
+
+<!--
 From this we can directly calculate the mean and variance for this process by multiplying by $n$ or $n^2$ and summing over all $n$.
@@
 We will illustrate how this is done in both the traditional diffusion approximation and then in the van Kampen system size expansion.
+-->
 
@@
 
-@Kramers1940 @Moyal1949, @Kurtz1978
-
-@vanKampen1961 @vanKampen1976 @Kurtz1972 @Kurtz1978
-
-A derivation of diffusion approximation in @Nisbet1982's classic textbook on fluctuating populations, or more recently in @Lande2003.
-This diffusion approximation is also given for a more mathematically inclined audience in @Gardiner2009 and in @vanKampen2007.
-The approach starts with the same master equation that defines the underlying dynamics of births and deaths (matching the process implemented in the exact Gillespie algorithm, @Gillespie1977, see Appendix A).
+A derivation of diffusion approximation is originally credited to @Kramers1940 and later improved by @Moyal1949, though Einstein makes implicit use of this approximation in his own work on Brownian Motion decades earlier [@Gardiner2009].
+The diffusion approximation has been a popular basis for ecological stochastic models as well, such as in @Nisbet1982's classic textbook on fluctuating populations, or more recently in @Lande2003 and work reviewed in @Ovaskeinen.
+This diffusion approximation is nicely summarized @Gardiner2009 and in @vanKampen2007, though I believe @Kurtz1978 provides a more formal proof that goes unmentioned by any of these references.
+Both the diffusion approximation and the van Kampen expansion start with the same master equation that defines the underlying dynamics of births and deaths (matching the process implemented in the exact Gillespie algorithm, @Gillespie1977, see Appendix A).
 
@@
 
-## The System size expansion
+## The system size expansion
 
- The method is based on derivation presented by the van Kampen [@vanKampen2007], though a formal proof can be found in earlier work of Thomas Kurtz [@Kurtz1970; @Kurtz1971; @Kurtz1972].
-
-The van Kampen system size expansion, shown for a general birth-death
-process in Appendix B, derives an *ordinary* differential equation (ODE)
-for the macroscopic process, coupled to a *linear* SDE which
-governs the *deviations* ($\xi$)  from the macroscopic equation:
+This method is based on derivation originally proposed by van Kampen @vanKampen1961 @vanKampen1976 and later grounded in more formal work of Thomas Kurtz, [@Kurtz1970; @Kurtz1971; @Kurtz1972; @Kurtz1978].
+In contrast to the diffusion approximation above, the system size expansion derives an *ordinary* differential equation (ODE) for the macroscopic process, coupled to a *linear* SDE which governs the *deviations* ($\xi$)  from the macroscopic equation:
 
@@
 
-### Introduction: a change of variables
+ Since this system of equations defines a Gaussian probability distribution, the main text summarizes this result as coupled ODEs for mean and variance, which should be familiar to readers not familiar with stochastic differential equation SDE formulations.
+I summarize the derivation here for the birth death process, though an accessible presentation can found for both this and the more general form of the master equation in van Kampen's popular text, @vanKampen2007.
 
@@
 
-We will change Eq \eqref{master1} into the variables $\phi$ and $\xi$.
+We will change Eq \eqref{master1} into the variables $\phi$ (the macrosopic value, i.e.
+density) and $\xi$ (i.e.
+fluctuations/deviations from the expected density).
 We begin with the step operator, which can be approximated by a Taylor expansion.To formulate a Taylor expansion of the step operator (where $k$ can be a positive or negative integer), first consider the steop operator under the original discrete variable $n$:
@@
 \begin{align}
-\frac{\ud \phi(t)}{\ud t} = b(\phi)-d(\phi) = \alpha_1(\phi) \label{macroscopic}
+\frac{\ud \phi(t)}{\ud t} = b(\phi)-d(\phi) = a_1(\phi) \label{macroscopic}
 \end{align}
@@
 This is known as the macroscopic equation, and corresponds to the density equations commonly written down.
-Following @vanKampen2007, we will call this difference $\alpha_1(\phi)$ as a shorthand\footnote{This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.
+Following @vanKampen2007, we will call this difference $a_1(\phi)$ as a shorthand.
+This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.
 It so happens this term will give the macroscopic law for any Markov process, not only a birth-death process.
 The second moment of the transition rates, $a_2$, will be for us simply the sum of the birth and death rates.
-We use this notation as it generalizes to processes that have a different master equation from \eqref{master1}, to describe a transition of arbitrary step size: $\alpha_{i,j} = \int r^i \Phi_j(r) \ud r$.}
+We use this notation as it generalizes to processes that have a different master equation from \eqref{master1}, to describe a transition of arbitrary step size:
 
-Collecting terms of order $\Omega^0$ we recover the diffusion equation:
+$$
+\begin{equation}
+a_i = \int r^i \Phi(r) \mathrm{d} r
+\end{equation}
+$$
+The first jump moment, $i == 1$ the two possible steps $r$ are $\pm 1$, and thus $a_1$ is sum of the two possible one-step transitions $+1 \cdot b(x) + -1 \cdot d(x)$.
+The second jump moment the $r$ term is raised to the second power, making both contributions positive such that $a_2(x) = b(x) + d(x)$
+
+Collecting terms of order $\Omega^0$ we recover the *linear* Fokker Planck equation:
 
 \begin{align}
-\frac{\partial \Pi}{\partial t} = - \alpha'_{1,0}(\phi) \frac{\partial}{\partial \xi} \xi \Pi + \tfrac{1}{2}\alpha_{2,0}(\phi) \frac{\partial^2}{\partial \xi^2} \Pi \label{FP}
+\frac{\partial \Pi}{\partial t} = - a'_1(\phi) \frac{\partial}{\partial \xi} \xi \Pi + \tfrac{1}{2}a_2(\phi) \frac{\partial^2}{\partial \xi^2} \Pi \label{FP}
 \end{align}
 
-Where we define $\alpha_2(\phi) = b(\phi)+d(\phi)$ as an analogously.
-From this it is a straight forward exercise to calculate the moments of the distribution (multiply by $\xi$, $\xi^2$ and integrate),
+Where we define $a_2(\phi) = b(\phi)+d(\phi)$ as an analogously.
+Contrast this to the Fokker Planck equation of the diffusion equation, where the birth and death terms appear *inside* the derivatives and the derivatives are with respect to the state variable $x$ rather than the fluctuations, $\xi$.
+
+From here it is a straight forward exercise to calculate the moments of the distribution (multiply by $\xi$, $\xi^2$ and integrate),
 
 \begin{align}
-\partial_t \langle \xi \rangle &= \alpha'_{1,0}(\phi) \langle \xi \rangle \\
-\partial_t  \langle \xi^2 \rangle &= 2\alpha'_{1,0}(\phi) \langle \xi \rangle + \alpha_{2,0}(\phi) \label{fluc} 
+\partial_t \langle \xi \rangle &= a'_1(\phi) \langle \xi \rangle \\
+\partial_t  \langle \xi^2 \rangle &= 2a'_1(\phi) \langle \xi \rangle + a_2(\phi) \label{fluc} 
 \end{align}
@@
 Eq \eqref{fluc} is particularly instructive.
-Note that $\alpha_2(\phi)$ is always positive, and hence will try to increase the fluctuations $\langle \xi^2 \rangle$.
-As this term increases, it increases the influence of its coefficient $\alpha_1'(\phi)$.
+Note that $a_2(\phi)$ is always positive, and hence will try to increase the fluctuations $\langle \xi^2 \rangle$.
+As this term increases, it increases the influence of its coefficient $a_1'(\phi)$.
 If this term is negative (as it must be near a stable equilibrium) then it serves to dissipate these fluctuations.
@@
 \begin{align}
-\langle \xi^2 \rangle = \frac{-\alpha_2(\phi)}{2 \alpha_1'(\phi)} \label{fluctuation dissipation}
+\langle \xi^2 \rangle = \frac{-a_2(\phi)}{2 a_1'(\phi)} \label{fluctuation dissipation}
 \end{align}
 
-However, $\alpha_1'(\phi)$ need not be negative everywhere, but will be negative for any stable point, $\alpha_1'(\phi^s) < 0$.
+However, $a_1'(\phi)$ need not be negative everywhere, but will be negative for any stable point, $a_1'(\phi^s) < 0$.
 Hence there will be a region around any stable point where this fluctuation-dissipation relationship given by \eqref{fluctuation dissipation} provides a good description of the fluctuations.
@@
 \begin{align}
-\alpha_1(\phi) &= c \phi(1- \phi) - e\phi \\
-\alpha_2(\phi) &= c \phi(1-\phi) + e\phi
+a_1(\phi) &= c \phi(1- \phi) - e\phi \\
+a_2(\phi) &= c \phi(1-\phi) + e\phi
 \end{align}
@@
 
-# An alternate derivation: the diffusion approximation
-
 # System size expansion in higher-dimensional systems: environmental noise example
@@
 \begin{align*}
-\frac{\ud \phi}{\ud t} &= \alpha_{1,0}(\phi, \psi) \\
-\frac{\ud \psi}{\ud t} &= \beta_{1,0}(\psi) 
+\frac{\ud \phi}{\ud t} &= a_1(\phi, \psi) \\
+\frac{\ud \psi}{\ud t} &= \beta_1(\psi) 
 \end{align*}
@@
 \begin{align}
-\frac{\ud \langle \xi^2 \rangle}{\ud t} &= 2 \frac{\partial \alpha_{1,0}}{\partial \phi} \langle \xi^2 \rangle + 2 \frac{\partial \alpha_{1,0} }{\partial \psi} \langle \xi \eta \rangle + \alpha_{2,0} \label{phivar} \\
-\frac{\ud \langle \xi \eta \rangle}{\ud t} &= \left( \frac{\partial \alpha_{1,0}}{\partial \phi} + \frac{\partial \beta_{1,0} }{\partial \psi} \right)\langle \xi \eta \rangle + \frac{\partial \alpha_{1,0} }{\partial \psi} \langle \eta^2 \rangle \label{covar}\\
-\frac{\ud \langle \eta^2 \rangle}{\ud t} &= 2 \frac{\partial \beta_{1,0}}{\partial \psi} \langle \eta^2 \rangle + \beta_{2,0} \label{psivar}
+\frac{\ud \langle \xi^2 \rangle}{\ud t} &= 2 \frac{\partial a_1}{\partial \phi} \langle \xi^2 \rangle + 2 \frac{\partial a_1 }{\partial \psi} \langle \xi \eta \rangle + a_2 \label{phivar} \\
+\frac{\ud \langle \xi \eta \rangle}{\ud t} &= \left( \frac{\partial a_1}{\partial \phi} + \frac{\partial \beta_1 }{\partial \psi} \right)\langle \xi \eta \rangle + \frac{\partial a_1 }{\partial \psi} \langle \eta^2 \rangle \label{covar}\\
+\frac{\ud \langle \eta^2 \rangle}{\ud t} &= 2 \frac{\partial \beta_1}{\partial \psi} \langle \eta^2 \rangle + \beta_2 \label{psivar}
 \end{align}
 
-Without an explicit model for the environmental variation process, we can merely summarize its role in terms of the autocorrelation, $\tau_c := \tfrac{\partial \beta_{1,0}}{\partial \psi}$ and its overall variance at steady-state, $\langle \eta^2 \rangle = \tfrac{\beta_{2,0}}{-2 \partial_{\psi} \beta{1,0}}$, which we will refer to as the environmental variation $\sigma^2_e := \langle \eta^2 \rangle$.
+Without an explicit model for the environmental variation process, we can merely summarize its role in terms of the autocorrelation, $\tau_c := \tfrac{\partial \beta_1}{\partial \psi}$ and its overall variance at steady-state, $\langle \eta^2 \rangle = \tfrac{\beta_2}{-2 \partial_{\psi} \beta{1,0}}$, which we will refer to as the environmental variation $\sigma^2_e := \langle \eta^2 \rangle$.
 A more detailed mechanistic description of how the environmental noise arises would permit a time-dependent solution, but the system size expansion tells us that these two summary statistics are all we need to express the steady-state solution above.
@@
 \begin{equation*}
-\langle xi^2 \rangle = \frac{ \left( \partial_{\psi} \alpha_{1,0}(\phi,\psi) \right)^2}{\left( \partial_{\phi} \alpha_{1,0}(\phi,\psi) \right)^2 + \partial_{\phi} \alpha_{1,0} (x,y) \tau_c } \sigma_e^2 + \sigma^2_d
+\langle xi^2 \rangle = \frac{ \left( \partial_{\psi} a_1(\phi,\psi) \right)^2}{\left( \partial_{\phi} a_1(\phi,\psi) \right)^2 + \partial_{\phi} a_1 (x,y) \tau_c } \sigma_e^2 + \sigma^2_d
 \end{equation*}
 
-From which we recover the environmental variation equation presented in the main text.
+<!-- FIXME -->
+
+$$
+\begin{align}
+\frac{\mathrm{d} x}{\mathrm{d} t} &= \alpha_1(x, y)  \\
+\frac{\mathrm{d} y}{\mathrm{d} t} &= \beta_1(y) 
+\end{align}
+$$
+
+while the fluctuation dynamics are now given by three coupled differential equations:
+
+For a generic birth death process that depends both on
+on population state $x$ but also on environmental
+factors $y$, where we write $a_1(x,y) := b(x,y) - d(x,y)$,
+sum $a_2(x,y) := b(x,y) + d(x,y)$
+
+For a generic population model
+that depends not only on population state $x$ but also on environmental
+factors $y$, it is possible to derive the steady-state variance
+(Appendix B) as:
+
+$$
+\begin{align}
+&\sigma_x^2  = \frac{  \partial_y a_1(x,y) ^2}{ \partial_x a_1(x,y) ^2 +  \tau_c \phantom\cdot \partial_x a_1(x,y)  } \sigma_e^2 + \frac{1}{2} \frac{a_2(x,y)}{\partial_x a_1(x,y)}
+\end{align}
+$$
+
+where $\sigma_e^2$ is the variance in the environmental noise and
+$\tau_c$ the auto-correlation of that noise.
 
@@
 @Ovaskainen2010 provides an overview literature in that addresses this limit and discusses the use of the WKB^[Wentzel, Kramers & Brillouin.
-Note that the physicist Gregor Wentzel is not the same as the mathematician Alexander Wentzell of the Fredilin-Wentzell theorem for large deviations years later, though I believe the former's result can be established more formally by the theorem from the latter.] method to address this case.
+Note that the physicist Gregor Wentzel is not the same as the mathematician Alexander Wentzell of the Fredilin-Wentzell theorem for large deviations years later, though I believe the former's result can be established more formally by the theorem from the latter.] method to address this case, which is of particular interest in the persistence / coexistence literature where stochastic extinction arises only from such large deviations.
 

### 2018-02-26 paper/paper.Rmd
WIP

@@
 will experience different noise intensities! In fact, this variance can only
-have an equilibrium size if the macroscopic equation has an equilibrium ($\partial_x  \left[ b(x)  - d(x) \right] = 0$)
+have an equilibrium size if the macroscopic equation has an equilibrium ($\partial_x  \left[ b(x)  - d(x) \right] < 0$)
 Third, this expansion shows the variance
 equation is also less precise, accurate only to order $1/\sqrt{N}$. 
+This variance equation will be important to us later as we examine the relative
+roles of noise and nonlinearity in noise-driven phenomena such as stochastic
+transitions between alternate stable states.
+
+The expected auto-correlation, $\langle x_0, x_t \rangle$ also follows from the
+system size expansion, e.g. at stationary state:
+
+$$\langle x_0, x_{t} \rangle = \frac{b(x) + d(x)}{2 \partial_x \left[ b(x) - d(x) \right]} \exp\lbrace  -t \left| \partial_x b(x) - \partial_x d(x) \right|  \rbrace $$
+which will also prove useful later on.  Note this expression simply states that the equilbrium
+variance is a balance between noise terms, $b(x)+ d(x)$ and the slope or eigenvalue of the
+deterministic skeleton, $2 \partial_x \left[ b(x) - d(x) \right]$, (a balance that forms the basis of the
+the fluctuation-disspation theorem) and that correlations fall off exponentially at a rate given by this slope.   
+We will see both autocorrelation and variance play similar roles to infer this eigenvalue when we 
+reach the role of noise as informer.
 
@@
 
-This result has several potential draw-backs.  First, as noted in @vanKampen2007 or
+This better-known approach has several potential draw-backs.  First, as noted in @vanKampen2007 or
 @Gardiner2009 (see Appendix B), the Kramer-Moyal expansion approach rests on rather
@@
 
-- Link back to Levins, show where same
-- IMPORTANCE of fluctuation dissipation theorem to the tranisition probability discussion!!
-- ENV noise and colored noise
-
-The change in variance can be seen
-as a balance between the stochastic contribution, $b(x) + d(x)$
-and the deterministic contribution, $\partial_x \left(b(x) - d(x)\right)$. 
-The deterministic contribution is the slope of the macroscopic
-equation, that is, the eigenvalue indicating the stability of the
-equilibrium point.  For stable points, this value must be negative
-and precisely balance the strictly positive stochastic contribution 
-of $b(x) + d(x)$ (the greater the rate of stochastic events involved, 
-the greater the fluctuations).  This balance at steady-state is known as the
-fluctuation-dissipation theorem [@vanKampen2007], an will prove a useful
-expression in understanding stochastic dynamics.  Because Eq \eqref{vanKampen}
-also describes the time-evolution of these fluctuations, we can similarly
-solve for the correlation between two points separated by time $\tau$.
-
-For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$,
-we find the average number of occupied patches 
+Applying the system size approximation to the example of the Levins model introduced
+previously lets us both recover the classic model while adding some nuance in 
+the interpretation.
+Taking $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$ and solving for the steady
+state to van Kampen's ODEs, we see the average number $\langle n \rangle$  of occupied patches is
 $\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of 
@@
 to the system size, $N$, which explains the standard deviation
-increases only as $\sqrt{N}$ while the steady state itself increases
-in direct proportion to $N$, resulting in the smaller relative size
-of fluctuations in the larger system in Figure \ref{gillespie}.  We can also see
-that different ways of partitioning the same macroscopic equation
+increases as $\sqrt{N}$, as in the canonical model and as predicted by Figure \ref{gillespie}.  
+We can also see that different ways of partitioning the same macroscopic equation
 into components of birth and death will create the same mean dynamics
-but with different levels of noise.  For instance, in a logistic
+but with *different* levels of noise.  For instance, in a logistic
 model with $b = rn$ and $d = rn^2/N$, we instead find the variance
@@
 
-Whereas demographic noise reflects processes at a lower scale (the
-individual, rather than the population), environmental stochasticity
-reflects processes whose dynamics we are not explicitly modelling.
-Once again, this can best be illustrated through an explicit
-formulation of the master equation.  For a generic population model
-that depends not only on population state $x$ but also on environmental
-factors $y$, it is possible to derive the steady-state variance
-(Appendix B) as:
-
-$$
-\begin{equation}
-\sigma_x^2  = \frac{ \left( \partial_y f(x,y) \right)^2}{\left( \partial_x f(x,y) \right)^2 + \partial_x f(x,y) \tau_c } \sigma_e^2 + \sigma^2_d 
-\label{env_noise}
-\end{equation}
-$$
-
-where $\sigma_e^2$ is the variance in the environmental noise and
-$\tau_c$ the auto-correlation of that noise, $\sigma_d^2$ the
-equilibrium variance contributed by demographic noise from before,
-($\tfrac{b + d}{-2\partial_x (b - d)}$) and where we have defined
-$f(x,y) = b(x,y) - d(x,y)$ for notational convenience.  For example,
-if extinction rate $e$ in our Levins' model is allowed to vary
+The van Kampen expansion can also be useful in illustrating how 
+environmental noise enters a model as a separate process or dimension.
+Like the expansion of demographic noise, this derivation leads to 
+environmental noise that is approximately Gaussian in distribution,
+but unlike the canonical model, leads to explicitly autocorrelated rather 
+than white noise (see derivation, Appendix B).  This result is intuitive:
+for continuous time models, all fluctations are positively auto-correlated
+by definition -- measure any continuous process at ever-closer intervals in
+time and these measurements must converge. As the derivation makes it clear,
+it is in fact the *relative* time scales of the environmental process
+and population level process that really matter.  The degree of coupling
+to environmental noise also depends on the derivative of the population 
+dynamics with respect to the stochastically varying environmental variable:
+the larger that derivative, the stronger the contribution of environmental
+fluctuations to population fluctuations. For example, if extinction
+rate $e$ in our Levins' model is allowed to vary
 stochastically as Gaussian white noise with variance $\sigma_e^2$
@@
 independent of system size and directly proportional to the variance
-in the environment.  This derivation provides a useful illustration
-of the difference between demographic noise and environmental noise:
-the latter simply describes the variation introduced by some other
-dynamic variable that we are not modelling explicitly, and whose
-dynamics are independent of the state of our system, $x$. This
+in the environment.
+
+This
 environmental variation no doubt also arises from some more detailed
@@
 
-The importance of $\tau_c$, or the degree of auto-correlation, which
-emerges from our derivation above, has also been and continues to be a subject of
+The importance of auto-correlation, has and continues to be a subject of
 significant interest in the study of stochastic population dynamics
@@
 Early work often assumed environmental noise would be
-uncorrelated as a mathematical convenience.  As the derivation of
-Eq \eqref{env_noise} makes clear, for a continuous-time process,
-the relevant metric is the timescale of auto-correlation in the
-environmental noise process relative to the auto-correlation timescale
-of the population dynamics.  The Gaussian noise that emerges out
-of the system-size approximation is always auto-correlated to some
-degree -- if we measure our environment at consecutive times
-infinitesimally far apart, we expect nearly identical values.  At
-intervals much longer than the correlation time, consecutive
-measurements are effectively independent (i.e. white noise). 
-The degree of auto-correlation in  noise is all about _relative_ timescales.
-
-## Other forms of stochasticity
+uncorrelated as a mathematical convenience.
+
+<!-- This can be condensed or dropped to save space -->
+
+## Origins of other forms of stochasticity
 

### 2018-02-27 paper/paper.Rmd
refs

@@
 can include transitions between stage classes or trait values, including
-mutations to continuously-valued traits in evolutionary dynamics [e.g. @Boettiger2009].
+mutations to continuously-valued traits in evolutionary dynamics [e.g. @Boettiger2010].
 
@@
 which is much more typical in the ecological literature [e.g. the books of @Nisbet1982 and @Lande2003
-or the review in @Ovaskeinen2010]. @Black2012 gives an excellent introduction 
+or the review in @Ovaskainen2010]. @Black2012 gives an excellent introduction 
 to the van Kampen expansion as a connection between individual-based and population-level
@@
 (equivalently, an SDE or PDE with linear noise term, see Appendix B), but rather
-a *non-linear* SDE of the form (e.g. @Ovaskeinen2010)
+a *non-linear* SDE of the form (e.g. @Ovaskainen2010)
 
@@
 a linear SDE for deviations).  Third, while both approximations break down
-in the case of large deviations [see @Ovaskeinen2010], the van Kampen expansion
+in the case of large deviations [see @Ovaskainen2010], the van Kampen expansion
 shows that the macroscopic equation will continue to hold even when the 
@@
 The loss of a species is without a doubt one of the most important
-possible consequences of noise, and unsurprisingly few topics have
+possible consequences of noise. Unsurprisingly few topics have
 recieved greater attention then the question of the persistence of
@@
 away from the well-trodden ground of persistence and coexistence to focus
-on other phenomena that
-
-here we will focus on a few examples of stochastic phenomena 
-that can emerge on shorter time scales and provide some intuition for 
-how the interaction between stochastic an nonlinear dynamics can 
-create these surprising results.
+on other phenomena that can emerge in stochastic models.  This will
+highlight the importance of noisy phenomena beyond the ability to obscure
+a deterministic skeleton of some underlying process or in driving extinctions,
+but rather in shaping and driving other interesting patterns we observe in
+natural systems. It is these phenomena that will also underpin the third
+transition in which noise can be turned to our advantage as a source
+of information about ecological processes.
 
@@
 
-**Bet hedging**, Lewontin & Cohen 1969; Levins 1969;
+**Bet hedging**, @Lewontin1969 & Cohen 1969; Levins 1969;
 

### 2018-02-27 paper/paper.Rmd
refs

@@
 $N$, illustrating the magnitude of these fluctuations indeed scales in proportion
-to $\sqrt{N}$ as postulated by the canonical equation, Eq \eqref{canoical}.
+to $\sqrt{N}$ as postulated by the canonical equation, Eq \eqref{canonical}.
 
@@
 (i.e. averaged over an esemble of replicate realizations of the Gillespie simulations),
-and with remarkable accuracy, up to order $1/N}$, i.e.
+and with remarkable accuracy, up to order $1/N$, i.e.
 a single individual.^[We can actually extend the van Kampen expansion to higher order and successfully approximate even this higher-order correction to the mean dynamics, recovering the phenomenon of stochastic inflation [@Gonzalez2002], where the average population size can be larger or smaller than predicted by the deterministic model depending on the curvature of the macroscopic equation].
@@
 and deaths add randomness, they both add to the growth of varation observed, i.e.
-$b(x) + d(x)$.  Second, we see that dynamics for the variance $\sigma_^2$ are *non-linear* whenever
+$b(x) + d(x)$.  Second, we see that dynamics for the variance $\sigma^2$ are *non-linear* whenever
 $b(x)$ or $d(x)$ are non-linear, so that different values of the population density
@@
 prove more difficult to work with than the convenient ODE expressions from van Kampen,
-Eq \eqref{\vanKampen} (which can be expressed as a coupled ODE for the mean and 
+Eq \eqref{vanKampen} (which can be expressed as a coupled ODE for the mean and 
 a linear SDE for deviations).  Third, while both approximations break down

### 2018-02-27 appendixB/appendixB.Rmd
environmental noise

@@
 
-preamble: "\\newcommand{\\ud}{\\mathrm{d}}"
+preamble: "\\newcommand{\ud}{\\mathrm{d}}"
 ---
@@
 \begin{align}
-\frac{\ud p_n}{\ud t} = (\mathbb E^{-1} - 1) b_n p_n + (\mathbb E -1) d_n p_n \label{master1}
+\frac{\mathrm{d} p_n}{\mathrm{d} t} = (\mathbb E^{-1} - 1) b_n p_n + (\mathbb E -1) d_n p_n \label{master1}
 \end{align}
@@
 \begin{align}
-\frac{\ud }{\ud t} \langle n \rangle &= \sum n (\mathbb E^{-1} - 1) b_n p_n + \sum n (\mathbb E -1) d_n p_n \nonumber \\ 
+\frac{\mathrm{d} }{\mathrm{d} t} \langle n \rangle &= \sum n (\mathbb E^{-1} - 1) b_n p_n + \sum n (\mathbb E -1) d_n p_n \nonumber \\ 
 &= \sum  b_n p_n(\mathbb E - 1)n  + \sum d_n p_n(\mathbb E^{-1} -1) n \nonumber \\
@@
 \begin{align}
-\frac{\ud }{\ud t} \langle n^2 \rangle &= 2 \langle n (b_n - d_n) \rangle + \langle b_n+d_n \rangle \label{var}
+\frac{\mathrm{d} }{\mathrm{d} t} \langle n^2 \rangle &= 2 \langle n (b_n - d_n) \rangle + \langle b_n+d_n \rangle \label{var}
 \end{align}
@@
 \begin{equation}
-\ud X_t = \left[ b(x) - d(x) \right] \ud t + \sqrt{b(X_t) + d(X_t)} \ud W_t
+\mathrm{d} X_t = \left[ b(x) - d(x) \right] \mathrm{d} t + \sqrt{b(X_t) + d(X_t)} \mathrm{d} W_t
 \end{equation}
@@
 \begin{align}
-\frac{\ud n}{\ud t} &= \Omega \frac{\ud \phi(t)}{\ud t}  + \Omega^{1/2}\frac{\ud \xi}{\ud t} = 0, \nonumber \\
+\frac{\mathrm{d} n}{\mathrm{d} t} &= \Omega \frac{\mathrm{d} \phi(t)}{\mathrm{d} t}  + \Omega^{1/2}\frac{\mathrm{d} \xi}{\mathrm{d} t} = 0, \nonumber \\
 \intertext{therefore}
-\frac{\ud \xi}{\ud t} &= -\Omega^{1/2}\frac{\ud \phi(t)}{\ud t}. \label{dxidt}
+\frac{\mathrm{d} \xi}{\mathrm{d} t} &= -\Omega^{1/2}\frac{\mathrm{d} \phi(t)}{\mathrm{d} t}. \label{dxidt}
 \end{align}
@@
 \begin{align*}
-\frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  + \frac{\partial \Pi}{\partial \xi}\frac{\ud \xi}{\ud t} 
+\frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  + \frac{\partial \Pi}{\partial \xi}\frac{\mathrm{d} \xi}{\mathrm{d} t} 
 \end{align*}
@@
 \begin{align}
-\frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  - \Omega^{1/2}\frac{\ud \phi}{\ud t}\frac{\partial \Pi}{\partial \xi} \label{Pi}
+\frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  - \Omega^{1/2}\frac{\mathrm{d} \phi}{\mathrm{d} t}\frac{\partial \Pi}{\partial \xi} \label{Pi}
 \end{align}
@@
 \begin{align}
-&\frac{\partial \Pi(\xi, t)}{\partial t}  - \Omega^{1/2}\frac{\ud \phi}{\ud t}\frac{\partial \Pi}{\partial \xi}  = \nonumber \\
+&\frac{\partial \Pi(\xi, t)}{\partial t}  - \Omega^{1/2}\frac{\mathrm{d} \phi}{\mathrm{d} t}\frac{\partial \Pi}{\partial \xi}  = \nonumber \\
 &\quad \left( -\Omega^{-1/2} \frac{\partial}{\partial \xi} +  \frac{\Omega^{-1}}{2} \frac{\partial^2}{\partial \xi^2} + \ldots\right)b(\phi\Omega+\xi\Omega^{1/2})\Pi(\xi, t)\nonumber \\
@@
 \begin{align}
-\frac{\ud \phi(t)}{\ud t} = b(\phi)-d(\phi) = a_1(\phi) \label{macroscopic}
+\frac{\mathrm{d} \phi(t)}{\mathrm{d} t} = b(\phi)-d(\phi) = a_1(\phi) \label{macroscopic}
 \end{align}
@@
 
-This expression has the wonderful properties of being both simple and general.
+This expression has the wonderful properties of being both simple and quite general.
 
@@
 \begin{align}
-\frac{\ud n}{\ud t} = c n \left(1 - \frac{n}{N}\right) - e n \label{levins}
+\frac{\mathrm{d} n}{\mathrm{d} t} = c n \left(1 - \frac{n}{N}\right) - e n \label{levins}
 \end{align}
@@
 
-$$\frac{\ud n}{\ud t} = r n \left( 1 - \frac{n}{K} \right)$$
+$$\frac{\mathrm{d} n}{\mathrm{d} t} = r n \left( 1 - \frac{n}{K} \right)$$
 though the partition into birth and death rates is not explicit.
@@
 
-We introduce environmental noise into the master equation by allowing the transition rates to depend explicitly on the macroscopic variable $\psi$ as well as the state variable $\phi$.
-It can be helpful to think of fluctuations in $\psi$ as also arising from some lower level transitions, but we need not model these explicitly.
-Following the expansion as before (see @vanKampen2007 for details), the mean dynamics are given by the macroscopic equations:
+Environmental noise arises not as a summary of lower-level processes, like in the case of intrinsic noise, but rather as a recognition that things we treat as model parameters are in fact dynamic variables themselves and subject to their own dynamics which we summarize statistically rather than model explicitly.
+We can recognize this as simply a special case of a multivariate system size expansion; e.g.
+where a parameter of our birth or death process varies according to some environmental dynamics (e.g.
+a death rate $e$ linked to fluctuations in temperature, $y$.)  In general, we can write this as a macroscopic equation in two variables:
 
+$$\frac{\mathrm{d} x}{\mathrm{d} t} = a_1(x, y) $$
+
+In general, we could imagine a specific equation for the dynamics of the environmental process $y$, perhaps also subject to a system size expansion.
+To keep things simple, we will adopt a linear OU process for the noise,
+
+$$
 \begin{align*}
-\frac{\ud \phi}{\ud t} &= a_1(\phi, \psi) \\
-\frac{\ud \psi}{\ud t} &= \beta_1(\psi) 
+\mathrm{d} y &= - \beta y \phantom \cdot \mathrm{d} t + \sigma_e \mathrm{d}W
 \end{align*}
-
-while the fluctuation dynamics are now given by three coupled differential equations:
-
-\begin{align}
-\frac{\ud \langle \xi^2 \rangle}{\ud t} &= 2 \frac{\partial a_1}{\partial \phi} \langle \xi^2 \rangle + 2 \frac{\partial a_1 }{\partial \psi} \langle \xi \eta \rangle + a_2 \label{phivar} \\
-\frac{\ud \langle \xi \eta \rangle}{\ud t} &= \left( \frac{\partial a_1}{\partial \phi} + \frac{\partial \beta_1 }{\partial \psi} \right)\langle \xi \eta \rangle + \frac{\partial a_1 }{\partial \psi} \langle \eta^2 \rangle \label{covar}\\
-\frac{\ud \langle \eta^2 \rangle}{\ud t} &= 2 \frac{\partial \beta_1}{\partial \psi} \langle \eta^2 \rangle + \beta_2 \label{psivar}
-\end{align}
-
-Without an explicit model for the environmental variation process, we can merely summarize its role in terms of the autocorrelation, $\tau_c := \tfrac{\partial \beta_1}{\partial \psi}$ and its overall variance at steady-state, $\langle \eta^2 \rangle = \tfrac{\beta_2}{-2 \partial_{\psi} \beta{1,0}}$, which we will refer to as the environmental variation $\sigma^2_e := \langle \eta^2 \rangle$.
-A more detailed mechanistic description of how the environmental noise arises would permit a time-dependent solution, but the system size expansion tells us that these two summary statistics are all we need to express the steady-state solution above.
-
-Setting the remaining equations to zero to solve for the stationary state we can write down a general formula for the variation in $\phi$:
-
-\begin{equation*}
-\langle xi^2 \rangle = \frac{ \left( \partial_{\psi} a_1(\phi,\psi) \right)^2}{\left( \partial_{\phi} a_1(\phi,\psi) \right)^2 + \partial_{\phi} a_1 (x,y) \tau_c } \sigma_e^2 + \sigma^2_d
-\end{equation*}
-
-<!-- FIXME -->
+$$
+which corresponds to mean-zero Gaussian noise with auto-correlation $\beta$, where the white-noise limit corresponds to $\beta \to 0$.
+Following the expansion as before (see @vanKampen2007 for details of the multi-variate case), the mean dynamics for the population size are once again given by the macroscopic equation, $\frac{\mathrm{d} x}{\mathrm{d} t} = a_1(x, y) $, while the fluctuation dynamics are now given by three coupled differential equations:
 
@@
 \begin{align}
-\frac{\mathrm{d} x}{\mathrm{d} t} &= \alpha_1(x, y)  \\
-\frac{\mathrm{d} y}{\mathrm{d} t} &= \beta_1(y) 
+\frac{\mathrm{d} \langle \xi^2 \rangle}{\mathrm{d} t} &= 2 \frac{\partial a_1}{\partial x} \langle \xi^2 \rangle + 2 \frac{\partial a_1 }{\partial y} \langle \xi \eta \rangle + a_2 \\
+\frac{\mathrm{d} \langle \xi \eta \rangle}{\mathrm{d} t} &= \left( \frac{\partial a_1}{\partial x} - \beta \right)\langle \xi \eta \rangle + \frac{\partial a_1 }{\partial y} \langle \eta^2 \rangle \\
+\frac{\mathrm{d} \langle \eta^2 \rangle}{\mathrm{d} t} &= -2 \beta \langle \eta^2 \rangle + \gamma_e^2
 \end{align}
+
 $$
 
-while the fluctuation dynamics are now given by three coupled differential equations:
-
-For a generic birth death process that depends both on
-on population state $x$ but also on environmental
-factors $y$, where we write $a_1(x,y) := b(x,y) - d(x,y)$,
-sum $a_2(x,y) := b(x,y) + d(x,y)$
-
-For a generic population model
-that depends not only on population state $x$ but also on environmental
-factors $y$, it is possible to derive the steady-state variance
-(Appendix B) as:
+Note that the fluctuations in population density $x$ depends as before on the fluctuation-disspation tradeoff, $2 \frac{\partial a_1}{\partial x} \langle \xi^2 \rangle + a_2$, but include the additonal term arising from a covariance with the environmental flucatuations.
+We can simplify this expression as before by considering the equilibrium noise dynamics
 
 $$
-\begin{align}
-&\sigma_x^2  = \frac{  \partial_y a_1(x,y) ^2}{ \partial_x a_1(x,y) ^2 +  \tau_c \phantom\cdot \partial_x a_1(x,y)  } \sigma_e^2 + \frac{1}{2} \frac{a_2(x,y)}{\partial_x a_1(x,y)}
-\end{align}
+\langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} + \underbrace{\frac{\left[\partial_{y} a_1\right] ^2}{\partial_x a_1 - \beta} \frac{\langle\eta^2\rangle}{2\partial_x a_1}}_{\textrm{environmental contribution}}
 $$
 
-where $\sigma_e^2$ is the variance in the environmental noise and
-$\tau_c$ the auto-correlation of that noise.
+Here we have the familiar demographic noise term, coupled to an environmental noise contribution which depends on the details of how the environmental noise enters the macroscopic model, $\partial_y a_1$, as well as the relative correlation times $\partial_x a_1$ and $\beta$ of the population process and environmental process.
+Note that we have left this expression in terms of $\langle \eta^2 \rangle$, rather than subsituting in the equilibrium expression
+
+$$\langle \eta^2 \rangle = \frac{\gamma_e^2}{2\beta},$$
+
+so that we can consider separately the limit of white environmental noise without also increasing the variance of the process.
+Changing the correlation while holding the total variation $\langle \eta^2\rangle$ constant thus amounds to increasing the intrinsic noise level $\gamma^2$ proportionally.
+In general, a more mechanistic description of what it means to change the autocorrelation of an environmental process for a given variance is needed.
+Nevertheless, considering the the white noise limit with some fixed variance, when $\beta \ll \partial_x \alpha_1$,  this expression simplifies as:
+
+$$
+\langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} +  \underbrace{\frac{1}{2} \left[\frac{\partial_{y} a_1}{ \partial_x a_1}\right] ^2}_{\textrm{environmental coupling}} \langle\eta^2\rangle
+$$
+Taking extinction as such a Gaussian white noise process with meant $\bar e$ and variance $\sigma_e^2$, we can plug in our birth-death equations for the Levins model (and rescaling by system size, from $x = n/N$ back to $n = xN$), into the above equation to find the expected fluctuations are:
+
+$$\sigma_n^2 = \frac{\bar e}{c}N + \frac{\left[(1-\frac{\bar{e}}{c}) N\right]^2}{(\bar e-c)^2} \sigma_e^2$$
+
+where we recognize the term $(1-\frac{\bar{e}}{c}) N$ as the equilibrium population size, $\bar n$.
+Meanwhile, if timescale of environmental fluctuations is much longer than population fluctuations (a red noise limit, $\beta \gg \partial_x \alpha_1$) this can instead be written as
+
+$$
+\langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} + \underbrace{- \frac{\left[\partial_{y} a_1\right] ^2}{2\partial_x a_1 \beta}}_{\textrm{environmental coupling}}  \langle\eta^2\rangle
+$$
+(Recall that we have taken these expressions at equilibrium, and at equilibrium the derivative of the macroscopic equation $\partial_x a_1 < 0$ by definition, which guarentees the variances above are strictly positive).
+In this limit ($\beta \gg \partial_x \alpha_1$), the overall noise decreases both with stronger dampening / higher autocorelation in the environmental process, (larger $\beta$) just as it does with increased dampening in population dynamics $\partial_x a_1$.
+
+<!--
+## Crowley model
+
+Consider the Crowley model \cite{Crowley05} of competition for space:
+
+\begin{align}
+\dot x &= b_1 x (1-x-y) - d_1 x + cxy \\
+\dot y &= b_2 y (1-x-y) - d_2 y - cxy
+\end{align}
+
+Where $x$ and $y$ are the respective densities of two species competing for space.
+In the model species $x$ is the better competitor, displacing species $y$ from patches at rate $cyx$, while species $y$ is the better colonizer when $b_2 > b_1$, taking $d_1 = d_2$ for simplicity.
+Under this competition-colinization trade-off, this the deterministic skeleton supports stable coexistence of both species.
+We intepret this as an individaul model analgous to the Levins model, with a finite number of patches $N$ that both species compete to occupy.
+Taking $N$ as our system size ($\Omega$), we have $n_x = \Omega x + \Omega^{1/2} \xi$ and $n_y = \Omega y + \Omega^{1/2} \eta$, we proceed much as before to derive the fluctuation equations:
+
+$$
+\begin{align*}
+\partial_t \sigma_{\xi}^2 &= 2( b_1(1-2x-y) + c y-d_1 )\sigma_{\xi}^2 \\ 
+& \quad + 2( -b_1x+cx )\langle x y \rangle		 \\
+& \quad + b_1 x(1-x-y) + c x y+d_2 x)\\
+\sigma_{\eta}^2 &= 2(b_2(1-2y-x) - c x-d_2)\sigma_{\eta}^2							\ \\
+& \quad + 2(-b_2 y-c y)\langle \xi \eta \rangle								\\
+& \quad + b_2 y(1-x-y) + c x y+d_2 y)\\
+\langle \xi \eta \rangle &= b_1(1-2x-y)+c y-d_1)\langle \xi \eta \rangle			 \\
+& \quad + (-b_1 x+c x) \sigma^2_{\eta}										 \\ 
+& \quad + (-b_2 y - c y)\sigma^2_{\xi}										 \\ 
+& \quad + (b_2(1-2 y-x) - c x - d_2)\langle \xi \eta \rangle - c x y;
+\end{align*}
+$$
+
+These three coupled equations can be solved by numerical integration, which we compare to individual-based simulation of the Crowley competition model, Fig~\ref{crowley}.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=.5\textwidth]{crowley_fluc.png}
+\caption{Fluctuations for species $x$ in black and $y$ in red.
+Smooth curves are theoretical predictions of variance within each population.
+Model parameters: $b_1 = 0.2$, $b_2 = 0.6$, $d_1 = d_2 = c = 0.1$, $N=1000$, plots based on exact Gillespie simulations averaged over an ensemble of $10^2$ realizations.}\label{crowley}
+\end{center}
+\end{figure}
+
+Mimicking the structure of the Levins model only in multiple dimensions, this model can also provide a simple mechanistic illustration of the impact of "environmental," e.g.
+exogenous, noise as well as demographic noise on a system.
+We can focus on the dynamics of a single species, say, $y$, treating the role of species $x$ as some external disturbance that can make open patches temporarily uninhabitable and displace inhabited patches.
+-->
 

### 2018-02-27 paper/paper.Rmd
environmental noise

@@
 stochastically as Gaussian white noise with variance $\sigma_e^2$
-and mean $e$, we find the expected variance introduced in the
+and mean $\bar e$, we find the expected variance introduced in the
 population to be:
 
-$$\sigma_n^2 = \frac{e^2}{(c+e)^2} \sigma_e^2 + \frac{e}{c}N $$
+$$\sigma_n^2 = \frac{\bar e}{c}N + \frac{\left[(1-\frac{\bar{e}}{c}) N\right]^2}{(\bar e - c)^2} \sigma_e^2$$
 
-Note that the contribution of environmental stochasticity is
-independent of system size and directly proportional to the variance
-in the environment.
+where $\left[(1-\frac{\bar{e}}{c}) N$ is the equilibrium population size.
+This reflects the assumption of the canonical equation, Eq \eqref{canonical}
+that the standard deviation scales in direct proportion to the system size $N$,
+while the demographic noise scales as the square root.  Consequently,
+for a large system, the demographic contribution becomes proportionally smaller
+while the environmental contribution remains the same.
 

### 2018-02-27 appendixB/appendixB.Rmd
WIP

@@
 A derivation of diffusion approximation is originally credited to @Kramers1940 and later improved by @Moyal1949, though Einstein makes implicit use of this approximation in his own work on Brownian Motion decades earlier [@Gardiner2009].
-The diffusion approximation has been a popular basis for ecological stochastic models as well, such as in @Nisbet1982's classic textbook on fluctuating populations, or more recently in @Lande2003 and work reviewed in @Ovaskeinen.
+The diffusion approximation has been a popular basis for ecological stochastic models as well, such as in @Nisbet1982's classic textbook on fluctuating populations, or more recently in @Lande2003 and work reviewed in @Ovaskainen2010.
 This diffusion approximation is nicely summarized @Gardiner2009 and in @vanKampen2007, though I believe @Kurtz1978 provides a more formal proof that goes unmentioned by any of these references.

### 2018-02-27 paper/paper.Rmd
WIP

@@
 
+The gap between theory-focused specialists and more empirically minded non-specialists
+is particularly acute in the ecological literature on stochastic processes, making
+a review of
+
 As the literature on stochastic modeling in ecology is both extensive and often
@@
 will focus on the simplest examples necessary to illustrate this broader trend
-in the changing way we think about the role of noise and the opportunity
-that shift can present to better bridge theoretical and empirical communities.
+in the changing way we think about the role of noise.
 To that end, I will focus on examples from simple population models, echoing
@@
 [e.g. @Ewens2004]. In an effort to make these themes accessible to more emprical
-readers, I will also focus on simplest examples necessary, with only passing
+readers, I will rely on the simplest examples necessary, with only passing
 mention of the rich literature extending these concepts to more realistic
@@
 
-where $\left[(1-\frac{\bar{e}}{c}) N$ is the equilibrium population size.
+where $(1-\frac{\bar{e}}{c}) N$ is the equilibrium population size.
 This reflects the assumption of the canonical equation, Eq \eqref{canonical}

### 2018-02-27 appendixB/appendixB.Rmd
WIP

@@
 
-Here we have the familiar demographic noise term, coupled to an environmental noise contribution which depends on the details of how the environmental noise enters the macroscopic model, $\partial_y a_1$, as well as the relative correlation times $\partial_x a_1$ and $\beta$ of the population process and environmental process.
+note that this process provides a justification for moving the noise from 'inside' the deterministic skeleton to merely an additive Gaussian random variable on the end of the determinstic process.
+The details of how the noise term enter the dynamics are reflected in the dervative of the population equation with respect to the environmental variable, $\partial_y a_1(x,y)$, as well as the relative correlation times $\partial_x a_1$ and $\beta$ of the population process and environmental process.
+This environmental contribution adds on to the variance contributed by the demographic noise.
+Both demographic and environmental noise contributions are 'damped' in proportion to the stability of the macroscopic equation, $\partial_x a_1$.
+
 Note that we have left this expression in terms of $\langle \eta^2 \rangle$, rather than subsituting in the equilibrium expression

### 2018-02-27 paper/paper.Rmd
WIP

@@
 
-The van Kampen expansion can also be useful in illustrating how 
-environmental noise enters a model as a separate process or dimension.
+The van Kampen expansion can also be useful in illustrating how extrinsic
+factors such as environmental noise enter into a stochastic model.
 Like the expansion of demographic noise, this derivation leads to 
 environmental noise that is approximately Gaussian in distribution,
-but unlike the canonical model, leads to explicitly autocorrelated rather 
-than white noise (see derivation, Appendix B).  This result is intuitive:
-for continuous time models, all fluctations are positively auto-correlated
-by definition -- measure any continuous process at ever-closer intervals in
-time and these measurements must converge. As the derivation makes it clear,
-it is in fact the *relative* time scales of the environmental process
-and population level process that really matter.  The degree of coupling
-to environmental noise also depends on the derivative of the population 
-dynamics with respect to the stochastically varying environmental variable:
-the larger that derivative, the stronger the contribution of environmental
-fluctuations to population fluctuations. For example, if extinction
-rate $e$ in our Levins' model is allowed to vary
-stochastically as Gaussian white noise with variance $\sigma_e^2$
-and mean $\bar e$, we find the expected variance introduced in the
-population to be:
+but unlike the canonical model, leads to explicitly autocorrelated noise 
+(see derivation, Appendix B).  Auto-correlated or "colored" environmental
+noise has been a topic of significant theoretical interest 
+[e.g. @Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015], but such 
+formulations start from the empirical observation of autocorrelation 
+in environmental time series rather than a process-based derivation.
+As the expansion clearly shows, any bounded continuous-time process will
+involve auto-correlation -- measure a continuous process at ever-closer intervals in
+time and these measurements must converge. As the derivation in Appendix B makes clear,
+it is rather the the *relative* time scales of the environmental process
+and population level process that really matter (more precisely, the relative 
+magnitudes of the autocorrelation time of the environmental dynamics and population dynamics.)
+An environment with comparatively short autocorrelation scale will act
+effectively as white noise upon the population dynamics.
+
+The van Kampen expansion for extrinsic factors (appendix B) also highlights
+the importance of how environmental noise couples to the population dynamics.
+A mechanistic description of environmental noise acts directly on parameters
+of the ecological model: for instance, in our Levins' model we may imagine
+that the extinction rate of patches does not occur at some fixed rate $e$
+but rather at some stochastically fluctuating rate of a given mean, variance,
+and autocorellation.  The system size expansion illustrates how we replace
+a model where stochasticity is nested inside some parameter into the simpler
+format of the canonical equation where this contribution can be merely added
+on as Gaussian noise at the end of a deterministic skeleton.  (The expansion
+derives that the environmental noise is proportional to the square of the derivative
+of the deterministic skeleton with respect to the environmental variable,
+see appendix B.) Thus for the Levin's model, we can capture the fluctuations
+introduced by demographic and environmental noise at equilibrium as
+a Gaussian distribution with variance given by the sum of of demographic
+and environmental contributions:
 
@@
 This reflects the assumption of the canonical equation, Eq \eqref{canonical}
-that the standard deviation scales in direct proportion to the system size $N$,
+that the standard deviation (e.g. square root of the variance) scales in direct proportion to the system size $N$,
 while the demographic noise scales as the square root.  Consequently,
@@
 
-This
-environmental variation no doubt also arises from some more detailed
-lower level process, but we need not be explicit about those dynamics
--- the expansion tells us that it is sufficient to know the overall
-variance and the degree of auto-correlation.  Recent work has also
-highlighted the importance of including both demographic and
-environmental noise [@Melbourne2008; @Bonsall2004].  The system
-size expansion still requires the assumption that the noise is small
-relative to the system size.  In the case of large fluctuations,
-this approximation must be replaced with a different approach, such
-as the WKB approximation, as discussed in @Ovaskainen2010.
-
-The importance of auto-correlation, has and continues to be a subject of
-significant interest in the study of stochastic population dynamics
-[e.g.@Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
-Early work often assumed environmental noise would be
-uncorrelated as a mathematical convenience.
+These examples of how a stochastic model arises from both intrinsic and extrinsic
+sources of noise through the use of the system size expansion serve both
+as critique and justification of the familiar notion of a stochastic model as 
+a deterministic skeleton plus some Gaussian white noise term.  Critique, because
+the expansions show that both "noise" and "deterministic skeleton" are the 
+result of lower-level mechanistic processes, whose precise form will 
+depend on the details of those processes.  Yet the expansion also shows us that
+this Gaussian noise description can be a surprisingly good approximation,
+that noise intensity will converge to a stationary value whenever the system
+dynamics have a stable point, and that the contribution of extrinsic
+(i.e. environmental) noise factors will dominate in large systems.  These
+justifications will allow us to pursue our analysis of noise-driven phenomena
+using simple models in which a nonlinear deterministic skeleton is driven by
+Gaussian environmental noise.
 
@@
 
-<!--  -->
+<!-- Justify and contrast models to individual noise -->
 
@@
 # Noise the informer: Noise can provide novel information
+
+<!-- More frank introduction to this section, see reply to Reviewer 1-->
 

### 2018-02-28 paper/paper.Rmd
refs, sp

@@
 
-To the empricical ecologist, stochasticity, or more simply, "noise,"
+To the empirical ecologist, stochasticity, or more simply, "noise,"
 is just that -- something which obscures patterns we are trying to infer [@Knape2011];
-and an ever richer batteries of stastical methods are developed largely in an
+and an ever richer batteries of statistical methods are developed largely in an
 attempt to strip away this undesirable randomness to reveal the patterns
 beneath [@Coulson2001].  Over the past several decades, theoretical ecology has transitioned
-from thinking of stochasticity in such terms; where noise is a nuciance
+from thinking of stochasticity in such terms; where noise is a nuisance
 that obscures the "deterministic skeleton" of the underlying mechanisms,
 to the recognition that stochasticity can itself be a mechanism for 
-driving many interesting phenomena [@Coulson2004].  Yet this transistion from "noise the nuisance"
+driving many interesting phenomena [@Coulson2004].  Yet this transition from "noise the nuisance"
 to "noise the creator" of ecological phenomena has had, with a few notable exceptions,
-relatively little impact in empricial thinking about stochasticity.  One of the
+relatively little impact in empirical thinking about stochasticity.  One of the
 most provocative of those exceptions has turned the classical notion of noise
@@
 this third shift in perspective offers an opportunity to better bridge the divide
-between respective primarily theoretical and primarily empricial communities by 
-seeing noise not as mathematical curiousity or statistical bugbear, but as a
+between respective primarily theoretical and primarily empirical communities by 
+seeing noise not as mathematical curiosity or statistical bugbear, but as a
 source for new opportunities for inference.
@@
 a review of 
+<!-- FIXME -->
 
@@
 stochastic models in evolution [@Lenormand2009] and population genetics
-[e.g. @Ewens2004]. In an effort to make these themes accessible to more emprical
+[e.g. @Ewens2004; @Gillespie2004]. In an effort to make these themes accessible to more empirical
 readers, I will rely on the simplest examples necessary, with only passing
@@
 
-In this spririt, I provide concise, detailed and commented
+In this spirit, I provide concise, detailed and commented
 code for simulating each of the models presented in Appendix A.
@@
 "skeleton" -- responsible for the phenomena of interest -- that is obscured
-from view by the nusiance of noise has slowly given way to recognition
+from view by the nuisance of noise has slowly given way to recognition
 that stochastic and deterministic elements not only interact in 
@@
 *intrinsic* factors: where stochasticity is introduced to reflect
-the reality that our variables of interest are intriniscally 
+the reality that our variables of interest are intrinsically 
 statistical averages of lower-level processes, and *extrinsic*
@@
 individuals [@Melbourne2012].  **Environmental stochasticity** is the primary example 
-of an extinsic factor, reflecting the fact that parameters we may
+of an extrinsic factor, reflecting the fact that parameters we may
 treat as constants in a model in fact depend on variables such as 
@@
 not to explicitly model -- an individual death occurs at rate $\lambda$.  
-This is in fact an extrinisic factor (variables not modeled explicitly),
+This is in fact an extrinsic factor (variables not modeled explicitly),
 which makes birth appear probabilistic; (just as we describe the toss of a coin
@@
 @Ovaskainen2010], and provides a clear illustration of the partition of separate deterministic and stochastic elements.
-Significant research continues to focus on extending the study of this model to consider more complex and higher-dimensional $f(n)$ (coexistence models, stage structured models) and consider the case of autocorrelated, rather than than white noise of $\textrm{d}B_t^{(e)}$.
+Significant research continues to focus on extending the study of this model to consider more complex and higher-dimensional $f(n)$ (coexistence models, stage structured models) and consider the case of auto-correlated, rather than than white noise of $\textrm{d}B_t^{(e)}$.
 
@@
 for simulating demographic stochasticity at an individual level.  
-The algorithm is merely a simple and direct implemtentation of the master
-equation, progressing in random step sizes deterimed
+The algorithm is merely a simple and direct implementation of the master
+equation, progressing in random step sizes determined
 by the waiting time until the next event.
@@
 [^2]: that is, free from the approximation made by SDE models as we see in the
-van Kampen example.  All models are, of course, only approixmations.
+van Kampen example.  All models are, of course, only approximations.
 
@@
 processes, it will be helpful to make the notion of a master equation concrete
-with a spefiic example.  We will focus on the classic case of @Levins1969 patch model,
+with a specific example.  We will focus on the classic case of @Levins1969 patch model,
 to illustrate the Gillespie algorithm and the van Kampen system size expansion
@@
 birth and death rate.  As we shall see, this is not merely a notational
-convience but leads to real differences in observed stochastic properties
+convenience but leads to real differences in observed stochastic properties
 of the model. Both the numerical approach (details in Appendix A) and the 
 analytical approach (Appendix B) illustrate that it is straight-forward
-to use this same approach in alternate formuluations.
+to use this same approach in alternate formulations.
 
@@
 van Kampen's approximation follows a central-limit theorem style argument established rigorously in
-the theorems of Kurtz at about the same time [@Kurtz1970 & @Kurtz1971, @Kurtz1972, in particular, see Thereom 4.4. @Kurtz1978].
+the theorems of Kurtz at about the same time [@Kurtz1970 & @Kurtz1971, @Kurtz1972, in particular, see Theorem 4.4. @Kurtz1978].
 The heart of the approximation is to imagine a population on some large landscape of which
@@
 system size, which as it expands includes more and more total individuals and
-experiences proportionally smaller fluctionations in population density.  This 
+experiences proportionally smaller fluctuations in population density.  This 
 results in the conclusion that the population density obeys a simple Gaussian
@@
 population count $n$ describes the *expected* density of the population
-(i.e. averaged over an esemble of replicate realizations of the Gillespie simulations),
+(i.e. averaged over an ensemble of replicate realizations of the Gillespie simulations),
 and with remarkable accuracy, up to order $1/N$, i.e.
@@
 illustrating the origins of this noise.  Because both individual events of births
-and deaths add randomness, they both add to the growth of varation observed, i.e.
+and deaths add randomness, they both add to the growth of variation observed, i.e.
 $b(x) + d(x)$.  Second, we see that dynamics for the variance $\sigma^2$ are *non-linear* whenever
@@
 This variance equation will be important to us later as we examine the relative
-roles of noise and nonlinearity in noise-driven phenomena such as stochastic
+roles of noise and non-linearity in noise-driven phenomena such as stochastic
 transitions between alternate stable states.
@@
 $$\langle x_0, x_{t} \rangle = \frac{b(x) + d(x)}{2 \partial_x \left[ b(x) - d(x) \right]} \exp\lbrace  -t \left| \partial_x b(x) - \partial_x d(x) \right|  \rbrace $$
-which will also prove useful later on.  Note this expression simply states that the equilbrium
+which will also prove useful later on.  Note this expression simply states that the equilibrium
 variance is a balance between noise terms, $b(x)+ d(x)$ and the slope or eigenvalue of the
 deterministic skeleton, $2 \partial_x \left[ b(x) - d(x) \right]$, (a balance that forms the basis of the
-the fluctuation-disspation theorem) and that correlations fall off exponentially at a rate given by this slope.   
-We will see both autocorrelation and variance play similar roles to infer this eigenvalue when we 
+the fluctuation-dissipation theorem) and that correlations fall off exponentially at a rate given by this slope.   
+We will see both auto-correlation and variance play similar roles to infer this eigenvalue when we 
 reach the role of noise as informer.
@@
 model descriptions and applications, but fails to point out these differences,
-mistakenly characterizing the van Kampen expansion as synonomous with the diffusion approximation.
+mistakenly characterizing the van Kampen expansion as synonymous with the diffusion approximation.
 In brief, the diffusion approximation does not lead to a Gaussian distribution
@@
 @Gardiner2009 (see Appendix B), the Kramer-Moyal expansion approach rests on rather
-a more dubious double limit that often has no clear biological intepretation and
+a more dubious double limit that often has no clear biological interpretation and
 can restrict its validity (though I believe theorem 3.3 of @Kurtz1978 establishes
@@
 environmental noise that is approximately Gaussian in distribution,
-but unlike the canonical model, leads to explicitly autocorrelated noise 
+but unlike the canonical model, leads to explicitly auto-correlated noise 
 (see derivation, Appendix B).  Auto-correlated or "colored" environmental
@@
 [e.g. @Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015], but such 
-formulations start from the empirical observation of autocorrelation 
+formulations start from the empirical observation of auto-correlation 
 in environmental time series rather than a process-based derivation.
@@
 and population level process that really matter (more precisely, the relative 
-magnitudes of the autocorrelation time of the environmental dynamics and population dynamics.)
-An environment with comparatively short autocorrelation scale will act
+magnitudes of the auto-correlation time of the environmental dynamics and population dynamics.)
+An environment with comparatively short auto-correlation scale will act
 effectively as white noise upon the population dynamics.
@@
 but rather at some stochastically fluctuating rate of a given mean, variance,
-and autocorellation.  The system size expansion illustrates how we replace
+and auto-correlation.  The system size expansion illustrates how we replace
 a model where stochasticity is nested inside some parameter into the simpler
@@
 of the deterministic skeleton with respect to the environmental variable,
-see appendix B.) Thus for the Levin's model, we can capture the fluctuations
+see appendix B.) Thus for the Levins' model, we can capture the fluctuations
 introduced by demographic and environmental noise at equilibrium as
@@
 possible consequences of noise. Unsurprisingly few topics have
-recieved greater attention then the question of the persistence of
+received greater attention then the question of the persistence of
 a given species [e.g. @Kendall1949;  @Bartlett1955; @Bartlett1960;
@@
 [@Ovaskainen2010 provides one excellent review], in
-particular addressing autocorrelated noise [e.g. @Ripa; @Ripa1999; 
+particular addressing auto-correlated noise [e.g. @Ripa; @Ripa1999; 
 @Petchey, @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure

### 2018-03-02 paper/paper.Rmd
intro!

@@
 
-<!--
-Directions:
-
-We hope that these papers will combine a mix of synthetic reviews of the current state of an area, as well as insights into how the field has developed during the last twenty years, discussion of current important questions, and developments and insights that would significantly advance the field.
-Although certainly not necessary, if there are any past Ecology Letters papers that have had a profound impact on the development of an area – be it beneficial or not – a brief discussion of that impact would be appropriate.
-We also hope that articles will be written in a thoughtful style, but also provocative if appropriate.
-
--->
-
 \newpage
-
-<!-- FIXME look at Coulson, Bjornstead, Ovaskeinen and others and see how they define scope to focus on "population ecology" -->
 
@@
 
-The gap between theory-focused specialists and more empirically minded non-specialists
-is particularly acute in the ecological literature on stochastic processes, making
-a review of 
-<!-- FIXME -->
-
-As the literature on stochastic modeling in ecology is both extensive and often
-technical, and so rather than attempt a comprehensive review, this synthesis
-will focus on the simplest examples necessary to illustrate this broader trend
-in the changing way we think about the role of noise.
-To that end, I will focus on examples from simple population models, echoing
-the motivation of @Bartlett1960 to focus on basic theoretical aspects common
-across models. In so doing, we will ignore the equally rich literature of 
-stochastic models in evolution [@Lenormand2009] and population genetics
-[e.g. @Ewens2004; @Gillespie2004]. In an effort to make these themes accessible to more empirical
-readers, I will rely on the simplest examples necessary, with only passing
-mention of the rich literature extending these concepts to more realistic
-models.  It is easy to critique this choice: indeed, one might summarize decades
-of research in noting the twin observations that (1)  ecology is complex and (2)
-that complexity is important. In the stochasticity literature alone, much work 
-has been done to understand and extend fundamental results to more complex models,
-including the importance of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
-@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g. @Durrett1994; @Dieckmann2000;
-@Kerr2002; @Schreiber2010], Individual heterogeneity [e.g. @Vindenes2015; @Hart2016]
-or the coexistence of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
-@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]; 
-correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010; @Lee2017; @Spanio2017] or periodic
-[@Coulson2001; @Bjornstad2001; @Keeling2001] structure in environmental noise, or
-the interaction of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
-@Lenormand2009; @Vindenes2015; @Schreiber2015]. In this we follow @Levins1966 in
-sacrificing realism for generality.
-
-In this spirit, I provide concise, detailed and commented
-code for simulating each of the models presented in Appendix A.
-Simulations permit poking and prodding of empirically-minded
+<!-- -->
+
+The gap between specialists and non-specialists is particularly acute in the
+ecological literature on stochastic processes.  This makes the potential benefit
+for new synthesis all the greater, but also poses an obvious challenge in 
+appealing to both audiences.  In recognition of this, I provide concise and
+commented code for simulating each of the models presented in the supplement,
+Appendix A, while more mathematical background and derivations are presented Appendix B.
+Numerical simulations permit poking and prodding of empirically-minded
 investigation unencumbered by either experimental design or mathematical formalism.
@@
 bug reports, suggestions or help requests are welcome through the
-issue tracker.  Appendix B provides more mathematical details
-for interested readers.  With these in place, we are ready to begin
-our tour through three major areas of noisy phenomena: Origins of
-noise, emergent phenomena, and noise-driven inference.
+issue tracker.
+
+To emphasize the underlying trend in the changing roles in which we
+see and understand noisy processes, I will also restrict my focus 
+to relatively simple models primarily from population ecology context.
+Simplicity not only makes examples (in equations and in code) more
+tractable but also allows us to focus on aspects that are germane to many
+contexts rather than unique to particular complexitites [@Levins1966; @Bartlett1960].
+Nevertheless, that complexity matters -- few themes have been better 
+emphasized in the theoretical literature [@Bjornstad2001].  Both
+the foundational literature and recent research continue to echo the
+theme of understanding the impact different real world complexities
+have in stochastic dynamics, including the importance
+of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
+@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g.
+@Durrett1994; @Dieckmann2000; @Kerr2002; @Schreiber2010], Individual
+heterogeneity [e.g. @Vindenes2015; @Hart2016] or the coexistence
+of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
+@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007;
+@Schreiber2017]; correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010;
+@Lee2017; @Spanio2017] or periodic [@Coulson2001; @Bjornstad2001;
+@Keeling2001] structure in environmental noise, or the interaction
+of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
+@Lenormand2009; @Vindenes2015; @Schreiber2015].  As such, we will 
+rely on both textbooks and recent reviews to provide a proper treatment
+of these issues, and focus on broader trends that may be less apparent
+to non-specialists.
+
+This review is structured into three sections: Origins of noise, emergent
+phenomena, and noise-driven inference.  The first section lays the conceptual
+groundwork we will need, while also highlighting a shift to more and more
+mechanistically rooted descriptions of noise.  We will see where the common
+formulation of 'deterministic skeleton plus noise term' comes from, how it 
+is best justified, and when it breaks down.  The second section introduces
+"noise the creator," showing examples of ecological phenomena generated by
+stochasticity.  These examples will be familiar to most specialists but
+illustrate a different way of thinking than held by most ecologists, where
+noise is only a nuciance to be filtered or averaged out.  The third and
+final section, "noise the informer," turns these examples back-to-front, 
+asking what noise can tell us about a system: such as its underlying resilience
+or stability, or the approach of a catastophic shift. Examples are fewer
+here, and have largely yet to benefit from the introduction of either
+the rigorous theorems or more complex models so plentiful in the previous
+sections.  Yet the promise of prediction, of early warning signs before
+tipping points, have spurred broad interest in such noise-based inference.
+This review is a call to both deepen the connection to mechanism and better
+formalize this thinking, but also look more broadly into other ways in which
+noisy phenomena can help inform and predict underlying processes.
 

### 2018-03-02 paper/paper.Rmd
intro to 'creator' section, working on intro/rewrite to 'informer'

@@
 it was first developed in the 70s, and remains an underutilized approach
-for writing simple and approximation-free[^2] stochastic ecological models.
-
-[^2]: that is, free from the approximation made by SDE models as we see in the
-van Kampen example.  All models are, of course, only approximations.
+for writing simple and approximation-free^[that is, free from the approximation
+made by SDE models as we see in the van Kampen example.  All models are, of course,
+only approximations.] stochastic ecological models.
 
@@
 
+Outside of the primarily theoretical literature on stochastic dynamics
+in ecology, noise is viewed largely as a nusiance: something entirely
+distinct from the processes we are interested in which invariably keeps
+getting in the way.  As we have seen in section on origins, such 
+characterizations are misleading: noise arises from the very same processes.
+While both empirical and statistical literature have frequently emphasized
+techniques to remove noise, a central theme of stochastic processes
+have emphasized an almost opposite conclusion: noise cannot be subtracted or
+averaged away, since tthe presence of noise can fundamentally alter the
+dynamics we observe.
 Stochasticity can give rise to a wide array of interesting phenomena
@@
 
-<!-- More frank introduction to this section, see reply to Reviewer 1-->
-
-One of the most interesting aspects of noise is as a source of
-additional information.  This can be realized in a variety of ways.
-For instance, in deriving expressions for the variance created by
-demographic noise above, we observed that different partitions of
-a logistic growth model can lead to the same macroscopic dynamics
-(same values for $b$ - $d$), while having different variance (due
-to different values for $b$+$d$).  This suggests that the variation
-observed in such a population could potentially help resolve
-how dynamics are partitioned between individual birth and death
-rates (though in practice one would have to first account for any
-contribution of environmental noise).
+In examining "noise the creator," we began with a tension between
+the intepretation of noise as it is usually presented in the context of
+statistical inference: a nusiance to be stripped away, and how it is
+viewed by theorist, as a potentially fundamental driver of patterns,
+including oscillations, regime shifts, extinction & coexistence.
+In this third section, we attempt to re-unite these two largely
+divergent views of the role of noise by highlighting how phenomena
+from the latter (noisy phenomena) can be used to address the questions
+of the former (inferring underlying process from pattern).
+
+The most provacative example of how noisy phenomena can act to reveal
+underlying processes comes from the literature on so-called 'early
+warning signs' of critical transitions [e.g. @Scheffer2009].  This approach
+exploits the ability of the equations we encountered in "origins of noise"
+to reveal subtle changes in underlying processes before those changes are
+visible at the macroscopic (deterministic) level.  This literature focuses
+on the phenomenon of a tipping point, in which gradual changes to the environment
+show essentially no changes to the behavior of a system (i.e. the abundance of a
+population or health of lake ecosystem) followed by a rapid change when the
+tipping point is reached that cannot be easily reversed. [@Scheffer2009;
+@Scheffer2015].  The fundamental insight of early warning signals has been
+to realize that though the average or deterministic dynamics of such a system
+do not change significantly prior to the tipping point, the dynamics of the
+noise must change in a consistent and predictable manner.
+
+This pattern, known as 'critical slowing down' for reasons that will become 
+obvious, was first proposed an an early warning sign for ecological
+transitions by @Wissel1984, but has received considerable attention only more recently,
+particularly owing to work in empirical systems [@Scheffer2015].
+Our analysis in the origins of noise provides sufficient framework to illustrate
+this idea completely.  The premise behind the tipping point just described
+is that of a fold bifurucation [which I will not detail here, but see @Holling1973; @Scheffer2009; @Strogatz2001].
+The key thing we need to know about this bifurcation is that it involves
+changes *derivative* of the deterministic skeleton ($\partial_x \left[ b(x) - d(x) \right]$ in
+the notation of our example), which can build up for some time without
+significant changes to the location of the equilibrium state $\hat x$.
+This derivative did not appear in our marcroscopic equation, $\tfrac{\mathrm{d}x}{\mathrm{d}t}} = b(x) - d(x)$,
+but featured regularly in our equations for the noise, such as the 
+expected variance, Eq \eqref{var} and autocorrelation, Eq \eqref{autocor}.
+This term represents the *stability*  of the equilibrium: how quickly
+deviations return to the stationary value.
+
+The trick of the early warning signals approach is to view noise as 
+a sequence of countless miniature perturbation experiments.
 
@@
 stability; possibly due to an approaching catastrophic transition
-or saddle-node bifurcation.  This pattern, known as critical slowing
+or saddle-node bifurcation.
+
+This pattern, known as critical slowing
 down (in reference to the auto-correlation time in the physics
@@
 as a whole.
+
+<!--
+One of the most interesting aspects of noise is as a source of
+additional information.  This can be realized in a variety of ways.
+For instance, in deriving expressions for the variance created by
+demographic noise above, we observed that different partitions of
+a logistic growth model can lead to the same macroscopic dynamics
+(same values for $b$ - $d$), while having different variance (due
+to different values for $b$+$d$).  This suggests that the variation
+observed in such a population could potentially help resolve
+how dynamics are partitioned between individual birth and death
+rates (though in practice one would have to first account for any
+contribution of environmental noise).
+-->
 

### 2018-03-02 paper/paper.Rmd
WIP

@@
 The key thing we need to know about this bifurcation is that it involves
-changes *derivative* of the deterministic skeleton ($\partial_x \left[ b(x) - d(x) \right]$ in
+changes *derivative* of the deterministic skeleton ($\partial_x \left[ b(x) - d(x) \right]_{x=\hat x}$) in
 the notation of our example), which can build up for some time without
@@
 expected variance, Eq \eqref{var} and autocorrelation, Eq \eqref{autocor}.
-This term represents the *stability*  of the equilibrium: how quickly
+This derivative represents the *stability*  of the equilibrium: how quickly
 deviations return to the stationary value.  
+The magnitude of both variance and autocorrelation must increase
+as the stability ($\partial_x \left[ b(x) - d(x) \right]_{x=\hat x}$) decreases
+prior to a critical tranisition.
 
@@
 
-Noise can provide novel information in other ways as well.  Stochastic
-fluctuations act like a continual miniature perturbation experiment,
-constantly pushing a system away from an equilibrium to reveal its
-behavior elsewhere in state space. Observing the size and
-auto-correlation of these fluctuations thus provides insight on the
-stability of an equilibrium point, (as can be seen, for instance
-in the fluctuation-dissipation result we found for generic birth
-death models).  This observation has been exploited with particular
-success as a potential *early warning sign* that a system is loosing
-stability; possibly due to an approaching catastrophic transition
-or saddle-node bifurcation.
-
-This pattern, known as critical slowing
-down (in reference to the auto-correlation time in the physics
-literature) was first proposed an an early warning sign for ecological
-systems by @Wissel1984, but has received considerable empirical as
-well as theoretical attention only more recently [@Scheffer2009;
-@Scheffer2015].  Increases in two of these early warning indicators,
+Increases in two of these early warning indicators,
 variance and auto-correlation, are illustrated along with the

### 2018-03-03 paper/paper.Rmd
refs, real data

@@
 ```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="cairo_pdf", fig.width=7, fig.height=3.5, cache = TRUE)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="cairo_pdf", fig.width=7, fig.height=3.5, cache = FALSE)
 library(tidyverse)
@@
 library(ggthemes)
+library(png)
 colours <- ptol_pal()(2)
+
+#https://stackoverflow.com/a/24241954/258662
+fancy_scientific <- function(l) { 
+     l <- format(l, scientific = TRUE)
+     l <- gsub("^(.*)e", "'\\1'e", l)
+     l <- gsub("e", "%*%10^", l)
+     l <- gsub("^'1'%\\*%", "", l)
+     l <- gsub("\\+", "", l)
+     parse(text=l)
+}
 ```
 
-```{r suggested_libs}
+```{r suggested_libs, cache = FALSE}
 library(hrbrthemes)
@@
 have in stochastic dynamics, including the importance
-of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
+of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989; @Tuljapurkar1997;
 @Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g.
@@
 \\label{quasicycles}"}
-read_csv("../appendixA/quasicycles.csv") %>%
+fig2 <- read_csv("../appendixA/quasicycles.csv") %>%
   gather(species, population, -t, -sigma) %>%
@@
   rename(time = t) %>%
-  ggplot(aes(time, population, col = species)) + geom_line() + 
-  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() 
+  ggplot(aes(time, population, col = species)) +
+  geom_line() + 
+  facet_wrap(~ sigma, ncol=2) + 
+  scale_color_ptol() +
+  theme(legend.position = "none")
+
+fig2c <- ggplot() + 
+  annotation_raster(readPNG("figures/resonance-bjornstad2004.png"), 
+                    xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
+  labs(subtitle = "C.")
+
+gridExtra::grid.arrange(fig2,fig2c, layout_matrix = rbind(c(1,1,2)))
+
 ```
@@
 Understanding the origin of sustained oscillations has long been a
-central question in ecology [@Hastings1996; @Bjornstad2001].  The
+central question in ecology [@Hastings1996; @Kendall1999;; @Bjornstad2001].  The
 simple Lotka Volterra model produces only neutrally stable centers,
@@
 sustained oscillations (Figure \ref{quasicycles}B) 
-[e.g. empirical examples in @Blarer1999; @Greenman2003; @Black2010].  
-In the example shown here, a
-simple predator-prey system is driven by environmental white noise
-$\xi_t$, with prey following the equation:
+[empirical discussed in @Greenman2003; @Bjornstad2004; @Black2010].
+
+@Nisbet1976 first identifies the potential for quasi-cycles to explain
+sustained oscillations in a simple predator prey model with stochastic
+forcing, which provides a familiar formulation for us to illustrate
+the phenomena. Consider a predator-prey system is driven by environmental 
+white noise $\xi_t$ (e.g. as we have seen arise from stochastic growth rates), 
+with prey following the equation:
 
@@
 
-$$y_{t+1} = y_t + c  x_t  y_t - d  y_t+ \xi_{y,t} $$
+$$y_{t+1} = y_t + c  x_t  y_t - d  y_t+ \xi_{y,t}$$
+
 with $c$ the conversion ratio and $d$ the predator mortality rate.
@@
 
-Early work by @Blarer1999 identified quasi-cycles as a possible
+Since @Nisbet1976, examples of stochastic resonance have been demonstrated in common
+models and implicated in empirical examples of increasing complexity [@Greenman2003].
+@Blarer1999 identified quasi-cycles as a possible
 explanation for oscillations and periodic outbreaks in pest ecosystems.
@@
 to drive cyclic behavior in population abundances of Atlantic bluefin
-tuna and Atlantic Cod, while @Black2010 illustrates the use of
-power-spectrum approaches to examine the behavior of quasi-cycles
-in common disease outbreak models, focusing on a case where noise
-arises explicitly from demographic stochasticity, rather than generic
-environmental noise as in the model above and elsewhere [@Kendall1999; @Pascual2003].  
-@Pineda-Krch2007 compare patterns from these stochastically generated cycles to
-stable limit cycles they resemble.  Stochastic resonance highlights
+tuna and Atlantic Cod.  @McKane2005, @Alonso2007, and @Black2010 illustrate
+the emergence of quasi-cycles in common disease outbreak models, including
+the use of power spectrum analysis.  
+@Pineda-Krch2007 compare patterns from stochastically generated cycles to
+stable limit cycles they resemble to ask how we can discriminate between
+these mechanims.  Stochastic resonance highlights
 one way in which even small noise can play an important role in
@@
 
-<!--Figure \ref{noisy_switch}: Noisy switch -->
+<!--Figure 3: Noisy switch -->
 ```{r noisy_switch, fig.width=7, fig.height=7, fig.cap="Stochastic switiching in May's model of alternative stable state dynamics.
@@
 
+p0 <- ggplot() + 
+  annotation_raster(readPNG("figures/switching-keeling2001.png"), 
+                    xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
+  labs(subtitle = "A.")
+
 p1 <- read_csv("../appendixA/noisy_switch.csv")  %>%
@@
   geom_line(lwd=.1, col=colours[[1]]) +
-  labs(subtitle = bquote(bold(A.)))
+  labs(subtitle = bquote(bold(B.)))
 
@@
   scale_color_ptol() + 
-  labs(x = bquote(X[t]), y = bquote(X[t+1]), subtitle= bquote(bold(B.))) +
+  labs(x = bquote(X[t]), y = bquote(X[t+1]), subtitle= bquote(bold(C.))) +
   theme(legend.position="top")
@@
   coord_cartesian(xlim=c(0,1.6), ylim=c(-1.15,-0.9)) + 
-  labs(subtitle = bquote(bold(C.)))
-
-gridExtra::grid.arrange(p1,p2,p3, layout_matrix = rbind(c(1,1), c(2,3)))
+  labs(subtitle = bquote(bold(D.)))
+
+gridExtra::grid.arrange(p0,p1,p2,p3, layout_matrix = rbind(c(1,2), c(3,4)))
 ```
@@
 significant changes to the location of the equilibrium state $\hat x$.
-This derivative did not appear in our marcroscopic equation, $\tfrac{\mathrm{d}x}{\mathrm{d}t}} = b(x) - d(x)$,
+This derivative did not appear in our marcroscopic equation, $\tfrac{\mathrm{d}x}{\mathrm{d}t} = b(x) - d(x)$,
 but featured regularly in our equations for the noise, such as the 
@@
 <!-- Figure 4: Early warning signals -->
-```{r tipping, fig.width=7, fig.height=7, fig.cap="Early warning signs of a critical transition in a simulation of increasing serial dilution of budding yeast.
-Panel A: Rolling variance   autocorrelation Time series is shown in panel C.
-Model details and simulation parameters in Appendix A.
-\\label{tipping}"}
+```{r tipping, fig.width=7, fig.height=10, fig.cap="Early warning signs of a critical transition in a simulation of increasing serial dilution of budding yeast.
+Panel A: Data from replicate manipulations by @Dai2015.
+Panel B: simulated single time series moel.
+Panel C, D: Rolling variance and autocorrelation over simulated data show an increase in advance of the transition.
+Simulation details in Appendix A.
+Warning signal statistics for experimental data not shown, see @Dai2015.\\label{tipping}"}
 tipping <- read_csv("../appendixA/tipping.csv")
 
-p1 <- tipping %>%
-  select(-x) %>%
-  rename("A. autocorrelation" = autocorrelation, "B. variance" = variance) %>%
+pop_dai <- ggplot() + 
+  annotation_raster(readPNG("figures/pop-dai2015.png"), 
+                    xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
+  labs(subtitle = "A.")
+
+pop_sim <- 
+  tipping %>% 
+  select(t, population = x, tip_time) %>%
+  ggplot(aes(t, population)) + geom_line(col=colours[1]) + 
+  geom_vline(aes(xintercept = tip_time), col=colours[2], lty=2) + 
+  scale_y_log10(labels=fancy_scientific)  + 
+  coord_cartesian(ylim=c(1e3, 5e5)) +
+  labs(subtitle = "B. Yeast population size", 
+       caption = "Vertical red dashed line indicates tipping point location")
+
+stat_sim <- tipping %>%
+  dplyr::select(-x) %>%
+  dplyr::rename("C. autocorrelation" = autocorrelation, "D. variance" = variance) %>%
   gather(series, value, -t, -tip_time) %>%
@@
   facet_wrap(~series, ncol = 2, scales="free_y")
-p2 <- tipping %>% 
-  select(t, population = x, tip_time) %>%
-  ggplot(aes(t, population)) + geom_line(col=colours[1]) + 
-  geom_vline(aes(xintercept = tip_time), col=colours[2], lty=2) + 
-  #scale_y_log10()  + 
-  labs(subtitle = "C. Yeast population size", 
-       caption = "Vertical red dashed line indicates tipping point location")
-
-gridExtra::grid.arrange(p1,p2, layout_matrix = rbind(c(1,1), c(2,2)))
+
+gridExtra::grid.arrange(pop_dai, cv_dai, pop_sim, stat_sim,
+                        layout_matrix = rbind(c(1,2), c(3,3), c(4,4)))
 
@@
 way to stable oscillations (limit cycle) of growing magnitude
-[e.g. @Fussmann2000; @Strogatz2001].
+[e.g. @Fussmann2000; @Nelson2013; @Strogatz2001].
 

### 2018-03-03 paper/paper.Rmd
fix fig

@@
 Panel A: Data from replicate manipulations by @Dai2015.
-Panel B: simulated single time series moel.
-Panel C, D: Rolling variance and autocorrelation over simulated data show an increase in advance of the transition.
+Panel B: Coefficient of variation for the same experimental data.
+Panel C: Simulated single time series moel.
+Panel D, E: Rolling variance and autocorrelation over simulated data show an increase in advance of the transition.
 Simulation details in Appendix A.
@@
   labs(subtitle = "A.")
-
+cv_dai <- ggplot() + 
+  annotation_raster(readPNG("figures/cv-dai2015.png"), 
+                    xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
+  labs(subtitle = "B.")
 pop_sim <- 
@@
   coord_cartesian(ylim=c(1e3, 5e5)) +
-  labs(subtitle = "B. Yeast population size", 
+  labs(subtitle = "C. Yeast population size", 
        caption = "Vertical red dashed line indicates tipping point location")
@@
   dplyr::select(-x) %>%
-  dplyr::rename("C. autocorrelation" = autocorrelation, "D. variance" = variance) %>%
+  dplyr::rename("D. autocorrelation" = autocorrelation, "E. variance" = variance) %>%
   gather(series, value, -t, -tip_time) %>%

### 2018-03-03 appendixA/appendixA.Rmd
WIP
WIP: brief mentions of emprical work; a bit on tipping point / informer section

@@
 ---
+
+\renewcommand{\thefigure}{A\arabic{figure}}
 
@@
 
-```{r libraries, cache=FALSE}
+```{r libraries, cache=FALSE, message=FALSE}
 library(nimble)
@@
 library(gridExtra)
-
+library(scales)
 ## Set a default color pallete
@@
 
-```{r echo=FALSE, cache=FALSE, eval=TRUE}
+```{r echo=FALSE, cache=FALSE, eval=TRUE, message=FALSE}
 ## plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
@@
 ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
+
+#https://stackoverflow.com/a/24241954/258662
+scientific <- function(l) {
+    if(max(l, na.rm=TRUE) < 1e3) return(l)
+     l <- format(l, scientific = TRUE)
+     l <- gsub("^(.*)e", "'\\1'e", l)
+     l <- gsub("e", "%*%10^", l)
+     l <- gsub("^'1'%\\*%", "", l)
+     l <- gsub("\\+", "", l)
+     parse(text=l)
+}
 ```
@@
   write_csv("tipping.csv")
+tipping
 ```
@@
   select(-x) %>%
-  rename("A. autocorrelation" = autocorrelation, "B. variance" = variance) %>%
+  filter(t < 1866) %>%
+  rename("A. autocorrelation" = autocorrelation, "variance" = variance) %>%
   gather(series, value, -t, -tip_time) %>%
@@
   geom_vline(aes(xintercept = tip_time), col=colours[[2]], lty=2) +
-  facet_wrap(~series, ncol = 2, scales="free_y")  
+  facet_wrap(~series, ncol = 2, scales="free_y") +
+  scale_y_continuous(labels=scientific)
 p2 <- tipping %>% 
   select(t, population = x, tip_time) %>%
-  ggplot(aes(t, population)) + geom_line(col=colours[1]) + 
+  ggplot(aes(t, population)) + 
+  geom_line(col=colours[1]) + 
   geom_vline(aes(xintercept = tip_time), col=colours[2], lty=2) +
-  #scale_y_log10()  + 
+  scale_y_log10(labels=scientific)  + 
   labs(subtitle = "C. Yeast population size", 
@@
 ```
+
+Note that @Dai2015 plots coefficient of variation, (standard deviation over mean), rather than the variance.
+This is a popular choice since the resulting indicator is a dimensionless measure of the noise relative to the mean, but also confounds changes in the mean dynamics with changes in the noise dynamics.
+This can give a misleading impression of the performance of the indicator: in many models, the mean (that is, the value of the equilibrium population size, $\hat x$) will also decrease prior to the transition, resulting in a more pronounced increase in the coefficent of variation than in the variance.
+However, changes in mean per se should are not a generic indicator of stability predicted by the theory -- indeed, the whole point of the theory was to detect changes that would not be manifested in the mean.
+In general, systems going through a saddle-node bifurcation can vary greatly in how much the mean changes.
+The mean will generally decrease, since it must eventually meet the 'saddle' point below it -- it is a question of whether the saddle point does most of the increasing or the node does most of the decreasing.
 

### 2018-03-03 paper/paper.Rmd
WIP: brief mentions of emprical work; a bit on tipping point / informer section

@@
 [empirical discussed in @Greenman2003; @Bjornstad2004; @Black2010].  
+Figure \ref{quasicycles}C shows empirical data on records of bluefin 
+tuna catch that @Bjornstad2004 analyze as a possible example for such
+quasi-cycle oscillations in an important marine predator.
 
@@
 
-We turn from stochastic resonance to oscillations of a very different
-sort, which do not resemble any deterministic process, as seen in
-Figure \ref{noisy_switch}A.  This pattern is characterized by long stretches during
+We turn from stochastic resonance to oscillations of a  different
+sort, illustrated in Figure \ref{noisy_switch} in empirical data on 
+rubella outbreaks in Copenhagen, analyzed by @Keeling2001, and in 
+simulations of a standard model shown in Figure \ref{noisy_switch}B.
+This pattern is characterized by long stretches during
 which fluctuations in the state variable remain concentrated around
@@
 arises through the interaction of stochasticity and alternative
-stable state dynamics.  The simulation shown in Figure \ref{noisy_switch}A for the
-is based on an resource consumption model first proposed by @May1977
-as one of the early examples of the potential for alternative stable
-state dynamics.  The resource grows according to a logistic growth
+stable state dynamics.
+
+Both for clarity of presentation and highlight the generality of 
+stochastic switching in diverse models, we will focus not specifically
+on the more complex disease model analyzed in @Kelling2001 but
+rather rely on the classic resource consumption model of @May1977
+which has been widely used to study alternative state dynamics.
+As we shall see, stochastic switching is a generic property of 
+models with alternative stable states which will have other implications
+later on.  In @May1977 model, a resource species grows according to a logistic growth
 model and is consumed according to a Holling Type-III functional
@@
 where $\xi_t$ is normally distributed environmental variation with
-mean zero and standard deviation $\sigma$. Figure \ref{noisy_switch}B shows separate
+mean zero and standard deviation $\sigma$. Figure \ref{noisy_switch}C shows separate
 curves corresponding the parts of the model describing vegetation
@@
 
-The concept of a potential well, Figure \ref{noisy_switch}C, is often invoked when
+The concept of a potential well, Figure \ref{noisy_switch}D, is often invoked when
 describing the behavior of alternative stable states.  The curve
@@
 of activation energy to grade-school students of chemistry).  This
-effect is visible even in the short simulation shown in 3A, which
+effect is visible even in the short simulation shown in Figure \ref{noisy_switch}B, which
 spends longer times in the lower state which has the deeper well.
@@
 least as much confusion as it is of insight.  While the curves in
-Figure \ref{noisy_switch}B can be clearly associated with underlying mechanisms of
+Figure \ref{noisy_switch}C can be clearly associated with underlying mechanisms of
 rates: logistic growth and consumption with handling time, the same
@@
 
+Because these transitions between alternative stable states appear as
+such a dramatic deviation from the previous pattern (as illustrated
+in Figure \ref{noisy_switch}A), it is  tempting to assume they
+result from some sudden external force or change in the
+environment. As the numerical examples clearly illustrate,
+these transitions can neverthless be explained entirely by chance
+[@Boettiger2012;  @Drake2013].
+
 Empirical evidence for stochastic switching in ecological timeseries
 is limited by the long time scale usually required 
-[but see @Keeling2001; @Clements2016; @Clements2017]. Because a single
-stochastic transition between alternative stable states appears as
-such a dramatic deviation from the previous pattern (as illustrated
-in Figure \ref{noisy_switch}A), it is often tempting to assume the change could only
-be the result of some sudden external force or change in the
-environment, so it is worth remembering that given long enough,
-these transitions can happen entirely by chance stochasticity
-[@Boettiger2012; @Drake2013]. However, these transitions are most
-likely to observed in systems with relatively shallow potential
-wells.  This is mostly likely to occur of some external change is
-slowly making the potential well shallower and thus transitions
-more likely -- a phenomenon that has become to be known as "flickering"
-and suggested as a potential early warning sign of an approaching
-transition [@Brock2010].  Examples of flickering between alternative
-stable states have been suggested in data ranging from geological
-timescales at the end of glaciation in the Pleistocene [@Taylor1993]
-to ecological scales over a few decades in lake ecosystems [@Wang2012].
+[but see @Keeling2001; @Taylor1993; @Wang2012; @Clements2016; @Clements2017]. 
+As our discussion highlights, we are most likely to observe the phenomenon when at
+least one of the potential wells is relatively shallow, which
+results in a pattern of long periods in a single state (the deeper well)
+punctuated by brief excursions into and back out of the shallow well.
+This drives the more spike-like pattern seen in the rubella outbreaks
+from @Keeling2001. 
 Stochastic transitions between bi-stable states are thought to play
@@
 in clonal microbial populations, which can act to buffer the
-population in a changing environment  [@Balaban2004].
-
-<!--
-**Further examples**
-- @Read2007  Stochasticity generates an evolutionary instability for infectious disease 
-- Spatial pattern formation: Turning patterns found in plankton-herbivore dynamics [@Levin1976]
-
--->
+population in a changing environment  [@Balaban2004]. This mechanism
+has also been explored in the emergence of multiple strategies
+under evolutionary dynamics in disease models [@Read2007].
 
 # Noise the informer: Noise can provide novel information
+
+Transitions from systems with particularly shallow wells have recently
+taken on new empirical significance.  Because such transitions become
+more likely as the potential well becomes less deep (i.e. the attractor
+has become less stable), the onset of such transitions can be intepreted
+as an indication of this loss of stability. This phenomenon has been
+dubbed "flickering" and suggested as a warning sign that a more
+permanent transition to the alternate stable state is approaching [@Brock2010].  
+Examples of flickering between alternative
+stable states have been suggested in data ranging from geological
+timescales at the end of glaciation in the Pleistocene [@Taylor1993]
+to ecological scales over a few decades in lake ecosystems [@Wang2012].
+Flickering is then our first example of "noise the informer" -- in
+which the emergence of a stochastic phenomenon is re-interpreted 
+in a way that sheds light on an underlying process.
 
@@
 
-The most provacative example of how noisy phenomena can act to reveal
-underlying processes comes from the literature on so-called 'early
-warning signs' of critical transitions [e.g. @Scheffer2009].  This approach
-exploits the ability of the equations we encountered in "origins of noise"
+Our most provacative example of how noisy phenomena can act to reveal
+underlying processes comes from the literature we have just touched upon
+in the case of flickering -- a growing set of so-called 'early
+warning signs' of critical transitions [e.g. @Scheffer2009].  The fundamental
+insight behind all of these signs comes from the exploiting the ability of
+equations we encountered in "origins of noise"
 to reveal subtle changes in underlying processes before those changes are
@@
 
-The trick of the early warning signals approach is to view noise as 
-a sequence of countless miniature perturbation experiments.
-
-<!-- fixme need more background? -->
-
 <!-- Figure 4: Early warning signals -->
@@
 
-Increases in two of these early warning indicators,
-variance and auto-correlation, are illustrated along with the
-timeseries of a simulation of the dynamics of yeast population
-growth in Figure \ref{tipping}.  This simulation is based on a mechanistic model
-developed by @Dai2012 that closely matches the population dynamics
+The trick of the early warning signals approach is to view noise as 
+a sequence of countless miniature perturbation experiments.  
+Figure \ref{tipping} illustrates early warning indicators in an
+experimental manipulation and corrsponding simulated model of
+@Dai2012 and @Dai2015.  Though early warning signals have been
+reported in a growing range of systems [@Scheffer2015], these
+laboratory experiments in population dynamics
 they observed in serial dilutions of budding yeast _Saccharomyces
-cerevisiae_ on a sucrose medium. Because the relevant proteins are
+cerevisiae_ on a sucrose medium provide the most convincing 
+mechanistic intepretation of the underlying fold bifurcation.
+Because the relevant proteins are
 located in the extra-cellular matrix rather than inside the cells,
@@
 Increasing the serial dilution rate slowly over time drives the
-stable population towards a fold bifurcation, which is accompanied
-by the patterns of increasing variance and increasing auto-correlation
-seen empirically in lab populations [@Dai2012] as well as the
-numerical simulation shown in Figure \ref{tipping}.
+stable population towards a fold bifurcation which is accompanied
+by the patterns of increasing variance and increasing auto-correlation.
+This pattern can be seen empirically in lab populations [@Dai2012; @Dai2015]
+and explored numerically in a simulation of the yeast dynamics shown in 
+Figure \ref{tipping}. As the mechanistic model of yeast population
+dynamics is already quite simple, the simulation shown here (code in Appendix B)
+corresponds directly to the model proposed by @Dai2015.
 
@@
 of an ensemble, and should be computed over replicate simulations.
-While the laboratory experiments of @Dai2012 do precisely this,
+While the laboratory experiments such as @Dai2015 do precisely this,
 conducting identical, independent replicate manipulations in separate
@@
 of replicates.
+
+In contrast to the examples we have considered previously, work that extends
+this analysis into more complex scenarios of structured populations, interacting species,
+autocorrelated noise etc are significantly underdeveloped.  Even the burgeoning area
+of spatial early warning indicators [e.g. @Guttal2009; @Dakos2011; @Bel2012; @Dai2013]
+remains idosynchratic, and rigorous theorems are few.
 
@@
 The greater our suite of stochastic indicators, the better our ability
-to pin down the processes involved.  Such examples suggest
+to pin down the processes involved.
+
+Such examples suggest
 that stochastic processes could play a similar role in identifying
@@
 
-Most importantly, the framing of stochastic processes as indicators of
-changes in an ecosystem that might otherwise go unnoticed has proven a
+Intepreting stochastic processes as indicators of subtle but ominous
+changes in an ecosystem has proven a
 rich catalyst for bringing these concepts beyond the realm of abstract
@@
 warning signals [_i.e._ @Carpenter2011; @Dai2012; @Clements2016; @Clements2017;
-and many others, _e.g._ see @Scheffer2015 review] has served no merely
+and many others, _e.g._ see @Scheffer2015 review] has served not merely
 to illustrate these phenomena can be seen in laboratory and field ecosystems,

### 2018-03-03 appendixA/appendixA.Rmd
fix typo

@@
   write_csv("tipping.csv")
-tipping
 ```

### 2018-03-03 paper/paper.Rmd
a bit more text

@@
 this analysis into more complex scenarios of structured populations, interacting species,
-autocorrelated noise etc are significantly underdeveloped.  Even the burgeoning area
-of spatial early warning indicators [e.g. @Guttal2009; @Dakos2011; @Bel2012; @Dai2013]
-remains idosynchratic, and rigorous theorems are few.
-
-The challenge of identifying warning signs that a system is approaching
-a tipping point has proven a rich area for applications of stochastic
-processes.  We have already seen how in addition to the patterns of
-increasing variance and auto-correlation, our survey of stochastic phenomena
-has highlighted the increased potential to observe stochastic switching,
-or flickering, as the system approaches the tipping point.  Challenges remain
+autocorrelated noise etc are significantly underdeveloped. Issues of spatial
+structure have recieved more attention [e.g. @Guttal2009; @Dakos2011; @Bel2012; @Dai2013]
+but still largely idosynchratic.  Significant challenges remain
 in distinguishing these patterns from other causes [@Boettiger2013a].
@@
 approaching a saddle-node bifurcation, but will not show flickering.
-The greater our suite of stochastic indicators, the better our ability
-to pin down the processes involved.
-
-Such examples suggest
+Teasing these issues apart in greater mathematical formalism and under more
+complex and realistic models remain largely open problems.
+
+More broadly, examples of early warning signals such as increasing variance,
+auto-correllation, and flickering all point to ways in which stochasticty
+and noise-driven phenomena can be transformed from something which obscures
+underlying processes to something which reveals them.  I believe this is
+role could apply far more broadly than in just the context of stability and
+tipping points.  The examples from fold bifurcations already suggest
 that stochastic processes could play a similar role in identifying
@@
 
-Intepreting stochastic processes as indicators of subtle but ominous
-changes in an ecosystem has proven a
+Intepreting stochastic processes as potential new sources of information
+has also proven to be a
 rich catalyst for bringing these concepts beyond the realm of abstract

### 2018-03-05 paper/paper.Rmd
WIP

@@
 tractable but also allows us to focus on aspects that are germane to many
-contexts rather than unique to particular complexitites [@Levins1966; @Bartlett1960].
+contexts rather than unique to particular complexities [@Levins1966; @Bartlett1960].
 Nevertheless, that complexity matters -- few themes have been better 
@@
 illustrate a different way of thinking than held by most ecologists, where
-noise is only a nuciance to be filtered or averaged out.  The third and
+noise is only a nuisance to be filtered or averaged out.  The third and
 final section, "noise the informer," turns these examples back-to-front, 
 asking what noise can tell us about a system: such as its underlying resilience
-or stability, or the approach of a catastophic shift. Examples are fewer
+or stability, or the approach of a catastrophic shift. Examples are fewer
 here, and have largely yet to benefit from the introduction of either
@@
 Outside of the primarily theoretical literature on stochastic dynamics
-in ecology, noise is viewed largely as a nusiance: something entirely
+in ecology, noise is viewed largely as a nuisance: something entirely
 distinct from the processes we are interested in which invariably keeps
@@
 have emphasized an almost opposite conclusion: noise cannot be subtracted or
-averaged away, since tthe presence of noise can fundamentally alter the
+averaged away, since the presence of noise can fundamentally alter the
 dynamics we observe.
@@
 
-Oscillations arise through the process of stochastic resonance.
+Oscillations arise through the process known as stochastic resonance [@Nisbet1976; @Greenman2003].
 The damped oscillations seen in Figure \ref{quasicycles}A are analgous to a child
@@
 squared norm of the Fourier transform, a measure of what frequencies
-are present; see, e.g. @Black2012) provides a convenient way to
+are present; see, e.g. @Nisbet1982 for background and examples in
+@McKane2005, @Alonso2007 or @Black2012) provides a convenient way to
 visualize the effects of these oscillations.  The power spectrum
@@
 shows a distribution of frequencies, centered at the the resonant
-frequency of the damped oscillator.
+frequency of the damped oscillator [@Nisbet1982].
 
 Since @Nisbet1976, examples of stochastic resonance have been demonstrated in common
-models and implicated in empirical examples of increasing complexity [@Greenman2003].
+models and implicated in empirical examples of increasing complexity, [@Greenman2003],
+particularly in higher-dimensional disease and outbreak models [@Blarer1999; @Greeman2003; @McKane2005; @Alonso2007; @Black2010].
 @Blarer1999 identified quasi-cycles as a possible
@@
 stable limit cycles they resemble to ask how we can discriminate between
-these mechanims.  Stochastic resonance highlights
+these mechanisms.  Stochastic resonance highlights
 one way in which even small noise can play an important role in
@@
 <!--Figure 3: Noisy switch -->
-```{r noisy_switch, fig.width=7, fig.height=7, fig.cap="Stochastic switiching in May's model of alternative stable state dynamics.
-r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000.
+```{r noisy_switch, fig.width=7, fig.height=7, fig.cap="Panel A: Data on prevelance of rubella in Copenhagen illustrating potential stochastic switching dynamics between two regimes, reproduced from @Keeling2001.
+B-D.
+Stochastic switiching in May's model of alternative stable state dynamics.
+Panel B: timeseries showning stochastic switching events between regimes.
+Panel C: separate nonlinear birth and death processes intersect at multiple points creating alternate stable states.
+Panel D: The corresponding potential well derived from the birth-death model: $U(x) = -\\int b(x) - d(x) \\ ud x $.
+Parameter values shown r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000, see code in Appendix A.
 \\label{noisy_switch}"}
@@
 result from some sudden external force or change in the
-environment. As the numerical examples clearly illustrate,
-these transitions can neverthless be explained entirely by chance
+environment. As the numerical examples illustrate,
+these transitions can be explained entirely by chance
 [@Boettiger2012;  @Drake2013].
@@
 more likely as the potential well becomes less deep (i.e. the attractor
-has become less stable), the onset of such transitions can be intepreted
+has become less stable), the onset of such transitions can be interpreted
 as an indication of this loss of stability. This phenomenon has been
@@
 In examining "noise the creator," we began with a tension between
-the intepretation of noise as it is usually presented in the context of
-statistical inference: a nusiance to be stripped away, and how it is
+the interpretation of noise as it is usually presented in the context of
+statistical inference: a nuisance to be stripped away, and how it is
 viewed by theorist, as a potentially fundamental driver of patterns,
@@
 
-Our most provacative example of how noisy phenomena can act to reveal
+Our most provocative example of how noisy phenomena can act to reveal
 underlying processes comes from the literature we have just touched upon
@@
 this idea completely.  The premise behind the tipping point just described
-is that of a fold bifurucation [which I will not detail here, but see @Holling1973; @Scheffer2009; @Strogatz2001].
+is that of a fold bifurcation [which I will not detail here, but see @Holling1973; @Scheffer2009; @Strogatz2001].
 The key thing we need to know about this bifurcation is that it involves
@@
 significant changes to the location of the equilibrium state $\hat x$.
-This derivative did not appear in our marcroscopic equation, $\tfrac{\mathrm{d}x}{\mathrm{d}t} = b(x) - d(x)$,
+This derivative did not appear in our macroscopic equation, $\tfrac{\mathrm{d}x}{\mathrm{d}t} = b(x) - d(x)$,
 but featured regularly in our equations for the noise, such as the 
-expected variance, Eq \eqref{var} and autocorrelation, Eq \eqref{autocor}.
+expected variance, Eq \eqref{var} and auto-correlation, Eq \eqref{autocor}.
 This derivative represents the *stability*  of the equilibrium: how quickly
 deviations return to the stationary value.  
-The magnitude of both variance and autocorrelation must increase
+The magnitude of both variance and auto-correlation must increase
 as the stability ($\partial_x \left[ b(x) - d(x) \right]_{x=\hat x}$) decreases
-prior to a critical tranisition.
+prior to a critical transition.
 
@@
 Panel B: Coefficient of variation for the same experimental data.
-Panel C: Simulated single time series moel.
+Panel C: Simulated single time series model of @Dai2015.
 Panel D, E: Rolling variance and autocorrelation over simulated data show an increase in advance of the transition.
@@
 Figure \ref{tipping} illustrates early warning indicators in an
-experimental manipulation and corrsponding simulated model of
+experimental manipulation and corresponding simulated model of
 @Dai2012 and @Dai2015.  Though early warning signals have been
@@
 cerevisiae_ on a sucrose medium provide the most convincing 
-mechanistic intepretation of the underlying fold bifurcation.
+mechanistic interpretation of the underlying fold bifurcation.
 Because the relevant proteins are
@@
 this analysis into more complex scenarios of structured populations, interacting species,
-autocorrelated noise etc are significantly underdeveloped. Issues of spatial
-structure have recieved more attention [e.g. @Guttal2009; @Dakos2011; @Bel2012; @Dai2013]
-but still largely idosynchratic.  Significant challenges remain
+auto-correlated noise etc are significantly underdeveloped. Issues of spatial
+structure have received more attention [e.g. @Guttal2009; @Dakos2011; @Bel2012; @Dai2013]
+but still largely idiosyncratic.  Significant challenges remain
 in distinguishing these patterns from other causes [@Boettiger2013a].
@@
 More broadly, examples of early warning signals such as increasing variance,
-auto-correllation, and flickering all point to ways in which stochasticty
+auto-correlation, and flickering all point to ways in which stochasticity
 and noise-driven phenomena can be transformed from something which obscures
@@
 
-Intepreting stochastic processes as potential new sources of information
+Interpreting stochastic processes as potential new sources of information
 has also proven to be a

### 2018-03-05 appendixB/appendixB.Rmd
fix typos

@@
 layout: 5p  
-bibliography: ../paper/refs.bib
+bibliography: "../paper/refs.bib"
 output: 
@@
     toc: true
-
-preamble: "\\newcommand{\ud}{\\mathrm{d}}"
+preamble: "\\newcommand{\\ud}{\\mathrm{d}}"
 ---
@@
 
-$$
 \begin{equation}
@@
 \end{equation}
-$$
 
@@
 \end{align}
+
 To change variables, recall the chain rule,
+
 \begin{align}
@@
 \end{align}
+
 Making this substitution to \eqref{En} we find:
@@
 Also by the chain rule, we have
+
 \begin{align*}
@@
 \end{equation}
-$$
+
 The first jump moment, $i == 1$ the two possible steps $r$ are $\pm 1$, and thus $a_1$ is sum of the two possible one-step transitions $+1 \cdot b(x) + -1 \cdot d(x)$.
@@
 
-$$\frac{\mathrm{d} n}{\mathrm{d} t} = r n \left( 1 - \frac{n}{K} \right)$$
+\begin{equation*}
+\frac{\mathrm{d} n}{\mathrm{d} t} = r n \left( 1 - \frac{n}{K} \right)
+\end{equation*}
+
 though the partition into birth and death rates is not explicit.
@@
 
-$$
 \begin{align*}
@@
 \end{align*}
-$$
+
 which corresponds to mean-zero Gaussian noise with auto-correlation $\beta$, where the white-noise limit corresponds to $\beta \to 0$.
-Following the expansion as before (see @vanKampen2007 for details of the multi-variate case), the mean dynamics for the population size are once again given by the macroscopic equation, $\frac{\mathrm{d} x}{\mathrm{d} t} = a_1(x, y) $, while the fluctuation dynamics are now given by three coupled differential equations:
+Following the expansion as before (see @vanKampen2007 for details of the multi-variate case), the mean dynamics for the population size are once again given by the macroscopic equation, $\tfrac{\mathrm{d} x}{\mathrm{d} t} = a_1(x, y)$, while the fluctuation dynamics are now given by three coupled differential equations:
 
-$$
 \begin{align}
@@
 
-$$
-
 Note that the fluctuations in population density $x$ depends as before on the fluctuation-disspation tradeoff, $2 \frac{\partial a_1}{\partial x} \langle \xi^2 \rangle + a_2$, but include the additonal term arising from a covariance with the environmental flucatuations.
@@
 
-$$
+\begin{align}
 \langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} + \underbrace{\frac{\left[\partial_{y} a_1\right] ^2}{\partial_x a_1 - \beta} \frac{\langle\eta^2\rangle}{2\partial_x a_1}}_{\textrm{environmental contribution}}
-$$
+\end{align}
 
@@
 
-$$\langle \eta^2 \rangle = \frac{\gamma_e^2}{2\beta},$$
+\begin{equation*}\langle \eta^2 \rangle = \frac{\gamma_e^2}{2\beta},\end{equation*}
 
@@
 
-$$
+\begin{equation}
 \langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} +  \underbrace{\frac{1}{2} \left[\frac{\partial_{y} a_1}{ \partial_x a_1}\right] ^2}_{\textrm{environmental coupling}} \langle\eta^2\rangle
-$$
+\end{equation}
+
 Taking extinction as such a Gaussian white noise process with meant $\bar e$ and variance $\sigma_e^2$, we can plug in our birth-death equations for the Levins model (and rescaling by system size, from $x = n/N$ back to $n = xN$), into the above equation to find the expected fluctuations are:
 
-$$\sigma_n^2 = \frac{\bar e}{c}N + \frac{\left[(1-\frac{\bar{e}}{c}) N\right]^2}{(\bar e-c)^2} \sigma_e^2$$
+\begin{equation}
+\sigma_n^2 = \frac{\bar e}{c}N + \frac{\left[(1-\frac{\bar{e}}{c}) N\right]^2}{(\bar e-c)^2} \sigma_e^2
+\end{equation}
 
@@
 
-$$
+\begin{equation}
 \langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} + \underbrace{- \frac{\left[\partial_{y} a_1\right] ^2}{2\partial_x a_1 \beta}}_{\textrm{environmental coupling}}  \langle\eta^2\rangle
-$$
+\end{equation}
+
 (Recall that we have taken these expressions at equilibrium, and at equilibrium the derivative of the macroscopic equation $\partial_x a_1 < 0$ by definition, which guarentees the variances above are strictly positive).

### 2018-03-05 appendixB/appendixB.Rmd
sp

@@
 Equations of this form are so-called "Langevin equation," popular in the physics literature [@vanKampen2007].
-This notation looks like a more familiar differential equation, but the resemblence is misleading as the standard calculus does not apply to stochastic equations.
+This notation looks like a more familiar differential equation, but the resemblance is misleading as the standard calculus does not apply to stochastic equations.
 
-A stochastic differntial equation (SDE) of the form
+A stochastic differential equation (SDE) of the form
 
@@
 
-The first integral (over $\textrm{d}t$) is the familiar Reimann integral of the classic calculus, but the second is an Itô integral, which integrates over the stochastic process $B_t$ (Brownian motion, also termed a Weiner process).
+The first integral (over $\textrm{d}t$) is the familiar Riemann integral of the classic calculus, but the second is an Itô integral, which integrates over the stochastic process $B_t$ (Brownian motion, also termed a Wiener process).
 The Itô calculus follows different rules from the classic calculus, largely due to the fact that an Itô integral has non-zero quadratic variation.
@@
 
-The intepretation of Langevin equation is less precise, as it can be read to correspond to one of two possible stochastic integrals, and is often associated with a Stratonovich integral.
+The interpretation of Langevin equation is less precise, as it can be read to correspond to one of two possible stochastic integrals, and is often associated with a Stratonovich integral.
 (Unlike the Itô integral, the Stratonovich integral has zero quadratic variation and so obeys the same chain rule as the classic calculus, but it is not a martingale, e.g.
@@
 Fortunately, the notion as a partial differential equation over probabilities does not have this ambiguity, which shows that @Ovaskainen2010 is referring to an Itô expression, just as I have written it more explicitly.
-See @vanKampen2007 for an excellent discussion of the Itô-Stratonovich dillema and its intepretation with respect to intrinsic vs environmental noise factors.
-To avoid this ambiguity, mathematical literature will frequently write the Statonovich-type SDE explicitly as $\mathrm{d} X_t = f(X_t) \mathrm{d}t +  g(X_t) \circ  B_t$
+See @vanKampen2007 for an excellent discussion of the Itô-Stratonovich dilemma and its interpretation with respect to intrinsic vs environmental noise factors.
+To avoid this ambiguity, mathematical literature will frequently write the Stratonovich-type SDE explicitly as $\mathrm{d} X_t = f(X_t) \mathrm{d}t +  g(X_t) \circ  B_t$
 
@@
 
-The master equation can be thought of as a more process-based description of the Chapman-Kolmorogov equation defining a stochastic process [@vanKampen2007].
+The master equation can be thought of as a more process-based description of the Chapman-Kolmogorov equation defining a stochastic process [@vanKampen2007].
 
@@
 
-The essence of the approach is to argue that in the limit of large populations, the discrete state for the number of indivduals in a population, $N$, can be replaced with a continuous variable, $n$, in which we can expand the master equation as a Taylor series and then truncate at second order.
+The essence of the approach is to argue that in the limit of large populations, the discrete state for the number of individuals in a population, $N$, can be replaced with a continuous variable, $n$, in which we can expand the master equation as a Taylor series and then truncate at second order.
 In contrast to the system size expansion of van Kampen, this approach does not make system size explicit, but instead hinges on taking a simultaneous limit of both short time and small step size.
@@
 
-We will change Eq \eqref{master1} into the variables $\phi$ (the macrosopic value, i.e.
+We will change Eq \eqref{master1} into the variables $\phi$ (the macroscopic value, i.e.
 density) and $\xi$ (i.e.
 fluctuations/deviations from the expected density).
-We begin with the step operator, which can be approximated by a Taylor expansion.To formulate a Taylor expansion of the step operator (where $k$ can be a positive or negative integer), first consider the steop operator under the original discrete variable $n$:
+We begin with the step operator, which can be approximated by a Taylor expansion.To formulate a Taylor expansion of the step operator (where $k$ can be a positive or negative integer), first consider the step operator under the original discrete variable $n$:
 
@@
 
-Note that the fluctuations in population density $x$ depends as before on the fluctuation-disspation tradeoff, $2 \frac{\partial a_1}{\partial x} \langle \xi^2 \rangle + a_2$, but include the additonal term arising from a covariance with the environmental flucatuations.
+Note that the fluctuations in population density $x$ depends as before on the fluctuation-dissipation trade-off, $2 \frac{\partial a_1}{\partial x} \langle \xi^2 \rangle + a_2$, but include the additional term arising from a covariance with the environmental fluctuations.
 We can simplify this expression as before by considering the equilibrium noise dynamics
@@
 
-note that this process provides a justification for moving the noise from 'inside' the deterministic skeleton to merely an additive Gaussian random variable on the end of the determinstic process.
-The details of how the noise term enter the dynamics are reflected in the dervative of the population equation with respect to the environmental variable, $\partial_y a_1(x,y)$, as well as the relative correlation times $\partial_x a_1$ and $\beta$ of the population process and environmental process.
+note that this process provides a justification for moving the noise from 'inside' the deterministic skeleton to merely an additive Gaussian random variable on the end of the deterministic process.
+The details of how the noise term enter the dynamics are reflected in the derivative of the population equation with respect to the environmental variable, $\partial_y a_1(x,y)$, as well as the relative correlation times $\partial_x a_1$ and $\beta$ of the population process and environmental process.
 This environmental contribution adds on to the variance contributed by the demographic noise.
@@
 
-Note that we have left this expression in terms of $\langle \eta^2 \rangle$, rather than subsituting in the equilibrium expression
+Note that we have left this expression in terms of $\langle \eta^2 \rangle$, rather than substituting in the equilibrium expression
 
@@
 so that we can consider separately the limit of white environmental noise without also increasing the variance of the process.
-Changing the correlation while holding the total variation $\langle \eta^2\rangle$ constant thus amounds to increasing the intrinsic noise level $\gamma^2$ proportionally.
-In general, a more mechanistic description of what it means to change the autocorrelation of an environmental process for a given variance is needed.
+Changing the correlation while holding the total variation $\langle \eta^2\rangle$ constant thus amounts to increasing the intrinsic noise level $\gamma^2$ proportionally.
+In general, a more mechanistic description of what it means to change the auto-correlation of an environmental process for a given variance is needed.
 Nevertheless, considering the the white noise limit with some fixed variance, when $\beta \ll \partial_x \alpha_1$,  this expression simplifies as:
@@
 
-Taking extinction as such a Gaussian white noise process with meant $\bar e$ and variance $\sigma_e^2$, we can plug in our birth-death equations for the Levins model (and rescaling by system size, from $x = n/N$ back to $n = xN$), into the above equation to find the expected fluctuations are:
+Taking extinction as such a Gaussian white noise process with meant $\bar e$ and variance $\sigma_e^2$, we can plug in our birth-death equations for the Levins model (and re-scaling by system size, from $x = n/N$ back to $n = xN$), into the above equation to find the expected fluctuations are:
 
@@
 
-(Recall that we have taken these expressions at equilibrium, and at equilibrium the derivative of the macroscopic equation $\partial_x a_1 < 0$ by definition, which guarentees the variances above are strictly positive).
-In this limit ($\beta \gg \partial_x \alpha_1$), the overall noise decreases both with stronger dampening / higher autocorelation in the environmental process, (larger $\beta$) just as it does with increased dampening in population dynamics $\partial_x a_1$.
+(Recall that we have taken these expressions at equilibrium, and at equilibrium the derivative of the macroscopic equation $\partial_x a_1 < 0$ by definition, which guarantees the variances above are strictly positive).
+In this limit ($\beta \gg \partial_x \alpha_1$), the overall noise decreases both with stronger dampening / higher auto-correlation in the environmental process, (larger $\beta$) just as it does with increased dampening in population dynamics $\partial_x a_1$.
 

### 2018-03-06 appendixA/appendixA.Rmd
tweaking refs

@@
 ```{r include=FALSE}
-## note: needs library Cairo
+# note: needs library Cairo
 knitr::opts_chunk$set(message=FALSE, cache = 1, warning = FALSE,  dev="cairo_pdf")
@@
 library(RcppRoll)
-## devtools::install_github("cboettig/regimeshifts")
+# devtools::install_github("cboettig/regimeshifts")
 library(regimeshifts) 
-library(adaptivetau)  ## For Gillespie simulation only
+library(adaptivetau)  # For Gillespie simulation only
 library(tidyverse)
@@
 library(scales)
-## Set a default color pallete
+# Set a default color pallete
 colours <- ptol_pal()(2)
@@
 ```{r echo=FALSE, cache=FALSE, eval=TRUE, message=FALSE}
-## plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
+# plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
 library(Cairo)
@@
 
-## One-time setup
+# One-time setup
 hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
@@
 
-## load fonts (needed each time)
+# load fonts (needed each time)
 extrafont::loadfonts()
@@
 
-e <- 0.2
-c <- 1
+e <- 0.2 # patch death rate
+c <- 1   # patch colonization rate
 
-## Run simulations
+# Run simulation for small and large N
 sim_small <- adaptivetau::ssa.exact(init.values = c(n = 50),
@@
 We use popular functions from the `tidyverse` suite of R packages for the manipulation of `data.frames`.
-Below, the `mutate` function is used to add new columns for plus and minuse one standard deviation based on theoretical calculation of $\sqrt{N\tfrac{e}{c}}$, and the `map_dfr` function is used to apply this to both `small` and `large` cases.
+Below, the `mutate` function is used to add new columns for plus and minus one standard deviation based on theoretical calculation of $\sqrt{N\tfrac{e}{c}}$, and the `map_dfr` function is used to apply this to both `small` and `large` cases.
 We use `tidyverse` syntax  [see @Ross2017] instead of base R in these examples for the following reasons:
@@
 ```{r}
-## combine data.frames, using column called "system_size"
-## to indicate large noise or small noise simulations
+# combine data.frames, using column called "system_size"
+# to indicate large noise or small noise simulations
 gillespie <- 
@@
 
-## Write data to file
+# Write data to file
 write_csv(gillespie, "gillespie.csv")
@@
 
-Example plotting command to generate the figure as shown in main text:
+Example plotting command to generate the figure as shown in main text, (using `ggplot2` plotting methods from `tidyverse`):
 
@@
 Change the values of the model and noise parameters below to explore these results.
-While the models below can easily be implemented in base R, the BUGS syntax for expressing a stochastic model provides a more generic formulation that can easily be used in parameter estimation as well; see @nimble.
-This can also imporve computational performance because @nimble compiles these models into C++ code.
+While the models below can easily be implemented in base R, the BUGS syntax for expressing a stochastic model provides a more generic formulation that can easily be used in parameter estimation as well; see @deValpine2017 for details.
+This can also improve computational performance because NIMBLE compiles these models into C++ code.
 
 ```{r}
-## Define the BUGS model:
+# Define the BUGS model:
 quasicycle  <- nimble::nimbleCode({
@@
   for(t in 1:(N-1)){
+    # Deterministic terms look like normal R code:
     mu_x[t] <- x[t] + x[t] * r * (1 - x[t] / K) - b * x[t] * y[t]
+    # Stochastic assignment uses ~ to indicate 'distributed as' normal (Gaussian)
     x[t+1] ~ dnorm(mu_x[t], sd = sigma_x)
@@
 
-## creates a data.frame of parameter combinations for each noise level
+# creates a data.frame of parameter combinations for each noise level
 p <- 
@@
 
-## define a function to perform a simulation given a set of parameters (constants)
+# define a function to perform a simulation given a set of parameters (constants)
 f <- function(constants){
@@
                                      inits = list(x0 = 1, y0 = 1)))
-  ## set random seed for reproducibility
+  # set random seed for reproducibility
   set.seed(123)
-  ## Run the actual simulation, using nimble
+  # Run the actual simulation, using nimble
   simulate(model)
 
-  ## Assemble the results into a data.frame (tibble).  
+  # Assemble the results into a data.frame (tibble).  
   tibble(t = seq_along(model$x), 
@@
 
-## Do this simulation for each set of parameters, given 
+# Do this simulation for each set of parameters, given 
 quasicycle_df <- p %>% rowwise() %>% do(f(.))
 
-## Write results to file for later reference and plotting
+# Write results to file for later reference and plotting
 write_csv(quasicycle_df, "quasicycles.csv")
@@
 
-From this quasicycle simulation data we can recreate the plot shown in Figure 2 as follows:
+From this quasi-cycle simulation data we can recreate the plot shown in Figure 2 as follows:
 
@@
 read_csv("quasicycles.csv") %>%
-  ## Data in long from
+  # Data in long from
   gather(species, population, -t, -sigma) %>%
-  ## Re-arrange and and re-lable for the plots:
+  # Re-arrange and and re-lable for the plots:
   mutate(sigma = as.factor(sigma)) %>%
@@
 
-Our example of a stochastic osccilato is based on May's model:
+Our example of a stochastic oscillator is based on May's model:
 
@@
 
+# Define stochastic model in BUGS notation
 may  <- nimble::nimbleCode({
@@
   for(t in 1:(N-1)){
+    # Determinstic mean looks like standard R
     mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)  - a * x[t] ^ Q / (x[t] ^ Q + H ^ Q)
+    # Note the use of ~ in BUGS to show 'distributed as normal' 
     y[t+1] ~ dnorm(mu[t], sd = sigma)
@@
 
-Seperately, we will also compute the growth and consumption curves and the potential function for comparison with theory:
+Separately, we will also compute the growth and consumption curves and the potential function for comparison with theory:
 
@@
   tibble(x= seq(0,2, length.out = 100)) %>%
-  ## As before, we add columns with mutate
+  # As before, we add columns with mutate
   mutate(g = x * p$r * (1 - x / p$K), 
@@
   mutate(potential = - cumsum(g - c)) %>%
-  ## assemble resulting data into "long" form with the
-  ## column headings as named here:
+  # assemble resulting data into "long" form with the
+  # column headings as named here:
   gather(curve, y, -x, -potential)
@@
 max_days <- 1000
-## Perform three simultaneous simulations, x,y,z, using
-## different dilution factors DF
+# Perform three simultaneous simulations, x,y,z, using
+# different dilution factors DF
 
-## Intialize variables
+# Intialize variables
 z <- y <- x <- numeric(max_days)
@@
 
-## set random seed for reproduciblity
+# set random seed for reproduciblity
 set.seed(123)
 
-## simulate for max_days
+# simulate for max_days
 for(day in 1:(max_days-1)){
@@
 
-## gather results into long data.frame 
+# gather results into long data.frame 
 df <- data.frame(t = 1:max_days, x = x, y = y, z = z) %>%
@@
 
-## and plot results
+# and plot results
 ggplot(df, aes(t, pop, col=series)) + 
@@
 
-We will also need a function for rolling autocorrelation for our warning sign calculation:
+We will also need a function for rolling auto-correlation for our warning sign calculation:
 
@@
 
-## Simulation of Dai model:
+### Simulation of Dai model:
 
 ```{r}
-## Slow environmental degredation through small stepwise changes
-## (increasing the serial dilution factor, as in the Dai et al experiment.)
+# Slow environmental degredation through small stepwise changes
+# (increasing the serial dilution factor, as in the Dai et al experiment.)
 DF <- as.numeric(sapply(seq(0, 2000, length=9), rep, 40))
@@
 
-## also compute warning signs, include in single table
+# also compute warning signs, include in single table
 tip_df %>% 
@@
 
-We are now ready to create the panels shown in Figure 4:
+We are now ready to create the panels shown in Figure 4.  A little data tidying and `ggplot` layering creates the final plot:
 
@@
   scale_y_log10(labels=scientific)  + 
+  coord_cartesian(ylim=c(1e3, 5e5)) + 
   labs(subtitle = "C. Yeast population size", 
@@
 This is a popular choice since the resulting indicator is a dimensionless measure of the noise relative to the mean, but also confounds changes in the mean dynamics with changes in the noise dynamics.
-This can give a misleading impression of the performance of the indicator: in many models, the mean (that is, the value of the equilibrium population size, $\hat x$) will also decrease prior to the transition, resulting in a more pronounced increase in the coefficent of variation than in the variance.
+This can give a misleading impression of the performance of the indicator: in many models, the mean (that is, the value of the equilibrium population size, $\hat x$) will also decrease prior to the transition, resulting in a more pronounced increase in the coefficient of variation than in the variance.
 However, changes in mean per se should are not a generic indicator of stability predicted by the theory -- indeed, the whole point of the theory was to detect changes that would not be manifested in the mean.
@@
 
-Analytical predicted population size, as a fraction of $K$, is $\tfrac{1}{2} + \tfrac{1}{2} \sqrt{1- 8\sigma^2/K^2}$.
+Analytically predicted population size, as a fraction of $K$, is $\tfrac{1}{2} + \tfrac{1}{2} \sqrt{1- 8\sigma^2/K^2}$.
 Note this is independent of $r$, and increases with $\sigma/K$, the larger the noise $\sigma_g$ as a fraction of the carrying capacity $K$.
@@
 
-## Parameters
+# Parameters
 p <- list(r = .4, K = 1, sigma = .15, N = 500)
@@
 
-## BUGS model
+# BUGS model
 logistic  <- nimble::nimbleCode({
@@
 
-## Compile and run model
+# Compile and run model
 model <- 
@@
 
-## Perform the simulations
+# Perform the simulations
 df <- map_dfr(1:10000, 
@@
 
+Note that the average across stochastic replicates is slightly less than K=1 (black line),
+and matches almost perfectly the equilibrium predicted by the higher order correction theory.
+The standard deviation of fluctuations shown for reference as well.
+
 ```{r}
@@
   geom_line(lwd=1) + 
+  geom_hline(yintercept = 1, col="black") +
   scale_color_ptol() + 
-  coord_cartesian(xlim=c(100,500), ylim=c(1.2, .5))
+  coord_cartesian(xlim=c(100,500), ylim=c(0.1, 1.2))
 ```

### 2018-03-06 paper/paper.Rmd
tweaking refs

@@
   # Word count in abstract: 
-  # Number of References: 
+  # Number of References: 117
   # Archive statement:
@@
 library(png)
+library(scales)
 colours <- ptol_pal()(2)
@@
 #https://stackoverflow.com/a/24241954/258662
-fancy_scientific <- function(l) { 
+scientific <- function(l) {
+    if(max(l, na.rm=TRUE) < 1e3) return(l)
      l <- format(l, scientific = TRUE)
@@
 source for new opportunities for inference.
-
-<!-- -->
 
@@
 models and implicated in empirical examples of increasing complexity, [@Greenman2003],
-particularly in higher-dimensional disease and outbreak models [@Blarer1999; @Greeman2003; @McKane2005; @Alonso2007; @Black2010].
+particularly in higher-dimensional disease and outbreak models [@Blarer1999; @Greenman2003; @McKane2005; @Alonso2007; @Black2010].
 @Blarer1999 identified quasi-cycles as a possible
@@
 Panel C: separate nonlinear birth and death processes intersect at multiple points creating alternate stable states.
-Panel D: The corresponding potential well derived from the birth-death model: $U(x) = -\\int b(x) - d(x) \\ ud x $.
+Panel D: The corresponding potential well derived from the birth-death model: $U(x) = -\\int b(x) - d(x) \\ud x$.
 Parameter values shown r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000, see code in Appendix A.
@@
   geom_vline(aes(xintercept = tip_time), col=colours[2], lty=2) + 
-  scale_y_log10(labels=fancy_scientific)  + 
+  scale_y_log10(labels=scientific)  + 
   coord_cartesian(ylim=c(1e3, 5e5)) +
@@
 
-More broadly, examples of early warning signals such as increasing variance,
-auto-correlation, and flickering all point to ways in which stochasticity
-and noise-driven phenomena can be transformed from something which obscures
-underlying processes to something which reveals them.  I believe this is
-role could apply far more broadly than in just the context of stability and
-tipping points.  The examples from fold bifurcations already suggest
+Interpreting stochastic processes as potential new sources of information
+has also proven to be an effective catalyst in bringing concepts from theory 
+to application. An array of experimental and observational tests of early
+warning signals [_i.e._ @Carpenter2011; @Dai2012; @Clements2016; @Clements2017; @Scheffer2015] 
+has illustrated where these phenomena can be seen in laboratory and field ecosystems,
+and also sparked discussion of
+a wide array of practical [@Boettiger2013b; @Scheffer2015], statistical
+[e.g. @Boettiger2012d; @Boettiger2012], and theoretical [e.g. @Dai2015;
+@Hastings2010] challenges that arise.  This creates a two-way dialog
+between empirical and theoretical work that does much to advance the field
+as a whole. Similar dialog between theory and experiment in other areas
+of stochastic dynamics could benefit both sides.
+
+The potential for stochastic phenomena to provide information about underlying
+processes goes well beyond avialable examples in stability and early warning signals.
+The examples from fold bifurcations already suggest
 that stochastic processes could play a similar role in identifying
@@
 way to stable oscillations (limit cycle) of growing magnitude
-[e.g. @Fussmann2000; @Nelson2013; @Strogatz2001].
-
-Interpreting stochastic processes as potential new sources of information
-has also proven to be a
-rich catalyst for bringing these concepts beyond the realm of abstract
-theory. A rich array of experimental and observational tests of early
-warning signals [_i.e._ @Carpenter2011; @Dai2012; @Clements2016; @Clements2017;
-and many others, _e.g._ see @Scheffer2015 review] has served not merely
-to illustrate these phenomena can be seen in laboratory and field ecosystems,
-but also to spark suggestions of additional possible indicators and surface
-a wide array of practical [@Boettiger2013b; @Scheffer2015], statistical
-[e.g. @Boettiger2012d; @Boettiger2012], and theoretical [e.g. @Dai2015;
-@Hastings2010] challenges to such application.  This creates a two-way dialog
-between empirical and theoretical work that does much to advance the field
-as a whole.
-
-<!--
-One of the most interesting aspects of noise is as a source of
-additional information.  This can be realized in a variety of ways.
-For instance, in deriving expressions for the variance created by
+[e.g. @Fussmann2000; @Nelson2013; @Strogatz2001].  Noise can provide
+information in unrelated ways as well. in deriving expressions for the variance created by
 demographic noise above, we observed that different partitions of
@@
 contribution of environmental noise).
--->
 
@@
 
+# Acknowledgements
+
+The author acknowledges feedback and advice from the editor, 
+Tim Coulson and two anonymous reviewers. This work was supported in
+part by  USDA National Institute of Food and Agriculture, Hatch
+project CA-B-INS-0162-H.
+
 # References

### 2018-03-06 paper/paper.Rmd
tweaks

@@
 ## Use trends style to quickly count refs
-csl: trendsjournals.csl
-#csl: ecology-letters.csl
+#csl: trendsjournals.csl
+csl: ecology-letters.csl
 bibliography: refs.bib
@@
   # Article Type: Review & Synthesis
-  # Running title: Noise phenomena
+  # Running title: From noise to knowledge
   # Figures: 4
   # Tables: 0
-  # Word count: 
-  # Word count in abstract: 
+  # Word count: 7430
+  # Word count in abstract: 181
   # Number of References: 117
@@
 ```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="cairo_pdf", fig.width=7, fig.height=3.5, cache = FALSE)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="cairo_pdf", fig.width=7, fig.height=3.5, cache = TRUE)
 library(tidyverse)
@@
 
-To the empirical ecologist, stochasticity, or more simply, "noise,"
+To the empirical ecologist, stochasticity, or more simply, noise,
 is just that -- something which obscures patterns we are trying to infer [@Knape2011];
@@
 from thinking of stochasticity in such terms; where noise is a nuisance
-that obscures the "deterministic skeleton" of the underlying mechanisms,
+that obscures the deterministic skeleton of the underlying mechanisms,
 to the recognition that stochasticity can itself be a mechanism for 
-driving many interesting phenomena [@Coulson2004].  Yet this transition from "noise the nuisance"
-to "noise the creator" of ecological phenomena has had, with a few notable exceptions,
+driving many interesting phenomena [@Coulson2004].  Yet this transition from *noise the nuisance*
+to *noise the creator* of ecological phenomena has had, with a few notable exceptions,
 relatively little impact in empirical thinking about stochasticity.  One of the
@@
 the nuisance on its head: recognizing that noise driven phenomena can become a tool
-to reveal underlying processes: to become "noise the informer."  Here I argue that
+to reveal underlying processes: to become *noise the informer*.  Here I argue that
 this third shift in perspective offers an opportunity to better bridge the divide
@@
 @Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007;
-@Schreiber2017]; correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010;
+@Schreiber2017]; correlated [@Ripa; @Ripa1999; @Petchey; @Schreiber2010;
 @Lee2017; @Spanio2017] or periodic [@Coulson2001; @Bjornstad2001;
@@
 mechanistically rooted descriptions of noise.  We will see where the common
-formulation of 'deterministic skeleton plus noise term' comes from, how it 
+formulation of "deterministic skeleton plus noise term" comes from, how it 
 is best justified, and when it breaks down.  The second section introduces
-"noise the creator," showing examples of ecological phenomena generated by
+*noise the creator*, showing examples of ecological phenomena generated by
 stochasticity.  These examples will be familiar to most specialists but
@@
 noise is only a nuisance to be filtered or averaged out.  The third and
-final section, "noise the informer," turns these examples back-to-front, 
+final section, *noise the informer*, turns these examples back-to-front, 
 asking what noise can tell us about a system: such as its underlying resilience
@@
 
-<!--
-I approach this synthesis in three parts.  First, I will sketch the ground-work
-for the origins of noise in any model as statistical summary of large number of
-additional processes from some lower level that are modeled only in aggregate.
-This will provide the fundamentals we need to illustrate several key examples
-of noise driven phenomena in ecological dynamics: noise the creator.  Finally,
-we will be able to invert some of these examples; instead of going from process
-to pattern, we will go from pattern to process: noise the informer.  
--->
-
 # Origins of noise
@@
 level in deterministic settings. The notion of a deterministic
-"skeleton" -- responsible for the phenomena of interest -- that is obscured
+skeleton -- responsible for the phenomena of interest -- that is obscured
 from view by the nuisance of noise has slowly given way to recognition
@@
 which reflects the fact that the continuous variable we often think of a
-"population size" or "population density" is an average over a 
+population size or population density is an average over a 
 lower level process of births and deaths among a finite number of 
@@
 
-It is often suggested that stochasticity is "inherent" or "fundamental,"
-e.g. "demographic stochasticity arises because the birth of an individual is probabilistic"
+It is often suggested that stochasticity is inherent or fundamental,
+e.g. demographic stochasticity arises because the birth of an individual is probabilistic
 [@Melbourne2012]. It is easy to forget that this is merely an assertion about a 
@@
 of these sources of noise more precise. Models in which stochasticity was
-essentially "tacked on" to a deterministic skeleton can be substituted for 
+essentially tacked on to a deterministic skeleton can be substituted for 
 approaches that permit both the skeleton and noise to emerge from underlying
@@
 ways to leave the state: a simple expression of probability balance
-known as the "master equation" [@vanKampen2007].  Note that in general
+known as the master equation [@vanKampen2007].  Note that in general
 this approach is equally applicable to stochastic transitions of any sort,
@@
 van Kampen's approximation follows a central-limit theorem style argument established rigorously in
-the theorems of Kurtz at about the same time [@Kurtz1970 & @Kurtz1971, @Kurtz1972, in particular, see Theorem 4.4. @Kurtz1978].
+the theorems of Kurtz at about the same time [e.g @Kurtz1970; @Kurtz1972; @Kurtz1978; theorem 4.4.].
 The heart of the approximation is to imagine a population on some large landscape of which
@@
 This result is striking for a variety of reasons.  First, we see the recovery of the 
-"deterministic skeleton," in which we naively replace population density $x/N$ for
+deterministic skeleton, in which we naively replace population density $x/N$ for
 population count $n$ describes the *expected* density of the population
@@
 
-$$\langle x_0, x_{t} \rangle = \frac{b(x) + d(x)}{2 \partial_x \left[ b(x) - d(x) \right]} \exp\lbrace  -t \left| \partial_x b(x) - \partial_x d(x) \right|  \rbrace $$
+\begin{equation}
+\langle x_0, x_{t} \rangle = \frac{b(x) + d(x)}{2 \partial_x \left[ b(x) - d(x) \right]} \exp\lbrace  -t \left| \partial_x b(x) - \partial_x d(x) \right|  \rbrace \label{autocor}
+\end{equation}
+
 which will also prove useful later on.  Note this expression simply states that the equilibrium
@@
 the diffusion approximation [or Kramer-Moyal expansion @Kramers1940; @Moyal1949; @Gardiner2009]
-which is much more typical in the ecological literature [e.g. the books of @Nisbet1982 and @Lande2003
-or the review in @Ovaskainen2010]. @Black2012 gives an excellent introduction 
+which is much more typical in the ecological literature [e.g. @Nisbet1982; @Lande2003; 
+@Ovaskainen2010]. @Black2012 gives an excellent introduction 
 to the van Kampen expansion as a connection between individual-based and population-level
@@
 but unlike the canonical model, leads to explicitly auto-correlated noise 
-(see derivation, Appendix B).  Auto-correlated or "colored" environmental
+(see derivation, Appendix B).  Auto-correlated or colored environmental
 noise has been a topic of significant theoretical interest 
@@
 
-$$\sigma_n^2 = \frac{\bar e}{c}N + \frac{\left[(1-\frac{\bar{e}}{c}) N\right]^2}{(\bar e - c)^2} \sigma_e^2$$
+\begin{equation}
+\sigma_n^2 = \frac{\bar e}{c}N + \frac{\left[(1-\frac{\bar{e}}{c}) N\right]^2}{(\bar e - c)^2} \sigma_e^2 \label{env-noise}
+\end{equation}
 
@@
 a deterministic skeleton plus some Gaussian white noise term.  Critique, because
-the expansions show that both "noise" and "deterministic skeleton" are the 
+the expansions show that both noise and deterministic skeleton are the 
 result of lower-level mechanistic processes, whose precise form will 
@@
 Gaussian environmental noise.
-
-<!-- This can be condensed or dropped to save space -->
-
-## Origins of other forms of stochasticity
-
-Stochasticity frequently enters our models in other ways as well.
-For instance, most population-level models treat all individuals
-within the population, or at least within a particular age or stage
-class, as identical.  *Individual heterogeneity* in relevant traits
-(growth rates, dispersal ability) can be represented probabilistically
-as a distribution rather than a fixed value, making it function as
-another form of random variation in a model [@Coulson2001;
-@Schreiber2010; @Vindenes2015].  *Spatial heterogeneity* can likewise be reflected
-in stochastic parameters of a spatially implicit model, or as
-spatially varying environmental stochasticity in a spatially explicit
-model.  As with other forms of noise [@Melbourne2008; @Bonsall2004],
-interaction between individual or spatial variation and other sources
-of noise can drive patterns including the potential for persistence,
-coexistence, and dispersal of species [@Schreiber2010; @Hart2016;
-@Schreiber2017].
-
-*Observational error* acts
-much like another form of stochasticity, arising from sampling
-effects or limitations of instruments, etc. Yet unlike the sources
-described above, this uncertainty does not impact the actual state
-of our system, and thus does not give rise to any additional phenomena
-or information about that process, though it can obscure it
-[@Knape2011; @deValpine2002].
-*Deterministic chaos* [@Desharnais2001] is another mechanism of introducing
-significant variation into a simple model. Chaos can a look a lot
-like noise: indeed, it can be observationally indistinguishable
-from a stochastic process [@Ellner2005; @Werndl2009].  Despite
-substantial interest in previous decades and the relative ease of
-encountering chaos in simple ecological models (particularly in
-differences equations), and convincing reconstructions of chaotic
-dynamics in laboratory manipulations [@Costantino1997] the prevalence
-of chaotic dynamics as a driving factor in most ecological systems
-remains unclear. After all, the great variation of natural populations
-is hardly surprising: the undeniable influences of variation created
-by all those things left out of our models, all that individual
-detail from intrinsic, lower level processes and external environmental
-processes that becomes aggregated in our simple models all but begs
-for a stochastic description.
 
@@
 particular addressing auto-correlated noise [e.g. @Ripa; @Ripa1999; 
-@Petchey, @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
+@Petchey; @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
 [e.g. @Durrett1994; @Hastings1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
@@
 
-<!--
-
-Insufficient space left for any treatment of these...
-Permenance, robust permenance
-
-**Bet hedging**, @Lewontin1969 & Cohen 1969; Levins 1969;
-
--->
-
 <!-- Figure 2: Quasi-cycles -->
@@
 
-$$x_{t+1} = x_t + x_t r \left(1 - \frac{x_t}{K}\right) - b x_t  y_t + \xi_{x,t}$$
+\begin{equation}
+x_{t+1} = x_t + x_t r \left(1 - \frac{x_t}{K}\right) - b x_t  y_t + \xi_{x,t} \label{quasi-cycle-prey}
+\end{equation}
+
 where $x_t$ is the prey density at time $t$, $r$ the growth rate, $K$ the carrying capacity, $b$ an encounter rate with the predator $y$, who obeys the equation:
 
-$$y_{t+1} = y_t + c  x_t  y_t - d  y_t+ \xi_{y,t}$$
+\begin{equation}
+y_{t+1} = y_t + c  x_t  y_t - d  y_t+ \xi_{y,t} \label{quasi-cycle-predator}
+\end{equation}
 
@@
 stochastic switching in diverse models, we will focus not specifically
-on the more complex disease model analyzed in @Kelling2001 but
+on the more complex disease model analyzed in @Keeling2001 but
 rather rely on the classic resource consumption model of @May1977
@@
 a convenient analogy to gravity, often implied by drawing a ball
-inside the curve and referring to this as a "ball-in-cup" diagram.
+inside the curve and referring to this as a ball-in-cup diagram.
 The potential function is particularly instructive for understanding
@@
 as an indication of this loss of stability. This phenomenon has been
-dubbed "flickering" and suggested as a warning sign that a more
+dubbed flickering and suggested as a warning sign that a more
 permanent transition to the alternate stable state is approaching [@Brock2010].  
@@
 to ecological scales over a few decades in lake ecosystems [@Wang2012].
-Flickering is then our first example of "noise the informer" -- in
+Flickering is then our first example of noise the informer -- in
 which the emergence of a stochastic phenomenon is re-interpreted 
@@
 
-In examining "noise the creator," we began with a tension between
+In examining noise the creator, we began with a tension between
 the interpretation of noise as it is usually presented in the context of
@@
 insight behind all of these signs comes from the exploiting the ability of
-equations we encountered in "origins of noise"
+equations we encountered in origins of noise
 to reveal subtle changes in underlying processes before those changes are
@@
 but featured regularly in our equations for the noise, such as the 
-expected variance, Eq \eqref{var} and auto-correlation, Eq \eqref{autocor}.
+expected variance, Eq \eqref{vanKampen} and auto-correlation, Eq \eqref{autocor}.
 This derivative represents the *stability*  of the equilibrium: how quickly

### 2018-03-06 paper/paper.Rmd
better creator section intro, new conclusion section

@@
 to use this same approach in alternate formulations.
-
 Figure \ref{gillespie} shows the results of two exact SSA simulations of this model with
@@
 for a generic birth death process and comparison with the more familiar diffusion 
-approximation are found in Appendix B.
-
-van Kampen's approximation follows a central-limit theorem style argument established rigorously in
-the theorems of Kurtz at about the same time [e.g @Kurtz1970; @Kurtz1972; @Kurtz1978; theorem 4.4.].
+approximation are found in Appendix B.  van Kampen's approximation follows a central-limit
+theorem style argument established rigorously in
+the theorems of Kurtz at about the same time [@Kurtz1970; @Kurtz1972; @Kurtz1978].
 The heart of the approximation is to imagine a population on some large landscape of which
@@
 roles of noise and non-linearity in noise-driven phenomena such as stochastic
-transitions between alternate stable states.
-
-The expected auto-correlation, $\langle x_0, x_t \rangle$ also follows from the
+transitions between alternate stable states. The expected auto-correlation,
+$\langle x_0, x_t \rangle$ also follows from the
 system size expansion, e.g. at stationary state:
@@
 variance does not; in contrast to the non-linear SDE from the diffusion approximation.
+Though the limit used in the diffusion approximation can break down in cases where the 
+van Kampen expansion is still valid [see @Gardiner2009], in most circumstances they
+produce similar results and both justify the postulates of the Eq \eqref{canonical}.
 
@@
 sources of noise through the use of the system size expansion serve both
-as critique and justification of the familiar notion of a stochastic model as 
-a deterministic skeleton plus some Gaussian white noise term.  Critique, because
-the expansions show that both noise and deterministic skeleton are the 
+as critique and justification of the familiar notion of a stochastic model.
+This approximation justifies pulling out the stochasticity and adding it on the end
+of a deterministic skeleton (van Kampen's macroscopic equation) as Gaussian 
+white noise term.  The expansion also serves as a reminder that
+both noise and deterministic skeleton are the 
 result of lower-level mechanistic processes, whose precise form will 
-depend on the details of those processes.  Yet the expansion also shows us that
-this Gaussian noise description can be a surprisingly good approximation,
+depend on the details of those processes. The Gaussian form, though
+convenient, is still an approximation, not suitable for small populations
+or rare, large deviations [see Appendix B and @Ovaskainen2010].
+The expansion has given us three things: (1) showing us that additive
+Gaussian noise can be a surprisingly good approximation, (2)
 that noise intensity will converge to a stationary value whenever the system
-dynamics have a stable point, and that the contribution of extrinsic
+dynamics have a stable point, and (3) that the contribution of extrinsic
 (i.e. environmental) noise factors will dominate in large systems.  These
@@
 
-Outside of the primarily theoretical literature on stochastic dynamics
-in ecology, noise is viewed largely as a nuisance: something entirely
-distinct from the processes we are interested in which invariably keeps
-getting in the way.  As we have seen in section on origins, such 
-characterizations are misleading: noise arises from the very same processes.
-While both empirical and statistical literature have frequently emphasized
-techniques to remove noise, a central theme of stochastic processes
-have emphasized an almost opposite conclusion: noise cannot be subtracted or
-averaged away, since the presence of noise can fundamentally alter the
-dynamics we observe.
-Stochasticity can give rise to a wide array of interesting phenomena
-that are often at odds with our intuition. As these examples will
-illustrate, such behavior arises through an interaction between
-stochastic and non-linear elements of a system, underscoring the
-importance of explicitly including stochasticity in ecological
-models [@Coulson2004; @Higgins1997].  Importantly, the noise itself
-does not have to be large to give rise to these phenomena.  In this
-section, I highlight several examples that have shown up frequently
-in the literature and prove sufficiently tractable to provide a
-simple intuition for the mechanisms involved.
+Most ecologists first encounter stochastic models in the context of statistics,
+where noise is just a nuisance to be stripped away [@Gotelli2004; @Bolker2007]. 
+In examining its origins, we were able to justify the simplistic
+view of noise as something merely added on to a deterministic skeleton
+to be less ad hoc than it appears; justified as it is by careful
+approximations and underlying theorems so long as our systems are 
+sufficiently large and the fluctuations sufficiently small.  If this
+appears to suggest that noise is somewhat negligible after all as long
+as we have the deterministic skeleton, then this section is all about 
+rejecting that notion.
+
+A major theme of the literature on stochastic models has been the illustration
+that, even in large systems with small noise, the qualitative behavior
+of a stochastic model can be entirely unlike the behavior of the deterministic
+skeleton [@Higgins1997; @Bjornstad2001; @Coulson2004].  Even with our
+focus on population dynamics alone, examples of these differences are too
+numerous to do justice to all of them here -- instead, we will focus on
+a handful of examples that show how noise interacts with non-linearity
+to create phenomena that are absent from the deterministic model. These examples
+will not only illustrate how ecological theory often sees noise in a
+very different role from ecological statistics, but will serve us well
+in our third section which seeks to turn that view of noise entirely on 
+it's head: not as something that obscures underlying processes but rather as
+a way of revealing them.
 
@@
 
-The loss of a species is without a doubt one of the most important
-possible consequences of noise. Unsurprisingly few topics have
-received greater attention then the question of the persistence of
-a given species [e.g. @Kendall1949;  @Bartlett1955; @Bartlett1960;
-@Lande1993; @Saether1998; @Coulson2001; @Schreiber2010] and coexistence
-of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
-@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017]
-in stochastic models.  Two important trends in this literature have
+Perhaps there is no starker example of differences between the behavior
+of a deterministic skeleton and a stochastic model than in the case
+of persistence of a species and the coexistence of multiple species.  
+Demographic noise guarantees extinction in the long-run for any finite
+population [e.g. @Kendall1949; @Lande2003;  @Bartlett1955; @Bartlett1960;
+@Lande1993; @Saether1998; @Coulson2001; @Schreiber2006;  @Schreiber2010; @Ovaskainen2010;
+@Schreiber2017]. Stochasticity can also
+do the opposite, allowing many species to coexist in scenarios where
+the deterministic skeleton would predict all but one of them to be
+doomed to extinction [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
+@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017].
+Two important trends in this literature have
 been establishing existing results in more general and precise language
@@
 [e.g. @Cohen1979; @Tuljapurkar1989; @Caswell2009; @Metcalf2015; @Vindenes2017b], 
-among other issues.
-
-Demographic noise in particular has fundamental implications for the long-term
-equilibrium: any finite population, however big, is guaranteed to
-eventually go extinct under demographic noise [e.g. @Kendall1949; @Lande2003].  
-In most models it is possible [@Schreiber2006] to define a notion of quasi-stationary 
-distribution concentrated around the deterministic attractor,
-and to prove the probability of extinction decreases exponentially with 
-system size [@Ovaskainen2010; @Schreiber2017]. We will turn our attention
+among other issues. We will turn our attention
 away from the well-trodden ground of persistence and coexistence to focus
-on other phenomena that can emerge in stochastic models.  This will
-highlight the importance of noisy phenomena beyond the ability to obscure
-a deterministic skeleton of some underlying process or in driving extinctions,
-but rather in shaping and driving other interesting patterns we observe in
-natural systems. It is these phenomena that will also underpin the third
-transition in which noise can be turned to our advantage as a source
-of information about ecological processes.
+on other phenomena that can emerge in stochastic models: quasi-cycles and
+stochastic switching.  Both examples will illustrate patterns that differ
+sharply from those of the deterministic skeleton, both for different underlying
+reasons.  Interesting in their own right, these examples will also provide
+intuition for our third section on noise the informer.
 
@@
 create a deeper well.
-
 Because these transitions between alternative stable states appear as
@@
 
-The trick of the early warning signals approach is to view noise as 
+The early warning signals approach views noise as 
 a sequence of countless miniature perturbation experiments.
@@
 The potential for stochastic phenomena to provide information about underlying
-processes goes well beyond avialable examples in stability and early warning signals.
+processes goes well beyond available examples in stability and early warning signals.
 The examples from fold bifurcations already suggest
@@
 
-The past two decades have seen widespread recognition of the central
-role that stochasticity plays through the use of mechanistic models
-can have in creating and understanding ecological phenomena
-[@Coulson2004; @Black2012]. We have seen increasing use of
-master-equation approaches to individual-based models to reflect
-demographic stochasticity that is amenable to analytical treatment
-[@Black2012], and much work in understanding the importance of
-environmental noise, particularly the more realistic but less
-mathematically convenient notion of colored or auto-correlated noise
-on population dynamics, persistence, and coexistence [@Ovaskainen2010;
-@Schreiber2017].  These deductive models can generate phenomena
-that resemble familiar patterns such as population cycles for which
-we have previously sought deterministic explanations, as well as
-stochastic phenomena with no such analog such as stochastic switching.
-Yet too many of these advances have remained limited to the domain
-of theory; limited to deductive models which may seek to mimic
-observed patterns but are rarely used to confront empirical data
-directly.  There is good reason for this: statistical inference of
-even relatively simple models with multiple sources of noise will
-always be difficult given limited ecological data [@Knape2011].
-However, many stochastic phenomena may suggest new ways to infer
-rather than merely obscure underlying processes.
-
-The eruption of interest in early warning signals for critical
+This review has highlighted contrasting views of the role of noise
+between specialists and non-specialists and pointed the way to bridging
+that divide by seeing noise not as a new source of information.
+Most ecologists are introduced to the concept of stochastic models 
+through the lens of statistical inference, in which noise is some
+relatively abstract nuisance to be stripped away by clever statistical
+inference [@Gotelli2004; @Bolker2007].  As we have seen, specialists have long viewed noise in a 
+different light, emerging from the same underlying mechanisms that
+create both what we think of as the deterministic processes of interest
+and the additional fog of noise [@vanKampen2007; @Black2012].  An intuitive approximation of
+larger system size both justifies the convenient partition between
+a so-called deterministic skeleton and noise components of a model,
+while also illustrating how different demographic and environmental
+processes will create different noise dynamics [@Coulson2004; @Ovaskainen2010;
+@Schreiber2017].  We have also seen the divide in how noise is perceived
+goes well beyond merely a more mechanistic origin story. Specialists
+have focused heavily on the role of noise as creator of phenomena:
+qualitative differences in the behavior of fully stochastic models and
+their deterministic skeletons [@Coulson2004].  Tipping our hat to 
+the extensive examples in persistence and coexistence, we explored 
+examples of quasi-cycles and stochastic switching showing patterns
+wholly unlike their deterministic skeletons. Finally, we endeavor to
+bring the divergent views of noise in empirical and theoretical communities
+together by turning the noise the nuisance on its head: drawing on
+examples from the early warning signals literature which have demonstrated
+how noisy phenomena can be used to reveal information about slow
+changes in the underlying deterministic skeleton.
+
+Divergent views of the role of stochasticity have separated our community
+for too long. The eruption of interest in early warning signals for critical
 transitions [@Scheffer2009] provides an excellent illustration of
-this potential.  A range of stochastic phenomena have been suggested
-and tested for their potential to reveal subtle signs of ecosystem
-change prior to a dramatic shift, and prompted a rich dialog between
-further theory and new experiment. Changes in stochastic properties
-such as variance or auto-correlation provide evidence of resilience
-loss that would be invisible to a deterministic model, and non-linear
-stochastic phenomena such as stochastic switching are observed as
-evidence of 'flickering' prior to a critical transition.  In this
-way, our understanding of stochastic dynamics is translated into
-novel sources of information about a system which has sparked a
-wealth of empirical investigation and further theoretical work to
-test the potential of this approach.  The study and application of
-stochastic phenomena to detect ecosystem change is unlikely to be
-limited to this simple scenario of a critical transition or fold
-bifurcation.
-
-Precisely because most stochastic phenomena arise through their
-interaction with non-linearity, the patterns that result can become
-a lens into better understanding those nonlinear processes and
-detecting how they may change.  Meanwhile, the widespread adoption
-of computational tools and the ability to express theoretical results
-in numerical algorithms can make any new approaches more readily
-accessible to application in empirical research. In the past two
+one way to bridge that divide and the value of greater dialogue between
+these two communities.  I hope this review challenges specialists to explore
+the potential of stochastic phenomena to inform on underlying processes --
+such results are more immediately applicable to empirical work that must
+infer process from pattern.  Meanwhile, non-specialists could also
+benefit from a more nuanced view of stochasticity as a part of the processes
+we seek to understand as well as a novel source of information.  All though
+mathematical complexity has frequently been a barrier, computational
+advances and the widespread
+adoption of computational tools such as the R language have made it possible
+to explore these phenomena numerically. Unlike static charts
+empirical data, simulations can be manipulated to build
+an intuition and generate sample data of these processes. To this end, I 
+have provided simple and documented code in Appendix B for each of
+examples discussed here.  In the past two
 decades, we have set to rest those deterministic skeletons that see

### 2018-03-06 paper/paper.Rmd
fix trailing whitespace

@@
 arises through the interaction of stochasticity and alternative
-stable state dynamics.
-
-Both for clarity of presentation and highlight the generality of 
-stochastic switching in diverse models, we will focus not specifically
-on the more complex disease model analyzed in @Keeling2001 but
-rather rely on the classic resource consumption model of @May1977
+stable state dynamics.  To sidestep some of the complexity of the disease
+model proposed in @Keeling2001 we will illustrate the same stochastic 
+switching phenomenon in the classic resource consumption model of @May1977
 which has been widely used to study alternative state dynamics.

### 2018-03-07 paper/paper.Rmd
submitted revisions

@@
   # Word count in abstract: 181
-  # Number of References: 117
+  # Number of References: 111
   # Archive statement:

### 2018-04-12 appendixB/appendixB.Rmd
WIP

@@
 
+## Summary of more complex models
+
+There
+
+Reference        |  Elements                   |  Description
+---------------- | ----------------------------|-----------------
+@Cohen1979       | Age/stage structure
+@Tuljapurkar1989 | Age/stage structure 
+@Tuljapurkar1997 | Age/stage structure 
+@Caswell2009     | Age/stage structure
+@Metcalf2015     | Age/stage structure 
+@Vindenes2017b   | Age/stage structure
+@Durrett1994     | Spatial structure
+@Dieckmann2000   | Spatial structure
+@Kerr2002        | Spatial structure
+@Schreiber2010   | Spatial structure
+@Vindenes2015    | Individual heterogeneity 
+@Hart2016        | Individual heterogeneity 
+@Caswell1978     | Coexistence
+@Tuljapurkar1980 | Coexistence
+@Chesson1981     | Coexistence
+@Chesson1982     | Coexistence
+@Chesson1985     | Coexistence
+@Chesson1989     | Coexistence
+@Melbourne2007   | Coexistence
+@Schreiber2017   | Coexistence
+@Ripa            | Colored noise
+@Roughgarden     | Colored noise
+@Ripa1999        | Colored noise
+@Petchey         | Colored noise
+@Schreiber2010   | Colored noise
+@Lee2017         | Colored noise 
+@Spanio2017      | Colored noise 
+@Coulson2001     | periodic/temporal structure
+@Bjornstad2001   | periodic/temporal structure
+@Keeling2001     | periodic/temporal structure
+@Saether1997     | eco-evolutionary
+@Ozgul2009       | eco-evolutionary
+@Lenormand2009   | eco-evolutionary
+@Vindenes2015    | eco-evolutionary
+@Schreiber2015   | eco-evolutionary
+
 # References

### 2018-04-12 paper/paper.Rmd
WIP

@@
 
-To the empirical ecologist, stochasticity, or more simply, noise,
+To many, stochasticity, or more simply, noise,
 is just that -- something which obscures patterns we are trying to infer [@Knape2011];
@@
 attempt to strip away this undesirable randomness to reveal the patterns
-beneath [@Coulson2001].  Over the past several decades, theoretical ecology has transitioned
+beneath [@Coulson2001].  Over the past several decades, literature in stochasticity has transitioned
 from thinking of stochasticity in such terms; where noise is a nuisance
@@
 to *noise the creator* of ecological phenomena has had, with a few notable exceptions,
-relatively little impact in empirical thinking about stochasticity.  One of the
+relatively little impact in broader thinking about stochasticity.  One of the
 most provocative of those exceptions has turned the classical notion of noise
@@
 
-The gap between specialists and non-specialists is particularly acute in the
-ecological literature on stochastic processes.  This makes the potential benefit
-for new synthesis all the greater, but also poses an obvious challenge in 
-appealing to both audiences.  In recognition of this, I provide concise and
-commented code for simulating each of the models presented in the supplement,
-Appendix A, while more mathematical background and derivations are presented Appendix B.
+This distinction between a paradigms of noise the "nuisance" versus noise the "creator"
+is necessarily a simplification across a spectrum of perspectives. These categories
+should not be mistaken for either a sharp dichotomy nor a reference to a stictly
+empirical-theoretical divide.  It is also essential to recognize that each paradigm
+expands upon rather than rejects the previous notion of noise: the recognition that
+noise can create novel phenomena does not mean that noise cannot also obscure the 
+signal of some process of interest.  Likewise, seeking to use noise as a novel source
+of information about underlying processes will be informed by both previous paradigms,
+as our discussion will illustrate.
+
+Accompanying this discussion, I provide concise and
+commented code for simulating each of the models we will discuss as
+Appendix A, while more mathematical background and derivations can be found in Appendix B.
 Numerical simulations permit poking and prodding of empirically-minded
@@
 under evolutionary dynamics in disease models [@Read2007].
+This further underscores that the collection of more long-term data
+will essential to our understanding noisy phenomena and their underlying mechanisms.
 

### 2018-04-14 appendixB/appendixB.Rmd
WIP

--- 2018-03-05 appendixB/appendixB.Rmd
@@
     address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
-layout: 5p  
+layout: 3p  
 bibliography: "../paper/refs.bib"
@@
 preamble: "\\newcommand{\\ud}{\\mathrm{d}}" 
+header-includes: |
+      \usepackage[table]{xcolor}
+      \definecolor{table-row-color}{HTML}{F5F5F5}
+      \definecolor{table-rule-color}{HTML}{999999}
+      \arrayrulecolor{table-rule-color}
+      \setlength\heavyrulewidth{0.3ex}
+      \renewcommand{\arraystretch}{1.3}
+      \rowcolors{3}{}{table-row-color!100}
+      \let\oldlongtable\longtable
+      \let\endoldlongtable\endlongtable
+      \renewenvironment{longtable}{\oldlongtable} {
+      \endoldlongtable
+      \global\rownum=0\relax}
+
 ---
@@
 
-\begin{figure}
+\begin{figure}{H}
 \begin{center}
@@
 
+\newpage
+
+# Stochasticity in more complex models
+
+The literature exploring how stochasticity plays out in more complex models, such as with age or stage structure, spatial structure, evolutionary dynamics and so forth is far more extensive than can be given justice here.
+Without any suggestion of being a comprehensive review, the following table merely highlights just a few of the classic textbooks, papers, and reviews for any reader looking to explore further, as well as a handful of more recent papers merely to illustrate that these all remain active areas of research.
+Addressing the intepretation, role, and consequences of stochasticity in each of these contexts would provide a more traditional introduction to the subject, such as a graduate course or seminar.
+In the review and synthesis of the main text I have endeavored instead to deviate from this structure and constrain the focus to simpler models to better underscore the paradigms in which we do and might think about noise throughout the field.
+
+Reference        |  Contexts          |  Description / Notes
+-----------------|--------------------|--------------------------------------------------
+@Cohen1979       | Age/stage structure         | Derives strong ergodic theorem for convergence to normalized age structure
+@Tuljapurkar1989 | Age/stage structure         | Classic review of demography in random environments
+@Tuljapurkar1997 | Age/stage structure         | Chapter on stochastic matrix models; from Tuljapurkar and Caswell's classic text on structured population modeling
+@Caswell2009     | Age/stage structure         | Nice recent paper exploring individual stochasticity in age and stage structured populations, with examples from right whales and a prairie plant. 
+@Metcalf2015     | Age/stage structure         | Excellent methods paper for using Integral Projection Models in stochastic population dynamics
+@Vindenes2017b   | Age/stage structure         | 
+@Dieckmann2000   | Spatial structure           | The book "The Geometry of Ecological Interactions" provides wide-ranging examples on approximations to spatially explicit dynamics, including stochastic context.
+@Durrett1994     | Spatial structure & Indiv heterogeneity       | Classic paper on "Importance of Being Discrete and Spatial", includes nice demonstration of how spatially explicit stochastic models with discrete individuals can give qualitatively different behavior from their associated limiting reaction-diffusion (i.e.
+a deterministic, continuous PDE) models.
+
+@Schreiber2010   | Spatial structure& Indiv heterogeneity       |
+@Vindenes2015    | Indiv heterogeneity         |
+@Hart2016        | Indiv heterogeneity         |
+@Caswell1978     | Coexistence                 |  
+@Tuljapurkar1980 | Coexistence                 |
+@Chesson1981     | Coexistence                 | Classic paper introducing what we now call the 'temporal storage effect' for coexistence of multiple species on the same resource through a varying (including but not limited to stochastic) environments.
+@Chesson1985     | Coexistence                 | Storage effect in time and space -- again, stochasticity can be a driver of the necessary variation across space as well as time.
+@Melbourne2007   | Coexistence                 |
+@Schreiber2017   | Coexistence                 | Excellent overview of modern coexistence theory under both cases of demographic and environmental noise.
+Includes several theorems showing how classic results (e.g.
+Chesson and others) can be extended to more general assumptions and also highlights some open questions and conjectures.
+@Roughgarden     | Colored noise               | Elegant simple model showing how autocorrelated environments differ from the dynamics predicted by classic white-noise approximations
+@Ripa            | Colored noise               | Another simple model of colored noise in which extinciton risk decreases with positively auto-correlated noise and decreases with negatively-autocorrelated noise
+@Schreiber2010   | Colored noise               | Excellent illustration of how the impact of temporal auto-correlation in noise will depend on spatial heterogeneity and dispersal.
+(In particular, proves that a metapopulation in which expected fitness in every patch is less than 1 can still persist(!) given positive temporal auto-correlation in relative fitness and weak spatial correlation.)
+@Lee2017         | Colored noise               | Recent numerical study which stands out from most other examples in this list by arguing that under a wide range of parameterizations they consider, ignoring autocorrelation has only a limited impact on estimating expected extinction times.
+@Coulson2001     | periodic/temporal           | 
+@Bjornstad2001   | periodic/temporal           | 
+@Keeling2001     | periodic/temporal           |
+@Saether1997     | eco-evolutionary            | 
+@Vindenes2015    | eco-evolutionary            | 
+@Schreiber2015   | eco-evolutionary            | Evolution of bet-hedging strategies in variable environments (a nice generalization of classic results from Gillespie 1973, 1974 through varying level of correlation within generations)
+@Lenormand2009   | evolutionary                | Excellent recent review of stochasticity in evolutionary models
+
+\pagebreak
+
 # References

### 2018-04-14 paper/paper.Rmd
WIP

@@
 @Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007;
-@Schreiber2017]; correlated [@Ripa; @Ripa1999; @Petchey; @Schreiber2010;
+@Schreiber2017]; correlated [@Roughgarden; @Ripa; @Petchey; @Schreiber2010;
 @Lee2017; @Spanio2017] or periodic [@Coulson2001; @Bjornstad2001;

### 2018-04-16 appendixB/appendixB.Rmd
nearly there

--- 2018-03-05 appendixB/appendixB.Rmd
@@
 
+\newpage
+
+# Stochasticity in more complex models
+
+The literature exploring how stochasticity plays out in more complex models, such as with age or stage structure, spatial structure, evolutionary dynamics and so forth is far more extensive than can be given justice here.
+Without any suggestion of being a comprehensive review, the following table merely highlights just a few of the classic textbooks, papers, and reviews for any reader looking to explore further, as well as a handful of more recent papers merely to illustrate that these all remain active areas of research.
+Addressing the interpretation, role, and consequences of stochasticity in each of these contexts would provide a more traditional introduction to the subject, such as a graduate course or seminar.
+In the review and synthesis of the main text I have endeavored instead to deviate from this structure and constrain the focus to simpler models to better underscore the paradigms in which we do and might think about noise throughout the field.
+
+Reference        |  Contexts          |  Description / Notes
+-----------------|--------------------|--------------------------------------------------
+@Cohen1979       | Age/stage structure         | Derives strong ergodic theorem for convergence to normalized age structure
+@Tuljapurkar1989 | Age/stage structure         | Classic review of demography in random environments
+@Tuljapurkar1997 | Age/stage structure         | Chapter on stochastic matrix models; from Tuljapurkar and Caswell's classic text on structured population modeling
+@Caswell2009     | Age/stage structure         | Nice recent paper exploring individual stochasticity in age and stage structured populations, with examples from right whales and a prairie plant. 
+@Metcalf2015     | Age/stage structure         | Excellent methods paper for using Integral Projection Models in stochastic population dynamics
+@Vindenes2017b   | Age/stage structure         | 
+@Dieckmann2000   | Spatial structure           | The book "The Geometry of Ecological Interactions" provides wide-ranging examples on approximations to spatially explicit dynamics, including stochastic context.
+@Durrett1994     | Spatial structure & Indiv heterogeneity       | Classic paper on "Importance of Being Discrete and Spatial", includes nice demonstration of how spatially explicit stochastic models with discrete individuals can give qualitatively different behavior from their associated limiting reaction-diffusion (i.e.
+a deterministic, continuous PDE) models.
+@Vindenes2015    | Indiv heterogeneity         | Very nice example of a recent paper incorporating individual heterogeneity into integral projection model methods for structured populations.
+@Hart2016        | Indiv heterogeneity         | Another elegant recent example of incorporating individual heterogeneity, this time in a competition model, exploring impact on coexistence
+@Tuljapurkar1980 | Coexistence                 | Proves convergence to the log-normal distribution for stochastic Leslie matrices
+@Chesson1981     | Coexistence                 | Classic paper introducing what we now call the 'temporal storage effect' for coexistence of multiple species on the same resource through a varying (including but not limited to stochastic) environments.
+@Chesson1985     | Coexistence                 | Storage effect in time and space -- again, stochasticity can be a driver of the necessary variation across space as well as time.
+@Schreiber2017   | Coexistence                 | Excellent overview of modern coexistence theory under both cases of demographic and environmental noise.
+Includes several theorems showing how classic results (e.g.
+Chesson and others) can be extended to more general assumptions and also highlights some open questions and conjectures.
+@Roughgarden     | Colored noise               | Elegant simple model showing how auto-correlated environments differ from the dynamics predicted by classic white-noise approximations
+@Ripa            | Colored noise               | Another simple model of colored noise in which extinction risk decreases with positively auto-correlated noise and decreases with negatively-auto-correlated noise
+@Schreiber2010   | Colored noise               | Excellent illustration of how the impact of temporal auto-correlation in noise will depend on spatial heterogeneity and dispersal.
+(In particular, proves that a metapopulation in which expected fitness in every patch is less than 1 can still persist(!) given positive temporal auto-correlation in relative fitness and weak spatial correlation.)
+@Lee2017         | Colored noise               | Recent numerical study which stands out from most other examples in this list by arguing that under a wide range of parameterizations they consider, ignoring auto-correlation has only a limited impact on estimating expected extinction times.
+@Paniw2018       | Colored noise               | Another recent numberical study demonstrating that autocorrelation most impacts stochastic population growth rates in populations with relatively fast life-histories.
+@Coulson2001     | periodic/temporal           | Landmark paper demonstrating the importance of interactions between population structure / individual heterogeneity and environmental stochasticity using long-term data from sheep on St.
+Kilda.
+@Bjornstad2001   | periodic/temporal           | Landmark paper highlighting the tension between 'deterministic' and 'stochastic' explanations for fluctuating populations.
+@Keeling2001     | periodic/temporal           | Nice example of seasonally forced switching between attractors, as discussed in Stochastic Switching section of main text.
+@Saether1997     | eco-evolutionary            | Great early review of environmental stochasticity in population dynamics emphasizing trait-based mechanisms and potential evolutionary consequences of that variation.
+@Vindenes2015    | eco-evolutionary            | Incorporating individual heterogeneity (i.e. individual traits) also allows the authors to explore evolutionary consequences on this variation
+@Schreiber2015   | eco-evolutionary            | Evolution of bet-hedging strategies in variable environments (a nice generalization of classic results from Gillespie 1973, 1974 through varying level of correlation within generations)
+@Lenormand2009   | evolutionary                | Excellent recent review of stochasticity in evolutionary models
+
+\pagebreak
+
 # References

### 2018-04-16 paper/paper.Rmd
nearly there

--- 2018-04-12 paper/paper.Rmd
@@
 
-This distinction between a paradigms of noise the "nuisance" versus noise the "creator"
-is necessarily a simplification across a spectrum of perspectives. These categories
-should not be mistaken for either a sharp dichotomy nor a reference to a stictly
-empirical-theoretical divide.  It is also essential to recognize that each paradigm
+In arguing for this shift, it essential to reconize this is a call for a bigger tent,
+not for the rejection of previous paradigms.  What I will characterize as 'noise the nuisance'
+reflects a predominately statistical approach, in which noise, almost by definition,
+represents all the processes we are not interested in that create additional variation
+which might obscure the pattern of interest.  By contrast, an extensive literature has
+long explored how noise itself can create patterns and explain processes from population
+cycling to coexistence.  These broad categories should be seen as a spectrum and not be
+mistaken for either a sharp
+dichotomy nor a reference to a stictly empirical-theoretical divide.  Each paradigm
 expands upon rather than rejects the previous notion of noise: the recognition that
@@
 @Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007;
-@Schreiber2017]; correlated [@Ripa; @Ripa1999; @Petchey; @Schreiber2010;
-@Lee2017; @Spanio2017] or periodic [@Coulson2001; @Bjornstad2001;
+@Schreiber2017]; correlated [@Roughgarden; @Ripa; @Petchey; @Schreiber2010;
+@Lee2017; @Spanio2017; @Paniw2018] or periodic [@Coulson2001; @Bjornstad2001;
 @Keeling2001] structure in environmental noise, or the interaction
@@
 rely on both textbooks and recent reviews to provide a proper treatment
-of these issues, and focus on broader trends that may be less apparent
-to non-specialists.
+of these issues, and focus on broader trends.
 
@@
 
-We choose this model because it has precisely same functional form
+We choose this model because it has the same functional form
 as the ubiquitous logistic growth model, $\frac{\ud n}{\ud t} = rn (1 - \tfrac{n}{K})$
 but is explicit about how this net growth rate is divided between
-birth and death rate.  As we shall see, this is not merely a notational
-convenience but leads to real differences in observed stochastic properties
+birth and death rate.  This leads to real differences in observed stochastic properties
 of the model. Both the numerical approach (details in Appendix A) and the 
@@
 Most ecologists first encounter stochastic models in the context of statistics,
-where noise is just a nuisance to be stripped away [@Gotelli2004; @Bolker2007]. 
-In examining its origins, we were able to justify the simplistic
+where noise is a nuisance to be stripped away [@Gotelli2004; @Bolker2007] to reveal
+deterministic mechanisms beneath, rather than being seen as a mechanism in and of itself. 
+In examining the origins of noise, we were able to justify the simplistic
 view of noise as something merely added on to a deterministic skeleton
@@
 the deterministic skeleton would predict all but one of them to be
-doomed to extinction [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
+doomed to extinction [@Tuljapurkar1980; @Chesson1981;
 @Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017].
@@
 
-This review has highlighted contrasting views of the role of noise
-between specialists and non-specialists and pointed the way to bridging
-that divide by seeing noise not as a new source of information.
-Most ecologists are introduced to the concept of stochastic models 
-through the lens of statistical inference, in which noise is some
-relatively abstract nuisance to be stripped away by clever statistical
-inference [@Gotelli2004; @Bolker2007].  As we have seen, specialists have long viewed noise in a 
-different light, emerging from the same underlying mechanisms that
-create both what we think of as the deterministic processes of interest
-and the additional fog of noise [@vanKampen2007; @Black2012].  An intuitive approximation of
-larger system size both justifies the convenient partition between
-a so-called deterministic skeleton and noise components of a model,
-while also illustrating how different demographic and environmental
-processes will create different noise dynamics [@Coulson2004; @Ovaskainen2010;
-@Schreiber2017].  We have also seen the divide in how noise is perceived
-goes well beyond merely a more mechanistic origin story. Specialists
-have focused heavily on the role of noise as creator of phenomena:
-qualitative differences in the behavior of fully stochastic models and
-their deterministic skeletons [@Coulson2004].  Tipping our hat to 
-the extensive examples in persistence and coexistence, we explored 
-examples of quasi-cycles and stochastic switching showing patterns
-wholly unlike their deterministic skeletons. Finally, we endeavor to
-bring the divergent views of noise in empirical and theoretical communities
-together by turning the noise the nuisance on its head: drawing on
-examples from the early warning signals literature which have demonstrated
-how noisy phenomena can be used to reveal information about slow
-changes in the underlying deterministic skeleton.
-
-Divergent views of the role of stochasticity have separated our community
-for too long. The eruption of interest in early warning signals for critical
-transitions [@Scheffer2009] provides an excellent illustration of
-one way to bridge that divide and the value of greater dialogue between
-these two communities.  I hope this review challenges specialists to explore
-the potential of stochastic phenomena to inform on underlying processes --
-such results are more immediately applicable to empirical work that must
-infer process from pattern.  Meanwhile, non-specialists could also
-benefit from a more nuanced view of stochasticity as a part of the processes
-we seek to understand as well as a novel source of information.  All though
-mathematical complexity has frequently been a barrier, computational
-advances and the widespread
-adoption of computational tools such as the R language have made it possible
-to explore these phenomena numerically. Unlike static charts
-empirical data, simulations can be manipulated to build
-an intuition and generate sample data of these processes. To this end, I 
-have provided simple and documented code in Appendix B for each of
-examples discussed here.  In the past two
-decades, we have set to rest those deterministic skeletons that see
-noise only as nuisance and embraced the central role noise plays
-as the creator of ecological phenomena. Going forward, I hope and
-predict that we will see greater discussion and use of stochastic
-phenomena to infer underlying processes in empirical patterns in
-similar interactions between theoretical and empirical research.
+This review has explored three paradigms in how noise is viewed throughout the 
+ecological literature, which I have dubbed respectively: noise the *nuisance*, 
+noise the *creator*, and noise the *informer*.  Noise can be seen as a nuisance
+almost by definition: in examining the origins of noise, we have seen how 
+stochasticity is introduced not because ecological processes are random in some fundamental
+sense, but rather, because those processes are influenced by a complex combination
+of forces we do not model explicitly.  In this view, noise captures all that additional
+variation that is separate from the process of interest, and a rich array of 
+statistical methods allow us to separate the one from the other in observations and
+experiments.  By examining the origins of noise, we have seen that despite the complex
+ways in this noise can enter a model, that a Gaussian white-noise approximation [@vanKampen2007; @Black2012] is often
+approriate given a limit of a large system size -- a fact often invoked implicitly but
+rarely derived explicitly from the theorems of @Kurtz1978 and others.
+
+Building on these foundations, we turned to noise the creator, illustrating how
+even small magnitude Gaussian noise could itself create and drive interesting phenomena.
+While a purely statistical paradigm might look to explain patterns such as oscillations
+or sudden transistions in terms of deterministic processes, this section highlighted how
+noise can create and sustain cycles [e.g. @Nisbet1976; @Bjornstad2004] and switches [@Keeling2001].  While our examples focused on the most
+tractable systems, a wealth of literature has explored such phenomena in ever more complex
+contexts.  These examples paint a very different picture of noise, one where "everything matters" [@Bjornstad2001],
+where it can be difficult to know what drives a pattern and where ommitting any of the
+complexity (age or spatial structure, autocorrelation, individual heterogeneity) can
+qualitatively alter the behavior of a model.
+
+Our third paradigm seeks a more optimistic middle ground of noise the informer.
+Here we saw the examples from an empricially driven literature on early warning signals
+[@Scheffer2009; @Dai2012] view noise as a source of countless miniature experiments
+which can reveal the underlying dynamics of a system and how they may be slowly changing.
+In this context, noise does not act to create phenomena of interest directly. The 
+sudden transistions we seek to anticipate are still explained by the deterministic part
+of the model -- bifurcations. But nor is noise a nuisance that merely cloaks this
+deterministic skeleton from plain view: rather, it becomes a novel source of information
+that would be inaccessible from a purely deterministic approach.  I believe more examples
+of how noise can inform on underlying processes is possible, but will require greater
+dialog between these world views.
 

### 2018-04-17 paper/paper.Rmd
submitted 2nd revisions

@@
     email: "cboettig@berkeley.edu"
-
 address:
@@
     address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
-
-## Use trends style to quickly count refs
 #csl: trendsjournals.csl
@@
     keep_tex: true
+  word_document: default
 layout:  3p # review = doublespace, 3p = singlespace, 5p = two-column
@@
   \doublespacing
-
 abstract: |
@@
   # Archive statement:
-
   All data (simulated data and code) is made available on GitHub (<https://github.com/cboettig/noise-phenomena>), and if accepted, will be archived with provided DOI in an appropriate scientific data repositoriy.
@@
   # Keywords: stochasticity, demographic noise, environmental noise, colored noise, quasi-cycles, tipping points
-
-  \newpage
 
@@
 
-#keywords:
-#  - stochasticity
-#  - demographic noise
-#  - environmental noise
-#  - colored noise
-#  - quasi-cycles
-#  - tipping points
+  \newpage
 
 ---
+
+<!--
+keywords:
+  - stochasticity
+  - demographic noise
+  - environmental noise
+  - colored noise
+  - quasi-cycles
+  - tipping points
+-->
 
@@
 
-In arguing for this shift, it essential to reconize this is a call for a bigger tent,
+In arguing for this shift, it essential to recognize this is a call for a bigger tent,
 not for the rejection of previous paradigms.  What I will characterize as 'noise the nuisance'
@@
 mistaken for either a sharp
-dichotomy nor a reference to a stictly empirical-theoretical divide.  Each paradigm
+dichotomy nor a reference to a strictly empirical-theoretical divide.  Each paradigm
 expands upon rather than rejects the previous notion of noise: the recognition that
@@
 commented code for simulating each of the models we will discuss as
-Appendix A, while more mathematical background and derivations can be found in Appendix B.
-Numerical simulations permit poking and prodding of empirically-minded
+Appendix A, and more mathematical background and derivations in Appendix B.
+Numerical simulations permit poking and prodding
 investigation unencumbered by either experimental design or mathematical formalism.
-Most stochastic models are expressed in the BUGS language, which
-may be more familiar to empirical readers than corresponding
-mathematical formulas. This also permits both efficient simulation
-and potential estimation of parameters given sample data using the
-R package, NIMBLE [@deValpine2017]; allowing the deductive models
-illustrated here to readily function as inductive models with
-parameters inferred from time-series data. A copy of this appendix
-is maintained at <https://github.com/cboettig/noise-phenomena>, and
-bug reports, suggestions or help requests are welcome through the
-issue tracker.
+A copy of these appendices
+is maintained at <https://github.com/cboettig/noise-phenomena>.
 
@@
 inherently probabilistic properties: births and deaths of organisms all have
-far more mechanistic explanations which we simply summarize in statistical terms:
-on average -- that is, summing over a wide range of possible mechanisms we have chosen
-not to explicitly model -- an individual death occurs at rate $\lambda$.
+far more mechanistic explanations which we simply summarize in statistical or average terms:
+an individual death occurs at rate $\lambda$.
 This is in fact an extrinsic factor (variables not modeled explicitly),
@@
 
-Where $\textrm{d}B_t^{(d)}$ and $\textrm{d}B_t^{(e)}$ refer to Brownian processes, (i.e.
-Gaussian white noise process) for demographic and environmental stochasticity respectively.
-When $f(n)$ is simply a Verhulst logistic function this is often referred to as the canonical model [e.g.
-@Ovaskainen2010], and provides a clear illustration of the partition of separate deterministic and stochastic elements.
-Significant research continues to focus on extending the study of this model to consider more complex and higher-dimensional $f(n)$ (coexistence models, stage structured models) and consider the case of auto-correlated, rather than than white noise of $\textrm{d}B_t^{(e)}$.
-
-Before we follow that path, we will examine how this canonical model can emerge from a lower-level description and how it can deviate from that description.
-This approach will also set the stage to see how our understanding of stochastic models can be extended to better reflect natural processes that shape not only for the so-called deterministic skeleton, but the noise as well.
-In so doing, I will highlight two tools fundamental to this analysis, one computational and one analytical: the exact simulation method of @Gillespie1977's stochastic simulation algorithm (SSA), and the van Kampen system size expansion [@vanKampen1976].
+Where $\textrm{d}B_t^{(d)}$ and $\textrm{d}B_t^{(e)}$ refer to Brownian processes, 
+(i.e. Gaussian white noise process) for demographic and environmental stochasticity
+respectively.  When $f(n)$ is simply a Verhulst logistic function this is often
+referred to as the canonical model [e.g. @Ovaskainen2010], and provides a clear
+illustration of the partition of separate deterministic and stochastic elements. 
+Significant research continues to focus on extending the study of this model to consider
+more complex and higher-dimensional $f(n)$ (coexistence models, stage structured models) and
+consider the case of auto-correlated, rather than than white noise of $\textrm{d}B_t^{(e)}$.
+
+Here, we will examine how this canonical model can emerge from a
+lower-level description and how it can deviate from that description. This approach will also
+set the stage to see how our understanding of stochastic models can be extended to better
+reflect natural processes that shape not only for the so-called deterministic skeleton, but the noise terms as well.
+In so doing, I will highlight two tools fundamental to this analysis, one computational and one analytical:
+the exact simulation method of @Gillespie1977's stochastic simulation algorithm (SSA), and the
+van Kampen system size expansion [@vanKampen1976].
 
@@
 The @Gillespie1977 provides an exact algorithm
-for simulating demographic stochasticity at an individual level.
-The algorithm is merely a simple and direct implementation of the master
+for simulating demographic stochasticity at an individual level, Figure \ref{gillespie}
+and Appendix A.
+The algorithm is a simple and direct implementation of the master
 equation, progressing in random step sizes determined
@@
 to use this same approach in alternate formulations.
-Figure \ref{gillespie} shows the results of two exact SSA simulations of this model with
-identical parameters except for the total number of available sites,
+
+Figure \ref{gillespie} shows the results of two exact SSA simulations of the classic patch model of @Levins1969,
+with identical parameters in both panels except for the total number of available sites,
 $N$, illustrating the magnitude of these fluctuations indeed scales in proportion
@@
 
-In the interests of a theoretical synthesis, I observe that this result
+Note that this result
 is *different* than the canonical equation, Eq \eqref{canonical}, or more generally
@@
 mistakenly characterizing the van Kampen expansion as synonymous with the diffusion approximation.
-In brief, the diffusion approximation does not lead to a Gaussian distribution
+The diffusion approximation does not lead to a Gaussian distribution
 (equivalently, an SDE or PDE with linear noise term, see Appendix B), but rather
@@
 in environmental time series rather than a process-based derivation.
-As the expansion clearly shows, any bounded continuous-time process will
+Any bounded continuous-time process will
 involve auto-correlation -- measure a continuous process at ever-closer intervals in
-time and these measurements must converge. As the derivation in Appendix B makes clear,
-it is rather the the *relative* time scales of the environmental process
+time and these measurements must converge. Appendix B shows that
+it is the *relative* time scales of the environmental process
 and population level process that really matter (more precisely, the relative 
@@
 the importance of how environmental noise couples to the population dynamics.
-A mechanistic description of environmental noise acts directly on parameters
-of the ecological model: for instance, in our Levins' model we may imagine
-that the extinction rate of patches does not occur at some fixed rate $e$
-but rather at some stochastically fluctuating rate of a given mean, variance,
-and auto-correlation.  The system size expansion illustrates how we replace
+The system size expansion illustrates how we can replace
 a model where stochasticity is nested inside some parameter into the simpler
@@
 where $(1-\frac{\bar{e}}{c}) N$ is the equilibrium population size.
-This reflects the assumption of the canonical equation, Eq \eqref{canonical}
+This justifies the assumption of the canonical equation, Eq \eqref{canonical}
 that the standard deviation (e.g. square root of the variance) scales in direct proportion to the system size $N$,
@@
 with $c$ the conversion ratio and $d$ the predator mortality rate.  
-Here we have assumed for the prey, $\xi_x \sim \mathcal{N(0, \sigma_x)}$, and similarly for the predator driven by $\xi_y$.
-Figure \ref{quasicycles}A shows examples for noise $\sigma_x = \sigma_y = 10^{-5}$, while Figure \ref{quasicycles}B shows $\sigma_x = \sigma_y = 0.01$.
+Here we have assumed for the prey, $\xi_x \sim \mathcal{N(0, \sigma_x)}$,
+and similarly for the predator driven by $\xi_y$. Figure \ref{quasicycles}A 
+shows examples for noise $\sigma_x = \sigma_y = 10^{-5}$, while Figure
+\ref{quasicycles}B shows $\sigma_x = \sigma_y = 0.01$.
 
@@
 ways in this noise can enter a model, that a Gaussian white-noise approximation [@vanKampen2007; @Black2012] is often
-approriate given a limit of a large system size -- a fact often invoked implicitly but
+appropriate given a limit of a large system size -- a fact often invoked implicitly but
 rarely derived explicitly from the theorems of @Kurtz1978 and others.
@@
 While a purely statistical paradigm might look to explain patterns such as oscillations
-or sudden transistions in terms of deterministic processes, this section highlighted how
+or sudden transitions in terms of deterministic processes, this section highlighted how
 noise can create and sustain cycles [e.g. @Nisbet1976; @Bjornstad2004] and switches [@Keeling2001].  While our examples focused on the most
@@
 contexts.  These examples paint a very different picture of noise, one where "everything matters" [@Bjornstad2001],
-where it can be difficult to know what drives a pattern and where ommitting any of the
-complexity (age or spatial structure, autocorrelation, individual heterogeneity) can
+where it can be difficult to know what drives a pattern and where omitting any of the
+complexity (age or spatial structure, auto-correlation, individual heterogeneity) can
 qualitatively alter the behavior of a model.
@@
 Our third paradigm seeks a more optimistic middle ground of noise the informer.
-Here we saw the examples from an empricially driven literature on early warning signals
+Here we saw the examples from an empirically driven literature on early warning signals
 [@Scheffer2009; @Dai2012] view noise as a source of countless miniature experiments
@@
 In this context, noise does not act to create phenomena of interest directly. The 
-sudden transistions we seek to anticipate are still explained by the deterministic part
+sudden transitions we seek to anticipate are still explained by the deterministic part
 of the model -- bifurcations. But nor is noise a nuisance that merely cloaks this

### 2019-10-17 paper/paper.Rmd
fix typo

@@
 state to van Kampen's ODEs, we see the average number $\langle n \rangle$  of occupied patches is
-$\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of 
+$\langle n \rangle = N\left(1 - \tfrac{e}{c}\right)$ with variance at steady-state of 
 $\sigma^2_n = N \tfrac{e}{c}$.  Observe that the variance scales proportionally
